---
title: "Import and tidy with plate2N"
author: "Morgane de Toeuf"
format: 
  gfm:
    toc: true
    #code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# Set up

```{r}
#| message: false
#| warning: false
#| output: false

rm(list = ls())

# once in a while, redownload
#pak::pak("mdetoeuf/plate2N”) # if doesn't work --> try pak::pak_cleanup()
library(plate2N)
library(tidyverse)
#library(janitor) # for row_to_names()
```

# 1 - Import data

All object names ending with "abs" contain absorbance data, all object names ending with "map" contain the mapping of the equivalent abs data

Here we import all our datasets, with various functions from the package `plate2N`. The output is displayed each time a new function is used.

```{r}
# Nmin for t1 and t2
(Nmin_t1t2_abs <- txt_to_tibble("raw_data/Nmin/"))
(Nmin_t1t2_map <- csv_to_tibble("raw_data/Nmin/Nmin_maps.csv"))

# Nmin for t3
Nmin_t3_abs <- csv_to_tibble("raw_data/Nmin_t3/Nmint3_data.csv")
Nmin_t3_map <- csv_to_tibble("raw_data/Nmin_t3/Nmint3_maps.csv")

# TDN (map csv not in the right format --> back to import_csv)
TDN_abs <- csv_to_tibble("raw_data/TDN/TDN_data.csv")
(TDN_map <- 
    read_csv(
      "raw_data/TDN/TDN_maps.csv", 
      col_names = FALSE, col_select = X14:X26, show_col_types = FALSE) |> 
    na.omit() |> 
    rename_with(~c("row", paste0("X", seq(1:12)))))

# Example for Skanit-formatted csv
(Skanit_abs <- skanit_to_tibble("raw_data/Skanit_example/MR_R1_t0.csv")$abs)
(Skanit_map <- skanit_to_tibble("raw_data/Skanit_example/MR_R1_t0.csv")$map)

```

::: callout-tip
## For users of the Skanit file format

The previous chunk (absorbance and plate map data) demonstrate the usage of the Skanit format version of the import function, but the rest of this pipeline does not require the Skanit data, so it is no longer used passed this import step. However, the output format should, in theory, be strictly identical to other format options, so that the object `Skanit_abs` should, in theory, be interchangeable with `TDN_abs` or `Nmin_abs`. Please reach out if you encounter any difficulty.
:::

# 2 - Verticalize plates

We do this separately for each data set. Because those distinct data sets will be combined eventually, we add a prefix to column names to reflect dataset + whether it is abs or map.

```{r}
# Nmint1t2
Nmin_t1t2_vert <- join_abs_map(Nmin_t1t2_abs, Nmin_t1t2_map, dataset = "Nmint1t2-")

# Nmint3
Nmin_t3_vert <- join_abs_map(Nmin_t3_abs, Nmin_t3_map, dataset = "Nmint3-")

# TDN
TDN_vert <- join_abs_map(TDN_abs, TDN_map, dataset = "TDN-")

## Check them out
Nmin_t1t2_vert; Nmin_t3_vert; TDN_vert
```

Then, we use the `dplyr::left_join()` function to join all plate data in a single data table. Note the distinct column names with their dataset and abs/map-related prefixes

```{r}
# join all 3 in a single table with all "plate data"
all_vert <- Nmin_t1t2_vert |> 
  left_join(Nmin_t3_vert, by = join_by(row, column)) |> 
  left_join(TDN_vert, by = join_by(row, column))

# check it out + look at variable names
all_vert; names(all_vert)[1:100]
```

# 3 - tidy table

We use pivot longer to get columns in a more compacted way. In the end, we will need to separate again absorbance data (numerical) from mapping data (strings), that will come in the next step

```{r}
all_vert_too_long <- all_vert |> 
  pivot_longer(
    cols = !any_of(c("row", "column")),
    names_to = c("dataset", "abs_map", "plate_id"),
    names_sep = "-",
    values_to = "value"
    )
all_vert_too_long
```

This went a bit too far. What we want, is to have one column for absorbance data, another one for the mapping. This is what we do with the call to `pivot_wider()` in the next chunk. The next steps in the pipeline (`relocate()` and `mutate()`) serve to improve the table, with additional columns that will prove useful later (well_id and unique_well_id).

```{r}
all_raw_abs_tidy <- all_vert_too_long |> 
  pivot_wider(
    names_from = "abs_map",
    values_from = "value"
  ) |> 
  relocate(map, .before = "abs" ) |> 
  mutate(
    well_id = paste0(row, column),
    unique_well_id = paste0(well_id, "_", plate_id),
    .before = 3
  ) 
all_raw_abs_tidy
```

# 4 - Add plate metadata

Now we have a tidy table that is, in theory, ready for export. In real life, this data cannot be used on its own, and plate metadata is necessary.

First, we import metadata for each dataset, and add a "dataset" column that will help with the next steps.

```{r}
# import csvs
(Nmin_metadata <- read_csv(
  "raw_data/Nmin/Nmin_metadata.csv", show_col_types = FALSE
  ) |>
  mutate(dataset = "Nmint1t2"))
str(Nmin_metadata)

Nmin_t3_metadata <- read_csv(
  "raw_data/Nmin_t3/Nmint3_metadata.csv", show_col_types = FALSE
  ) |> 
  mutate(dataset = "Nmint3")

TDN_metadata <- read_csv(
  "raw_data/TDN/TDN_metadata.csv", show_col_types = FALSE
  ) |> 
  mutate(std_column = as.character(std_column)) |> 
  mutate(dataset = "TDN")
```

Then we join them in one big metadata file. ! this only works because the 3 metadata files have stricly the same stucture (same column names in the same order, containing the same data type (string, ...)).

```{r}
# join csv's
(all_plate_metadata <- bind_rows(Nmin_metadata, Nmin_t3_metadata, TDN_metadata))
```

We could already here, join both metadata and tidy data, but this would create a much larger file to save on the computer than the 2 files separately (because rows of the metadata would be repeated 96 times, one per well of the plate.

# 5 - Export

```{r}
all_raw_abs_tidy |> write_rds("output/data/1.1_all_raw_abs_tidy.rds")
all_plate_metadata |> write_rds("output/data/1.1_all_plate_metadata.rds")
```
