---
title: "V. Pipeline for Absorbance data"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
message: false
warning: false
echo: true
output: false
toc: true
bibliography: references.bib
---

# To Do

-   bla

# Algorithm in natural language

-   Plate info list:

    -   plate number - element \<chr\>

    -   column(s) with std curve(s) - vector \<chr\>

    -   standard identity & unit - vector \<chr\> (element1 = name, element2 = unit)

    -   range of std concentration - vector \<num\>

    -   column(s) with blanc - vector \<chr\>

    -   blanc identity & unit of concentration - vector \<chr\> (element1 = name, element2 = unit)

    -   concentration of blanc - element \<num\>

    -   timestamp - date-time format (?)

    -   wavelength in nm - element \<int\>

-   vectorization of absorbance data, vectors are:

    -   plate number

    -   row

    -   column

    -   raw absorbance

    -   legend (or ID)

-   Correct absorbances for blanc

    -   Return a warning message

        -   when absorbances are below 0.05 or above 1.5

        -   give the number of wells concerned,

        -   give the max and min value of those wells

    -   case when row concerns std curve: correct std absorbances for std blanc

        -   find rows where concentration is zero (slice.min?)

        -   take average of absorbance from those rows

        -   substract to all absorbances of the std the average "zero" value and store it in a new column = "corrected_abs"

    -   case when row concerns samples: Correct sample absorbances for blanc

        -   find rows with blanc

        -   compute variation coefficient

            -   return a warning message when var.coeff \> 30%?

            -   exclude wells based on this? Or human decision?

        -   compute average of blancs

        -   substract that value to all non-standard rows and store it in column "corrected_abs"

-   Compute standard curve

    -   Operation per plate

    -   in plate df, filter rows corresponding to the std

    -   

# Code

## 1 - Set up

Loading packages

```{r}
library(tidyverse)
```

Prep template data: fake tables for the sake of building the next steps of the code. Once that code is working, then we can figure out how to extract real plate data instead of this fake model one.

```{r}
# create an empty table with NAs
matrix <- matrix(NA, nrow = 8, ncol = 12)
# give it names 1 to 12
colnames(matrix) <- as.character(c(1:12))

# turn it into a tibble and add column with letters
plate_empty <- as_tibble(matrix) |> 
  mutate(row = LETTERS[1:8], .before = 1)

# use this template to create a plate with random numbers
plate_abs <- plate_empty
for (col in 2:13) {
  plate_abs[col] = sample(1:100, 8) / 100
}

# same but create random id's for the plate map (whole columns, who cares...)
plate_map <- plate_empty
for (col in 2:13) {
  plate_map[col] = rep(str_flatten(sample(letters, 3)), 8)
}

# create fake plate information, this info could be extracted from real data set, for ex each variable here becomes a column name, and we record one row per plate
plate_id <- "plate_id"
std_column <- c("1", "12")
std_id <- c("no3", "mgN_per_L", "H2O") # n species, unit, prepared in
std_conc <- c(1,2,3,4,5,6,7,8) # we need numerics
blanc_column <- "7"
blanc_id <- c("K2SO4", "M") # extractant, unit
blanc_conc <- 0.5 # we need numeric
timestamp <- timestamp() # see if the format of this is important
wavelength <- "540 nm" # this is just to store info and will not be used for calculations, can be any format 
delay_min <- 30 # also just to store info, it is the incubation time of the manip (plate left under the hood before measurement). Called delay to not confuse with incubation times of PMN and PNR 
```

## 2 - Vectorize plate data

We use pivot longer

```{r}
# vectorize absorbance data
abs_longer <- plate_abs |> 
  pivot_longer(cols = `1`:`12`, names_to = "column", values_to = "abs") 

# vectorize well ids
map_longer <- plate_map |> 
  pivot_longer(cols = `1`:`12`, names_to = "column", values_to = "well_content")

# join both according to info stored in "row" and in "column"
abs_data <- 
  left_join(abs_longer, map_longer, by = c("column", "row")) |> 
  # add info on plate id and well id
  mutate(
    plate_id = rep(plate_id),
    well_id = paste0(row, column),
    .before = 1) |> 
  # sort according to column
  arrange(column)
```

## 3 - Correct absorbance values

### 3.1 - QC - warning if absorbances not in \[0.05,1.5\]

The ideal range for absorbance readings (Beer-lambert in linear range of relationship between concentration and absorbance) is between 0.1 and 1. But these are not super strict borders. I don't want to send out a warning message too soon, so we take higher values.

This chunk filters out only rows where absorbance is out of range, and returns either a warning (when there are out-of-range values) or a happy message (when there are none). In case of a warning, it also shares the table with suspicious wells, so that the user can take an informed decision.

[**To be thought through:**]{.underline}

-   What are options then? Remove suspicious wells (replace by NAs?) –\> deal with it if we are confronted with it

-   another option could be, instead of returning a table, to return only the min and max values of absorbance and/or the number of wells that are out of range

-   We could also visualize a histogram of absorbance with threshold values marked on the graph

```{r}
#** Make sure empty wells contain NA, otherwise, lots of warning messages? To be tested *
#*

# initiate data frame that will contain suspicious well ids
suspicious_wells <- c()
for (i in 1:nrow(abs_data)) {
  if (abs_data$abs[i] < 0.05 || abs_data$abs[i] > 1.5) {
    suspicious_wells <- append(suspicious_wells, abs_data$well_id[i])
    }
}

# Send a warning message
if (!is.null(suspicious_wells)) {
  warning("Some wells are out of range for absorbance, i.e., not in [0.05,1.5] \nSee table hereabove to identify suspicious wells")
  abs_data |> filter(well_id %in% suspicious_wells)
} else {
  message("°^° !! YAY !! °^° All wells are in range for absorbance")
}
```

### 3.2 - Correct std curves for blanc

-   case when row concerns std curve: correct std absorbances for std blanc

    -   find rows where concentration is zero (slice.min?)

    -   take average of absorbance from those rows

    -   substract to all absorbances of the std the average "zero" value and store it in a new column = "corrected_abs"

-   case when row concerns samples: Correct sample absorbances for blanc

    -   find rows with blanc

    -   compute variation coefficient

        -   return a warning message when var.coeff \> 30%?

        -   exclude wells based on this? Or human decision?

    -   compute average of blancs

    -   substract that value to all non-standard rows and store it in column "corrected_abs"
