---
title: "Import and tidy raw data"
format: gfm
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
toc: true
---

# To Do

-   retrieve all raw data, figure out formats and to do

-   For raw "spectro-like" txt files:

    -   Harmonize file names, containing plate nb and N sp

    -   encode a table / tables with plate map infos

    -   encode a metadata table: 1 row per plate, info

# Intro

Unfortunately, absorbance data has been saved in many different formats (txt, csv, xlsx), and files have a diversity of content (how many plates are in 1 file), of naming rules, and of structure (how many rows and columns before the cell A1 appears), ...

So I will have to either record raw data in a new form / rename files, or code multiple ways of importing data. I will first try the latter

# Set up

Load packages

```{r}
library(tidyverse)
library(janitor) # for row_to_names()

```

# 1 - Import spectro-like txt

I will start with the files that are as delivered by the specrto when we work with a USB key (so we only get absorbance data, no plate map or plate info).

## 1.1 - import absorbance data (plate_abs)

In the folder "raw_data", I copied randomly such a file.

```{r}

# have a look at files in the folder
all_files <- list.files("raw_data", full.names = TRUE)
all_files

# I'll take the first one (for now there is only one, this might change...)
plate_abs <- 
  read_tsv(all_files[1], col_names = TRUE, skip = 5, show_col_types = FALSE) |> 
  rename(row = `...1`)


```

## 1.2 - import plate data (plate_map + infos)

From same txt file (from the spectro), we can extract some metadata. The rest will have to be supplied by extra table to be imported. Also, We could get some info from the file name, but that is not a general code, it will depend on each pesron and experiment

```{r}
# 
meta_data <- read_tsv(all_files[1], col_names = FALSE, n_max = 3, show_col_types = FALSE)
#problems(meta_data)

# !! The next line extracts something that looks like a N species, but it is just the spectro protocol, with only 2 possible values: no3 (for nitrate, nitrite, TDN) or nh4 (nh4+)
protocol <- strsplit(meta_data$X2[1], split = " ")[[1]][1]
date <- strsplit(meta_data$X2[1], split = " ")[[1]][2]
time <- strsplit(meta_data$X2[1], split = " ")[[1]][3]

wavelength <- strsplit(meta_data$X1[3], split = ": ")[[1]][2]


# Just as a failsafe, here is everything that we wanted to collect ...
plate_id <- "plate_id"
std_column <- c("1", "12")
std_id <- c("no3", "mgN_per_L", "H2O") # n species, unit, prepared in
std_conc <- c(0,1,2,4,8,16, 24, 32) # we need numerics
extractant_column <- "7"
extractant_id <- c("K2SO4", "M") # extractant, unit
extractant_conc <- 0.5 # we need numeric
timestamp <- timestamp() # see if the format of this is important
wavelength <- "540 nm" # this is just to store info and will not be used for calculations, can be any format 
delay_min <- 30 # 
```

[**To be thought through:**]{.underline}

-   For the rest of the plate info, including a plate map, a file will have to be encoded!

# 2 - Import TDN data from csv (Sang style)

Actually the original file is an xlsx file. But relevant sheets are saved as csv for easier manipulation.

Everything comes from raw data csv that originates from an xlsx file : /Users/Admin/Nextcloud/PhD/2024_trial/Fab_Mo/1_Data/TDN_data.xlsx

## 2.1 - Import plate absorbance data

Now, working for first plate

```{r}
tdn_abs_raw <- read_csv("../raw_data/TDN_data.csv", col_names = FALSE)

i = 1
plate_id <- tdn_abs_raw$X1[i]

plate_abs <- tdn_abs_raw[i:(i+8),] |> 
  row_to_names(row_number = 1) |> 
  rename(row = 1)


```

## 2.2 - import plate map

From the sample list we can create plate maps. The next chunk does it for one plate. —\> make it a loop or a function or something...

```{r}
tdn_samples <- read_csv("../raw_data/TDN_tubes.csv")

std_column <- 1
extractant_column <- 8
plate_id <- tdn_samples$Plate_id_NO3[11]

# For NO3
#tubes <- 
 # (tdn_samples |> filter(Plate_id_NO3 == plate_id))$TN_sample_nb

dilution_up <- "2Xdiluted"
dilution_down <- "nodilution"

tubes_up <- paste0(
  (tdn_samples |> filter(Plate_id_NO3 == plate_id))$TN_sample_nb, 
  "_", dilution_up)

tubes_down <- paste0(
  (tdn_samples |> filter(Plate_id_NO3 == plate_id))$TN_sample_nb, 
  "_", dilution_down)

plate_map <- rbind(tubes_up, tubes_up, tubes_up, tubes_up, 
                tubes_down, tubes_down, tubes_down, tubes_down) |>
  as.matrix() |> as_tibble() |> 
  mutate(std = rep("std"), .before = std_column) |> 
  mutate(extr = rep("extractant"), .before = extractant_column) |> 
  mutate(row = LETTERS[1:8], .before = 1) #|> 

colnames(plate_map) <- c("row", as.character(c(1:12)))

plate_map

```

## 2.3 - import plate metadat

this is strictly speaking the import. Then we can set for each row i the individual parameters

```{r}
meta_data <- read_csv("../raw_data/TDN_metadata.csv")
```

# 3 - import student csv (from xlsx)
