---
title: "Import and tidy raw data"
format: gfm
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
toc: true
---

# To Do

-   For TDN: !! [**If we do the NO2**]{.underline}, then need to deal with the fact that in plate NO2_TDN_18, there is another blanc for some samples (KCl 1M), but that is only in 4 wells...

-   TDN:

    -   Then add it to plate info of Nmin

-   For raw "spectro-like" txt files:

    -   Harmonize file names, containing plate nb and N sp

    -   encode a table / tables with plate map infos

    -   encode a metadata table: 1 row per plate, info

# Intro

Unfortunately, absorbance data has been saved in many different formats (txt, csv, xlsx), and files have a diversity of content (how many plates are in 1 file), of naming rules, and of structure (how many rows and columns before the cell A1 appears), ...

So I will have to either record raw data in a new form / rename files, or code multiple ways of importing data. I will first try the latter

# Set up

Load packages

```{r}
library(tidyverse)
library(janitor) # for row_to_names()

# load functions
source("functions/import_abs_txt.R")
source("functions/import_abs_csv.R")
source("functions/join_maps_abs.R")

```

# 1 - Import Nmin data (t1, t2, t3)

The Nmin absorbance data is in the "raw" format of a txt file as it comes directly from the spectrophotometer

Here we use the import functions to get txt-based data and csv-based data stored in variables. Because the data set is partly in txt (t1, t2, absorbance data), and partly in csv format (t1, t2 maps + t3 absorbance data and maps), we need to go through import steps separately.

```{r}
#| warning: false

#** !! Now turned off warnings --> but should check why it gives such a message and if can really be ignored... !! *

# set file path for data in .TXT format (t1, t2)
filepath <- "raw_data/Nmin/"
# import csv for data in csv format (t3)
csv_file <- read_csv("raw_data/Nmin_t3/Nmint3_data.csv", col_names = FALSE)

# import & shape data
# from .txt format
Nmin_t1t2_abs <- import_abs_txt(filepath)

# from .csv format
Nmin_t3_abs <- import_abs_csv(csv_file)

# look at structure of both data frames
Nmin_t1t2_abs$abs_data_df 
Nmin_t3_abs$abs_data_df
```

Now that all data is in the same format, we can regroup data in two general data frames: one for the absorbance data, one for the "maps" data.

```{r}
# join both data frames
Nmin_all_abs <- left_join(Nmin_t1t2_abs$abs_data_df, Nmin_t3_abs$abs_data_df)
Nmin_all_abs

# look at plate maps
# for Nmin t3
maps_t3_file <- read_csv("raw_data/Nmin_t3/Nmint3_maps.csv", col_names = FALSE)
Nmin_t3_maps <- import_abs_csv(maps_t3_file)
Nmin_t3_maps$abs_data_df 

# For Nmin t1 and t2
maps_t1t1_file <- read_csv("raw_data/Nmin/Nmin_maps.csv", col_names = FALSE)
Nmin_t1t1_maps <- import_abs_csv(maps_t1t1_file)
Nmin_t1t1_maps$abs_data_df

# join both data frames
Nmin_all_maps <- left_join(Nmin_t1t1_maps$abs_data_df, Nmin_t3_maps$abs_data_df)
Nmin_all_maps
```

So now we have those 2 data frames that have strictly the same structure, with `r nrow(Nmin_all_maps)` rows and `r ncol(Nmin_all_maps)` columns, with 2 columns attributed to the well identifier ("row" and "column"), and the remaining `r ncol(Nmin_all_maps) - 2` columns representing the 96-well plates.

## 1.1 - Plate metadata

This section need to be improved, based on files to import. (now it is manual encoding...)

```{r}


# Just as a failsafe, here is everything that we wanted to collect ...

std_column <- c("1", "12")
std_id <- c("no3", "mgN_per_L", "H2O") # n species, unit, prepared in
std_conc <- c(0,1,2,4,8,16, 24, 32) # we need numerics
extractant_column <- "7"
extractant_id <- c("K2SO4", "M") # extractant, unit
extractant_conc <- 0.5 # we need numeric
timestamp <- timestamp() # see if the format of this is important
#wavelength <- "540 nm" # this is just to store info and will not be used for calculations, can be any format 
delay_min <- 30 # 
```

# 2 - Join absorbance and map data

First we need to use `pivot_longer()` to get absorbance data / plate map data in a single column, before we can join the data in a single data frame. The target structure for the data frame would be to have the following columns: "row" (from the plate), "column" (from the plate), "well_id" (= concatenation of row and column), plate_map (sample name or extractant or std curve), absorbance.

The steps for this are recorded into a function `join_maps_abs()`, so we can easily repeat it

```{r}

Nmin_data <- join_maps_abs(maps_df = Nmin_all_maps, abs_df = Nmin_all_abs)
Nmin_data

```

Now that we have our final raw data frame, we can export it to save it as a document to use for downstream analysis

```{r}
write_rds(Nmin_data, file = "output/data/Nmin_tidy.rds")
```

# °°° — Below this, draft, to be picked up — °°°

# 3 - Import TDN data from csv (Sang style)

Actually the original file is an xlsx file. But relevant sheets are saved as csv for easier manipulation.

Everything comes from raw data csv that originates from an xlsx file : /Users/Admin/Nextcloud/PhD/2024_trial/Fab_Mo/1_Data/TDN_data.xlsx

## 2.1 - Import plate absorbance data

Now, working for first plate

```{r}
tdn_abs_raw <- read_csv("raw_data/TDN/TDN_data.csv", col_names = FALSE)

i = 1
plate_id <- tdn_abs_raw$X1[i]

plate_abs <- tdn_abs_raw[i:(i+8),] |> 
  row_to_names(row_number = 1) |> 
  rename(row = 1)


```

## 2.2 - import plate map

From the sample list we can create plate maps. The next chunk does it for one plate. —\> make it a loop or a function or something...

```{r}
#| eval: false
#| include: false
tdn_samples <- read_csv("raw_data/TDN/TDN_tubes.csv")

std_column <- 1
extractant_column <- 8
plate_id <- tdn_samples$Plate_id_NO3[11]

# For NO3
#tubes <- 
 # (tdn_samples |> filter(Plate_id_NO3 == plate_id))$TN_sample_nb

dilution_up <- "2Xdiluted"
dilution_down <- "nodilution"

tubes_up <- paste0(
  (tdn_samples |> filter(Plate_id_NO3 == plate_id))$TN_sample_nb, 
  "_", dilution_up)

tubes_down <- paste0(
  (tdn_samples |> filter(Plate_id_NO3 == plate_id))$TN_sample_nb, 
  "_", dilution_down)

plate_map <- rbind(tubes_up, tubes_up, tubes_up, tubes_up, 
                tubes_down, tubes_down, tubes_down, tubes_down) |>
  as.matrix() |> as_tibble() |> 
  mutate(std = rep("std"), .before = std_column) |> 
  mutate(extr = rep("extractant"), .before = extractant_column) |> 
  mutate(row = LETTERS[1:8], .before = 1) #|> 

colnames(plate_map) <- c("row", as.character(c(1:12)))

plate_map

```

## 2.3 - import plate metadat

this is strictly speaking the import. Then we can set for each row i the individual parameters

```{r}
#| eval: false
#| include: false
meta_data <- read_csv("raw_data/TDN/TDN_metadata.csv")
```

# 3 - import student csv (from xlsx)
