---
title: "1 Wet-lab data"
author: "Morgane de Toeuf"
format: 
  gfm:
    toc: true
    code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# To Do

# Set up

```{r}
#| message: false

rm(list = ls())

library(tidyverse)
```

# 1 - Import raw data

```{r}
# import other "wet lab" raw data
raw_data_field <- read_csv(
  "../raw_data/2024_raw.csv", show_col_types = TRUE,
  col_types = list(
    Soil = col_factor(),
    crop_diversity = col_factor(),
    CS = col_factor(),
    bloc = col_factor()
  ),
  na = c("", "NA", "Na")
  ) |> clean_names() |> 
  rename(sample_short = short) |> 
  filter(expe == "Field")
```

# 2 - Lab data on a per plot basis

(biological units = plots of t2, with 3 zones per plot)

```{r}
raw_field_t2_lab <- raw_data_field |> 
  filter(sampling_time == "t2") |> 
  arrange(biol_unit_nb, zone) |> 
  # remove useless columns
  select(
    biol_unit_nb:sample_short, 
    soil:zone,
    sample_name, #useful?
   starts_with(c("whc", "run", "flush", "yd_rs", "rt"))
    )

# Check it out
raw_field_t2_lab
```

# 3 - Lab data to get water content needed for PMN.

Unfortunately, WC was not computed separately, but can only be derived from the WHC manip.

! Although WC was computed per bloc (4 blocs per soil), PMN was done on a pooled sample from the 4 blocs. So we compute here only one mean value of wc per soil, that is the mean of the 4 blocs.

```{r}
pmn_wc <- raw_data_field |> 
  select(biol_unit_nb, expe, soil, bloc, sampling_time, starts_with("whc")) |> 
  filter(sampling_time == "t0", !is.na(whc_tare_tube_g)) |> 
  #dm = g dry soil / g fresh soil       
  # wc = 1 - dm
  mutate(
    dm = (whc_g_dw_g - whc_tare_dish_g) / (whc_g_fw_g - whc_tare_tube_g),
    wc = 1-dm,
    .keep = "unused",
    .after = sampling_time
  ) |> 
  group_by(soil) |> 
  summarize(
    dm = mean(dm),
    wc = mean(wc)
  )
```

# 4 - Lab data to get water content and fresh weight for PNR

# 5 - Correspondence between biol_unit_nb, cs, soil and bloc

```{r}
cs_map <- raw_field_t2_lab |> 
  select(biol_unit_nb, cs, soil, bloc) |> 
  unique() 

cs_map
```
