---
title: "1 Wet-lab data"
author: "Morgane de Toeuf"
format: 
  gfm:
    toc: true
    code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# To Do

# Set up

```{r}
#| message: false

rm(list = ls())

library(tidyverse)
library(janitor)
```

# 1 - Import raw data

```{r}
# import other "wet lab" raw data
raw_data <- read_csv(
  "../raw_data/2024_raw.csv", show_col_types = TRUE,
  col_types = list(
    Soil = col_factor(),
    crop_diversity = col_factor(),
    CS = col_factor(),
    bloc = col_factor()
  ),
  na = c("", "NA", "Na")
  ) |> clean_names() |> 
  rename(sample_short = short) 
```

# 2 - Extract data for various pipelines

## 2.1 - Water content for PMN

Unfortunately, WC was not computed separately, but can only be derived from the WHC manip.

! Field: Although WC was computed per bloc (4 blocs per soil), PMN was done on a pooled sample from the 4 blocs. So we compute here only one mean value of wc per soil, that is the mean of the 4 blocs.

! Pot: there was one pot per soil for PMN

```{r}
pmn_wc <- raw_data |> 
  select(biol_unit_nb, expe, soil, sampling_time, starts_with("whc")) |> 
  filter(sampling_time == "t0", !is.na(whc_tare_tube_g)) |> 
  #dm = g dry soil / g fresh soil       
  # wc = 1 - dm
  mutate(
    dm = (whc_g_dw_g - whc_tare_dish_g) / (whc_g_fw_g - whc_tare_tube_g),
    wc = 1-dm,
    .keep = "unused",
    .after = sampling_time
  ) |> 
  summarize(
    .by = c("expe", "soil"),
    dm = mean(dm),
    wc = mean(wc)
  )

# For field
pmn_wc_field <- pmn_wc |>  filter(expe == "Field") 

# For greenhouse
pmn_wc_pot <- pmn_wc |> 
  filter(expe == "Pot") |> 
  mutate(biol_unit_nb = paste0(expe, "_", soil))
```

## 2.2 - Field data vs greenhouse data

```{r}
# raw_data_field <- raw_data |> filter(expe == "Field")
# raw_data_pot <- raw_data |> filter(expe == "Pot")
```

## 2.3 - Data on a per plot / pot basis

Field: biological units = plots of t2, with 3 zones per plot

Greenhouse: biological units = pots

```{r}
raw_t2_lab <- raw_data |> 
  filter(sampling_time == "t2") |> 
  # remove bare soil
  filter_out(cs == "B") |> 
  # remove bloc B1 from field data
  filter_out(expe == "Field" & bloc == "B1") |> 
  # correct biological unit
  mutate(biol_unit_nb = case_when(
    biol_unit_nb < 200 ~ biol_unit_nb,
    biol_unit_nb > 200 ~ biol_unit_nb - 200
  )) |> 
  arrange(biol_unit_nb, zone) |> 
  # remove useless columns
  select(
    biol_unit_nb:sample_short, 
    soil:zone,
    sample_name, #useful?
   starts_with(c("whc", "run", "flush", "yd_rs", "rt"))
    )
# 
# raw_field_t2_lab <- raw_t2_lab |> filter(expe == "Field")
# raw_greenhouse_t2_lab <- raw_t2_lab |> filter(expe == "Pot")
# 
# # Field, t2
# raw_field_t2_lab <- raw_data_field |> 
#   filter(sampling_time == "t2") |> 
#   arrange(biol_unit_nb, zone) |> 
#   # remove useless columns
#   select(
#     biol_unit_nb:sample_short, 
#     soil:zone,
#     sample_name, #useful?
#    starts_with(c("whc", "run", "flush", "yd_rs", "rt"))
#     )

# Greenhouse, t2
# raw_greenhouse_t2_lab <- raw_data_pot |> 
#   filter(sampling_time == "t2") |> 
#   # remove bare soil
#   filter_out(cs == "B") |> 
#   # correct biological unit
#   mutate(biol_unit_nb = case_when(
#     biol_unit_nb < 200 ~ biol_unit_nb,
#     biol_unit_nb > 200 ~ biol_unit_nb - 200
#   )) |> 
#   arrange(biol_unit_nb) |> 
#   # remove useless columns
#   select(
#     biol_unit_nb:sampling_time, 
#     sample_name, #useful?
#    starts_with(c("whc", "flush", "yd_rs"))
#     )
```

Check it out

```{r}
# 
# raw_field_t2_lab
# raw_greenhouse_t2_lab
```

### 2.1.1 - Yield data

```{r}
# Greenhouse
yield_data_greenhouse <- raw_t2_lab |> 
  filter(expe == "Pot") |> 
  select(biol_unit_nb:sampling_time, starts_with("yd")) 
```

### 2.1.2 - Flush data (from K2SO4 extraction) for Npools

```{r}
raw_flush <- raw_t2_lab |> 
  # remove clutter
  select(!starts_with(c("yd", "whc", "rt")) & !run_id_mr)

# raw_flush_field <- raw_flush |> filter(expe == "Field")
# raw_flush_greenhouse <- raw_flush |> filter(expe == "Pot")

# Field
# raw_flush_field <- raw_field_t2_lab |> 
#   select(!starts_with(c("yd", "whc", "rt")) & !run_id_mr) |> 
#   # remove bloc 1
#   filter_out(bloc == "B1")

# Greenhouse
# raw_flush_greenhouse <- raw_greenhouse_t2_lab |> 
#   select(
#     !starts_with(c("yd", "whc")) &
#       !cra_trial:sd_c)
```

## 2.4 - Correspondence table: biol_unit_nb, cs, soil and bloc

```{r}
cs_map <- raw_t2_lab |> 
  select(expe, biol_unit_nb, cs, soil, bloc) |> 
  unique() 

# cs_map_field <- cs_map |> filter(expe == "Field")
# cs_map_pot <- cs_map |> filter(expe == "Pot")

# Field
# cs_map_field <- raw_field_t2_lab |> 
#   select(biol_unit_nb, cs, soil, bloc) |> 
#   unique() 
# 
# # Greenhouse
# cs_map_pot <- raw_greenhouse_t2_lab |> 
#   select(biol_unit_nb, cs, soil, bloc) |> 
#   unique()
# 
# cs_map_field; cs_map_pot
```

## 2.5 - Water content and fresh weight for PNR

# 3 - Compute new variables

## 3.1 - Dry matter content from flush data for Npools

### 3.1.1 - Principle and equations

In this data set, dry matter and water content have not yet been computed. The raw data with 3 technical replicates is encoded, in separate columns starting with `flush_dm...`\`. In this section, we compute average dry matter and water content for the sample (across technical replicates). We add additional columns with intermediary results:

- FW~neat~ = FW~gross~ - tare ; DW~neat~ = DW~gross~ - tare

- average (FW, then DW) = sum(3 technical replicates) / 3

- DM = avg~DW~ / avg~FW~

- WC = 1 - DM

- With

  - FW = fresh weight \[g\]

  - DW = dry weight \[g\],

  - DM = dry matter content \[g~dry soil~ / g~fresh soil~\],

  - WC = water content \[g~water~ / g~fresh soil~\] [**!! UNIT: per g fresh soil !!**]{.underline}

- ratio = fresh weight of soil added into tubes for extraction (\~10g) / volume of extractant added (=20ml)

Then, we can finally convert concentrations in mg/L into concentrations in ppm (mg/kg dry soil)

- see <https://mycloud.ulb.be/index.php/s/9ZKaJg9rXJg4iB7> for equation

### 3.1.2 - Pivot data so that technical replicates are above each other, not next to each other

First, we select relevant columns and pivot to get

- one column respectively for tare, gross fresh weight (g_fw) and gross dry weight (g_dw)

- 1 row per technical replicate

```{r}
raw_subset_wc <- raw_flush |> 
  select(expe, sample_short, flush_dm_tare_tr1:flush_dm_g_dw_tr3) |> 
  pivot_longer(
    cols = !c(expe, sample_short),
    names_pattern = "flush_dm_(g_fw|g_dw|tare)_tr(\\d)",
    names_to = c(".value", "tech_rep"),
    values_to = "weight") 

# raw_subset_wc_field <- raw_subset_flush |> filter(expe == "Field")
# raw_subset_wc_greenhouse <- raw_subset_flush |> filter(expe == "Pot")
# 

#Field
# raw_subset_wc_field <- raw_flush_field |> 
#   select(sample_short, flush_dm_tare_tr1:flush_dm_g_dw_tr3) |> 
#   pivot_longer(
#     cols = !sample_short,
#     names_pattern = "flush_dm_(g_fw|g_dw|tare)_tr(\\d)",
#     names_to = c(".value", "tech_rep"),
#     values_to = "weight") 
# 
# # check it out
# raw_subset_wc_field 
# 
# # Greenhouse
# raw_subset_wc_greenhouse <- raw_flush_greenhouse |> 
#   select(sample_short, flush_dm_tare_tr1:flush_dm_g_dw_tr3) |> 
#   pivot_longer(
#     cols = !sample_short,
#     names_pattern = "flush_dm_(g_fw|g_dw|tare)_tr(\\d)",
#     names_to = c(".value", "tech_rep"),
#     values_to = "weight") 
# 
# # check it out
# raw_subset_wc_greenhouse
```

Then, we

- compute the neat fresh weight and dry weight for each row, and the dry matter content and water content

- Compute the average per sample (over the technical triplicate)

```{r}
# To simplify computation, bind both data
# raw_subset_wc <- bind_rows(
#   raw_subset_wc_field |> mutate(expe = "Field"),
#   raw_subset_wc_greenhouse |> mutate(expe = "Greenhouse")
# )

wc <- raw_subset_wc |> 
  # compute neat values
  mutate(
    n_fw = g_fw - tare,
    n_dw = g_dw - tare,
    dm = n_dw / n_fw,
    wc = 1 - dm,
    .keep = "unused"
  ) |> 
  # compute mean per sample
  summarize(
    .by = c(expe, sample_short),
    dm = mean(dm),
    wc = mean(wc)
  ) 

# wc_field <- wc |> filter(expe == "Field")
# 
# wc_greenhouse <- wc |> filter(expe == "Pot")
# 
# # check it out
# wc_field 
# wc_greenhouse 
```

Now that we have that clean data, we can

- remove useless columns (raw data used above)

- rejoin it to the complete greenhouse data (with all columns)

- compute each sample's individual "ratio" (\~10g/20ml)

```{r}
# Volume of extractant used in the K2SO4 extraction
vol_extr <- 20 

flush_clean <- raw_flush |> 
  select(
    !flush_dm_tare_tr1:flush_dm_g_dw_tr3 & 
      !ends_with("comment")) |> 
  left_join(wc) |> 
  mutate(
    ratio_nf = flush_fw_nf_g / vol_extr,
    ratio_cfe = flush_fw_cfe_g / vol_extr,
    .keep = "unused") 

#flush_clean_field <- flush_clean |> filter(expe == "Field")
#flush_clean_greenhouse <- flush_clean |> filter(expe == "Pot")

# flush_clean_field <- raw_flush_field |> 
#   select(
#     !flush_dm_tare_tr1:flush_dm_g_dw_tr3 & 
#       !ends_with("comment")) |> 
#   left_join(wc_field) |> 
#   mutate(
#     ratio_nf = flush_fw_nf_g / vol_extr,
#     ratio_cfe = flush_fw_cfe_g / vol_extr,
#     .keep = "unused") 
# 
# flush_clean_greenhouse <- raw_flush_greenhouse |> 
#   select(
#     !flush_dm_tare_tr1:flush_dm_g_dw_tr3 & 
#       !ends_with("comment")) |> 
#   left_join(wc_greenhouse) |> 
#   mutate(
#     ratio_nf = flush_fw_nf_g / vol_extr,
#     ratio_cfe = flush_fw_cfe_g / vol_extr,
#     .keep = "unused") 
 
# Check it out
# flush_clean_field
# flush_clean_greenhouse
```

Now, we divide the data in 2 subsets: sample data and standard soil data (several reps, messes up the analysis (for now))

[**!! TODO: decide if export sample & std grouped or separated**]{.underline}

```{r}
flush_sample_greenhouse <- flush_clean |> filter(biol_unit_nb < 100 & expe == "Pot")
flush_std_greenhouse <- flush_clean |> filter(biol_unit_nb >= 100 & expe == "Pot")

flush_sample_field <- flush_clean |> filter(biol_unit_nb < 112 & expe == "Field")
flush_std_field <- flush_clean |> filter(biol_unit_nb >= 112 & expe == "Field")
```

## 3.2 - Develop yield data (greenhouse)

The per-plant yield data for the field was deemed not trustworthy. In this section, there is thus only greenhouse yield data. Harvest yield data for the field was imported and developed in the pipeline `1_CRA_data`.

### 3.2.1 - Human error check

First, we compute the number of plants of each species in each pot

```{r}
nb_plant <- yield_data_greenhouse |> 
  rowwise() |> 
  relocate(starts_with(c("yd_rs_f_h", "yd_rs_f_s", "yd_rs_w_h", "yd_rs_w_ti"))) |> 
  mutate(
    # compute nb of plants of each sp
    nb_fb = case_when(
      !is.na(yd_rs_f_stem1) & !is.na(yd_rs_f_stem2) ~ 2,
      !is.na(yd_rs_f_stem1) & is.na(yd_rs_f_stem2) ~ 1
    ),
    nb_w = case_when(
      !is.na(yd_rs_w_till1) & !is.na(yd_rs_w_till2) ~ 2,
      !is.na(yd_rs_w_till1) & is.na(yd_rs_w_till2) ~ 1
    ))
```

First, we check the comments written during experimentation

```{r}
nb_plant |> 
  select(biol_unit_nb, soil, cs, crop_diversity, nb_w, nb_fb, yd_rs_comment) |> 
  filter(!is.na(yd_rs_comment)) |> print(width = 110)
```

And we correct the data accordingly

- pot #14: change nb of wheat plants to 1

- pot #19: on n'en fait rien de cette info...?

- Pot #27: Changer IC en W pour cs, et IC en SC pour crop_diversity

- Pot #30: on ne fait rien de cette info, mais si nb est outlier, on sait pourquoi...

- Pot #55: changer IC en F (cs) et en SC (crop_diversity)

```{r}
nb_plant_clean <- nb_plant |> 
  mutate(
    nb_w = case_when(biol_unit_nb == 14 ~ 1, .default = nb_w),
    cs = case_when(
      biol_unit_nb == 27 ~ "W",
      biol_unit_nb == 55 ~ "F",
      .default = cs
    ),
    crop_diversity = case_when(
      biol_unit_nb %in% c(27, 55) ~ "SC",
      .default = crop_diversity
    )
  )
```

Check the modifications

```{r}
nb_plant_clean |> 
  select(biol_unit_nb, soil, cs, crop_diversity, nb_w, nb_fb, yd_rs_comment) |> 
  filter(!is.na(yd_rs_comment)) |> print(width = 110)
```

All good.

### 3.2.2 - Compute per plant clean yield data

Compute per plant versions of weight, plant height and stem/till number

```{r}
yield_per_plant <- nb_plant_clean |> 
  rowwise() |> 
  mutate(
    height_per_fb = mean(c_across(starts_with("yd_rs_f_h")), na.rm = TRUE),
    stem_per_fb = mean(c_across(starts_with("yd_rs_f_s")), na.rm = TRUE),
    height_per_w = mean(c_across(starts_with("yd_rs_w_h")), na.rm = TRUE),
    till_per_w = mean(c_across(starts_with("yd_rs_w_ti")), na.rm = TRUE),
    
    fw_fb_per_pot = yd_rs_f_biom_1_2plants_raw_fw_g - yd_rs_f_tare_g,
    fw_w_per_pot = yd_rs_w_biom_1_2plants_raw_fw_g - yd_rs_w_tare_g,
    dw_fb_per_pot = yd_rs_f_biom_1_2plants_raw_dw_g - yd_rs_f_tare_g,
    dw_w_per_pot = yd_rs_w_biom_1_2plants_raw_dw_g - yd_rs_w_tare_g,
    
    fw_fb_per_plant = fw_fb_per_pot / nb_fb,
    fw_w_per_plant = fw_w_per_pot / nb_w,
    dw_fb_per_plant = dw_fb_per_pot / nb_fb,
    dw_w_per_plant = dw_w_per_pot / nb_w,
    
    .keep = "unused"
    ) 

# check it out
yield_per_plant
```

Save it in clean data set to join and export

```{r}
yield_clean_greenhouse <- yield_per_plant
```

# Export

```{r}
cs_map_field <- cs_map |> filter(expe == "Field")
cs_map_pot <- cs_map |> filter(expe == "Pot")
cs_map_field |> write_rds("output/data/1_field_cs_map.rds")
cs_map_pot |> write_rds("output/data/1_greenhouse_cs_map.rds")

pmn_wc_field |> write_rds("output/data/1_field_pmn_wc.rds")
pmn_wc_pot |> write_rds("output/data/1_greenhouse_pmn_wc.rds")

flush_clean_field <- flush_clean |> filter(expe == "Field")
flush_clean_greenhouse <- flush_clean |> filter(expe == "Pot")
flush_clean |> write_rds("output/data/1_flush_clean.rds")
flush_clean_greenhouse |>  write_rds("output/data/1_greenhouse_flush_clean.rds")
flush_clean_field |>  write_rds("output/data/1_field_flush_clean.rds")
 # or flush_sample_greenhouse # and flush_std_greenhouse 

yield_clean_greenhouse |> write_rds("output/data/1_greenhouse_yield_clean.rds")
```

Safekeep from other scripts. Consider whether this export is needed

```{r}
#| eval: false
#| include: false
raw_field_t2_lab |> write_rds("output/data/3_field_t2_raw_lab.rds")
raw_greenhouse_t2 |> write_rds("output/data/3_greenhouse_t2_raw_lab.rds")
```
