---
title: "MicroResp Data, from absorbance to indices"
author: "Morgane de Toeuf"
format: 
  gfm:
    toc: true
    #code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
bibliography: references.bib
---

# TO DO

# Set up

```{r}
rm(list = ls())

library(plate2N)
library(tidyverse)

# Functions
source("functions/MR_to_tibble.R")
```

# 1 - Import raw data

## 1.1 - Import raw absorbance data

First, import absorbance data (t0 = pre-incubation, t5 = after 5h of incubation). Plate ids correspond to sample names (equivalence table stored somewhere else)

```{r}

(MR_greenhouse_abs_t0 <- MR_to_tibble(
  filepath = "../raw_data/MR_Pot_abs.csv", 
  column_range = c(1:13)))

MR_greenhouse_abs_t5 <- MR_to_tibble(
  filepath = "../raw_data/MR_Pot_abs.csv", 
  column_range = c(17:29))

(MR_field_abs_t0 <- MR_to_tibble(
  filepath = "../raw_data/MR_Field_abs.csv", 
  column_range = c(1:13)))

MR_field_abs_t5 <- MR_to_tibble(
  filepath = "../raw_data/MR_Field_abs.csv", 
  column_range = c(16:28))

```

## 1.2 - prepare the mapping

```{r}
MR_columns <- c(
  "Std_Glu", "Std_H2O", 
  "H2O", "OA", "Glu", "Lgn", "NAG", "gABA", "Ala", "Urea")

nb_plates_greenhouse <- MR_greenhouse_abs_t0 |> 
  filter(row %ni% LETTERS) |> 
  nrow()

nb_plates_field <- MR_field_abs_t0 |> 
  filter(row %ni% LETTERS) |> 
  nrow()

MR_greenhouse_map <- map_plates(
  plate_ids = paste0("P", sprintf("%02d", seq(01:nb_plates_greenhouse))), 
  samples = rep(MR_columns, nb_plates_greenhouse),
  n_samples_per_plate = 10,
  column_curves = c(), column_blank = c(), column_empty = c(1,12),
  n_wells_samples = 8)

MR_field_map <- map_plates(
  plate_ids = paste0("P", sprintf("%02d", seq(01:nb_plates_field))), 
  samples = rep(MR_columns, nb_plates_field),
  n_samples_per_plate = 10,
  column_curves = c(), column_blank = c(), column_empty = c(1,12),
  n_wells_samples = 8)
```

## 1.3 - Import metadata

First import

```{r}
MR_greenhouse_metadata <- read_csv(
  "../raw_data/MR_Pot_metadata.csv", show_col_types = FALSE) |> 
  as_tibble() |> 
  select(!c(1,2)) |> 
  janitor::clean_names() |>
  drop_na(plate_id)

MR_field_metadata <- read_csv(
  "../raw_data/MR_Field_metadata.csv", show_col_types = FALSE) |> 
  as_tibble() |> 
  select(!c(1,2)) |>
  janitor::clean_names() |>
  drop_na(plate_id) 

```

Then, join greenhouse and field metadata into 1 object and add equivalence between dataset and expe (needed later)

```{r}
all_MR_metadata <- bind_rows(MR_greenhouse_metadata, MR_field_metadata) |> 
  # add column dataset into metadat to use for left_join() later
  mutate(dataset = case_when(
    expe == "Pot" ~ "MR_greenhouse",
    expe == "Field" ~ "MR_field",
    .default = "oups"),
    .before = expe)
```

# 2 - From plate layout to tidy data

## 2.1 - Verticalize all plate data

Verticalize, 1 column per plate, 96 rows (1 per well position)

```{r}
MR_greenhouse_vertical <- join_abs_map( tibble_list = list( MR_greenhouse_abs_t0, MR_greenhouse_abs_t5, MR_greenhouse_map), abs_map = c("abs_t0-", "abs_t5-", "map-"), coerce_numeric = FALSE, dataset = "MR_greenhouse-" )

MR_field_vertical <- join_abs_map( tibble_list = list( MR_field_abs_t0, MR_field_abs_t5, MR_field_map), abs_map = c("abs_t0-", "abs_t5-", "map-"), coerce_numeric = FALSE, dataset = "MR_field-" )

# Check it out
MR_greenhouse_vertical ; MR_field_vertical

# Join them in one
all_vertical_MR <- MR_greenhouse_vertical |> 
  left_join(MR_field_vertical, by = join_by(row, column))

# Check out the structure of the column names (one per plate)
sample(names(all_vertical_MR),size = 30)

# Check out data
all_vertical_MR
```

## 2.2 - Elongate data into tidy format

Tidy the table, 1 row per unique well (96 x nb of physical plates), and 3 important columns: `abs_t0`, `abs_t5`, `map`.

```{r}
(all_raw_abs_tidy_MR <- vertical_to_tidy(all_vertical_MR, column_def = c("abs_t0", "abs_t5", "map")))
```

Optional: export the data

```{r}
#| eval: false
#| include: false
# all_raw_abs_tidy_MR |> write_rds("output/data/1_all_raw_abs_tidy_MR.rds")
# all_MR_metadata |> write_rds("output/data/1_all_MR_metadata.rds")

```

## 2.3 - Join data with metadata

Optional: load data

```{r}
#all_raw_abs_tidy_MR <- read_rds("output/data/1_all_raw_abs_tidy_MR.rds")
#all_MR_metadata <- read_rds("output/data/1_all_MR_metadata.rds") 

```

Join data with metadata, and add "dataset" column (derived from "expe")

```{r}
MR <- all_raw_abs_tidy_MR |> left_join(all_MR_metadata) 
```

# 3 - clean data

## 3.1 - QC - suspicious wells

Check first t0

```{r}
MR |> 
  mutate(abs = abs_t0) |> 
  qc_raw_abs(
    min_abs = 0.9, max_abs = 2.5, 
    plot_col_facet = "dataset", 
    show_plot = TRUE, export_plot = "abs_t0_distrib") 
abs_t0_distrib + labs(title = "Distribution of absorbance at t0, all data")
```

For now, it looks fine for both datasets for t0. Let's check t5

```{r}
MR |> 
  mutate(abs = abs_t5) |> 
  qc_raw_abs(
    min_abs = 0.1, max_abs = 2.5, 
    plot_col_facet = "dataset", 
    show_plot = TRUE,
    export_plot = "abs_t5_distrib") 
abs_t5_distrib + labs(title = "Distribution of absorbance at t5, all data")
```

They also look good, no outlier removal needed so far

## 3.2 - remove "empty" and NAs

With MicroResp experiments, edge effects are not rare, so we exclude any data from columns 1 and 12 of the 96-well plate. Those have been encoded in the mapping as "empty" (though they were not physically).

Also, the Excel template was not 100% full, which means that some plates are without absorbance data

```{r}
MR_clean <- MR |> 
  filter_out(map == "empty" | is.na(abs_t0)) |> 
  mutate(abs_t0 = as.numeric(abs_t0), abs_t5 = as.numeric(abs_t5))

MR_clean
```

# 4 - Normalisation

In preliminary works we noticed that `detection_plate_id` was the biggest driver of variability in our dataset. Let's visualize t0 absorbance readings (see plot)

```{r}
unnormed_plot <- MR_clean |> ggplot(aes(x = detection_plate_id, y = as.numeric(abs_t0))) +
  theme_minimal() +
  geom_boxplot() +
  facet_wrap(~dataset, scales = "free_x") +
  labs(title = "Plate-level normalisation")

unnormed_plot
```

So, instead of a "per-plate" normalization, we propose a "per-dataset" normalisation. I.e., instead of

- $normalized_{abs} = \frac{abs_{i,t5}}{abs_{i,t0}} \cdot mean(abs_{plate,t0})$

We get

- $normalized_{abs} = \frac {abs_{i,t5}}{abs_{i,t0}} \cdot mean(abs_{dataset,t0})$

## 4.1 - Compute per-dataset mean of t0

```{r}
avg_t0 <- MR_clean |> 
  # option: keep only "best" plate: M1 and
 # filter(detection_plate_id %in% c("M1", "M28")) |> 
  # remove first and last row for this (bc edge effects)
  filter_out(row %in% c("A", "H")) |> 
  group_by(dataset) |> 
  summarise(dataset_avg_t0 = mean(abs_t0))
```

## 4.2 - Normalize data

Using formula above. First, join data with computed average, then compute normalisation

```{r}
MR_norm <- MR_clean |> 
  left_join(avg_t0, by = join_by(dataset)) |> 
  relocate(dataset_avg_t0, .after = abs_t5) |> 
  mutate(
    norm = (abs_t5 / abs_t0) * dataset_avg_t0,
    .after = dataset_avg_t0
  )
```

Redo the same plot to confirm the improvement

```{r}
normed_plot <- MR_norm |> 
  ggplot(aes(x = detection_plate_id, y = as.numeric(norm))) +
  theme_minimal() +
  geom_boxplot() +
  facet_wrap(~dataset, scales = "free_x") +
  labs(title = "Dataset-level normalisation")

unnormed_plot / normed_plot
```

# 5 - Convert to %CO2

Those are the coefficients as given in the MicroResp manual. Pending homemade calibration.

%Co2 = -0.2265 + (-1.606)/(1+(-6.771)\*norm)

```{r}
MR_percent_co2 <- MR_norm |> 
  mutate(
    percent_co2 = 
      -0.2265 + 
      ((-1.606) / 
         (1 + 
            ((-6.771) * norm))
       ),
    .after = norm
    )
```

# 6 - Convert to g CO2 / g dry soil

[**!! TODO: this is taken from Excel document –\> check again in Manual that it is the right version**]{.underline}

co2_emitted = (((percent_co2/100)x850x(44/22.4)x*(12/44)x*(273/(273+25)))/(soil_per_well_gx(soil_dw/100)))/5

```{r}
MR_co2_g_h <- MR_percent_co2 |> 
  mutate(
    co2_g_h = (
      (
        (percent_co2 / 100) * 850 * (44/22.4) * (12/44) * (273 / (273+25))
        ) / 
        case_when(
          # for std soils
          column %in% c(2,3) ~ 
            (std_soil_well_g * (std_soil_dm_content_percent / 100)),
          column %in% c(4:11) ~ 
            (soil_well_g * (soil_dm_content_percent / 100)))
      ) / 5
  )

MR_co2_g_h
```

Check-out values

```{r}
MR_co2_g_h |> 
  ggplot(aes(x = co2_g_h)) + 
  theme_minimal() +
  #geom_histogram() +
  geom_density() +
  facet_grid(map~dataset, scales = "free_y")
```

::: callout-warning
## Calibration or correction needed !

Until we find a way to correct / calibrate this data, we are probably still going to have some issues with the negative values (we need to compute the log to compute the Shannon Index)
:::

# 7 - Per plate & per substrate outlier removal

Subset data.

!! A preliminary run of the following chunks (up to saving the plots) showed that while they are not the only outliers, wells from row A and H are outliers on many plates, possibly a majority of plates.

So I choose to follow the recommendation of our colleagues from the FiBL and remove those wells already in the subset now. If needed, just "comment" the filtering line to check out plots with all data again

```{r}

# MR_greenhouse_raw <- MR_co2_g_h |> 
#   filter_out(row %in% c("A", "H")) |> 
#   filter(dataset == "MR_greenhouse")
# 
# MR_field_raw <- MR_co2_g_h |> 
#   filter_out(row %in% c("A", "H")) |> 
#   filter(dataset == "MR_field")
```

## 7.1 - create QC plots

Create one list of QC plots per dataset. But that would be too many plots to assess –\> Take a subset with coefficient of variation at 15%?

```{r}
MR_avg_prelim <- MR_co2_g_h |> 
  filter_out(row %in% c("A", "H")) |> 
  group_by(dataset, plate_id, map) |> 
  summarize(
    substrate_avg = mean(co2_g_h),
    std_dev = sd(co2_g_h)
  ) |> 
  mutate(coef_var_percent = 100 * std_dev / substrate_avg) |> 
  arrange(desc(coef_var_percent))

MR_avg_prelim |> ggplot(aes(x = coef_var_percent)) +
  theme_minimal() +
  geom_histogram() +
  xlim(-100, 100) +
  facet_wrap(~dataset, ncol = 1)

suspicious_substrates <- MR_avg_prelim |> 
  filter(abs(coef_var_percent) > 15) |> 
  select(plate_id, map) |> unique()

suspicious_MR <- MR_co2_g_h |> 
  filter_out(row %in% c("A", "H")) |> 
  right_join(suspicious_substrates) 


# ## Repeat for field
# MR_field_avg <- MR_field_raw |> 
#   group_by(plate_id, map) |> 
#   summarize(
#     substrate_avg = mean(co2_g_h),
#     std_dev = sd(co2_g_h)
#   ) |> 
#   mutate(coef_var_percent = 100 * std_dev / substrate_avg) |> 
#   arrange(desc(coef_var_percent))
# 
# MR_field_avg |> ggplot(aes(x = coef_var_percent)) +
#   geom_histogram() +
#   xlim(-100, 100)
# 
# suspicious_substrates_field <- MR_field_avg |> 
#   filter(abs(coef_var_percent) > 15) |> 
#   select(plate_id, map) |> unique()
# 
# suspicious_MR_field <- MR_field_raw |> 
#   right_join(suspicious_substrates_field) 
```

Now, with a threshold of 15% as max coef of variation, we are left with only \~200 combinations of plate x substrate to look at (instead of \~5000).

[**!! This means we are ok with not looking at all data !!**]{.underline}

```{r}

source("functions/plot_qc_sample_conc.R")

# qc_plots_greenhouse <- plot_list_qc_MR(MR_greenhouse_raw)
# 
# qc_plots_field <- plot_list_qc_MR(MR_field_raw)

qc_plots_greenhouse <- plot_list_qc_MR(
  suspicious_MR |> filter(dataset == "MR_greenhouse"))

qc_plots_field <- plot_list_qc_MR(
  suspicious_MR |> filter(dataset == "MR_field"))

```

Save QC plots for review outside of this Markdown document

!! The plotting chunk is slow and is thus deactivated. Manually run it to save plots again.

For the sake of illustration, we just plot one of the plots (one per dataset per substrate): it facets the plots per run, and only displays data that is suspicious (coefficient of variation above threshold defined above)

```{r}
qc_plots_field$Glu 
```

Loop to save the plots

```{r}
#| eval: false
#| include: true

# For greenhouse
for (i in 1:length(qc_plots_greenhouse)) {
  
  plot <- qc_plots_greenhouse[i]
  
  filepath <- paste0(
      "output/figures/QC/greenhouse_MR/",
      names(plot), "_", today(), ".pdf")
  
  ggsave(
    filename = filepath, 
    plot = plot, width = 18, height = 10)
}

# For field
for (i in 1:length(qc_plots_field)) {
  
  plot <- qc_plots_field[i]
  
  filepath <- paste0(
      "output/figures/QC/field_MR/",
      names(plot), "_", today(), ".pdf")
  
  ggsave(
    filename = filepath, 
    plot = plot, width = 18, height = 10)
}
```

## 7.2 - identifying outliers + removal

Then, by visual assessment of the plots one by one, we create a table of wells to remove, and clean the dataset from obvious outliers

Starting with greenhouse outliers

```{r}
to_remove_greenhouse <- tribble(
  ~ plate_id, ~well_id, 
  # Alanine
  "P08", "C10",    "P19", "B10",    "P28", "C10",
  "P35", "F10",    "P43", "G10",    "P50", "B10",
  "P70", "C10",    "P96", "B10",
  # gABA
  "P01", "F9",    "P15", "G9",    "P31", "D9",
  "P35", "G9",    "P39", "E9",    "P43", "G9",
  "P51", "D9",    "P56", "F9",    "P66", "B9",
  "P67", "F9",    "P69", "D9",    "P70", "D9",
  "P84", "D9",    "P93", "D9",
  #Glu
  "P24", "B6",    "P43", "B6",    "P43", "G6",
  "P56", "G6",    "P81", "B6",    "P82", "D6",
  # H2O
  "P02", "B4",    "P06", "C4",    "P08", "B4",
  "P17", "B4",    "P18", "C4",    "P25", "G4",
  "P41", "B4",    "P44", "G4",    "P45", "G4",
  "P46", "B4",    "P51", "C4",    "P60", "C4",
  "P62", "G4",    "P85", "G4",    "P85", "F4",
  "P88", "F4",
  # Lgn
  "P08", "B7",    "P24", "B7",    "P35", "G7",
  "P43", "G7",    "P46", "G7",    "P52", "G7",
  "P70", "G7",    "P83", "B7",    "P88", "C7",
  # NAG
  "P12", "B8",    "P12", "G8",    "P16", "B8",
  "P28", "E8",    "P31", "B8",    "P35", "G8",
  "P52", "E8",    "P56", "B8",    "P67", "E8",
  "P70", "E8",
  # OA
  "P25", "C5",    "P46", "B5",    "P49", "C5",
  "P54", "C5",    "P64", "C5",    "P81", "B5",
  "P90", "G5",
  #Std-Glu
  "P08", "F2",    "P44", "B2",    "P46", "D2",
  "P66", "G2",    
  # Std-H2O
  "P01", "B3",    "P12", "D3",    "P16", "B3",
  "P24", "B3",    "P24", "G3",    "P25", "E3",
  "P29", "D3",    "P33", "D3",    "P44", "G3",
  "P62", "B3",    "P64", "F3",    "P86", "G3",
  "P91", "F3",    "P93", "E3",    
  # Urea
  "P30", "C11",    "P35", "F11",    "P35", "G11",
  "P51", "B11",    "P52", "B11",    "P56", "B11",
  "P78", "D11",    "P92", "E11"
  
) |> 
  mutate(dataset = "MR_greenhouse")

to_remove_greenhouse
```

Repeat it with Field outliers

```{r}
to_remove_field <- tribble(
  ~ plate_id, ~well_id, 
  #Ala
  "P18", "B10",    "P22", "E10",    "P26", "B10",
  "P27", "B10",    "P52", "G10",    "P66", "B10",
  "P66", "G10",    "P70", "B10",    "P72", "B10",
  # gABA
  "P10", "B9",    "P31", "C9",    "P34", "C9",
  "P57", "G9",    "P57", "F9",    "P62", "B9",
  "P62", "C9",    "P65", "G9",    "P66", "G9",
  "P74", "B9",
  # Glu
  "P29", "G6",    "P66", "G6",    "P75", "B6",
  "P76", "G6",    "P79", "G6",
  # H2O
  "P04", "D4",    "P37", "G4",    "P38", "D4",
  # Lgn
  "P50", "B7",    "P60", "B7",    "P62", "B7",
  "P75", "B7",    "P80", "C7",
  # NAG
  "P02", "B8",    "P28", "B8",    "P31", "B8",
  "P37", "B8",    "P42", "C8",    "P57", "E8",
  # OA
  "P12", "B5",    "P14", "C5",    "P14", "D5",
  "P17", "B5",    "P26", "B5",    "P31", "B5",
  "P34", "E5",    "P35", "E5",    "P52", "E5",
  "P62", "B5",    "P66", "G5",    "P73", "E5",
  # Std-Glu
  "P50", "G2",    "P61", "B2",    "P74", "B2",
  "P75", "G2",    "P79", "C2",
  # Std-H2O # NADA
  # Urea
  "P23", "F11",    "P33", "D11",    "P46", "C11",
  "P52", "F11",    "P71", "F11",    "P77", "F11"
) |> 
  mutate(dataset = "MR_field")

to_remove_field
```

Remove those outliers

```{r}
MR_wash1 <- MR_co2_g_h |> 
  remove_wells(to_remove_greenhouse) |> 
  remove_wells(to_remove_field) 

# 
# MR_greenhouse_wash1 <- MR_greenhouse_raw |> remove_wells(to_remove_greenhouse)
# 
# MR_field_wash1 <- MR_field_raw |> remove_wells(to_remove_field)
```

It is possible to run once more through the quality check before deciding to move on, and possibly create a "wash2", "wash3", etc. version, though I don't find it totally relevant

Store last version in a new variable (easier for edits...)

```{r}
MR_clean <- MR_wash1
#MR_field_clean <- MR_field_wash1
```

# 8 - Average per plate per substrate

## 8.1 - Compute sample\*substrate average

Now that outliers are removed, we can compute the average value per plate per substrate.

```{r}
#MR_clean <- bind_rows(MR_greenhouse_clean, MR_field_clean)

MR_avg <- MR_clean |> 
  group_by(dataset, plate_id, map) |> 
  summarize(
    co2_g_h_avg = mean(co2_g_h)) #|> 

```

Plot distribution of values

```{r}
distrib_uncorrected <- MR_avg |> 
  ggplot(aes(x = co2_g_h_avg)) +
  theme_minimal() +
  geom_histogram() +
  labs(
    title = "Distribution of per plate per substrate average\nof CO2 emission rate",
    x = "CO2 emission rate [gCO2 / g dry soil / hour]") + facet_wrap(~dataset)
distrib_uncorrected
```

We have values below zero that will cause problems in the computation of the Shannon index.

## 8.2 - 1st correction of averages - deal with negative values

We know that our regression equation is probably off somehow because the detection plates were too concentrated. I find that the computed values are acceptable (cf normalisation step), also because values for standard soils are quite stable, but still we will hit a mathematical problem.

When dealing with log transformation of data, it is quite frequent to use log(x + 1) instead of log(x) to account for the deformation of values that occur for very low x. On a logarithmic scale, a translation by 1 is quite small. Moreover, if we would simply transform our data through a linear translation, similar to adding 1, although absolute numbers would be off, correlations between variables would hold. So, to deal with our negative values, I propose a simple correction: adding to all values a constant value, i.e., the absolute value of the smallest negative value within the data set, rounded up a bit so that the smallest value becomes a positive number, and not just zero.

Let's compute it

```{r}
MR_avg_shifted <- MR_avg |> 
  ungroup() |> 
  mutate(
    min_value = min(co2_g_h_avg),
    co2_g_h_shifted = co2_g_h_avg - floor(min_value*1000)/1000,
    # add info whether Std soil or sample
    std_sample = case_when(
      map %in% c("Std_H2O", "Std_Glu") ~ "std_soil",
      .default = "sample"
    ))
```

Now, let's look at it graphically

```{r}
distrib_corrected <- MR_avg_shifted |> 
  ggplot(aes(x = co2_g_h_shifted)) +
  theme_minimal() +
  geom_histogram() +
  labs(
    title = "Distribution of per plate per substrate average\nof CO2 emission rate - Corrected values",
    subtitle = paste0("Correction : shifting by adding the rounded\nminimum value = ", floor(unique(MR_avg_shifted$min_value)*1000)/1000),
    x = "CO2 emission rate [µgCO2-C / g dry soil / hour]") + facet_wrap(~dataset)

distrib_uncorrected / distrib_corrected
#distrib_corrected + facet_wrap(dataset~map, scales = "free_x")
```

It should now be discussed whether this correction is acceptable or not. For example, the issue with current values is that, to compute Shanon index, we need to take the log of those values. by taking x/ln(x). But for other indices we could just take the raw values as is. I have a feeling that it is best to keep the same correction for all data, though I cannot argue it.

I suggest to accept this "translation" correction. I find that it is honest, at least for now, until a calibration is possible. I propose to keep that correction for all computations.

## 8.3 - 2nd correction of averages - remove basal respiration

[@bongiorno2020] write: *Thereafter, we corrected for the average respiration rate of soil to which the deionized water was added.*

Although not explicit, I believe that what they say if that they consider that substrate-induced respirations as computed so far are indeed the sum of basal respiration + the "real" part of the respiration that is substrate induced. So probably they do a subtraction?

So, now we correct SIR's for basal respiration (subtract H2O-resp from substrate-induced resp)

First, extract basal respiration

```{r}
# Extract basal resp data only
basal_resp <- MR_avg_shifted |> 
  filter(map %in% c("H2O", "Std_H2O")) |> 
                                  #*** If want unshifted data, change here !! *
  select(dataset, plate_id, std_sample, co2_g_h_shifted) |> 
  rename(basal_resp = co2_g_h_shifted)

# Check it out
basal_resp
```

Now we can join this info to the rest of the data and compute H2O-corrected data

```{r}
H2O_corrected <- MR_avg_shifted |> 
  left_join(basal_resp) |> 
  mutate(H2O_corrected = case_when(
    map %in% c("H2O", "Std_H2O") ~ basal_resp,
                                #*** If want unshifted data, change here !!  *
    .default = co2_g_h_shifted - basal_resp
  ))
```

In the following sections, I follow computations as in [@bongiorno2020a], i.e.:

- Compute MSIR, MBC, metabolic quotient

- Compute relative contributions of SIR's (SIR / MSIR), refered to as rSIR's

- Derive from this relative contribution an intermediary value needed for Shanon (x\*ln(x)), refered to as sSIR's

- Compute the Shannon index

Then, we can still discuss some N-based index

# 9 - MBC, basal respiration, metabolic quotient, MSIR

Now, we compute the MBC, qCO2 and MSIR (here, we use H2O-corrected data, and don't keep water in the computation. So it is only a sum of non-H2O substrates. Also, we don't keep std soil data from here on, so we start by storing it somewhere (it will be needed for QC)

[**!! TO DO: check where computation of MBC and qCO2 come from + units !!**]{.underline}

```{r}
# pivot to get one column per substrate
wider_data <- H2O_corrected |> 
  pivot_wider(
    names_from = map, 
    values_from = H2O_corrected, id_cols = c(dataset, plate_id))

# extract std_soil_data and store it in an object
std_soil_data <- wider_data |> select(dataset, plate_id, starts_with("Std"))

# compute first round of indicators
MSIR <- wider_data |> 
  # exclude std soil data
  select(!starts_with("Std")) |> 
  # compute Basal resp, MBC, metabolic quotient
  rename(basal_resp = "H2O") |> relocate(basal_resp, .after = everything()) |> 
  mutate(
    MBC = (Glu * 40.04) + 0.37, # not sure where those numbers come from!
    qCO2 = basal_resp * 1000 / MBC # same here
  ) |> 
  # compute MSIR
  rowwise() |> mutate(MSIR = sum(c_across(Ala:gABA)))
```

# 10 - Shannon index

Now, we re-pivot longer to compute on a per-substrate basis again, and directly compute relative SIRs (rSIR) and -x\*ln(x) intermediary value

```{r}
rSIRs <- MSIR |> 
  pivot_longer(cols = c(Ala:gABA), names_to = "substrate", values_to = "SIR") |> 
  mutate(
    rSIR = SIR / MSIR,
    sSIR = -rSIR * log(rSIR))

rSIRs
```

! There is still one warning: we are loosing one data point (possibly still a negative or null value given to the log?)

[**!! TODO : Deal with this later !!**]{.underline}

Now, we can pivot wider again to sum sSIRs on a per sample basis

```{r}
shannon <- rSIRs |> 
  pivot_wider(
    names_from = substrate, names_prefix = "sSIR_",
    values_from = sSIR,
    id_cols = c(dataset, plate_id, basal_resp, MBC, qCO2, MSIR)
  ) |> 
  rowwise() |> mutate(shannon = sum(c_across(sSIR_Ala:sSIR_gABA)))

# Check it out
shannon
```

Gather all data (except std soil data)

```{r}
MR_indices <- MR_clean |> 
  #take relevant categorical variables (no quantitative data)
  select(dataset, plate_id, run_id:short) |> 
  # reduce it to one row per plate
  unique() |> 
  # join it to the quantitative, plate-wise (=sample-wise) agregated quantitative data
  left_join(shannon)

MR_indices
```

We lost

- the data on raw SIR and relative SIR (rSIR), which is stored in rSIRs (just in case)

- and std soil data

```{r}
rSIRs
std_soil_data
```

# 11 - QC - check standard soils

I'm actually not sooo happy about this. FYI: here, I took H2O-corrected data for the glucose

```{r}
std_soil_longer <- std_soil_data |> 
  pivot_longer(
    values_to = "SIR", 
    names_to = "substrate", 
    cols = starts_with("Std")) |> 
  # add categorical data
  left_join(
    MR_clean |> 
      select(dataset, plate_id, run_id, detection_plate_id) |> 
      unique())

std_soil_longer |> 
  ggplot(aes(x = run_id, y = SIR)) +
  theme_minimal() + 
  geom_boxplot() +
  facet_wrap(substrate~dataset, scales = "free")
```

# 12 - Last data wrangling

For our Field experiment, some modalities are to be removed: Bloc 1 is not a real bloc, and the faba bean did not take in the Ref treatment, so I remove it from MR data

[**!! TO DO: modalities not yet removed from rSIRs, get to it if relevant**]{.underline}

```{r}
MR_export_ready <- MR_indices |> 
  filter_out(expe == "Field" & bloc == "B1") |> 
  filter_out(expe == "Field" & soil == "Ref" & cs == "IC")
```

- Get rSIRs export ready ?

# 11 - Export

```{r}
MR_export_ready |> write_rds("output/data/1_MR_data.rds")
```
