---
title: "MicroResp Data, from absorbance to indices"
author: "Morgane de Toeuf"
format: 
  gfm:
    toc: true
    #code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
bibliography: references.bib
---

# TO DO

# Set up

```{r}
#| include: false
rm(list = ls())

library(plate2N)
library(tidyverse)
library(roperators)

# Functions
source("functions/MR_to_tibble.R")
source("functions/plot_qc_sample_conc.R")

```

# 1 - Import raw data

## 1.1 - Import raw absorbance data

First, import absorbance data (t0 = pre-incubation, t5 = after 5h of incubation). Plate ids correspond to sample names (equivalence table stored somewhere else)

::: callout-tip
## Homemade function

I use here a function that I did not put in plate2N as I found that it will not be useful longterm. To access it, go in the folder "functions" and get the file called "MR_to_tibble.R". It is just a variant on read_csv() to avoid copy-pasting. It is adapted to (my version of) the Excel template that we received from the FiBL.

To import raw data from the plate-reader, use `plate2N::txt_to_tibble()` or `plate2N::skanit_to_tibble()`.
:::

```{r}

MR_greenhouse_abs_t0 <- MR_to_tibble(
  filepath = "../raw_data/MR_Pot_abs.csv", 
  column_range = c(1:13))

MR_greenhouse_abs_t5 <- MR_to_tibble(
  filepath = "../raw_data/MR_Pot_abs.csv", 
  column_range = c(17:29))

MR_field_abs_t0 <- MR_to_tibble(
  filepath = "../raw_data/MR_Field_abs.csv", 
  column_range = c(1:13))

MR_field_abs_t5 <- MR_to_tibble(
  filepath = "../raw_data/MR_Field_abs.csv", 
  column_range = c(16:28))
```

Check out just one of them for illustration

```{r}
MR_greenhouse_abs_t0
```

## 1.2 - prepare the mapping

Plate mapping is strictly the same for each plate throughout the experiment, so I propose to take advantage of the function `plate2N::map_plates()` to create this mapping data in R.

First, create a vector with the names of the substrates (in the order of the plate map)

```{r}
# vector of substrates
MR_columns <- c(
  "Std_Glu", "Std_H2O", 
  "H2O", "OA", "Glu", "Lgn", "NAG", "gABA", "Ala", "Urea")
```

Then, compute the number of plates required per dataset (field vs greenhouse)

```{r}
# nb of plates
nb_plates_greenhouse <- MR_greenhouse_abs_t0 |> 
  filter(row %ni% LETTERS) |> 
  nrow()

nb_plates_field <- MR_field_abs_t0 |> 
  filter(row %ni% LETTERS) |> 
  nrow() 
```

Finally, run the mapping function with

- plate_id that pastes "P" with numbers from 01 to the nb of plates

- "sample list" = repetition of the vector of substrates x the nb of plates

- no std curves or blank, and 2 empty columns (1 and 12)

- 8 wells per "sample" (in this case: substrates)

```{r}
# compute the mapping
MR_greenhouse_map <- map_plates(
  plate_ids = paste0("P", sprintf("%02d", seq(01:nb_plates_greenhouse))), 
  samples = rep(MR_columns, nb_plates_greenhouse),
  n_samples_per_plate = 10,
  column_curves = c(), column_blank = c(), column_empty = c(1,12),
  n_wells_samples = 8)

MR_field_map <- map_plates(
  plate_ids = paste0("P", sprintf("%02d", seq(01:nb_plates_field))), 
  samples = rep(MR_columns, nb_plates_field),
  n_samples_per_plate = 10,
  column_curves = c(), column_blank = c(), column_empty = c(1,12),
  n_wells_samples = 8)
```

Check one out

```{r}
MR_greenhouse_map
```

## 1.3 - Import plate metadata

First import: this is a custom import to fit my data

```{r}
MR_greenhouse_metadata <- read_csv(
  "../raw_data/MR_Pot_metadata.csv", show_col_types = FALSE) |> 
  as_tibble() |> 
  select(!c(1,2)) |> 
  janitor::clean_names() |>
  drop_na(plate_id)

MR_field_metadata <- read_csv(
  "../raw_data/MR_Field_metadata.csv", show_col_types = FALSE) |> 
  as_tibble() |> 
  select(!c(1,2)) |>
  janitor::clean_names() |>
  drop_na(plate_id) 
```

Then, join greenhouse and field metadata into 1 object and add equivalence between dataset and expe (needed later)

```{r}
all_MR_metadata <- bind_rows(MR_greenhouse_metadata, MR_field_metadata) |> 
  # add column dataset into metadat to use for left_join() later
  mutate(dataset = case_when(
    expe == "Pot" ~ "MR_greenhouse",
    expe == "Field" ~ "MR_field",
    .default = "oups"),
    .before = expe)
```

Check it out

```{r}
all_MR_metadata
```

# 2 - From plate layout to tidy data

## 2.1 - Verticalise all plate data

Verticalise, 1 column per plate, 96 rows (1 per well position)

```{r}
MR_greenhouse_vertical <- join_abs_map(
  tibble_list = list( MR_greenhouse_abs_t0, MR_greenhouse_abs_t5, MR_greenhouse_map), 
  abs_map = c("abs_t0-", "abs_t5-", "map-"), 
  coerce_numeric = FALSE, 
  dataset = "MR_greenhouse-" )

MR_field_vertical <- join_abs_map(
  tibble_list = list( MR_field_abs_t0, MR_field_abs_t5, MR_field_map), 
  abs_map = c("abs_t0-", "abs_t5-", "map-"), 
  coerce_numeric = FALSE, 
  dataset = "MR_field-" )
```

Check one out

```{r}
MR_greenhouse_vertical
```

Join field and greenhouse data in one

```{r}
# Join them in one
all_vertical_MR <- MR_greenhouse_vertical |> 
  left_join(MR_field_vertical, by = join_by(row, column))
```

Check out the structure of column names, with `dataset-data_type-plate_id`.

```{r}
# Check out the structure of the column names (one per plate)
sample(names(all_vertical_MR),size = 30)
```

## 2.2 - Elongate data into tidy format

Tidy the table –\> 1 row per unique well (96 x nb of physical plates), and 3 important columns: `abs_t0`, `abs_t5`, `map`.

```{r}
all_raw_abs_tidy_MR <- vertical_to_tidy(
  all_vertical_MR, 
  column_def = c("abs_t0", "abs_t5", "map"))
```

Check it out

```{r}
all_raw_abs_tidy_MR
```

```{r}
#| eval: false
#| include: false
# Optional: export the data
# all_raw_abs_tidy_MR |> write_rds("output/data/1_all_raw_abs_tidy_MR.rds")
# all_MR_metadata |> write_rds("output/data/1_all_MR_metadata.rds")

```

## 2.3 - Join data with metadata

```{r}
# Optional: load data
# all_raw_abs_tidy_MR <- read_rds("output/data/1_all_raw_abs_tidy_MR.rds")
# all_MR_metadata <- read_rds("output/data/1_all_MR_metadata.rds") 

```

Join data with metadata

```{r}
#| output: false
MR <- all_raw_abs_tidy_MR |> left_join(all_MR_metadata) 
```

Check it out

```{r}
MR
```

# 3 - clean data

## 3.1 - QC - suspicious wells

`plate2N::qc_raw_abs()` checks whether absorbance data is within a range or not. Check the documentation on this function for more details. Here, I stop the output from the next chunk to appear, to avoid cluttering the document.

Check first t0

```{r}
#| output: false
MR |> 
  mutate(abs = abs_t0) |> 
  qc_raw_abs(
    min_abs = 0.9, max_abs = 2.5, 
    plot_col_facet = "dataset", 
    show_plot = TRUE, 
    export_plot = "abs_t0_distrib") 
```

Check out the outputted plot

```{r}
abs_t0_distrib + labs(title = "Distribution of absorbance at t0, all data")
```

For now, it looks fine for both datasets for t0. Let's check t5

```{r}
#| output: false
MR |> 
  mutate(abs = abs_t5) |> 
  qc_raw_abs(
    min_abs = 0.1, max_abs = 2.5, 
    plot_col_facet = "dataset", 
    show_plot = TRUE,
    export_plot = "abs_t5_distrib") 
```

Check the plot

```{r}
abs_t5_distrib + labs(title = "Distribution of absorbance at t5, all data")
```

They also look good, no outlier removal needed so far

## 3.2 - remove "empty" and NAs

With MicroResp experiments, edge effects are not rare, so we exclude any data from columns 1 and 12 of the 96-well plate. Those have been encoded in the mapping as "empty" (though they were not physically).

Also, the Excel template was not 100% full, which means that some plates are without absorbance data.

I also change data format so that all absorbance data becomes numeric (until now stored as character)

```{r}
MR_clean <- MR |> 
  filter_out(map == "empty" | is.na(abs_t0)) |> 
  mutate(abs_t0 = as.numeric(abs_t0), abs_t5 = as.numeric(abs_t5))

MR_clean
```

# 4 - Normalisation

In preliminary works we noticed that `detection_plate_id` was the biggest driver of variability in our dataset. This is due to some lab issues and should not be a recurring problem. Let's visualize t0 absorbance readings for standard soils and glucose, supposedly the same (see plot)

```{r}
unnormed_plot <- MR_clean |> 
  filter(map == "Std_Glu") |> 
  ggplot(aes(x = detection_plate_id, y = as.numeric(abs_t5))) +
  theme_minimal() +
  geom_boxplot() +
  facet_wrap(~dataset, scales = "free_x") +
  labs(title = "Raw data at t5, no normalisation") +
  ylim(0.3,2.1)

unnormed_plot
```

So, instead of a "per-plate" normalization, we propose a "per-dataset" normalisation. I.e., instead of

- $normalized_{abs} = \frac{abs_{i,t5}}{abs_{i,t0}} \cdot mean(abs_{plate,t0})$

We get

- $normalized_{abs} = \frac {abs_{i,t5}}{abs_{i,t0}} \cdot mean(abs_{dataset,t0})$

## 4.1 - Compute mean of t0

Per dataset (here, not only Glu of std soil, but all data)

```{r}
avg_t0_dataset <- MR_clean |> 
  # remove first and last row for this (bc edge effects)
  filter_out(row %in% c("A", "H")) |> 
  group_by(dataset) |> 
  summarise(dataset_avg_t0 = mean(abs_t0))
```

For comparison, original version, per plate

```{r}
avg_t0_plate <- MR_clean |> 
  # remove first and last row for this (bc edge effects)
  filter_out(row %in% c("A", "H")) |> 
  group_by(dataset, plate_id) |> 
  summarise(plate_avg_t0 = mean(abs_t0))
```

## 4.2 - Normalise data

### 4.2.1 - Compute normalisation (2 options)

We compute both normalizations, using both formulae above.

First, join data with computed average, then compute normalisation (on all data, not just Glucose of std soil)

```{r}
# our new norm version
MR_norm_dataset <- MR_clean |> 
  left_join(avg_t0_dataset, by = join_by(dataset)) |> 
  relocate(dataset_avg_t0, .after = abs_t5) |> 
  mutate(
    norm = (abs_t5 / abs_t0) * dataset_avg_t0,
    .after = dataset_avg_t0
  )

# original norm version
MR_norm_plate <- MR_clean |> 
  left_join(avg_t0_plate, by = join_by(dataset, plate_id)) |> 
  relocate(plate_avg_t0, .after = abs_t5) |> 
  mutate(
    norm = (abs_t5 / abs_t0) * plate_avg_t0,
    .after = plate_avg_t0
  )
```

### 4.2.2 - Choice of normalisation - graphical

Redo the same plots to confirm the improvement, and look at plots with original vs new normalisation

```{r}
dataset_normed_plot <- MR_norm_dataset |> 
  filter(map == "Std_Glu") |> 
  ggplot(aes(x = detection_plate_id, y = as.numeric(norm))) +
  theme_minimal() +
  geom_boxplot() +
  facet_wrap(~dataset, scales = "free_x") +
  labs(title = "Dataset-level normalisation") +
  ylim(0.3,2.1)

plate_normed_plot <- MR_norm_plate |> 
  filter(map == "Std_Glu") |> 
  ggplot(aes(x = detection_plate_id, y = as.numeric(norm))) +
  theme_minimal() +
  geom_boxplot() +
  facet_wrap(~dataset, scales = "free_x") +
  labs(title = "Plate-level normalisation") +
  ylim(0.3,2.1)

unnormed_plot / plate_normed_plot / dataset_normed_plot
```

Each step, from raw to plate-level to dataset-level is an improvement on the data. This confirms the proposition, and we keep the dataset-level normalised data for downstream steps

```{r}
MR_norm <- MR_norm_dataset
```

# 5 - Convert to %CO2

Those are the coefficients as given in the MicroResp manual. Pending homemade calibration.

%Co2 = -0.2265 + (-1.606)/(1+(-6.771)\*norm)

```{r}
MR_percent_co2 <- MR_norm |> 
  mutate(
    percent_co2 = 
      -0.2265 + 
      ((-1.606) / 
         (1 + 
            ((-6.771) * norm))
       ),
    .after = norm
    )
```

# 6 - Convert to µg CO2-C / g dry soil / h

[**!! TODO: this is taken from Excel document –\> check again in Manual that it is the right version**]{.underline}

co2_emitted = (((percent_co2/100)x850x(44/22.4)x*(12/44)x*(273/(273+25)))/(soil_per_well_gx(soil_dw/100)))/5

```{r}
MR_co2_g_h <- MR_percent_co2 |> 
  mutate(
    co2_g_h = (
      (
        (percent_co2 / 100) * 850 * (44/22.4) * (12/44) * (273 / (273+25))
        ) / 
        case_when(
          # for std soils
          column %in% c(2,3) ~ 
            (std_soil_well_g * (std_soil_dm_content_percent / 100)),
          column %in% c(4:11) ~ 
            (soil_well_g * (soil_dm_content_percent / 100)))
      ) / 5
  )

MR_co2_g_h
```

Check-out value distribution

```{r}
MR_co2_g_h |> 
  ggplot(aes(x = co2_g_h)) + 
  theme_minimal() +
  ggridges::geom_density_ridges(aes(y = map)) +
  facet_wrap(~dataset)
```

::: callout-warning
## Calibration or correction needed !

Until we find a way to correct / calibrate this data, we are probably still going to have some issues with the negative values (we need to compute the log to compute the Shannon Index)
:::

# 7 - Per plate & per substrate outlier removal

## 7.1 - Compute preliminary average (pre-outlier removal)

!! A preliminary run of the following chunks (up to saving the plots) showed that while they are not the only outliers, wells from row A and H are outliers on many plates, possibly a majority of plates. So I choose to follow the recommendation of our colleagues from the FiBL and remove those wells already in the subset hereunder, before computation of the average. If needed, just "comment" the filtering line to check out plots with all data again.

We then compute coefficient of variation which will serve as a threshold to identify potential outliers

```{r}
MR_avg_prelim <- MR_co2_g_h |> 
  filter_out(row %in% c("A", "H")) |> 
  group_by(dataset, plate_id, map) |> 
  summarize(
    substrate_avg = mean(co2_g_h),
    std_dev = sd(co2_g_h)
  ) |> 
  mutate(coef_var_percent = 100 * std_dev / substrate_avg) |> 
  arrange(desc(coef_var_percent))
```

Check out the distribution of the coefficient of variation

```{r}
MR_avg_prelim |> ggplot(aes(x = coef_var_percent)) +
  theme_minimal() +
  geom_histogram() +
  xlim(-100, 100) +
  facet_wrap(~dataset, ncol = 1)
```

We have quite high values, which is expected with this experiment: there are many sources of noise

I propose a threshold of 15%, i.e., data where the coefficient of variation is under 15% will not be checked (would be too much!)

```{r}
suspicious_substrates <- MR_avg_prelim |> 
  filter(abs(coef_var_percent) > 15) |> 
  select(plate_id, map) |> unique()

suspicious_MR <- MR_co2_g_h |> 
  filter_out(row %in% c("A", "H")) |> 
  right_join(suspicious_substrates) 
```

Now, with a threshold of 15% as max coef of variation, we are left with only \~200 combinations of plate x substrate to look at (instead of \~5000).

[**!! This means we are ok with not looking at all data !!**]{.underline}

## 7.2 - create QC plots

Create one list of QC plots per dataset.

```{r}
qc_plots_greenhouse <- plot_list_qc_MR(
  suspicious_MR |> filter(dataset == "MR_greenhouse"))

qc_plots_field <- plot_list_qc_MR(
  suspicious_MR |> filter(dataset == "MR_field"))
```

Save QC plots for review outside of this Markdown document

!! The plotting chunk is slow and is thus deactivated. Manually run it to save plots again.

For the sake of illustration, we just plot one of the plots (one per dataset per substrate): it facets the plots per run, and only displays data that is suspicious (coefficient of variation above threshold defined above)

```{r}
qc_plots_field$Glu 
```

Loop to save the plots on disk

```{r}
#| eval: false
#| include: true

# For greenhouse
for (i in 1:length(qc_plots_greenhouse)) {
  
  plot <- qc_plots_greenhouse[i]
  
  filepath <- paste0(
      "output/figures/QC/greenhouse_MR/",
      names(plot), "_", today(), ".pdf")
  
  ggsave(
    filename = filepath, 
    plot = plot, width = 18, height = 10)
}

# For field
for (i in 1:length(qc_plots_field)) {
  
  plot <- qc_plots_field[i]
  
  filepath <- paste0(
      "output/figures/QC/field_MR/",
      names(plot), "_", today(), ".pdf")
  
  ggsave(
    filename = filepath, 
    plot = plot, width = 18, height = 10)
}
```

## 7.3 - identifying outliers + removal

Then, by visual assessment of the plots one by one, we create a table of wells to remove, and clean the dataset from obvious outliers

Starting with greenhouse outliers

```{r}
to_remove_greenhouse <- tribble(
  ~ plate_id, ~well_id, 
  # Alanine
  "P08", "C10",    "P19", "B10",    "P28", "C10",
  "P35", "F10",    "P43", "G10",    "P50", "B10",
  "P70", "C10",    "P96", "B10",
  # gABA
  "P01", "F9",    "P15", "G9",    "P31", "D9",
  "P35", "G9",    "P39", "E9",    "P43", "G9",
  "P51", "D9",    "P56", "F9",    "P66", "B9",
  "P67", "F9",    "P69", "D9",    "P70", "D9",
  "P84", "D9",    "P93", "D9",
  #Glu
  "P24", "B6",    "P43", "B6",    "P43", "G6",
  "P56", "G6",    "P81", "B6",    "P82", "D6",
  # H2O
  "P02", "B4",    "P06", "C4",    "P08", "B4",
  "P17", "B4",    "P18", "C4",    "P25", "G4",
  "P41", "B4",    "P44", "G4",    "P45", "G4",
  "P46", "B4",    "P51", "C4",    "P60", "C4",
  "P62", "G4",    "P85", "G4",    "P85", "F4",
  "P88", "F4",
  # Lgn
  "P08", "B7",    "P24", "B7",    "P35", "G7",
  "P43", "G7",    "P46", "G7",    "P52", "G7",
  "P70", "G7",    "P83", "B7",    "P88", "C7",
  # NAG
  "P12", "B8",    "P12", "G8",    "P16", "B8",
  "P28", "E8",    "P31", "B8",    "P35", "G8",
  "P52", "E8",    "P56", "B8",    "P67", "E8",
  "P70", "E8",
  # OA
  "P25", "C5",    "P46", "B5",    "P49", "C5",
  "P54", "C5",    "P64", "C5",    "P81", "B5",
  "P90", "G5",
  #Std-Glu
  "P08", "F2",    "P44", "B2",    "P46", "D2",
  "P66", "G2",    
  # Std-H2O
  "P01", "B3",    "P12", "D3",    "P16", "B3",
  "P24", "B3",    "P24", "G3",    "P25", "E3",
  "P29", "D3",    "P33", "D3",    "P44", "G3",
  "P62", "B3",    "P64", "F3",    "P86", "G3",
  "P91", "F3",    "P93", "E3",    
  # Urea
  "P30", "C11",    "P35", "F11",    "P35", "G11",
  "P51", "B11",    "P52", "B11",    "P56", "B11",
  "P78", "D11",    "P92", "E11"
  
) |> 
  mutate(dataset = "MR_greenhouse")

to_remove_greenhouse
```

Repeat it with Field outliers

```{r}
to_remove_field <- tribble(
  ~ plate_id, ~well_id, 
  #Ala
  "P18", "B10",    "P22", "E10",    "P26", "B10",
  "P27", "B10",    "P52", "G10",    "P66", "B10",
  "P66", "G10",    "P70", "B10",    "P72", "B10",
  # gABA
  "P10", "B9",    "P31", "C9",    "P34", "C9",
  "P57", "G9",    "P57", "F9",    "P62", "B9",
  "P62", "C9",    "P65", "G9",    "P66", "G9",
  "P74", "B9",
  # Glu
  "P29", "G6",    "P66", "G6",    "P75", "B6",
  "P76", "G6",    "P79", "G6",
  # H2O
  "P04", "D4",    "P37", "G4",    "P38", "D4",
  # Lgn
  "P50", "B7",    "P60", "B7",    "P62", "B7",
  "P75", "B7",    "P80", "C7",
  # NAG
  "P02", "B8",    "P28", "B8",    "P31", "B8",
  "P37", "B8",    "P42", "C8",    "P57", "E8",
  # OA
  "P12", "B5",    "P14", "C5",    "P14", "D5",
  "P17", "B5",    "P26", "B5",    "P31", "B5",
  "P34", "E5",    "P35", "E5",    "P52", "E5",
  "P62", "B5",    "P66", "G5",    "P73", "E5",
  # Std-Glu
  "P50", "G2",    "P61", "B2",    "P74", "B2",
  "P75", "G2",    "P79", "C2",
  # Std-H2O # NADA
  # Urea
  "P23", "F11",    "P33", "D11",    "P46", "C11",
  "P52", "F11",    "P71", "F11",    "P77", "F11"
) |> 
  mutate(dataset = "MR_field")

to_remove_field
```

Remove those outliers

```{r}
MR_wash1 <- MR_co2_g_h |> 
  remove_wells(to_remove_greenhouse) |> 
  remove_wells(to_remove_field) 
```

It is possible to run once more through the quality check before deciding to move on, and possibly create a "wash2", "wash3", etc. version, though I don't find it totally relevant

Store last version in a new variable (easier for edits...)

```{r}
MR_clean <- MR_wash1
```

# 8 - Average per plate per substrate

## 8.1 - Compute sample\*substrate average

Now that outliers are removed, we can compute the average value per plate per substrate.

```{r}
#| output: false
MR_avg <- MR_clean |> 
  group_by(dataset, plate_id, map) |> 
  summarize(
    co2_g_h_avg = mean(co2_g_h)) #|> 

```

Plot distribution of values

```{r}
distrib_uncorrected <- MR_avg |> 
  ggplot(aes(x = co2_g_h_avg)) +
  theme_minimal() +
  geom_histogram() +
  labs(
    title = "Distribution of per plate per substrate average\nof CO2 emission rate",
    x = "CO2 emission rate [gCO2 / g dry soil / hour]") + facet_wrap(~dataset)

distrib_uncorrected
```

We have values below zero that will cause problems in the computation of the Shannon index.

## 8.2 - 1st correction of averages - deal with negative values

We know that our regression equation is probably off somehow because the detection plates were too concentrated. I find that the computed values are acceptable (cf normalisation step), also because values for standard soils are quite stable, but still we will hit a mathematical problem.

When dealing with log transformation of data, it is quite frequent to use log(x + 1) instead of log(x) to account for the deformation of values that occur for very low x. On a logarithmic scale, a translation by 1 is quite small. Moreover, if we would simply transform our data through a linear translation, similar to adding 1, although absolute numbers would be off, correlations between variables would hold. So, to deal with our negative values, I propose a simple correction: adding to all values a constant value, i.e., the absolute value of the smallest negative value within the data set, rounded up a bit so that the smallest value becomes a positive number, and not just zero.

Let's compute it

```{r}
MR_avg_shifted <- MR_avg |> 
  ungroup() |> 
  mutate(
    min_value = min(co2_g_h_avg),
    co2_g_h_shifted = co2_g_h_avg - floor(min_value*1000)/1000,
    # add info whether Std soil or sample
    std_sample = case_when(
      map %in% c("Std_H2O", "Std_Glu") ~ "std_soil",
      .default = "sample"
    ))
```

Now, let's look at it graphically

```{r}
distrib_corrected <- MR_avg_shifted |> 
  ggplot(aes(x = co2_g_h_shifted)) +
  theme_minimal() +
  geom_histogram() +
  labs(
    title = "Distribution of per plate per substrate average\nof CO2 emission rate - Corrected values",
    subtitle = paste0("Correction : shifting by adding the rounded\nminimum value = ", floor(unique(MR_avg_shifted$min_value)*1000)/1000),
    x = "CO2 emission rate [µgCO2-C / g dry soil / hour]") + facet_wrap(~dataset)

distrib_uncorrected / distrib_corrected
```

It should now be discussed whether this correction is acceptable or not. For example, the issue with current values is that, to compute Shanon index, we need to take the log of those values. by taking x/ln(x). But for other indices we could just take the raw values as is. I have a feeling that it is best to keep the same correction for all data, though I cannot argue it.

I suggest to accept this "translation" correction. I find that it is honest, at least for now, until a calibration is possible. I propose to keep that correction for all computations.

## 8.3 - 2nd correction of averages - remove basal respiration

[@bongiorno2020] write: *Thereafter, we corrected for the average respiration rate of soil to which the deionized water was added.*

Although not explicit, I believe that what they say if that they consider that substrate-induced respirations as computed so far are indeed the sum of basal respiration + the "real" part of the respiration that is substrate induced. So probably they do a subtraction?

So, now we correct SIR's for basal respiration (subtract H2O-resp from substrate-induced resp)

First, extract basal respiration

```{r}
# Extract basal resp data only
basal_resp <- MR_avg_shifted |> 
  filter(map %in% c("H2O", "Std_H2O")) |> 
                                  #*** If want unshifted data, change here !! *
  select(dataset, plate_id, std_sample, co2_g_h_shifted) |> 
  rename(basal_resp = co2_g_h_shifted)

# Check it out
basal_resp
```

Now we can join this info to the rest of the data and compute H2O-corrected data

```{r}
#| output: false
H2O_corrected <- MR_avg_shifted |> 
  left_join(basal_resp) |> 
  mutate(H2O_corrected = case_when(
    map %in% c("H2O", "Std_H2O") ~ basal_resp,
                                #*** If want unshifted data, change here !!  *
    .default = co2_g_h_shifted - basal_resp
  ))
```

Check it out

```{r}
H2O_corrected
```

In the following sections, I follow computations as in [@bongiorno2020a], i.e.:

- Compute MSIR, MBC, metabolic quotient

- Compute relative contributions of SIR's (SIR / MSIR), refered to as rSIR's

- Derive from this relative contribution an intermediary value needed for Shanon (x\*ln(x)), refered to as sSIR's

- Compute the Shannon index

Then, we can still discuss some N-based index

# 9 - MBC, basal respiration, metabolic quotient, MSIR

Now, we compute the MBC, qCO2 and MSIR (here, we use H2O-corrected data, and don't keep water in the computation. So it is only a sum of non-H2O substrates. Also, we don't keep std soil data from here on, so we start by storing it somewhere (it will be needed for QC)

[**!! TO DO: check where computation of MBC and qCO2 come from + units !!**]{.underline}

```{r}
# pivot to get one column per substrate
wider_data <- H2O_corrected |> 
  pivot_wider(
    names_from = map, 
    values_from = H2O_corrected, id_cols = c(dataset, plate_id))

# extract std_soil_data and store it in an object
std_soil_data <- wider_data |> select(dataset, plate_id, starts_with("Std"))

# compute first round of indicators
MSIR <- wider_data |> 
  # exclude std soil data
  select(!starts_with("Std")) |> 
  # compute Basal resp, MBC, metabolic quotient
  rename(basal_resp = "H2O") |> relocate(basal_resp, .after = everything()) |> 
  mutate(
    MBC = (Glu * 40.04) + 0.37, # not sure where those numbers come from!
    qCO2 = basal_resp * 1000 / MBC # same here
  ) |> 
  # compute MSIR
  rowwise() |> mutate(MSIR = sum(c_across(Ala:gABA)))
```

# 10 - Shannon index

Now, we re-pivot longer to compute on a per-substrate basis again, and directly compute relative SIRs (rSIR) and -x\*ln(x) intermediary value

```{r}
rSIRs <- MSIR |> 
  pivot_longer(cols = c(Ala:gABA), names_to = "substrate", values_to = "SIR") |> 
  mutate(
    rSIR = SIR / MSIR,
    sSIR = -rSIR * log(rSIR))
```

! There is still one warning: we are loosing one data point (possibly still a negative or null value given to the log?)

[**!! TODO : Deal with this later !!**]{.underline}

Still, check out the output

```{r}
rSIRs
```

Now, we can pivot wider again to sum sSIRs on a per sample basis

```{r}
shannon <- rSIRs |> 
  pivot_wider(
    names_from = substrate, names_prefix = "sSIR_",
    values_from = sSIR,
    id_cols = c(dataset, plate_id, basal_resp, MBC, qCO2, MSIR)
  ) |> 
  rowwise() |> mutate(shannon = sum(c_across(sSIR_Ala:sSIR_gABA)))

# Check it out
shannon
```

Gather all data (except std soil data)

```{r}
#| output: false
MR_indices <- MR_clean |> 
  #take relevant categorical variables (no quantitative data)
  select(dataset, plate_id, run_id:short) |> 
  # reduce it to one row per plate
  unique() |> 
  # join it to the quantitative, plate-wise (=sample-wise) agregated quantitative data
  left_join(shannon)
```

Check it out

```{r}
MR_indices
```

We lost

- the data on raw SIR and relative SIR (rSIR), which is stored in rSIRs (just in case)

- and std soil data

```{r}
rSIRs
std_soil_data
```

# 11 - QC - check standard soils

I'm actually not sooo happy about this. FYI: here, I took H2O-corrected data for the glucose

```{r}
# prepare data for the graph
std_soil_longer <- std_soil_data |> 
  pivot_longer(
    values_to = "SIR", 
    names_to = "substrate", 
    cols = starts_with("Std")) |> 
  # add categorical data
  left_join(
    MR_clean |> 
      select(dataset, plate_id, run_id, detection_plate_id) |> 
      unique(),
    by = join_by(dataset, plate_id))

# plot it
std_soil_longer |> 
  ggplot(aes(x = run_id, y = SIR)) +
  theme_minimal() + 
  geom_boxplot() +
  facet_wrap(substrate~dataset, scales = "free")
```

# 12 - Last data wrangling

For our Field experiment, some modalities are to be removed: Bloc 1 is not a real bloc, and the faba bean did not take in the Ref treatment, so I remove it from MR data

[**!! TO DO: modalities not yet removed from rSIRs, get to it if relevant**]{.underline}

```{r}
MR_export_ready <- MR_indices |> 
  filter_out(expe == "Field" & bloc == "B1") |> 
  filter_out(expe == "Field" & soil == "Ref" & cs == "IC")
```

- Get rSIRs export ready ?

# 11 - Export

```{r}
MR_export_ready |> write_rds("output/data/1_MR_data.rds")
```
