---
title: "Import and tidy Absorbance Data with plate2N"
author: "Morgane de Toeuf"
format: 
  gfm:
    toc: true
    #code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

To Do

- Add PNR data as well (Cloé)

# Set up

```{r}
#| message: false
#| warning: false
#| output: false

rm(list = ls())

# once in a while, redownload
#pak::pak("mdetoeuf/plate2N”) # if doesn't work --> try pak::pak_cleanup()
library(plate2N)
library(tidyverse)
```

# 1 - Import data

All object names ending with "abs" contain absorbance data, all object names ending with "map" contain the mapping of the equivalent abs data (sample id, or one of three options: "empty", "Std" or "extr" for empty wells, standard curve and extractant (K2SO4) respectively)

Here we import all our datasets, with various functions from the package `plate2N`. The output is displayed each time a new function is used.

## 1.1 - Nmin, TDN

```{r}
# Nmin for t1 and t2
(Nmin_t1t2_abs <- txt_to_tibble("raw_data/Nmin/"))
(Nmin_t1t2_map <- csv_to_tibble("raw_data/Nmin/Nmin_maps.csv"))

# Nmin for t3
Nmin_t3_abs <- csv_to_tibble("raw_data/Nmin_t3/Nmint3_data.csv")
Nmin_t3_map <- csv_to_tibble("raw_data/Nmin_t3/Nmint3_maps.csv")

# TDN (map csv not in the right format --> back to import_csv = manual import)
TDN_abs <- csv_to_tibble("raw_data/TDN/TDN_data.csv")
(TDN_map <- 
    read_csv(
      "raw_data/TDN/TDN_maps.csv", 
      col_names = FALSE, col_select = X14:X26, show_col_types = FALSE) |> 
    na.omit() |> 
    rename_with(~c("row", paste0("X", seq(1:12)))))

# Example for Skanit-formatted csv
(Skanit_abs <- skanit_to_tibble("raw_data/Skanit_example/MR_R1_t0.csv")$abs)
(Skanit_map <- skanit_to_tibble("raw_data/Skanit_example/MR_R1_t0.csv")$map)

```

::: callout-tip
## For users of the Skanit file format

The previous chunk (absorbance and plate map data) demonstrates the usage of the Skanit format version of the import, but the rest of this pipeline does not require the Skanit data, so it is no longer used passed this import step. However, the output format should, in theory, be strictly identical to other format options, so that the object `Skanit_abs` should, in theory, be interchangeable with `TDN_abs` or `Nmin_abs`. Please reach out if you encounter any difficulty.
:::

## 1.2 - PMN

For PMN, import will be similar, but we generate the plate map from a list of samples

```{r}
sample_list <- read_csv("raw_data/PMN/PMN_sample_list.csv") |> 
  select(plate_id, short)
samples <- sample_list$short 
plate_ids <- sample_list$plate_id |> unique()
n_samples_per_plate <- length(samples) / length(plate_ids)

tibble_map_pmn <- map_plates(
  samples, 
  n_samples_per_plate, 
  plate_ids,column_curves = c(1,12), column_blank = 8)

#check it out
tibble_map_pmn

tibble_map_pmn |> write_csv("../raw_data/PMN_map_fromR_to_check.csv")
```

After a manual check that all plate maps are correct (it was, but if not, it is easy to move one or two cells in excel), we saved the file under a different name for re-import. It seems that the headers were not identified as such, so we filter out the first row.

```{r}
PMN_map <- csv_to_tibble("raw_data/PMN/PMN_map_manually_checked.csv") |> 
  filter_out(row == "row")
PMN_abs <- txt_to_tibble("raw_data/PMN")
```

It appears that the plates are not in the same order between map and abs, but this should not be a problem. Check it out anyway:

```{r}
PMN_map ; PMN_abs
```

## 1.3 - PNR

### 1.3.1 - PNR mapping

For the slurry test, we also generate the plate map from a list of samples. First, import it and extract relevant info

```{r}
soil_sample_list <- read_csv("raw_data/PNR/PNR_sample_list.csv", show_col_types = FALSE) |> 
  mutate(
    zone = str_extract(samples, ".*_(\\w\\d)$", group = 1),
    biol_unit_nb = str_extract(samples, "(\\w*)_(\\w*)", group = 1),
    biol_unit_nb = as.double(case_when(biol_unit_nb == "Std_soil" ~ "210", .default = biol_unit_nb)),
    batch = rep(paste0("R", seq(1:8)),10) |> sort()
  ) 

# check it out
soil_sample_list
```

Now we only have once every sample. We need to duplicate them for each incubation time: T0, T2, T4, T6, T24, T26 and 1x for ctrl, then pivot it into 1 long list, and attribute to each slurry sample its extractant identifier

```{r}
slurry_samples <- soil_sample_list |> 
  mutate(
    T0 = paste0(soil_sample_list$samples, "_T0"),
    T6 = paste0(soil_sample_list$samples, "_T6"),
    T2 = paste0(soil_sample_list$samples, "_T2"),
    T24 = paste0(soil_sample_list$samples, "_T24"),
    T4 = paste0(soil_sample_list$samples, "_T4"),
    T26 = paste0(soil_sample_list$samples, "_T26"),
    ctrl = paste0(soil_sample_list$samples, "_ctrl")
  ) |> 
  pivot_longer(cols = T0:ctrl, names_to = "incubation",values_to = "slurry_sample") |> 
  mutate(
    extr_id = rep(c(rep("extr",6), "blank_ctrl"), 80)) 

# check it out
slurry_samples
```

Then we create a table with "fake samples" = blank controls (will take the spot of 2 samples in the plate mapping, but will later be handled as extractant).

```{r}
blank_ctrl <- slurry_samples |> 
  select(batch) |> 
  unique() |> 
  mutate(slurry_sample = "blank_ctrl", reshuffle = 11)

```

We extract only "control" slurry samples (those that got the negative control buffer, and did not undergo the incubation, add the "fake samples" (blanks for control) then reorder them to fit the plate mapping.

```{r}
ctrl <- slurry_samples |> 
  filter(incubation == "ctrl") |> 
  mutate(
    reshuffle = rep(c(seq(1,9,2),seq(2,10,2)),8),
    col_to_order = "end"
  ) |> 
  bind_rows(blank_ctrl) |> bind_rows(blank_ctrl) |> 
  arrange(batch,reshuffle) 
```

Then we extract only the samples that received ammonium (the complement to the negative controls) and format the table to fit the one of the negative controls

```{r}
nh4 <- slurry_samples |> 
  filter(incubation != "ctrl") |> 
  mutate(col_to_order = "begin", reshuffle = 0)
```

Finally, we reassemble all subsets: real, incubated samples with negative control (including the "fake sample" bkanks)

```{r}
map_list <- nh4 |> 
  bind_rows(ctrl) |> 
  arrange(batch,reshuffle) |> 
  mutate(
    plate_nb_per_batch = rep(sort(rep(1:4, 18)), 8),
    plate_id_generic = paste0(batch, "_", plate_nb_per_batch),
    plate_id_NO2 = paste0("NO2_", plate_id_generic),
    plate_id_NO3 = paste0("NO3_", plate_id_generic))

# check it out
map_list
```

Now, we can finally construct the plate mapping

```{r}
samples <- map_list$slurry_sample
plate_ids <- map_list$plate_id_generic |> unique()
#plate_ids_NO2 <- map_list$plate_id_NO2 |> unique()
#plate_ids_NO3 <- map_list$plate_id_NO3 |> unique()

n_samples_per_plate <- length(samples) / length(plate_ids)

tibble_map_pnr <- map_plates(
  samples = rep(samples,2), 
  n_samples_per_plate, 
  plate_ids = c(
    paste0("NO3_",plate_ids), 
    paste0("NO2_",plate_ids)), 
  column_curves = c(1,12), 
  column_blank = 2, rename_na = "blank_ctrl")

#check it out
tibble_map_pnr
```

Only for double-checking: export it to csv for easier reading

```{r}
tibble_map_pnr |> write_csv("../raw_data/PNR_map_fromR_to_check.csv")
```

Double-checked, all good

!! Uncomment if need to re-import!!

```{r}
#(PNR_map <-  read_csv("raw_data/PNR/PNR_map_fromR_checked.csv", show_col_types = FALSE) |> rename(plate_id = row))

PNR_map <- tibble_map_pnr
```

### 1.3.2 - PNR absorbance data

```{r}
PNR_raw_abs <- read_csv("raw_data/PNR/PNR_abs.csv", show_col_types = FALSE, skip = 5, comment = "INPUT") |> 
  filter(!if_all(everything(), is.na))

# replace names with something more handy
names(PNR_raw_abs) <- names(tibble_example)

# create table to construct plate_ids in the order of the file
plate_ids <- PNR_raw_abs |> 
  filter(str_extract(X5, "NO") == "NO") |> 
  select(X3, X5) |> 
  mutate(plate_ids = paste0(X5, "_", X3))

# cf structure: plate ids should appear from row 3, every 11th row
row_id <- seq(3,nrow(PNR_raw_abs), 11)

# replace those cells by plate ids
PNR_raw_abs$row[row_id] <- plate_ids$plate_ids

# then remove useless rows
PNR_abs <- PNR_raw_abs |> 
  filter(row %in% c(LETTERS[1:8], plate_ids$plate_ids))

# check it out
PNR_abs
```

# 2 - Verticalize plates

Here, we transform data from plate format to "verticalized" format: 1 row per well position (well_id –\> 96 rows) and 1 column per "plate" (actually 2 columns per physical plate: absorbance data plate and mapping plate).

We do this separately for each data set. Because those distinct data sets will be combined eventually, a prefix is added to column names to reflect dataset + whether it is abs or map. Notice the "-" at the end of the dataset prefix.

```{r}
# Nmint1t2
Nmin_t1t2_vertical <- join_abs_map(
  Nmin_t1t2_abs, Nmin_t1t2_map, dataset = "Nmint1t2-")

# Nmint3
Nmin_t3_vertical <- join_abs_map(Nmin_t3_abs, Nmin_t3_map, dataset = "Nmint3-")

# TDN
TDN_vertical <- join_abs_map(TDN_abs, TDN_map, dataset = "TDN-")

# PMN
PMN_vertical <- join_abs_map(PMN_abs, PMN_map, dataset = "PMN-")

# PNR
PNR_vertical <- join_abs_map(PNR_abs, PNR_map, dataset = "PNR-")

## Check them out
Nmin_t1t2_vertical; Nmin_t3_vertical; TDN_vertical; PMN_vertical; PNR_vertical
```

Then, we use the `dplyr::left_join()` function to join all plate data in a single data table. Note the distinct column names with their dataset and abs/map-related prefixes

```{r}
# join all 3 in a single table with all "plate data"
all_vertical <- Nmin_t1t2_vertical |> 
  left_join(Nmin_t3_vertical, by = join_by(row, column)) |> 
  left_join(TDN_vertical, by = join_by(row, column)) |> 
  left_join(PMN_vertical, by = join_by(row, column)) |> 
  left_join(PNR_vertical, by = join_by(row, column))

# look at the structure of variable names
sample(names(all_vertical),size = 30)

# check it out  
all_vertical

```

# 3 - tidy table

What we want is to have one column for absorbance data (all plates), another one for the mapping (all plates), i.e., a much longer table (number of rows = 96 \* nb of plates).

`verticalize_to_tidy()` brings the verticalized plates into a tidy format, extracting from column names dataset, plate_id , mapping data and absorbance data into their own column, thus creating a very long tidy table where each row corresponds to a unique well reported in a new column called `unique_well_id`.

```{r}
(all_raw_abs_tidy <- vertical_to_tidy(all_vertical))
```

# 4 - Add plate metadata

Now we have a tidy table that is, in theory, ready for export. In real life, this data cannot be used on its own, and plate metadata is necessary.

First, we import metadata for each dataset, and, if it is not already in it, add a "dataset" column that will help with the next steps.

```{r}
# import csvs
(Nmin_metadata <- read_csv(
  "raw_data/Nmin/Nmin_metadata.csv", show_col_types = FALSE
  ) |>
  mutate(dataset = "Nmint1t2"))

# check it out
str(Nmin_metadata)

Nmin_t3_metadata <- read_csv(
  "raw_data/Nmin_t3/Nmint3_metadata.csv", show_col_types = FALSE
  ) |> 
  mutate(dataset = "Nmint3")

TDN_metadata <- read_csv(
  "raw_data/TDN/TDN_metadata.csv", show_col_types = FALSE
  ) |> 
  mutate(std_column = as.character(std_column)) |> 
  mutate(dataset = "TDN")

PMN_metadata <- read_csv("raw_data/PMN/PMN_metadata.csv", show_col_types = FALSE) |> 
  mutate(dataset = "PMN")

PNR_metadata <- read_csv("raw_data/PNR/PNR_metadata.csv", show_col_types = FALSE) |> 
  mutate(dataset = "PNR")
```

Then we join them in one big metadata file. ! this only works because the 3 metadata files have strictly the same structure (same column names in the same order, containing the same data type (string, numeric, ...)).

```{r}
# join csv's
(all_plate_metadata <- bind_rows(Nmin_metadata, Nmin_t3_metadata, TDN_metadata, PMN_metadata, PNR_metadata))
```

Keep only relevant columns.

```{r}
all_plate_metadata_keep <- all_plate_metadata |> 
  select(dataset, plate_id, std_sp, std_conc, std_unit, sample_dilution, date) 
```

We could already here, join both metadata and tidy data, but this would create a much larger file to save on the computer than the 2 files separately (because rows of the metadata would be repeated 96 times, one per well of the plate.

# 5 - Export

Some steps will be specific of TDN (because we work with much higher concentrations), so we separate the raw data according to TDN or not TDN[^1].

[^1]: TDN stands for Total Dissolved Nitrogen, i.e., NO3- is dosed after an oxidation of all N compounds to NO3-

First, prepare those subsets

```{r}
all_raw_abs_TDN <- all_raw_abs_tidy |> filter(dataset == "TDN")
all_raw_abs_noTDN <- all_raw_abs_tidy |> filter_out(dataset == "TDN")

all_plate_metadata_TDN <- all_plate_metadata_keep |> 
  filter(dataset == "TDN")
all_plate_metadata_noTDN <- all_plate_metadata_keep |> 
  filter_out(dataset == "TDN")
```

Then, export them as .rds

```{r}

#all_raw_abs_tidy |> filter(dataset == "PNR")

all_raw_abs_TDN |> write_rds("output/data/1_all_raw_abs_TDN.rds")
all_raw_abs_noTDN |> write_rds("output/data/1_all_raw_abs_noTDN.rds")

all_plate_metadata_TDN |> write_rds("output/data/1_all_plate_metadata_TDN.rds")
all_plate_metadata_noTDN |> write_rds("output/data/1_all_plate_metadata_noTDN.rds")

```

Also export info concerning PNR that will be needed downstream

```{r}
slurry_samples |> write_rds("output/data/1_slurry_sample_PNR.rds")
```
