---
title: "2.1.2 - From Absorbance to Concentration"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Set up

```{r}
rm(list = ls())

library(tidyverse)
library(plate2N)
```

# 1 - Import data

```{r}
lm_output <- read_rds("output/data/2.1.1_lm_output.rds")
```

# 2 - Apply regression equation

Check that we are now left with only one curve per plate

```{r}
if (
  (lm_output$std_data_clean |> group_by(plate_id) |>  n_groups()) == 
  (lm_output$std_data_clean |> group_by(unique_curve_id) |>  n_groups())
) {message("All good: there is exactly one curve per plate")} else {
  warning("Warning: there is at least one plate with several curves")
}
```

Regression equation is Abs = slope \* Concentration

```{r}
reg_data <- lm_output$lm_table_clean |> 
  select(plate_id, slope, adj_r_squared, lm_p)

abs_data <- lm_output$abs_corrected |> 
  select(dataset, plate_id, map, abs_corrected, std_sp, std_unit) |> 
  left_join(reg_data, by = join_by(plate_id)) |> 
  mutate(target_sp = rep("N"), .after = std_sp)

conc_data <- 
  abs_data |> 
  # compute concentration in mg Nsp per L
  mutate(conc_mgNsp_L = abs_corrected / slope)
```

```{r}
# TEST CHUNK

```

# TO DO

Other, coding

- Apply linear model on sample data to get raw concentration data

  - std_regression()

  - abs_to_mgN_L()

- Bullet-point fill the vignette (basically copy-paste of this script)

- Start new script for data transformation (or not if too late in the week)
