---
title: "2.1.2 - From Absorbance to Concentration"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Set up

```{r}
rm(list = ls())

library(tidyverse)
library(plate2N)
```

# 1 - Import data

```{r}
lm_output <- read_rds("output/data/2_lm_output.rds")
```

# 2 - Apply regression equation

Check that we are now left with only one curve per plate

```{r}
if (
  (lm_output$std_data_clean |> group_by(plate_id) |>  n_groups()) == 
  (lm_output$std_data_clean |> group_by(unique_curve_id) |>  n_groups())
) {message("All good: there is exactly one curve per plate")} else {
  warning("Warning: there is at least one plate with several curves")
}
```

Regression equation is Abs = slope \* Concentration

Here, we

- connect regression data to sample absorbance data

- apply the regression equation to go from absorbance to concentration in mg N-sp per L

- convert unit to mg N per L

```{r}
data_mg_N_L <- 
  # add slope + info regression (p-val and R2) to absorbance data
  reg_join_abs(lm_output$lm_table_clean, lm_output$abs_corrected, target_sp = "N") |> 
  # compute concentration from absorbance
  mutate(conc_mgNsp_L = abs_corrected / slope) |> 
  convert_molec(masses = molar_masses)
```

```{r}

```

# TO DO

Other, coding

- Bullet-point fill in the vignette

  - And create intermediate data for the package

- Start new script for data transformation (or not if too late in the week)
