---
title: "2.1.1 Absorbance Data QC"
author: "Morgane de Toeuf"
format: 
  gfm:
    toc: true
    #code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# TO DO

- Check concentrations courbe std de Cloé

# Set up

Loading packages

```{r}

rm(list = ls())

library(tidyverse)
library(plate2N)
library(roperators)
```

Loading data

```{r}
all_raw_abs_tidy <- read_rds("output/data/1.1_all_raw_abs_tidy.rds")
all_plate_metadata <- read_rds("output/data/1.1_all_plate_metadata.rds")
```

Joining plate data and metadata

```{r}
raw_meta <- all_raw_abs_tidy |> left_join(all_plate_metadata, by = join_by(dataset, plate_id))
#raw_meta |> filter(plate_id == string, map == "Std")
```

# 1 - Suspicious wells removal

## 1.1 - Manual records

This section allows the removal of wells that "we know" are failed wells (e.g., something went wrong during pipetting...).

First, we create a template document to record identifiers of failed wells. To unequivocally identify a well, 3 pieces of info are needed: dataset, plate id, well id.

```{r}
(template <- failed_wells_template(nrow = 30))
```

Then we export it as a csv for manual encoding

```{r}
#| eval: false
#| include: true
write_csv(template, file = "output/template/failed_wells_30.csv")
#write_excel_csv(template, file = "output/template/failed_wells_30.csv")
```

After filling that file manually, we re-import it and correct the tidy data table

```{r}
(failed_wells <- read_csv("raw_data/failed_wells.csv", show_col_types = FALSE))

raw_abs_tidy <- raw_meta |> remove_wells(failed_wells)
#raw_abs_tidy |> filter(plate_id == string, map == "Std")
```

## 1.2 - Suspicious absorbance values (automated)

Observe values for absorbance (iteratively)

```{r}
suspicious_wells <- raw_abs_tidy |> 
  qc_raw_abs(
    min_abs = 0.03, max_abs = 4, 
    plot_col_facet = "std_sp", 
    export_plot = "none") 

suspicious_wells |> slice_max(abs, n = 10)
```

For now, I decide to remove those wells. To be reviewed

```{r}
raw_abs_ok <- raw_abs_tidy |> remove_wells(suspicious_wells)
#raw_abs_ok|> filter(plate_id == string, map == "Std")
```

Check the QC once more

```{r}
raw_abs_ok |> 
  qc_raw_abs(min_abs = 0.03, max_abs = 3, 
    plot_col_facet = "std_sp", 
    export_plot = "none") #|> 
  #select(well_id, map) |> unique()

# Once validated, store last version in a "validated" data
raw_abs_clean <- raw_abs_ok
```

It appears that only the most concentrated wells in the standard curve for TDN (well H1) show absorbance levels above 3. We can later look at those curves and see whether those points are outside of the linear range. Not to worry now, though

# 2 - Correction for blank

## 2.1 - Standard curve

Obtain curve concentrations from metadata

```{r}
curve_concentration <- extract_curve(all_plate_metadata)
```

Extract Std wells, add unique curve ID, then add curve_concentration

```{r}
std_data <- raw_abs_clean |> 
  extract_std_data() |> 
  select(!std_conc) |> 
  left_join(curve_concentration, by = join_by(row, dataset, plate_id))
```

Check unstrusted blanks (where the smallest value for a given curve is not in row A (top_down pipetting) or in row H (bottom_up pipetting)

```{r}
std_blank <- raw_abs_clean |> plate2N::extract_std_blank()
std_blank$untrusted
#blank$all |> filter(plate_id == "NO3_R2R3_1")
```

Check it out graphically.

```{r}
# get unique curve id for the curves of interest (containing the suspicious wells)
#curve <- blank$untrusted$unique_curve_id

# all data --> too messy. 
std_data |> 
  plot_std() +
  facet_grid(dataset~std_sp)

# Subset: look at suspicious blanks
std_data |> 
  filter(unique_curve_id %in% std_blank$untrusted$unique_curve_id) |> 
  plot_std() +
  facet_wrap(~plate_id, scales = "free")
```

In this case, all "untrusted wells" are to be removed, as the values in A1 are clearly out of the curve and probably come from a pipetting error (failure to eject liquid with the automated pipette).

Should there be a choice, where only some of those wells need to be removed, but not all, the function `remove_wells()` can be used on `blank$all` to remove the selection of really-untrusted wells, which would generate a new version of "trusted wells", from which the average values need to be recomputed manually.

```{r}
#| eval: false
#| include: true
# EXAMPLE, NOT NEEDED IN THIS PIPELINE
# get info on wells to remove

      # (suspicious_wells <- std_blank$untrusted |> 
      #   ungroup() |> 
      #   select(dataset, plate_id, well_id))

# Here, those wells are in blank$all, which we can use to remove the wells
      # blank <- std_blank$all |> ungroup() |> remove_wells(suspicious_wells)

# In this case, std_blank is the same as std_blank$trusted, but we could decide to keep some
      # sum(blank != std_blank$trusted)
      # 
      # std_blank_avg <- blank |>
      #     dplyr::summarise(
      #       .by = c(dataset, plate_id),
      #       blank_avg = mean(abs),
      #       blank_sdev = stats::sd(abs)
      #     ) |>
      #     dplyr::mutate(blank_coeff_var_percent = 100 * blank_sdev / blank_avg)


```

Checking how many blanks have been removed

```{r}

nrow(std_blank$all) ; nrow(std_blank$untrusted) ; nrow(std_blank$trusted) 

```

::: {#loss-of-std-blank .callout-caution}
## CAUTION

We are here removing wells with the blank for the standard curve. This works because we had 2 curves per plate in those plates.

- Should there have only been 1 curve per plate, it would be more complex, as we cannot afford to have zero value for the blank of the standard curve.

- An option would be to see whether the inter-plate variation in absorbance values for the standard curves is sufficiently small. If so, then maybe the blank value of one plate could be replaced by the mean of other plates.

- In doing so, watch out for batch effect. Maybe inter-plate variation is smallest during a single day of experimentation, etc.

- To be tested and implemented in coding.
:::

Now that we have all the trusted wells with blank values, we can finally correct absorbance values for the standard curves

```{r}
# # 2,302 rows
# std_corrected <- raw_abs_clean |> 
#   correct_std_blank(std_blank_average = blank$average)

std_corrected <- 
  blank_correct_abs(
    raw_wells_data = std_data|> 
      ungroup() |> 
      filter_out(row == "A"),
    per_plate_avg_blank = std_blank$average |> rename(blank_avg = blank_avg),
    map_to_exclude = ""
  )
```

Because the logic is similar, we will first go into blank-correction of sample data before finalizing work on the standard curves (applying linear regression model)

## 2.2 - Sample wells

First, extract data for wells containing extractant and have a look at its variation

```{r}
extr_data <- extract_extractant(raw_abs_clean)
(blank_avg <- extractant_average(raw_abs_clean) |> 
  arrange(desc(blank_coeff_var_percent)))

plot_blank_var_distrib(blank_avg)
```

We see that a few plates have a very high coefficient of variation, we will have to look at them individually. Let's set the threshold for the coefficient of variation at 5% (default)

```{r}
threshold <- 5

suspicious_plates <- raw_abs_clean |> 
  qc_raw_extr(suppress_warning = TRUE, max_coeff = threshold)

suspicious_extr <- suspicious_extr(
  raw_abs_clean, suspicious_plate_id = suspicious_plates, max_coeff = threshold)

# check it out
suspicious_extr

# plot outliers
suspicious_extr |> boxplot_outlier_extr(max_coeff = threshold)

```

We have 5 plates that each have one or more obvious outlier well. We will need to remove them manually.

First, we create a small tibble that will serve to construct the tibble of wells to remove

```{r}
plate_ids <- suspicious_extr |> 
  ungroup() |> 
  select(dataset, plate_id) |> unique() 
plate_ids <- plate_ids |> # save numbers for plate order in the plot
  mutate(plate_order = seq(1, nrow(plate_ids)))
```

Then, we create a vector with wells to remove (going through boxplots from top to bottom).

:::: callout-tip
## Manually remove outliers

::: {#tip-input-threshold-extr .callout-tip}
In the following chunk, you need to manually decide which wells to remove, based on the boxplots produced above.

- Make sure to deal appropriately with plates that require 2 outliers or no outlier to be removed (see example below)
:::
::::

```{r}
#** !!! MANUAL INPUT !!! *

# Which plate needs 2 outliers removed?
plate_with_2_outliers <- 5
plate_without_outliers <- 9 # use a number > nb of plates if there is no such plate

# Which wells are outliers? 
well_ids <- c("H8", "C8", "H3", "B5", "C8", "D8", "D2", "E8") # use NA for plates without outliers
```

Then we finish constructing the tibble of wells to be removed

```{r}
#nb_plates <- plate_ids |> nrow()

to_remove <- plate_ids |> 
  bind_rows(plate_ids |> filter(plate_order == plate_with_2_outliers)) |> 
  arrange(plate_order) |> 
  mutate(well_id = well_ids) |> 
  filter(plate_order != plate_without_outliers) |> #remove plate without outliers
  select(!plate_order)

```

And we remove it from extractant data

```{r}
extr_data_clean <- extr_data |> 
  remove_wells(to_remove) 

blank_avg_clean <- extractant_average(extractant_data = extr_data_clean) 
```

Check that biggest coeff_var indeed below threshold

```{r}
blank_avg_clean |> arrange(desc(blank_coeff_var_percent)) |> head()
```

Now that we are confident in the per-plate average value of raw absorbance of extractant wells, we can finally blank-correct all sample data

```{r}

abs_corrected <- 
  blank_correct_abs(
    raw_wells_data = raw_abs_clean, 
    per_plate_avg_blank = blank_avg_clean |> rename(blank_avg = blank_avg),
    map_to_exclude = c("empty","Std","extr")) 

```

## 2.3 - All corrected data

Let's just recall all corrected data. We have 2 separate tibbles (because the experimental design was arranged to have separate blanks for the curve and the samples)

```{r}
# Standard curve, blank-corrected and clean
std_corrected 

# Samples, blank-corrected and clean
abs_corrected
```

# 3 - Compute regression equation (per plate)

## 3.1 - QC standard curves - round 1

Assumptions of a linear model: (taken [here](https://towardsdatascience.com/all-the-statistical-tests-you-must-do-for-a-good-linear-regression-6ec1ac15e5d4/), apparently from [Spanish book](https://periodicos.ufpe.br/revistas/politicahoje/article/download/3808/31622))

- The residuals must follow a normal distribution.

- The residuals are homogeneous, there’s homoscedasticity.

- There’s no outliers in the errors.

- There’s no autocorrelation in the errors.

- There’s no multicolinearity between the independent variables.

::: callout-warning
## ATTENTION - ISSUE

TO DO:

- Figure out why the grouped table given as input to `lm_std_curve()` has 329 groups, but the output table has 331 rows!

- And why `lm_table_suspicious` has 47 rows, but `suspicious_lm_plotlist` has 45 elements
:::

!!! FORMULATE better and split in 2 chunks :-)

First, we perform a linear model on each curve individually (i.e., possibly several curves per plate).

Then we take a subset to examine individually: those curves where the linear model doesn't seem to perform ideally (e.g., non-significant model (p-value \> 0.05), residuals not normally distributed, or heteroscedasticity)

```{r}
lm_table_raw <- lm_std_curve(std_corrected |> group_by(plate_id, column))

# extract all plates where "something" is not perfect 
lm_table_suspicious <- lm_table_raw |> 
  suspicious_lm()
```

Then, for visual support, we create a list of plots where we store each individual plot of "suspicious" standard curves

```{r}
suspicious_lm_plotlist <- plot_list_lm(
  lm_data = lm_table_suspicious,
  std_data = std_corrected)

# check one plot out
suspicious_lm_plotlist$NH4_1F1_col1

```

Now we can look at each plot individually. Because there are still 45 plots to review, we will look through them in 3 batches of 15 plots.

```{r}
n_plots <- length(suspicious_lm_plotlist)
batch <- 15

batch_1 <- suspicious_lm_plotlist |> head(n = batch)
batch_2 <- suspicious_lm_plotlist |> tail(n = n_plots-batch) |> head(n = batch)
batch_3 <- suspicious_lm_plotlist |> tail(n = n_plots-(2*batch)) 

patchwork::wrap_plots(batch_1, axis_titles = "keep") +
     patchwork::plot_annotation(title = "Plots of suspicious Standard curves")
patchwork::wrap_plots(batch_2, axis_titles = "keep") +
     patchwork::plot_annotation(title = "Plots of suspicious Standard curves")
patchwork::wrap_plots(batch_3, axis_titles = "keep") +
     patchwork::plot_annotation(title = "Plots of suspicious Standard curves")


```

Most plates are from NH4 and NO2. Because most measurement of NH4 and NO2 are close to zero, it might be relevant to just remove highest points of the curve to get a better fit.

Let's check how high the absorbance goes with these data sets, for each N species. When comparing those highest values with the plots above, we see that the highest absorbance for NO2 and NH4 is far below the most concentrated point of the standard curve –\> we can remove row H for sure, but row G probably not.

This is not the case for NO3, or at least not for the TDN dataset, or part of it.

```{r}
abs_corrected |> filter(std_sp == "NO2") |> 
  arrange(desc(abs_corrected)) |> slice_head(n = 3)
abs_corrected |> filter(std_sp == "NH4") |> 
  arrange(desc(abs_corrected)) |> slice_head(n = 3)
```

Let's first list wells to remove the H row of NO2 and NH4 std.

```{r}
to_remove_nh4_no2_h <- std_corrected |> 
  filter(
    std_sp %in% c("NH4", "NO2"),
    row == "H") |> 
  select(dataset, plate_id, well_id)
```

Then, let's look at suspicious plots for NO3 only, so we can see the plots better

```{r}
lm_table_suspicious_NO3 <- lm_table_suspicious |> 
  # extract N_sp from plate name. May be done differently, this is just one of many possible approaches
  separate_wider_delim(
    cols = unique_curve_id, delim = "_", 
    names = c("N_sp", "rest"), 
    too_many = "merge", cols_remove = FALSE) |> 
  select(!rest) |> 
  filter(N_sp == "NO3")

# Create plot list
lm_plot_list_NO3 <- plot_list_lm(
  lm_data = lm_table_suspicious_NO3,
  std_data = std_corrected)

patchwork::wrap_plots(lm_plot_list_NO3, axis_titles = "keep") +
     patchwork::plot_annotation(title = "Plots of suspicious Standard curves")
```

For the NO3 plates, the issue has to be solved differently, but those are just a few –\> individual appraisal. We see here that, even out of TDN, we cannot remove the H row, because some points seem to have an absorbance close to that of the H row of the standard.

```{r}
abs_corrected |> filter(std_sp == "NO3", dataset != "TDN") |> 
  arrange(desc(abs_corrected)) |> slice_head(n = 3)
std_corrected |> filter(std_sp == "NO3", dataset == "TDN") |> 
  group_by(plate_id) |> slice_max(abs_corrected) # |> view()

```

For NO3_2F1 and NO3_1F1: there were probably 2 columns, so I can look at both columns together

```{r}
lm_table_raw |> 
  separate_wider_delim(
    cols = unique_curve_id, delim = "_", 
    names = c("n_sp", "plate", "rest"), 
    too_many = "merge", cols_remove = FALSE) |> 
  filter(n_sp == "NO3", plate %in% c("1F1", "2F1")) |> 
  plot_list_lm(std_data = std_corrected) |> 
  patchwork::wrap_plots()
```

Honestly, they look ok-ish. Maybe after computing the average of both curves the issue is solved!

Then re-compute linear model table as "clean" version.

Then apply it

```{r}
std_corrected_wash1 <- 
  std_corrected |> remove_wells(to_remove_nh4_no2_h)
```

Run the quality check once more

```{r}
lm_table_wash1 <- lm_std_curve(std_corrected_wash1 |> group_by(plate_id, column))

# extract all plates where "something" is not perfect 
lm_table_suspicious_wash1 <- lm_table_wash1 |> suspicious_lm()

suspicious_lm_plotlist_wash1 <- plot_list_lm(
  lm_data = lm_table_suspicious_wash1,
  std_data = std_corrected_wash1)

n_plots <- length(suspicious_lm_plotlist_wash1)
batch <- 16

batch_1 <- suspicious_lm_plotlist_wash1 |> head(n = batch)
batch_2 <- suspicious_lm_plotlist_wash1 |> tail(n = n_plots-batch) |> head(n = batch)

patchwork::wrap_plots(batch_1, axis_titles = "keep") +
     patchwork::plot_annotation(title = "Plots of suspicious Standard curves")
patchwork::wrap_plots(batch_2, axis_titles = "keep") +
     patchwork::plot_annotation(title = "Plots of suspicious Standard curves")

```

We still get quite some curves. But when we look at the p-values to reject normality or heteroscedasticity, they are indeed below 0.05, but it's also not 10\^-5 or something like that. The smallest p-value (shapiro) is at 0.007.

Let's compute the means per dilution and see...

## 3.2 - Compute per-dilution averages

Most plates in our dataset have 2 columns with the standard curves. It seems that the \~1min delay between the 2 (column 1 and column 12 of the plate) are responsible for a slight shift (example in plot below)

```{r}
std_corrected_wash1 |> filter(plate_id == "NO3_2F3_1") |> rename(abs = abs_corrected) |> 
  plot_std()
```

So we will now compute the mean for same row (e.g., mean of H1 and H12)

::: callout-warning
## WARNING

The next step computes per plate per row means for the standard curves.

If some wells have been swapped in some plates, this may cause problems. Make sure there was no pipetting issue, or correct raw data or solve it through code
:::

```{r}

std_dilution_avg <- std_dilution_average(std_corrected_wash1)

```

## 3.3 - QC standard curves - round 2

We repeat same steps as above: computation of linear model, identification of suspicious curves and plotting

```{r}
lm_std_avg <- lm_std_curve(std_dilution_avg |> rename(abs_corrected = abs_mean))
lm_suspicious_avg <- lm_std_avg |> suspicious_lm()

lm_plots_avg <- lm_suspicious_avg |> plot_list_lm(
  std_data = std_dilution_avg |> rename(abs_corrected = abs_mean))

lm_plots_avg |> patchwork::wrap_plots()
```

Now, we decide to get rid of just a few wells that really stick out, for the following plates:

NH4_1G2 (E13)

NH4_2F4_1 and NH4_2F4_2 (C13)

And the TDN issues will have to be resolved later (when we know the range of std concentration)

Let's compose a "to_remove" tibble

```{r}
to_remove <- std_dilution_avg |> 
  ungroup() |> 
  select(dataset, plate_id, well_id) |> 
  filter(
    ((plate_id == "NH4_1G2") & (well_id == "E13")) |
    (plate_id %in% c("NH4_2F4_1", "NH4_2F4_2") & (well_id == "C13"))
  )

std_corrected_wash2 <- std_dilution_avg |> 
  remove_wells(to_remove) |> 
  rename(abs_corrected = abs_mean)
```

One more run of QC to check if we are satisfied with the resulting curves

```{r}
lm_wash2 <- lm_std_curve(std_corrected_wash2)
lm_suspicious_wash2 <- lm_wash2 |> suspicious_lm()

lm_plots_wash2 <- plot_list_lm(lm_suspicious_wash2, std_corrected_wash2)
lm_plots_wash2 |> patchwork::wrap_plots()  
```

Ok, good enough!

Let's store the last correction into a clean variable name to reduce possible confusion, and let's compute all the plots in a big list, for storage purposes. We then export this as one output data in a single list

```{r}
std_data_clean <- std_corrected_wash2
lm_table_clean <- lm_wash2
lm_plots_clean <- plot_list_lm(lm_table_clean, std_data_clean)

lm_output <- list(
  "std_data_clean" = std_data_clean,
  "lm_table_clean" = lm_table_clean,
  "lm_plots_clean" = lm_plots_clean,
  "abs_corrected" = abs_corrected
)

lm_output |> write_rds("output/data/2.1.1_lm_output.rds")
```

## 3.4 - Multiple curve QC

First, let's look at the distribution of p-values of the std curve regressions

```{r}
p_threshold <- 0.05

lm_output$lm_table_clean |> 
  separate_wider_delim(
    cols = unique_curve_id, delim = "_", 
    names = c("n_sp", "rest"), too_many = "merge") |>
  ggplot(aes(x = -log(lm_p))) + 
  theme_minimal() +
  # geom_histogram() +
  geom_density() +
  geom_vline(aes(xintercept = -log(p_threshold)), linetype = 2, colour = "purple") +
  annotate(
    geom = "label", label = paste0("p-val = ", p_threshold),
    x = -log(p_threshold), y = 0.15, hjust = 0.25, colour = "purple" ) +
  facet_wrap(~n_sp, nrow = 3) +
  xlim(0,max(-log(lm_output$lm_table_clean$lm_p))) +
  xlab("-log(p-value of linear model)") +
  labs(
    title = "Distribution of p-values of the linear model",
    subtitle = "logarithmic scale"
  )
```

Then, same with R_squared (or adjusted?)

```{r}
threshold <- 95

lm_output$lm_table_clean |> 
  separate_wider_delim(
    cols = unique_curve_id, delim = "_", 
    names = c("n_sp", "rest"), too_many = "merge") |>
  ggplot(aes(x = adj_r_squared, colour = n_sp, fill = n_sp)) + 
  theme_minimal() +
  #geom_histogram() +
  geom_density(alpha = 0.3) +
  geom_vline(aes(xintercept = 0.95), linetype = 2, colour = "purple") +
  annotate(
    geom = "label", label = paste0("adjusted R2 = ", threshold, "%"),
    x = threshold/100, y = 500, hjust = 0.25, colour = "purple" ) +
  #facet_wrap(~n_sp, nrow = 3, scales = "free_y") +
  xlim(
    0.945,
    max(lm_output$lm_table_clean$adj_r_squared)) +
  xlab("Adjusted R2") +
  labs(
    title = "Distribution of adjusted R2 values of the linear model"
  )
```

Now we plot all curves on same plot

- for now: exclude TDN because concentrations of standard curves are wrong

- To Do : check std curve concentrations in Cloé's notebook!

```{r}
colors <- c("#7FC97F", "#BEAED4", "#FDC086")

lm_output$std_data_clean |> 
  filter_out(dataset == "TDN") |> 
  ggplot(aes(x = as.numeric(std_conc), y = abs_corrected, groups = plate_id, colour = dataset, fill = dataset)) +
  theme_minimal()+
  geom_smooth(
    formula = y~x-1, method = "lm", se = TRUE, 
    alpha = 0.05) +
  geom_point() +
  facet_wrap(~std_sp, scales = "free") +
  scale_color_discrete(palette = colors[1:2]) +
  scale_fill_discrete(palette = colors[1:2]) 
```

Now, finally, I decide that I am happy with my standard curves, so I can move on to apply the equations on my data

# TO DO

!!! FOR TDN !!!

- CHECK Std curve concentrations !!!

- Check how high is the highest absorbance value of 2x diluted samples for NO3 –\> can we remove H row?

!!! FOR Nmin t3 !!!

- Check in lab notebook of Cloé: which dilutions for NH4? Courbes sont bizarrement décalées (facteur 2 quasi! )
