---
title: "2.1.1 Absorbance Data QC"
author: "Morgane de Toeuf"
format: 
  gfm:
    toc: true
    #code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# Set up

Loading packages

```{r}

rm(list = ls())

library(tidyverse)
library(plate2N)
```

Loading data

```{r}
all_raw_abs_tidy <- read_rds("output/data/1.1_all_raw_abs_tidy.rds")
all_plate_metadata <- read_rds("output/data/1.1_all_plate_metadata.rds")
```

Joining plate data and metadata

```{r}
raw_meta <- all_raw_abs_tidy |> left_join(all_plate_metadata, by = join_by(dataset, plate_id))
```

# 1 - Suspicious wells removal

## 1.1 - Manual records

This section allows the removal of wells that "we know" are failed wells (e.g., something went wrong during pipetting...).

First, we create a template document to record identifiers of failed wells. To unequivocally identify a well, 3 pieces of info are needed: dataset, plate id, well id.

```{r}
(template <- failed_wells_template(nrow = 30))
```

Then we export it as a csv for manual encoding

```{r}
#| eval: false
#| include: true
write_csv(template, file = "output/template/failed_wells_30.csv")
#write_excel_csv(template, file = "output/template/failed_wells_30.csv")
```

After filling that file manually, we re-import it and correct the tidy data table

```{r}
(failed_wells <- read_csv("raw_data/failed_wells.csv", show_col_types = FALSE))

raw_abs_tidy <- raw_meta |> remove_wells(failed_wells)
```

## 1.2 - Suspicious absorbance values (automated)

Observe values for absorbance (iteratively)

```{r}
suspicious_wells <- raw_abs_tidy |> 
  qc_raw_abs(
    min_abs = 0.03, max_abs = 4, 
    plot_col_facet = "std_sp", 
    export_plot = "none") 

suspicious_wells |> slice_max(abs, n = 10)
```

For now, I decide to remove those wells. To be reviewed

```{r}
raw_abs_trusted <- raw_abs_tidy |> remove_wells(suspicious_wells)
```

# 2 - Correction for blank

## 2.1 - Standard curve
