---
title: "V. Pipeline for Absorbance data"
format: 
  gfm:
    toc: true
    code-fold: true
# format: 
#   html:
#     embed-resources: true
#     toc: true
#     toc_float: true
#     code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# To Do

-   [**!! 2 PLATES WHERE I WOULD WANT TO HAVE A LOWER THRESHOLD !! SO UPDATE PIPELINE CODE TO BE MORE FLEXIBLE IN FUNCTION CALL !!**]{.underline}
-   !! Figure out wh[**y some final concentrations are equal to zero!**]{.underline}
-   Improve visualization for multiple plots –\> make a loop to produce several pages?
-   improve rendering of @lst-qc4
    -   have the table(s) called in successive chunks with labels –\> can call them in comments above
    -   change the printing of warning / message? (seems to appear twice)
-   Change function for superposed plots so that it is easy to modulate which column of metadata goes into the color-coding of the curves

# Intro

To keep this pipeline easily readable and reproducible, most operations are "hidden" in functions. Until I learn how to create a package (upcoming...), you'll have to download the folder containing functions. It should be in the current working directory. If not, update the path below.

In this pipeline we refer to

-   "curve" = all wells and associated data that refer to one standard curve (typically a whole column on the plate).

-   "plate" = one 96-well plate used for the experiment. It is, together with rows and columns, one of the smallest scales of extracting data. Although data at this stage is no longer formated in the shape of a plate, the plate still constitutes a unit. All wells on that plate share the same blanc and the same standard curve, they will thus be treated as one for many computation steps.

-   "row" can be used to designate the row (i.e., observation) of a table or data set. But it also designates the row of a plate and, as such, becomes data recorded in a column called "row". This dual meaning can lead to some confusion. I try to use "row-plate" when refering to the rows (A, B, C...) of a 96-well plate. But it is possible that I forgot to do that sometimes and simply used "row". Context should help.

-   "suspicious". We refer to suspicious curves or suspicious wells or suspicious plates: they need to be examined and may contain / constitute outliers

-   "(un)trusted": this is one step further than "suspicion". A decision has been taken. This refers to wells that have been or will be removed (untrusted) or kept (trusted)

::: {#tip-input .callout-tip}
## Your input is needed!

The pipeline is highly automated. Nevertheless, as a user, you'll have to take a few decisions:

-   adapting path for import of data (@lst-load)
-   Playing around with the range of absorbance to accept (@lst-qc1-initial-range)
-   Playing around with max acceptable coefficient of variation for extractant (@lst-hist-extr-coeff-var)
-   Based on observations from @fig-suspicious-plates, thresholds can be set in @lst-cut-threshold-extractant to exclude outlier wells from extractant data
-   Based on visual appraisal of graphs @fig-unsorted-curves, encode identifier of outlier wells @lst-outlier-std
-   adapting path for export of data (@lst-export)
:::

# 1 - Set up

## 1.1 - Loading packages and homemade functions

```{r}
#| label: load-packages-functions
#| message: false
#| warning: false
#| output: false
library(tidyverse)
library(roperators) # to be able to add %ni% for "not in"
library(patchwork) # for wrap_plots and wrap_tables
library(RColorBrewer) # to find and set color palette

source("functions/extract_curve.R")
source("functions/plot_qc_std_all.R")
source("functions/qc1_initial_range_abs.R")
source("functions/qc2_plot_range_abs.R")
source("functions/qc3_nb_curves_per_plate.R")
source("functions/subset_data.R")
source("functions/qc4_std_blanc_variation.R")
source("functions/correct_abs_std.R")
source("functions/qc5_extr_blanc_variation.R")
source("functions/qc6_extr_trusted.R")
source("functions/qc7_std_find_outlier.R")
source("functions/correct_abs_samples.R")
source("functions/plot_qc_std_multiple.R")
source("functions/std_regression.R")
source("functions/abs_to_mgN_L.R")
```

## 1.2 - Loading data

::: callout-tip
## Your input is needed!

The paths in the next chunk should be replaced by your own data, possibly generated in previous pipeline
:::

```{r}
#| lst-label: lst-load
#| lst-cap: "Load data (possibly generated in previous pipeline)"

# import tidy data and metadata
#Nmin_data <- read_rds("output/data/Nmin_tidy.rds")
#Nmin_metadata <- read_rds("output/data/Nmin_metadata.rds")

N_all_plate <- read_rds("output/data/N_all_plate_tidy.rds")
N_all_metadata <- read_rds("output/data/N_all_metadata.rds")

```

# 2 - preliminary QC : suspicious wells

The ideal range for absorbance readings (Beer-Lambert in linear range of relationship between concentration and absorbance) is between 0.1 and 1. But these are not super strict borders. I don't want to send out a warning message too soon, so we take higher values as a default, especially because with NO2-, we expect virtually zero in a lot of wells.

The following chunk filters out only rows where absorbance is out of range, and returns either a warning (when there are out-of-range values) or a happy message (when there are none). In case of a warning, it also shares the table with suspicious wells, so that the user can take an informed decision.

The min and max need to be adapted also because some manips may allow more open numbers or protocol testing would not want to exclude them...

::: callout-tip
## To be thought through

If wells are out of range of "acceptable" absorbance values, you'll get a warning.

-   You can play around with the min and max of the "acceptable" range (@lst-qc1-initial-range)

-   You can also evaluate this range with having a look at @fig-plot_QC_wells

<!-- -->

-   What if some wells are out of range? Ware options then? Remove suspicious wells (replace by NAs?)
:::

```{r}
#| lst-label: lst-qc1-initial-range
#| lst-cap: "Here you can play around with the range of `acceptable` absorbance values"
qc1_initial_range_abs(
  N_all_plate, min_abs = 0.03, max_abs = 2.5)

#> ok with a threshold min of 0.03, but more than 4000 when threshold of 0.05
#out_of_range <- Nmin_full |> filter(row_number() %in% suspicious_rows) 

#> but they're all NH4 or NO2 --> acceptable!
# out_of_range |> filter(
#   str_split_i(plate_id, pattern = "_", 1) %ni% c("NH4", "NO2")
#   )
```

```{r}
#| label: fig-plot_QC_wells
#| fig-cap: "QC of Suspicious wells - Distribution of raw absorbance per N species"
#| fig-subcap: "This figures shows raw numbers of uncorrected absorbance (no blanc correction). It contains all values, including values of blanc wells and values of the standard curve. Low values are expected for NO2-, less so for NH4+"

qc2_plot_range_abs(N_all_plate) +
  labs(subtitle = "Many low values, especially for NH4+ and NO2-")
```

# 3 - Blanc-correction of absorbance values

Now we correct absorbance values by subtracting blanc values from raw values (absorbance of the light by the solution = absorbance by the blank solution + absorbance by the substance to be quantified)

## 3.1 - Correct standard curves for blanc

For now, this is a separate process (std curves vs samples) to account for the fact that the standard curve was prepared in H2O, not in the extractant (K2SO4 or KCl), so it needs its own blanc-correction.

[**To be thought through:**]{.underline}

-   If it becomes relevant: make some sort of if condition, based on plate information (`blanc_id` and `std_id`)

A typical pipetting error with the automated pipette is to forget to expell the first bit (containing air) before the "real" pipetting starts. In this case, the first well to be pipetted (typically well A1) will receive a wrong amount of reagent, which in turn may impact stoechiometry and volume, thus absorbance reads. The next chunk allows the identification of minimum values within a standard curve that are not situated in the first or last row of the plate (usually the standard curve is pipetted in ascending or descending order).

With this information, for example, we can exclude wells where the first row shows a higher absorbance than the second row.

--\> In the chunk below we get suspicious curves. So we can exclude those combinations of plate and column from the computation of the average blanc. So we will only take the value from the remaining std curve. Here, we always pipetted 2 per plate (yay!).

--\> This option of course is not valid in the case where only one standard curve is pipetted per plate. In that case, one option is to check whether absorbance values are fairly constant between plates. If so, it is a fair correction to take the inter-plate average value (or a standardized version of it... we'll cross that bridge when we get to it)

Once we disregard those suspicious wells, we can compute the average blanc values.

First, we check that we indeed have 2 columns with std curve on every plate

```{r}

# check that we have 2 columns with Std per plate --> option to remove suspicious blancs
qc3_nb_curve_per_plate(N_all_plate, nb_std = 2)

```

Second, we can take a subset of `std_data` that contains only the rows with blancs, and only those that we trust (normally row A or H only)

```{r}
#| lst-label: lst-qc4
#| lst-cap: "QC 4 - Variation of Standard blanc"

QC4 <- qc4_std_blanc_variation(N_all_plate, nb_std = 2)
lapply(QC4, print)


#** DECIDE WHAT TO DO WITH THIS INFO. SO FAR, I HAVEN'T HAD THE CASE THAT I NEED TO REMOVE FURTHER WELLS BEYOND THE "UNTRUSTED" ONES. SO CODE FOR THIS STILL REMAINS TO BE WRITTEN, ALTHOUGH INSPIRATION CAN BE TAKEN FROM FURTHER QC STEPS * 
```

Third, we compute the blanc value (average) and return a warning if blanc values show too much variation (in the case of several plate-columns with standard curves)

If we are troubled by the big variation within plate, we can check out the identified suspicious plates. In this case, I find it not so dramatic. We are just dealing with small values which tend to artificially increase the coefficient of variation (division by small numbers...).

We can look at suspicious blancs in their full plate context to decide how bad the situation is (see printed table).

Now we can blanc-correct the absorbance values for the whole standard curves.

```{r}
std_corrected <- correct_abs_std(N_all_plate)
```

We could add those corrected values back into the main data table, but actually those numbers are only useful to compute the regression equation between corrected absorbance and concentration. For thematic clarity purpose, this will be done in a later section (to keep all work on blancs in one place). So we have now just save the output into the object `std_corrected` so we can get back to it later.

## 3.2 - Correcting samples for blanc

While the mathematical approach is similar to correcting the blanc of the standard curve, the computational steps are somewhat different because the data is organised differently on the plates.

First, we extract the plate-rows containing extractant, then we do some quality check: how big is the variation? Do we have suspicious wells? All this happens with "QC 5" in the following chunk.

Looking at the distribution of the coefficient of variation per plate, it appears that some plates have very different scoring, see @fig-hist-extr-coeff-var

```{r}
#| lst-label: lst-hist-extr-coeff-var
#| lst-cap: "Playing around with maximum acceptable coefficient of variation for the absorbance of extractant (blanc)"
#| label: fig-hist-extr-coeff-var
#| fig-cap: "Distribution of coefficient of variation of absorbance of extractant (blanc)"

#** Look at this (and the next chunk) iteratively a couple of times to decide where to put the threshold. *
  QC5 <- qc5_extr_blanc_variation(N_all_plate,max_coeff = 3)
  
  QC5$distrib_coeff +
  labs(subtitle = "anything above 5% seems to be an outlier (even above 3%)")

```

Let's now have a look at those suspicious plates

```{r}
#| label: fig-suspicious-plates
#| fig-cap: "Distribution of absorbance in suspicious plates. From this we can manually identify then remove outliers"

#** To play around with the maximum coefficient of variation, change argument max_coeff in the function call *
#*
QC5$multiple_plot 

```

We have to remove outliers manually: it is really a visual appreciation that works (see @fig-suspicious-plates). Only "suspicious" plates from above are displayed.

-   Watch out, in the case of plates with several blancs like here (coming from several batches of soil extraction), that a bimodal distribution might not necessary be an issue.

-   Also look at the scale of the axes: bins will always spread over the whole x-axis, even if the distance is only of 0.001...

Based on visual appreciation, here is the list of plates that we want to correct. One way to do it is to impose, for each plate, a threshold value that we can later use to filter out outlier wells (wells above the threshold value).

::: {#tip-input-threshold-extr .callout-tip}
## Manually remove outliers!

In the following chunk, you need to input threshold values as a vector with one number per "suspicious" plate. If you decide to keep all values, you can set a threshold of 1.00.

-   look at @fig-suspicious-plates to decide on threshold values. By default, the function `wrap_plots()` orders plots by row, so that the order of plates in `extr_suspicious` corresponds to the plots read from left to right, then next row, etc.

-   The function call only allows to exclude values above the threshold for now. If your outliers are rather the lower values, you'll need to have a look in the source code of the function `qc6_extr_trusted()`, and update it somehow.
:::

**!! In case you want to exclude lower values only, then just change the `>` into `<`. But if you want to exclude some upper values, and some lower values, consider updating the code.**

[**!! 2 PLATES WHERE I WOULD WANT TO HAVE A LOWER THRESHOLD !! SO UPDATE PIPELINE CODE TO BE MORE FLEXIBLE IN FUNCTION CALL !!**]{.underline}

```{r}
#| lst-label: lst-cut-threshold-extractant
#| lst-cap: "Set up threshold and remove outlier wells for absorbance of extractant"

#** !! Manually input the threshold values of your choosing (read plots from left to right, then from up to down (rowise reading)) *

# print it to check in which order the plates are
# suspicious_plates

# Or manual input
cut_threshold <- c(
  0.040, 0.041, 0.039, 0.038, 
  0.038, 0.041, 0.042, 0.037,
  0.072, 0.09, 0.09, 1.00,
  2.00, 1.00, 1.00, 0.08)


QC6 <- qc6_extr_trusted(QC5, cut_threshold = cut_threshold)


```

Now we can use the list of untrusted wells to filter them out of the extractant data, and look at the improved distribution of intra-plate variation.

```{r}
#| label: fig-distrib-variation-extr-improved
#| fig-cap: "Distribution of variation of absorbance of extractant (blanc) after removal of outliers"

QC6$extr_distrib_coeff +
  labs(subtitle = "1 plate is still outlier (above 3%), but much less so than before")
```

Now that we have computed a trusted version of the average of absorbance per plate per "blanc", we can correct sample absorbance values.

```{r}
#N_all_plate |> filter(plate_map == "16_t2_CFE.1x")
#QC6$extr_avg_trusted

corrected_data <- correct_abs_samples(N_all_plate, QC6 = QC6)

```

# °^°°°^ Milestone : blanc-corrected data °^°°°^

# 4 - Compute regression equation btw absorbance and concentration

## 4.1 - Quality check of standard curves

The next chunk will first perfom a basic quality check of the standard curve

-   checking that metadata and data have the same nb of plates

-   check that there are no negative values for the corrected absorbance

### 4.1.1 - removing outliers

Then, it will look for curves that are not monotonous, i.e., not strictly increasing (or decreasing, depending on pipetting). Non-monotonous curves are referred to in the code as "unsorted_curves" (absorbance values are not sorted). Aberrant wells responsible for the non-monotony can be spotted on the plots in @fig-unsorted-curves

```{r}
#| label: fig-unsorted-curves
#| fig-cap: "Non-monotonous curves - helps to identify potential outliers"

QC7.1 <- qc7_std_find_outlier(std_corrected = std_corrected, metadata = N_all_metadata)

#** Same principle as above: visually identify obvious outliers, and list them in the vector below. !! Only works for one well per plate for now. If several --> work iteratively or update code *

QC7.1$multiplot + plot_annotation(subtitle = "1st iteration, no outlier removed yet")

```

::: callout-tip
## Manually remove outliers!

In the following chunk, we can manually encode a vector containing all outlier wells, based on visual appraisal of graphs @fig-unsorted-curves

-   For the plates where you want to keep all wells, encode NA

-   For the plates where you want to remove [**one**]{.underline} well, encode its identifier (e.g., "E12")

-   Currently, you have to encode exactly one value per suspicious curve. If you want to remove 2 wells, proceed either iteratively or go to the source code of the function `qc7_std_find_outlier()`.
:::

Once outlier wells are encoded, the following chunk will update input data (`std_tidy` will replace `std_corrected`). You can then look again at the remaining "suspicious" curves and see the improvement, then decide to proceed for one more iteration or keep the tidy data as is.

```{r}
#| lst-label: lst-outlier-std
#| lst-cap: "Manually encode identifier of outlier wells"

#** !!! MANUALLY ENCODE THE OBJECT CALLED outlier_wells base on last plot - read it from left to right then from up to down *

# Default just to be sure that the code can work, create an empty vector of the correct length
outlier_wells <- rep(NA, nrow(QC7.1$unsorted_curves))

# Or Manual encoding
outlier_wells <- c(NA, "E12", "C1", "C1", "E12", NA, NA, NA)

# connect id of problematic curves (plate id of "unsorted" curves) to ourlier_wells
outlier_curves <- 
  QC7.1$unsorted_curves |> 
    ungroup() |> 
    mutate(
      outliers = outlier_wells,
      unique_well_id = case_when(
        is.na(outliers) ~ NA,
        .default = paste0(plate_id, "_", outliers))
    )
#outlier_curves

# tidy std data with ourlier wells removed
std_tidy <- std_corrected |> 
  filter(unique_well_id %ni% outlier_curves$unique_well_id)

# Run it once more, approve of wells... evtl re-run previous bit
QC7.2 <- qc7_std_find_outlier(std_corrected = std_tidy, metadata = N_all_metadata)

# Look at plos and decide if happy
QC7.2$multiplot + plot_annotation(subtitle = "2nd iteration, outliers removed. Curves satisfactory now")

```

### 4.1.2 - Inter-curve variability

At this stage, we should be satisfied with the curves.

We can visualize all curves for a given N species, to have an idea of inter-curve variability.

The next chunk works, but it produces a very difficult to read plot with multiple panels (`r nrow(N_all_metadata)` actually).

```{r}

plot_qc_std_multiple(metadata = N_all_metadata, std_data = std_tidy)

```

Let's find a neater way to look at it by overplotting, using the function `plot_qc_std_all`.

First, for NH4+ (@fig-QC-std-all-nh4), then for NO2- (@fig-QC-std-all-no2) and for NO3- (@fig-QC-std-all-no3).

For now, the visual is made to attribute color-coding to the factor "sampling-time" which is related to lab experimental batches, so I expected it to show some variability. To change the attribution of colors, either go to the source code to improve options, or rename whatever column containing the factor of interest into "sampling_time" prior to sending the metadata into the pipeline.

```{r}
#| label: fig-QC-std-all-nh4
#| fig-cap: "QC for Standard curves. We se low intra-batch but higher inter-batch variability. Seeing this, I'd actually recommend considering increasing incubation time or concentrations: absorbance is very low"

# Choice of color palette

#display.brewer.all(n = 3)
color_time <- brewer.pal(n = 3, "Accent")
names(color_time) <- c("t1", "t2", "t3")

plot_qc_std_all(
  data = std_tidy |> filter(N_sp == "NH4"),
  metadata = N_all_metadata |> filter(std_sp == "NH4"),
  color_time = color_time,
  pipetting_direction = "top_down")

```

```{r}
#| label: fig-QC-std-all-no2
#| fig-cap: "QC for Standard curves. We se low intra- and interbatch variability. Seeing this, I'd actually recommend considering increasing incubation time or concentrations: absorbance is very low"


plot_qc_std_all(
  data = std_tidy |> filter(N_sp == "NO2"),
  metadata = N_all_metadata |> filter(std_sp == "NO2"),
  color_time = color_time,
  pipetting_direction = "top_down")
```

```{r}
#| label: fig-QC-std-all-no3
#| fig-cap: "QC for Standard curves. We se low intra- and interbatch variability. Seeing this, I'd actually recommend considering increasing incubation time or concentrations: absorbance is very low"


plot_qc_std_all(
  data = std_tidy |> filter(N_sp == "NO3"),
  metadata = N_all_metadata |> filter(std_sp == "NO3"),
  color_time = color_time,
  pipetting_direction = "top_down")
```

## 4.2 - Perform linear model and infer slope

The next chunk computes the regression on the data from the standard curve, to obtain, for each plate: the slope, the multiple R-squared and the p-value associated to the linear model. It also plots all curves in a series of successive multi-plots (\~pages) that are saved individually into a pdf (if the argument `save_pdf` is set to `TRUE`.). Only one of those plots is rendered here for illustration. The nb of plots per page can be modulated with the argument `max_nb_plots` (set to 16 by default).

```{r}

std_reg <- std_regression(
  data = corrected_data, 
  metadata = N_all_metadata,
  std_data = std_tidy,
  pipetting_direction = "top_down",
  include_density = TRUE,
  signif_p_val = 0.05,
  max_nb_plots = 9,
  save_pdf = TRUE, 
  filepath = "output/figures/QC/")

# Here we have the simplified output of the linear model
std_reg$lm_output

# plots of all regressions are saved as pdf. But just for illustration, let's look at one
std_reg$multi_plots[[1]]

# we can also see a density curve for the p-values
std_reg$density_p_val + 
  labs(title = "Density curve of p-values from all regression curves from all plates")

```

## 4.3 - Compute concentrations in N species

Finally, we can transform the data, using the slope obtained above to compute the concentration. For now, this pipeline was meant for dosage of N fractions, all going through the dosage of NO2-, NO3- or NH4+.

--\> For dosing anything else, the source code will need to be adapted, but it should be fairly simple.

The function `abs_to_mgN_L` first transforms corrected absorbances into concentration in mg of N species per L (e.g., mg NH4+/L), which is then converted to the next unit: mg N/L. The output of the function is multiple: it contains the transformed data in the same vertical format as has been used throughout this pipeline. Finally, the function also outputs 2 plot formats to have a rough visualization of the range of the transformed data.

::: {#tip-gross-no3 .callout-tip}
## NO3 not final yet

The concentration in mg N/L for NO3 is at this stage still a gross measurement that also contains the amouns of NO2 that was present in the sample but was oxidised to NO3. In theory we have to make a substraction (NO3 neat = NO3 gross - NO2). But

-   we can only do this once we've agregated data on each sample (so an average of the 4 wells = technical replicates)

-   in practice we see that concentrations in NO2 are so low that it's ok, in first approximation, to have a look at this value for now.
:::

To finally convert these numbers from mg N /L into mg N / g dry soil, we need to integrate the 2 variables from external data sets: soil dry matter and the soil:exctractant ratio. This is thus something for another script. We can nevertheless have a first look at plots and see that indeed NO2 is mostly at 0, there is some noise in NH4, and clear variations in NO3.

```{r}
#corrected_data |> filter(plate_map == "16_t2_CFE.1x")

data_transformed <- abs_to_mgN_L(
  data = corrected_data,
  metadata = N_all_metadata,
  lm_output = std_reg$lm_output)

# check out the first result
data_transformed$data

# the next 2 will be pu in separate chunks to get a figure nb
```

```{r}
#|label: fig-all-conc-boxplot
#|fig-cap: "Overview of all transformed data per N species - boxplot"

data_transformed$boxplot

```

```{r}
#|label: fig-all-conc-density
#|fig-cap: "Overview of all transformed data per N species - density curves"

data_transformed$density
```

Let us now format then export the table for later use. We will need the last step of computation to happen per sample, so that technical reps (wells) should be pivotted onto a single line. We'll have a lot of NAs bc each plate will only have values at either A, B, C, D or E, F, G, H. We could find a way to fuse it probably...

```{r}

# test <- data_transformed$data |>
#     mutate(
#       rep_tech = case_when(
#         row %in% c("A", "E") ~ "rt1",
#         row %in% c("B", "F") ~ "rt2",
#         row %in% c("C", "G") ~ "rt3",
#         row %in% c("D", "H") ~ "rt4"
#       )
#     ) |>
#     #filter(plate_id == "NH4_1F1") |>
#     select(!(c(well_id:unique_well_id, conc_mgNsp_L)))
# 
# bad <- test |> ungroup() |>
#   dplyr::summarise(n = dplyr::n(), .by = c(plate_id, N_sp, plate_map, rep_tech)) |>
#   dplyr::filter(n > 1L) |> arrange(plate_map)

# pivot for export
  data_export <- data_transformed$data |> 
    mutate(
      rep_tech = case_when(
        row %in% c("A", "E") ~ "rt1",
        row %in% c("B", "F") ~ "rt2",
        row %in% c("C", "G") ~ "rt3",
        row %in% c("D", "H") ~ "rt4"
      )
    ) |> 
    #filter(plate_id == "NH4_1F1") |> 
    select(!(c(well_id:unique_well_id, conc_mgNsp_L))) |> ungroup() |> 
    
    # the next 3 lines should be activated in case we get an error saying that there are duplicates observations (typically there are either 2x the same sample in one plate, or there is a mistake in the plate map). Once the problem is solved, these 3 lines can be deactivated again.
    # ungroup() |> 
    # dplyr::summarise(n = dplyr::n(), .by = c(plate_id, N_sp, plate_map, rep_tech)) |>
    # dplyr::filter(n > 1L)
    
    pivot_wider(
      id_cols = c(plate_id, N_sp, plate_map),
      names_from = rep_tech,
      values_from = conc_N_L,
      names_prefix = "conc_N_L"
      
    )
#data_export
#data_export |> ungroup() |> select(N_sp) |> unique()

```

# 5 - Export

```{r}
#| lst-label: lst-export
#| lst-cap: "Encode path and file name for export"

std_reg$lm_output |> write_rds("output/data/Nmin_std_curves_lm.rds")
data_export |> write_rds("output/data/Nmin_conc.rds")

```

# °^°°°^ Milestone : all data ready for downstream analysis °^°°°^
