---
title: "2.2 From Absorbance to Concentration with a Polynomial Model (incl QC) "
author: "Morgane de Toeuf"
format: 
  gfm:
    toc: true
    #code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# TO DO

- Check how high is the highest absorbance value of 2x diluted samples for NO3 –\> can we remove H row?

- Check why NO2_TDN_19, wells H6, H7 and H9 have abs = 0

  - not important wells, but if it is due to a failure in raw data, there may be other consequences

- Consider adding a function to plate2N to question the best std model

- Make a function to plot residuals of linear vs polynomial model: copy code already there, put it in a loop and save plots in a list –\> return list and use multiplot approach

# Set up

Loading packages

```{r}

rm(list = ls())

library(tidyverse)
library(plate2N)
library(roperators)
library(patchwork)
```

Loading data

```{r}
all_raw_abs_tidy <- read_rds("output/data/1_all_raw_abs_TDN.rds")
all_plate_metadata <- read_rds("output/data/1_all_plate_metadata_TDN.rds")
```

Joining plate data and metadata

```{r}
raw_meta <- all_raw_abs_tidy |> 
  left_join(all_plate_metadata, by = join_by(dataset, plate_id)) |> 
  mutate(date = as.Date(date, tryFormats = c("%d/%m/%Y")))

```

# 1 - Suspicious wells removal

## 1.1 - Manual records

This section allows the removal of wells that "we know" are failed wells (e.g., something went wrong during pipetting...).

For now, there are no such wells for the TDN dataset

```{r}
#| eval: false
#| include: false
#(failed_wells <- read_csv("raw_data/failed_wells.csv", show_col_types = FALSE))

#raw_abs_tidy <- raw_meta |> remove_wells(failed_wells)
#raw_abs_tidy |> filter(plate_id == string, map == "Std")
```

To keep consistent with object names, we create a new tidy table

```{r}
raw_abs_tidy <- raw_meta
```

## 1.2 - Suspicious absorbance values (automated)

Observe values for absorbance (iteratively)

```{r}
suspicious_wells <- raw_abs_tidy |> 
  qc_raw_abs(
    min_abs = 0.03, max_abs = 4, 
    plot_col_facet = "std_sp", 
    show_plot = TRUE) 

suspicious_wells |> slice_max(abs, n = 10)
```

For now, I decide to remove those wells as they were just meant for testing anyway. To be reviewed

```{r}
raw_abs_ok <- raw_abs_tidy |> remove_wells(suspicious_wells)
#raw_abs_ok|> filter(plate_id == string, map == "Std")
```

Check the QC once more

```{r}
raw_abs_ok |> 
  qc_raw_abs(min_abs = 0.03, max_abs = 3, 
    plot_col_facet = "std_sp", 
    export_plot = "none") 

# Once validated, store last version in a "validated" data
raw_abs_clean <- raw_abs_ok
```

It appears that only the most concentrated wells in the standard curve for TDN (well H1) show absorbance levels above 3. We can later look at those curves and see whether those points are outside of the linear range. Not to worry now, though

# 2 - Correction for blank

## 2.1 - Standard curve

Obtain curve concentrations from metadata

```{r}
curve_concentration <- extract_curve(all_plate_metadata)
```

Extract Std wells, add unique curve ID, then add curve_concentration

```{r}
(std_data <- raw_abs_clean |> 
  extract_std_data() |> 
  select(!std_conc) |> 
  left_join(curve_concentration, by = join_by(row, dataset, plate_id)))
```

Check unstrusted blanks (where the smallest value for a given curve is not in row A (top_down pipetting)

```{r}
std_blank <- raw_abs_clean |> extract_std_blank()
std_blank$untrusted
#blank$all |> filter(plate_id == "NO3_R2R3_1")
```

There are no untrusted wells, yay :-)

```{r}
#| eval: false
#| include: false
# get unique curve id for the curves of interest (containing the suspicious wells)
#curve <- blank$untrusted$unique_curve_id

# all data --> too messy. 
# std_data |> group_by(plate_id, column) |> 
#   plot_std() +
#   facet_grid(dataset~std_sp)
# 
# # Subset: look at suspicious blanks
# std_data |>
#   filter(unique_curve_id %in% std_blank$untrusted$unique_curve_id) |>
#   plot_std(through_origin = FALSE) +
#   facet_wrap(~plate_id, scales = "free")
```

Now that we have all the trusted wells with blank values, we can finally correct absorbance values for the standard curves

```{r}
#| include: false
std_corrected <- 
  blank_correct_abs(
    raw_wells_data = std_data|> 
      ungroup() |> 
      filter_out(row == "A"),
    per_plate_avg_blank = std_blank$average |> rename(blank_avg = blank_avg),
    map_to_exclude = ""
  ) |> 
  # take subset (to access plates according to their nb)
  separate_wider_delim(cols = plate_id, delim = "_", names = c("n_sp", "TDN", "nb", "bla"), too_few = "align_start", cols_remove = FALSE) |> 
  mutate(
    nb = as.integer(nb)) 
```

Because the logic is similar, we will first go into blank-correction of sample data before finalizing work on the standard curves (applying linear regression model)

## 2.2 - Sample wells

First, extract data for wells containing extractant and have a look at its variation

```{r}
extr_data <- extract_extractant(raw_abs_clean)
(blank_avg <- extractant_average(raw_abs_clean) |> 
  arrange(desc(blank_coeff_var_percent)))

plot_blank_var_distrib(blank_avg)
```

We see that one plate has a high coefficient of variation, we will have to look at it individually. Let's set the threshold for the coefficient of variation at 5% (default)

```{r}
threshold <- 5

suspicious_plates <- raw_abs_clean |> 
  qc_raw_extr(suppress_warning = TRUE, max_coeff = threshold)

suspicious_extr <- suspicious_extr(
  raw_abs_clean, suspicious_plate_id = suspicious_plates, max_coeff = threshold)

# check it out
suspicious_extr

# plot outliers
suspicious_extr |> boxplot_outlier_extr(max_coeff = threshold)

```

We have 1 plate that has one obvious outlier well. We will need to remove it manually.

First, we create a small tibble that will serve to construct the tibble of wells to remove

```{r}
plate_ids <- suspicious_extr |> 
  ungroup() |> 
  select(dataset, plate_id) |> unique() 
plate_ids <- plate_ids |> # save numbers for plate order in the plot
  mutate(plate_order = seq(1, nrow(plate_ids)))
```

Then, we create a vector with wells to remove (going through boxplots from top to bottom).

:::: callout-tip
## Manually remove outliers

::: {#tip-input-threshold-extr .callout-tip}
In the following chunk, you need to manually decide which wells to remove, based on the boxplots produced above.

- Make sure to deal appropriately with plates that require 2 outliers or no outlier to be removed (see example below)
:::
::::

```{r}
#** !!! MANUAL INPUT !!! *

# Which plate needs 2 outliers removed?
plate_with_2_outliers <- 3
plate_without_outliers <- 9 # use a number > nb of plates if there is no such plate

# Which wells are outliers? 
well_ids <- c("E8") # use NA for plates without outliers
```

Then we finish constructing the tibble of wells to be removed

```{r}
#nb_plates <- plate_ids |> nrow()

to_remove <- plate_ids |> 
  bind_rows(plate_ids |> filter(plate_order == plate_with_2_outliers)) |> 
  arrange(plate_order) |> 
  mutate(well_id = well_ids) |> 
  filter(plate_order != plate_without_outliers) |> #remove plate without outliers
  select(!plate_order)

```

Checking that we didn't get confused: look at `to_remove` in parallel to the boxplot

```{r}
to_remove
```

Looks good, so we remove it from extractant data and recompute the average

```{r}
extr_data_clean <- extr_data |> 
  remove_wells(to_remove) 

blank_avg_clean <- extractant_average(extractant_data = extr_data_clean) 
```

Check that biggest coeff_var indeed below threshold

```{r}
blank_avg_clean |> arrange(desc(blank_coeff_var_percent)) |> head()
```

Now that we are confident in the per-plate average value of raw absorbance of extractant wells, we can finally blank-correct all sample data

```{r}
#| include: false

samples_corrected <- 
  blank_correct_abs(
    raw_wells_data = raw_abs_clean, 
    per_plate_avg_blank = blank_avg_clean |> rename(blank_avg = blank_avg),
    map_to_exclude = c("empty","Std","extr")) 

```

## 2.3 - All corrected data

Let's just recall all corrected data. We have 2 separate tibbles (because the experimental design was arranged to have separate blanks for the curve and the samples)

```{r}
# Standard curve, blank-corrected and clean
std_corrected 

# Samples, blank-corrected and clean
samples_corrected 
```

# 3 - Compute regression equation (per plate)

::: callout-tip
## Polynomial model for high concentrations

For the TDN data set, standard curve concentrations have been increased ten-fold. This resulted in highly concentrated solutions generating absorbance values above 3.

- To be seen if we can remove the H row, but still, G rows are pretty high as well

- It appears that a polynomial model is more appropriated in this case, as can be seen in the the next chunks (example of a single curve)
:::

```{r}
# take data for a single curve and format it for the plotting
curve <- (std_corrected |> 
            group_by(plate_id, column) |> 
            filter(std_sp == "NO3") |> 
            rename(abs = abs_corrected) |> 
            group_split()
          )[[9]]

# check it out
curve

# compute both models
lm_linear <- lm(abs ~ 0 + std_conc, data = curve)
lm_poly <- lm(abs ~ 0 + std_conc + I(std_conc^2), data = curve)

(sum_linear <- summary(lm_linear))
(sum_poly <- summary(lm_poly))
```

Seeing the summary of both models: both are significant, but the p-value of the coefficient for the second degree term (b in bx\^2) in the polynomial model is \<0.05, which indicates that that term significantly contributes to the model. This becomes also very obvious when we look at the plots (the polynomial model fits a lot better)

```{r}
p_linear <- plot_std(curve, through_origin = TRUE, model = "linear") + 
  theme(legend.position = "none") + labs(title = "Linear model")
p_poly <- plot_std(curve, through_origin = TRUE, model = "poly") + 
  theme(legend.position = "none") + labs(title = "Polynomial model")

p_linear + p_poly
```

Let's look at the Residual plot to confirm this intuition

```{r}
res_linear <- residuals(sum_linear)
res_poly <- residuals(sum_poly)

# plot residuals for linear model
plot(curve$std_conc, res_linear, main = "Residuals Analysis", xlab = "Concentration", ylab = "Residuals", col = "grey30", pch = 16)
# add red line at y = 0
abline(h = 0, col = "red", lty = 2)
# add residuals for polynomial model
points(curve$std_conc, res_poly,  col = "magenta", pch = 15)
# add legend
points(0, y = -0.1, col = "grey30", pch = 16)
text(x = 6, y = -0.1, labels = "linear\nmodel", col = "grey30", adj = 0)
points(0, y = -0.15, col = "magenta", pch = 15)
text(x = 6, y = -0.15, labels = "polynomial\nmodel", col = "magenta", adj = 0)

```

So, we will adopt the polynomial model for the TDN dataset

## 3.1 - QC standard curves - round 1

First, we perform a polynomial model on each NO3 curve individually

(! not sure that NO2 is very relevant, but for NO2, linear model is fine)

```{r}
# Select NO3 & group
grouped_data_NO3 <- std_corrected |> 
  group_by(unique_curve_id) |> 
  filter(std_sp == "NO3")

grouped_data_NO2 <- std_corrected |> 
  group_by(unique_curve_id) |> 
  filter(std_sp == "NO2")

lm_NO3_raw <- lm_std_curve(grouped_data_NO3, model = "poly") 
lm_NO2_raw <- lm_std_curve(grouped_data_NO2, model = "linear")
```

Then we take a subset to examine individually: those curves where the linear model doesn't seem to perform ideally (e.g., non-significant model (p-value \> 0.05), residuals not normally distributed, or heteroscedasticity, or, for polynomial model, p-value of one of the coefficients \> 0.05)

```{r}
# extract all plates where "something" is not perfect 
lm_suspicious_NO3 <- lm_NO3_raw |> suspicious_lm(model = "poly")
lm_suspicious_NO2 <- lm_NO2_raw |> 
  suspicious_lm()

nrow(lm_suspicious_NO3) ; nrow(lm_suspicious_NO2)
```

There are 7 suspicious curves for NO3 and none for NO2. So we just accept NO2 curves without check. Just a quick look at \`lm_NO2_raw"

```{r}
lm_NO2_raw
```

For NO3, we go on with the QC of the 7 suspicious curves. For visual support, we create a list of plots where we store each individual plot of "suspicious" standard curves

```{r}
suspicious_plots_NO3 <- plot_list_lm(
  lm_suspicious_NO3, 
  std_data = std_corrected, 
  model = "poly")

# check one plot out
#suspicious_plots_NO3[[1]]
```

Now we can look at each plot individually.

```{r}
patchwork::wrap_plots(suspicious_plots_NO3, axis_titles = "keep") +
     patchwork::plot_annotation(title = "Plots of suspicious Standard curves")
```

I'm not quite sure that I can or should remove wells here. Indeed, with only 7 points, assessing normality is overstretching. It is probably best to rely on observation?

I decide to keep all points for all curves in this dataset.

But I do want to look at the curves from the last plates (those I did where the concentration was wrong)

```{r}
plot_Mo_curves <- plot_list_lm(
  lm_data = lm_NO3_raw |> tail(n = 7), std_data = std_corrected, model = "poly"
)

patchwork::wrap_plots(plot_Mo_curves, axis_titles = "keep") +
     patchwork::plot_annotation(title = "Plots of suspicious Standard curves")
```

Interestingly, the models are all good, but we are in an upward-facing parabole (which would change the computation of the solution (looking for x based on y –\> take higher value: also the only one that is \> 0)

Still, extrapolation well past those concentrations seem unlikely. Let's see the overplotting later, but probably we have to ignore those plates

## 3.2 - Multiple curve QC

First, let's look at the distribution of p-values and adjusted R\^2 of the std curve regressions

```{r}
NO3_p <- density_lm_param(
  lm_NO3_raw, 
  p_or_r = "p", threshold = 0.05, 
  facetting_std_sp = TRUE, color_std_sp = FALSE)

NO3_adjR2 <- density_lm_param(
  lm_NO2_raw, "adjR2", 0.95, color_std_sp = FALSE 
)

NO2_p <- density_lm_param(
  lm_NO3_raw, 
  p_or_r = "p", threshold = 0.05, 
  facetting_std_sp = TRUE, color_std_sp = FALSE)

NO2_adjR2 <- density_lm_param(
  lm_NO2_raw, "adjR2", 0.95, color_std_sp = FALSE 
)
  

(NO3_p + NO2_p) / (NO3_adjR2 + NO2_adjR2)
```

Now we plot all curves on same plot. I removed "my plates" bc there were polluting the plot. Anyway, we will get rid of those data points!

```{r}
colors <- c("#7FC97F", "#BEAED4", "#FDC086")

std_Sang <- std_corrected |> 
  # take only NO3 (bc polanomial model
  filter(std_sp == "NO3")  |> 
  filter_out(nb >32) 
  

annotation <- std_Sang |> 
  select(plate_id, nb, abs_corrected, std_conc, date) |> 
  slice_max(std_conc, with_ties = TRUE)

# plot
std_Sang |> 
  ggplot(aes(
    x = as.numeric(std_conc), 
    y = abs_corrected, groups = plate_id, 
    colour = date, fill = date)) +
  theme_minimal() + 
  theme(legend.position = "right") +
  geom_smooth(
    formula = y ~ 0 + x + I(x^2), method = "lm", se = TRUE, 
    alpha = 0, linetype = 1, linewidth = 0.15) +
  geom_point(alpha = 0.5) +
 # facet_wrap(~std_sp, scales = "free") +
#  scale_color_discrete(palette = colors[1:2]) +
 # scale_fill_discrete(palette = colors[1:2]) +
  xlab("Standard Concentration [mg N-species / L]") +
  ylab("Blank-corrected absorbance") +
  labs(title = "Inter-plate variability of the Standard Curves") +
  xlim(c(0, 300)) +
  annotate(
    geom = "text",
    x = annotation$std_conc*1.01, 
    y = annotation$abs_corrected, 
    label = annotation$date, size = 2,
    hjust = 0,
    color = annotation$date
  ) 


```

There is some batch effect, which can be due to several things:

- An error in the dilution of the stock solution cannot be excluded, though this alone would result in

  - clear, non-overlapping groups of curves done on the same day (dilution only done 1x per day)

  - no particular gradient based on date (random error in pipetting –\> random pattern)

- The gradient in date can only be explained by a gradient in reaction conditions which should act the same way on samples and standard curves, such as:

  - temperature

  - degradation of reagents (bulk preparation)

- The within day noise is possibly due also to slight changes in conditions

  - incubation time slightly different

  - delay since dilution and reagent preparation

    - is there some altering of the solution once it is thawed?

    - How long since the Grieß reagent has been mixed?

    - Degradation of VCl3?

    - etc.

For now I don't see reasons to worry about those curves, we can proceed to apply the regression equations

## 3.3 - Confirm data and export

Let's store all necessary data into a clean variable name to reduce possible confusion, and let's compute all the plots in a big list, for storage purposes. We then export this as one output data in a single list

```{r}
#| include: false

# remove latest plates with wrong std curves
std_data_clean <- std_corrected |> filter_out(nb > 32)

# Run once more lm with this smaller data
grouped_data_NO3 <- std_data_clean |> 
  group_by(unique_curve_id) |> 
  filter(std_sp == "NO3")

# actually unchanged for NO2, but keep it for consistency
grouped_data_NO2 <- std_data_clean |> 
  group_by(unique_curve_id) |> 
  filter(std_sp == "NO2")

# recompute the model
lm_NO3_clean <- lm_std_curve(grouped_data_NO3, model = "poly") 
lm_NO2_clean <- lm_std_curve(grouped_data_NO2, model = "linear")

# create a full list of plots (not only suspicious plots)
lm_plots_NO3_clean <- plot_list_lm(lm_NO3_clean, std_data_clean, model = "poly")
lm_plots_NO2_clean <- plot_list_lm(lm_NO2_clean, std_data_clean, model = "linear")

lm_output <- list(
  "std_data_clean" = std_data_clean,
  "lm_NO3_clean" = lm_NO3_clean,
  "lm_NO2_clean" = lm_NO2_clean,
  "lm_plots_NO3_clean" = lm_plots_NO3_clean,
  "lm_plots_NO2_clean" = lm_plots_NO2_clean,
  "samples_corrected" = samples_corrected
)


```

Now, we save this into an rds file, clean up the environment and re-import the clean data, so no mistake is possible

```{r}
#| eval: true
#| include: true
lm_output |> write_rds("output/data/2_lm_output_TDN.rds")

rm(list = ls())
lm_output <- read_rds("output/data/2_lm_output_TDN.rds")
```

# 4 - From absorbance to concentration

## 4.1 - Theoretical considerations - polynomial model

Regression equation is `Abs = poly_a * Concentration^2 + poly_b * Concentration`. There is no additional term because we constrained the model to be fitted through the origin. This equation can be transformed into `poly_a * Concentration^2 + poly_b * Concentration - Abs = 0`.

The function `polyroot()` computes the roots to a second (or higher) degree equation of the form `ax^2 + bx + c`. It takes as argument a vector `z` of the type `z = c(c, b, a)`. Its output is a vector with the roots of the equation expressed as a complex number. To extract the real component, we use the function `Re()`.

A second degree equation, by definition, accepts 2 solutions for most values of y and its shape is a parable. If a is negative, it is an downward-facing parable (see plotting above). In the case of our model, we already know that the smallest of the 2 solutions for x will be the one we are looking for, for several reasons:

- we constrained the curve to go through the origin,

- all our points will be in the ascending part of the curve (before its maximum)

- So that if we draw a horizontal line at any absorbance value, its first intersection with the parable (smallest of x1, x2) will be the one within the range of our standard curves (the second one would be in the other half, descending part of the parable)

!!! This applies because we have a downward facing parable. We are anyway in an ascending part. Should the parable, in another dataset, have an upward-facing shape (e.g., very small concentrations), then the reasoning is reverse: we would be in the right-hand half of the parable, thus the highest solution from x1 and x2.

## 4.2 - Application of the model

Polyroot is difficult to apply rowwise for now... Until I find a more elegant way, let's just input the standard definition of the solution to a second degree equation for the polynomial model:

`x = (-b +- sqrt(b^2 - 4*ac)) / (2*a)`, with c = -y

Here, we

- connect regression data to sample absorbance data

- apply the regression equation to go from absorbance to concentration in mg N-sp per L

- convert unit to mg N per L

```{r}
# For NO3
data_NO3 <- lm_output$lm_NO3_clean |> 
  filter(std_sp == "NO3") |> 
  reg_join_abs(
    lm_output$samples_corrected |> filter(std_sp == "NO3"), 
    target_sp = "N") 

data_NO2 <- lm_output$lm_NO2_clean |> 
  filter(std_sp == "NO2") |> 
  reg_join_abs(
    lm_output$samples_corrected |> filter(std_sp == "NO2"), 
    target_sp = "N") 

data_all <- 

data_mg_N_L <- full_join(data_NO3, data_NO2) |> 
  rowwise() |> 
  mutate(
    conc_mgNsp_L = case_when(
      std_sp == "NO2" ~ (abs_corrected / slope),
      std_sp == "NO3" ~ min(c(
        (-poly_b + sqrt(poly_b^2 - 4*poly_a*(-abs_corrected))) / (2*poly_a),
        (-poly_b - sqrt(poly_b^2 - 4*poly_a*(-abs_corrected))) / (2*poly_a)
      ))
    )
  ) |> 
  convert_molec(masses = molar_masses)
  
# Check it out
data_mg_N_L
```

# 5 - Export

```{r}

data_mg_N_L |> write_rds("output/data/2_mgNL_TDN.rds")
```
