---
title: "II.2 Raw data pipeline"
format: 
  gfm:
    toc: true
    code-fold: true
# format: 
#   html:
#     embed-resources: true
#     toc: true
#     toc_float: true
#     code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# Set up

```{r}
library(tidyverse)
```

# 1 - import data

```{r}
# import other "wet lab" raw data
raw_data <- read_csv("../raw_data/2024_raw.csv", show_col_types = FALSE) |> clean_names() |> 
  rename(sample_short = short) 
```

# 2 - Compute new variables

## 2.1 - dry matter content

In this data set, dry matter and water content have not yet been computed. The raw data with 3 technical replicates is still encoded, in separate columns `flush_dm...`\`. In this section, we compute average dry matter and water content for the sample (across technical replicates). We add additional columns with intermediary results:

-   FW~neat~ = FW~gross~ - tare; DW~neat~ = DW~gross~ - tare

-   average (FW, then DW) = sum(3 technical replicates) / 3

-   DM = avg~DW~ / avg~FW~

-   WC = 1 - DM

-   With

    -   FW = fresh weight \[g\]

    -   DW = dry weight \[g\],

    -   DM = dry matter content \[g dry soil / g fresh soil\],

    -   WC = water content \[g water / g [**fresh**]{.underline} soil\]

-   ratio = fresh wait of soil added into tubes for extraction (\~10g) / volume of extractant added (=20ml)

Then, we can finally convert concentrations in mg/L into concentrations in ppm (mg/kg dry soil)

-   see <https://mycloud.ulb.be/index.php/f/30359374> for equation

```{r}
# Volume of extractant used in the K2SO4 extraction
vol_extr <- 20

# 
lab_data <- raw_data |> 
  
  #** °°°°° first, update format of variables °°°°° *
  # mutate(
  #   soil = factor(soil, levels = c("Conv", "Ref", "Auto", "ABC")),
  #   crop_diversity = factor(crop_diversity, levels = c("SC", "IC")),
  #   cs = factor(cs, levels = c("F", "IC", "W")),
  #   bloc = as.factor(bloc)
  # ) |> 
  #** °°°°° Compute dm °°°°° *
  mutate(
    # subtract tare from gross fresh weight
    flush_dm_neat_fw_tr1 = flush_dm_g_fw_tr1 - flush_dm_tare_tr1,
    flush_dm_neat_fw_tr2 = flush_dm_g_fw_tr2 - flush_dm_tare_tr2,
    flush_dm_neat_fw_tr3 = flush_dm_g_fw_tr3 - flush_dm_tare_tr3,
    # subtract tare from gross dry weight
    flush_dm_neat_dw_tr1 = flush_dm_g_dw_tr1 - flush_dm_tare_tr1,
    flush_dm_neat_dw_tr2 = flush_dm_g_dw_tr2 - flush_dm_tare_tr2,
    flush_dm_neat_dw_tr3 = flush_dm_g_dw_tr3 - flush_dm_tare_tr3,
    # compute average fresh and dry weight
    flush_dm_avg_fw = (
      flush_dm_neat_fw_tr1 + flush_dm_neat_fw_tr2 + flush_dm_neat_fw_tr3
      ) / 3,
    flush_dm_avg_dw = (
      flush_dm_neat_dw_tr1 + flush_dm_neat_dw_tr2 + flush_dm_neat_dw_tr3) / 3,
    # compute dry matter content
    dm = flush_dm_avg_dw / flush_dm_avg_fw,
    # compute water content
    wc = 1- dm,
    
    #** °°°°° compute each sample's individual ratio (~10g moist per 20 ml) °°°°° *
    ratio_nf = flush_fw_nf_g / vol_extr,
    ratio_cfe = flush_fw_cfe_g / vol_extr,
    .keep = "unused"
  ) 
 

```

!! For t0, for PMN, water content can only be obtained from the WHC experiment (CHECK!)

# 3 - Export

```{r}
lab_data |> write_rds("output/data/2_lab_data.rds")
```
