---
title: "2.3 Converting MicroResp Data"
author: "Morgane de Toeuf"
format: 
  gfm:
    toc: true
    #code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
bibliography: references.bib
---

# TO DO

# Set up

```{r}
rm(list = ls())

library(plate2N)
library(tidyverse)
```

Load data

```{r}
MR_data <- read_rds("output/data/1_all_raw_abs_tidy_MR.rds")
MR_metadata <- read_rds("output/data/1_all_MR_metadata.rds") |> 
  # add column dataset into metadat to use for left_join()
  mutate(dataset = case_when(
    expe == "Pot" ~ "MR_greenhouse",
    expe == "Field" ~ "MR_field",
    .default = "oups"),
    .before = expe)

```

Join data

```{r}
MR <- MR_data |> left_join(MR_metadata)
```

# 1 - clean data

## 1.1 - QC - suspicious wells

Check first t0

```{r}
MR |> 
  mutate(abs = abs_t0) |> 
  qc_raw_abs(
    min_abs = 0.9, max_abs = 2.5, 
    plot_col_facet = "dataset", 
    show_plot = TRUE, export_plot = "abs_t0_distrib") 
abs_t0_distrib + labs(title = "Distribution of absorbance at t0, all data")
```

For now, it looks fine for both datasets for t0. Let's check t5

```{r}
MR |> 
  mutate(abs = abs_t5) |> 
  qc_raw_abs(
    min_abs = 0.1, max_abs = 2.5, 
    plot_col_facet = "dataset", 
    show_plot = TRUE,
    export_plot = "abs_t5_distrib") 
abs_t5_distrib + labs(title = "Distribution of absorbance at t5, all data")
```

They also look good, no outlier removal needed so far

## 1.2 - remove "empty" and NAs

With MicroResp experiments, edge effects are not rare, so we exclude any data from columns 1 and 12 of the 96-well plate. Those have been encoded in the mapping as "empty" (though they were not physically).

Also, the Excel template was not 100% full, which means that some plates are without absorbance data

```{r}
MR_clean <- MR |> 
  filter_out(map == "empty" | is.na(abs_t0)) |> 
  mutate(abs_t0 = as.numeric(abs_t0), abs_t5 = as.numeric(abs_t5))

MR_clean
```

# 2 - Normalisation

In preliminary works we noticed that `detection_plate_id` was the biggest driver of variability in our dataset. Let's visualize t0 absorbance readings (see plot)

```{r}
MR_clean |> ggplot(aes(x = detection_plate_id, y = as.numeric(abs_t0))) +
  theme_minimal() +
  geom_boxplot() +
  facet_wrap(~dataset, scales = "free_x")
```

So, instead of a "per-plate" normalization, we propose a "per-dataset" normalisation. I.e., instead of

- $normalized_{abs} = \frac{abs_{i,t5}}{abs_{i,t0}} \cdot mean(abs_{plate,t0})$

We get

- $normalized_{abs} = \frac {abs_{i,t5}}{abs_{i,t0}} \cdot mean(abs_{dataset,t0})$

## 2.1 - Compute per-dataset mean of t0

```{r}
avg_t0 <- MR_clean |> 
  # option: keep only "best" plate: M1 and
 # filter(detection_plate_id %in% c("M1", "M28")) |> 
  # remove first and last row for this (bc edge effects)
  filter_out(row %in% c("A", "H")) |> 
  group_by(dataset) |> 
  summarise(dataset_avg_t0 = mean(abs_t0))
```

## 2.2 - Normalize data

```{r}
MR_norm <- MR_clean |> 
  left_join(avg_t0, by = join_by(dataset)) |> 
  relocate(dataset_avg_t0, .after = abs_t5) |> 
  mutate(
    norm = (abs_t5 / abs_t0) * dataset_avg_t0,
    .after = dataset_avg_t0
  )
```

# 3 - Convert to %CO2

%Co2 = -0.2265 + (-1.606)/(1+(-6.771)\*norm)

```{r}
MR_percent_co2 <- MR_norm |> 
  mutate(
    percent_co2 = 
      -0.2265 + 
      ((-1.606) / 
         (1 + 
            ((-6.771) * norm))
       ),
    .after = norm
    )
```

# 4 - Convert to g CO2 / g dry soil

co2_emitted = (((percent_co2/100)x850x(44/22.4)x*(12/44)x*(273/(273+25)))/(soil_per_well_gx(soil_dw/100)))/5

```{r}
MR_co2_g_h <- MR_percent_co2 |> 
  mutate(
    co2_g_h = (
      (
        (percent_co2 / 100) * 850 * (44/22.4) * (12/44) * (273 / (273+25))
        ) / 
        case_when(
          # for std soils
          column %in% c(2,3) ~ 
            (std_soil_well_g * (std_soil_dm_content_percent / 100)),
          column %in% c(4:11) ~ 
            (soil_well_g * (soil_dm_content_percent / 100)))
      ) / 5
  )

MR_co2_g_h
```

Check-out values

```{r}
MR_co2_g_h |> 
  ggplot(aes(x = co2_g_h)) + 
  theme_minimal() +
  #geom_histogram() +
  geom_density() +
  facet_grid(map~dataset, scales = "free_y")
```

::: callout-warning
## Calibration or correction needed !

Until we find a way to correct / calibrate this data, we are probably still going to have some issues with the negative values (we need to compute the log)
:::

# 5 - Per plate & per substrate outlier removal

Subset data.

!! A preliminary run of the following chunks (up to saving the plots) showed that while they are not the only outliers, wells from row A and H are outliers on many plates, possibly a majority of plates.

So I choose to follow the recommendation of our colleagues from the FiBL and remove those wells already in the subset now. If needed, just "comment" the filtering line to check out plots with all data again

```{r}

MR_greenhouse_raw <- MR_co2_g_h |> 
  filter_out(row %in% c("A", "H")) |> 
  filter(dataset == "MR_greenhouse")

MR_field_raw <- MR_co2_g_h |> 
  filter_out(row %in% c("A", "H")) |> 
  filter(dataset == "MR_field")
```

## 5.1 - create QC plots

Create one list of QC plots per dataset

Take a subset with coefficient of variation at 5%?

```{r}
MR_greenhouse_avg <- MR_greenhouse_raw |> 
  group_by(plate_id, map) |> 
  summarize(
    substrate_avg = mean(co2_g_h),
    std_dev = sd(co2_g_h)
  ) |> 
  mutate(coef_var_percent = 100 * std_dev / substrate_avg) |> 
  arrange(desc(coef_var_percent))

MR_greenhouse_avg |> ggplot(aes(x = coef_var_percent)) +
  geom_histogram() +
  xlim(-100, 100)

suspicious_substrates_greenhouse <- MR_greenhouse_avg |> 
  filter(abs(coef_var_percent) > 15) |> 
  select(plate_id, map) |> unique()

suspicious_MR_greenhouse <- MR_greenhouse_raw |> 
  right_join(suspicious_substrates_greenhouse) 


## Repeat for field
MR_field_avg <- MR_field_raw |> 
  group_by(plate_id, map) |> 
  summarize(
    substrate_avg = mean(co2_g_h),
    std_dev = sd(co2_g_h)
  ) |> 
  mutate(coef_var_percent = 100 * std_dev / substrate_avg) |> 
  arrange(desc(coef_var_percent))

MR_field_avg |> ggplot(aes(x = coef_var_percent)) +
  geom_histogram() +
  xlim(-100, 100)

suspicious_substrates_field <- MR_field_avg |> 
  filter(abs(coef_var_percent) > 15) |> 
  select(plate_id, map) |> unique()

suspicious_MR_field <- MR_field_raw |> 
  right_join(suspicious_substrates_field) 
```

Now, with a threshold of 15% as max coef of variation, we are left with only \~200 combinations of plate x substrate to look at (instead of \~5000).

[**!! This means we are ok with not looking at all data !!**]{.underline}

```{r}

source("functions/plot_qc_sample_conc.R")

# qc_plots_greenhouse <- plot_list_qc_MR(MR_greenhouse_raw)
# 
# qc_plots_field <- plot_list_qc_MR(MR_field_raw)

qc_plots_greenhouse <- plot_list_qc_MR(suspicious_MR_greenhouse)

qc_plots_field <- plot_list_qc_MR(suspicious_MR_field)

```

Save QC plots for review outside of this Markdown document

!! The following chunk is slow and is thus deactivated. Manually run it to save plots again

```{r}
#| eval: false
#| include: true

# For greenhouse
for (i in 1:length(qc_plots_greenhouse)) {
  
  plot <- qc_plots_greenhouse[i]
  
  filepath <- paste0(
      "output/figures/QC/greenhouse_MR/",
      names(plot), "_", today(), ".pdf")
  
  ggsave(
    filename = filepath, 
    plot = plot, width = 18, height = 10)
}

# For field
for (i in 1:length(qc_plots_field)) {
  
  plot <- qc_plots_field[i]
  
  filepath <- paste0(
      "output/figures/QC/field_MR/",
      names(plot), "_", today(), ".pdf")
  
  ggsave(
    filename = filepath, 
    plot = plot, width = 18, height = 10)
}
```

## 5.2 - identifying outliers + removal

Then we create a table of wells to remove, and clean the dataset from obvious outliers

Starting with greenhouse outliers

```{r}
to_remove_greenhouse <- tribble(
  ~ plate_id, ~well_id, 
  # Alanine
  "P08", "C10",    "P19", "B10",    "P28", "C10",
  "P35", "F10",    "P43", "G10",    "P50", "B10",
  "P70", "C10",    "P96", "B10",
  # gABA
  "P01", "F9",    "P15", "G9",    "P31", "D9",
  "P35", "G9",    "P39", "E9",    "P43", "G9",
  "P51", "D9",    "P56", "F9",    "P66", "B9",
  "P67", "F9",    "P69", "D9",    "P70", "D9",
  "P84", "D9",    "P93", "D9",
  #Glu
  "P24", "B6",    "P43", "B6",    "P43", "G6",
  "P56", "G6",    "P81", "B6",    "P82", "D6",
  # H2O
  "P02", "B4",    "P06", "C4",    "P08", "B4",
  "P17", "B4",    "P18", "C4",    "P25", "G4",
  "P41", "B4",    "P44", "G4",    "P45", "G4",
  "P46", "B4",    "P51", "C4",    "P60", "C4",
  "P62", "G4",    "P85", "G4",    "P85", "F4",
  "P88", "F4",
  # Lgn
  "P08", "B7",    "P24", "B7",    "P35", "G7",
  "P43", "G7",    "P46", "G7",    "P52", "G7",
  "P70", "G7",    "P83", "B7",    "P88", "C7",
  # NAG
  "P12", "B8",    "P12", "G8",    "P16", "B8",
  "P28", "E8",    "P31", "B8",    "P35", "G8",
  "P52", "E8",    "P56", "B8",    "P67", "E8",
  "P70", "E8",
  # OA
  "P25", "C5",    "P46", "B5",    "P49", "C5",
  "P54", "C5",    "P64", "C5",    "P81", "B5",
  "P90", "G5",
  #Std-Glu
  "P08", "F2",    "P44", "B2",    "P46", "D2",
  "P66", "G2",    
  # Std-H2O
  "P01", "B3",    "P12", "D3",    "P16", "B3",
  "P24", "B3",    "P24", "G3",    "P25", "E3",
  "P29", "D3",    "P33", "D3",    "P44", "G3",
  "P62", "B3",    "P64", "F3",    "P86", "G3",
  "P91", "F3",    "P93", "E3",    
  # Urea
  "P30", "C11",    "P35", "F11",    "P35", "G11",
  "P51", "B11",    "P52", "B11",    "P56", "B11",
  "P78", "D11",    "P92", "E11"
  
  
) |> 
  mutate(dataset = "MR_greenhouse")

to_remove_greenhouse
```

Repeat it with Field outliers

```{r}
to_remove_field <- tribble(
  ~ plate_id, ~well_id, 
  #Ala
  "P18", "B10",    "P22", "E10",    "P26", "B10",
  "P27", "B10",    "P52", "G10",    "P66", "B10",
  "P66", "G10",    "P70", "B10",    "P72", "B10",
  # gABA
  "P10", "B9",    "P31", "C9",    "P34", "C9",
  "P57", "G9",    "P57", "F9",    "P62", "B9",
  "P62", "C9",    "P65", "G9",    "P66", "G9",
  "P74", "B9",
  # Glu
  "P29", "G6",    "P66", "G6",    "P75", "B6",
  "P76", "G6",    "P79", "G6",
  # H2O
  "P04", "D4",    "P37", "G4",    "P38", "D4",
  # Lgn
  "P50", "B7",    "P60", "B7",    "P62", "B7",
  "P75", "B7",    "P80", "C7",
  # NAG
  "P02", "B8",    "P28", "B8",    "P31", "B8",
  "P37", "B8",    "P42", "C8",    "P57", "E8",
  # OA
  "P12", "B5",    "P14", "C5",    "P14", "D5",
  "P17", "B5",    "P26", "B5",    "P31", "B5",
  "P34", "E5",    "P35", "E5",    "P52", "E5",
  "P62", "B5",    "P66", "G5",    "P73", "E5",
  # Std-Glu
  "P50", "G2",    "P61", "B2",    "P74", "B2",
  "P75", "G2",    "P79", "C2",
  # Std-H2O # NADA
  # Urea
  "P23", "F11",    "P33", "D11",    "P46", "C11",
  "P52", "F11",    "P71", "F11",    "P77", "F11"
) |> 
  mutate(dataset = "MR_field")

to_remove_field
```

Remove those outliers

```{r}
MR_greenhouse_wash1 <- MR_greenhouse_raw |> remove_wells(to_remove_greenhouse)

MR_field_wash1 <- MR_field_raw |> remove_wells(to_remove_field)
```

Store last version in a new variable (easier for edits...)

```{r}
MR_greenhouse_clean <- MR_greenhouse_wash1
MR_field_clean <- MR_field_wash1
```

# 6 - Average per plate per substrate

```{r}
MR_clean <- bind_rows(MR_greenhouse_clean, MR_field_clean)

MR_avg <- MR_clean |> 
  group_by(dataset, plate_id, map) |> 
  summarize(
    co2_g_h_avg = mean(co2_g_h)) |> 
  ungroup() |> 
  mutate(
    min_value = min(co2_g_h_avg),
    co2_g_h_shifted = co2_g_h_avg - round(min_value, digits = 3),
    # add info whether Std soil or sample
    std_sample = case_when(
      map %in% c("Std_H2O", "Std_Glu") ~ "std_soil",
      .default = "sample"
    ))
```

[@bongiorno2020] write: *Thereafter, we corrected for the average respiration rate of soil to which the deionized water was added.*

Although not explicit, I believe that what they say if that they consider that substrate-induced respirations as read are indeed the sum of basal respiration + the "real" part of the respiration that is substrate induced. So probably they do a subtraction?

Plot its distribution

```{r}
distrib_uncorrected <- MR_avg |> 
  ggplot(aes(x = co2_g_h_avg)) +
  theme_minimal() +
  geom_histogram() +
  labs(
    title = "Distribution of per plate per substrate average\nof CO2 emission rate",
    x = "CO2 emission rate [gCO2 / g dry soil / hour]")

distrib_corrected <- MR_avg |> 
  ggplot(aes(x = co2_g_h_shifted)) +
  theme_minimal() +
  geom_histogram() +
  labs(
    title = "Distribution of per plate per substrate average\nof CO2 emission rate - Corrected values",
    subtitle = paste0("Correction : shifting by adding the rounded\nminimum value = ", round(unique(MR_avg$min_value), digits = 3)),
    x = "CO2 emission rate [µgCO2-C / g dry soil / hour]")

distrib_uncorrected / distrib_corrected
distrib_corrected + facet_wrap(dataset~map, scales = "free_x")
```

It should now be discussed whether this correction is acceptable or not. We could also use it for some computation but not others.

For example, the issue with current values is that, to compute Shanon index, we need to take the log of those values. by taking x/ln(x). But for other indices we could just take the raw values as is. Also for x/ln(x), would it be an option to take x/ln(corrected_x)?

Let's take a few random values for x, and see how the correction affects the computation

```{r}
#| include: false
(x <- c(seq(0,1.5,0.1)))
x <- x + 0.001
#y <- x - round(unique(MR_avg$min_value), digits = 3)
y <- x + 1 - round(unique(MR_avg$min_value), digits = 3)

df <- tibble(x,y)
log(x)
x/log(x)
y/log(y)
y

df |> ggplot(aes(x = x, y = y)) + geom_point()
df |> ggplot(aes(x = x*log(x), y = y*log(y))) + geom_point()
df |> ggplot(aes(x = x*log(x), y = x*log(y))) + geom_point()
df |> ggplot(aes(x = x, y = x*log(x))) + geom_point()
df |> ggplot(aes(x = x, y = x*log(y))) + geom_point()

```

## 6.1 - Correction of raw averages

I suggest to accept the "translation" correction, i.e., shifting all values by adding the absolute value of the smallest value, rounded up a little so all values are indeed above zero. I find that it is honest, at least for now, until a calibration is possible. I propose to take that correction for all computations.

Then, I follow computations as in [@bongiorno2020a], i.e.:

- Correct SIR's for basal respiration (subtract H2O-resp from substrate-induced resp)

- Compute MSIR

- Compute relative contributions of SIR's (SIR / MSIR)

- Derive from this relative contribution an intermediary value needed for Shanon (x\*ln(x))

- Compute the Shanon index

Then, we can still discuss some N-based index

# 7 - MBC, basal respiration, metabolic quotient, MSIR

First, basal-respiration correction

```{r}
# Extract basal resp data only
basal_resp <- MR_avg |> 
  filter(map %in% c("H2O", "Std_H2O")) |> 
                                  #*** If want unshifted data, change here !! *
  select(dataset, plate_id, std_sample, co2_g_h_shifted) |> 
  rename(basal_resp = co2_g_h_shifted)

# Check it out
basal_resp
```

Now we can join this info to the rest of the data and compute H2O-corrected data

```{r}
H2O_corrected <- MR_avg |> 
  left_join(basal_resp) |> 
  mutate(H2O_corrected = case_when(
    map %in% c("H2O", "Std_H2O") ~ basal_resp,
                                #*** If want unshifted data, change here !!  *
    .default = co2_g_h_shifted - basal_resp
  ))
```

Then we compute the MBC, qCO2 and MSIR (here, we use H2O-corrected data, and don't keep water in the computation. So it is only a sum of non-H2O substrates. Also, we don't keep std soil data from here on, so we start by storing it somewhere (it will be needed for QC)

[**!! TO DO: check where computation of MBC and qCO2 come from + units !!**]{.underline}

```{r}
# pivot to get one column per substrate
wider_data <- H2O_corrected |> 
  pivot_wider(
    names_from = map, 
    values_from = H2O_corrected, id_cols = c(dataset, plate_id))

# extract std_soil_data and store it in an object
std_soil_data <- wider_data |> select(dataset, plate_id, starts_with("Std"))

# compute first round of indicators
MSIR <- wider_data |> 
  # exclude std soil data
  select(!starts_with("Std")) |> 
  # compute Basal resp, MBC, metabolic quotient
  rename(basal_resp = "H2O") |> relocate(basal_resp, .after = everything()) |> 
  mutate(
    MBC = (Glu * 40.04) + 0.37, # not sure where those numbers come from!
    qCO2 = basal_resp * 1000 / MBC # same here
  ) |> 
  # compute MSIR
  rowwise() |> mutate(MSIR = sum(c_across(Ala:gABA)))
```

# 8 - Shannon index

Now, we re-pivot longer to compute on a per-substrate basis again, and directly compute relative SIRs (rSIR) and -x\*ln(x) intermediary value

```{r}
rSIRs <- MSIR |> 
  pivot_longer(cols = c(Ala:gABA), names_to = "substrate", values_to = "SIR") |> 
  mutate(
    rSIR = SIR / MSIR,
    sSIR = -rSIR * log(rSIR))
```

! There is still one warning: we are loosing one data point (possibly still a negative or null value given to the log?)

[**!! TODO : Deal with this later !!**]{.underline}

Now, we can pivot wider again to sum sSIRs on a per sample basis

```{r}
shannon <- rSIRs |> 
  pivot_wider(
    names_from = substrate, names_prefix = "sSIR_",
    values_from = sSIR,
    id_cols = c(dataset, plate_id, basal_resp, MBC, qCO2, MSIR)
  ) |> 
  rowwise() |> mutate(shannon = sum(c_across(sSIR_Ala:sSIR_gABA)))

# Check it out
shannon
```

Gather all data (except std soil data)

```{r}
MR_indices <- MR_clean |> 
  #take relevant categorical variables (no quantitative data)
  select(dataset, plate_id, run_id:short) |> 
  # reduce it to one row per plate
  unique() |> 
  # join it to the quantitative, plate-wise (=sample-wise) agregated quantitative data
  left_join(shannon)
```

We lost

- the data on raw SIR and relative SIR (rSIR), which is stored in rSIRs (just in case)

- and std soil data

```{r}
rSIRs
std_soil_data
```

# °°° !!! START HERE !!! °°°

# 9 - QC - check standard soils

# 10 - Last data wrangling

- remove bloc 1, Ref-IC

- Get rSIRs export ready

- Sep all by field and greenhouse

# 11 - Export

```{r}
MR_indices |> write_rds("output/data/2.3_MR_indices.rds")
```
