---
title: "V. Pipeline for Absorbance data"
format: gfm
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
toc: true
---

# To Do

-   Turn as much as possible into functions

# Intro

For an explanation of the pipeline in plain English, see last section "Algorithm in natural language". It is not 100% up to date, but it shows the main steps of the pipeline

# Code

## 1 - Set up

Loading packages and homemade functions

```{r}
#| output: false
#| message: false
#| warning: false
library(tidyverse)
library(roperators) # to be able to add %ni% for "not in"
library(patchwork) # for wrap_plots and wrap_tables
library(RColorBrewer) # to find and set color palette

source("functions/extract_curve.R")
source("functions/plot_qc_std_all.R")
source("functions/qc1_initial_range_abs.R")
source("functions/qc2_plot_range_abs.R")
source("functions/qc3_nb_curves_per_plate.R")
source("functions/subset_data.R")
source("functions/qc4_std_blanc_variation.R")
source("functions/correct_abs_std.R")
source("functions/qc5_extr_blanc_variation.R")
source("functions/qc6_extr_trusted.R")
```

Loading data

```{r}

# import tidy data and metadata
Nmin_data <- read_rds("output/data/Nmin_tidy.rds")
Nmin_metadata <- read_rds("output/data/Nmin_metadata.rds")

# remove empty wells
Nmin_full <- Nmin_data |> filter(plate_map != "empty")

# extracting data for the standard curves only
#std_data <- extract_std_data(Nmin_data)


```

Set up pipetting direction for the std curve.

[**!! THIS IS SOMETHING TO CHECK / UPDATE BY THE USER**]{.underline}

(although failing to do so will become obvious latest with the std curves that will be plotted as decreasing curves)

```{r}
#| eval: false
#| include: false
# define both options
top_down_pipetting <- LETTERS[2:8]
bottom_up_pipetting <- LETTERS[8:2]

# choose one option (should the whole script become a function, this could be a default argument)
pipetting_direction <- top_down_pipetting
```

## 2 - QC suspicious wells

**--\> Issues a warning if absorbances not in specified range, e.g. \[0.03,1.1\]**

The ideal range for absorbance readings (Beer-Lambert in linear range of relationship between concentration and absorbance) is between 0.1 and 1. But these are not super strict borders. I don't want to send out a warning message too soon, so we take higher values.

This chunk filters out only rows where absorbance is out of range, and returns either a warning (when there are out-of-range values) or a happy message (when there are none). In case of a warning, it also shares the table with suspicious wells, so that the user can take an informed decision.

[**To be thought through:**]{.underline}

-   What are options then? Remove suspicious wells (replace by NAs?) –\> deal with it if we are confronted with it

-   another option could be, instead of returning a table, to return only the min and max values of absorbance and/or the number of wells that are out of range

```{r}

qc1_initial_range_abs(Nmin_data, min_abs = 0.03, max_abs = 1.1)

#> ok with a threshold min of 0.03, but more than 4000 when threshold of 0.05
#out_of_range <- Nmin_full |> filter(row_number() %in% suspicious_rows) 

#> but they're all NH4 or NO2 --> acceptable!
# out_of_range |> filter(
#   str_split_i(plate_id, pattern = "_", 1) %ni% c("NH4", "NO2")
#   )
```

Other approach: Report on min and max well values + distribution so user can evaluate, see @fig-plot_QC_wells

```{r}
#| label: fig-plot_QC_wells
#| fig-cap: "QC of Suspicious wells - Distribution of raw absorbance per N species"
#| fig-subcap: "This figures shows raw numbers of uncorrected absorbance (no blanc correction). It contains all values, including values of blanc wells and values of the standard curve. Low values are expected for NO2-, less so for NH4+"

qc2_plot_range_abs(Nmin_data) +
  labs(subtitle = "Many low values, especially for NH4+ and NO2-")
```

## 3 - Correct absorbance values

Now we correct absorbance values by subtracting blanc values from raw values (absorbance of the light by the solution = absorbance by the blank solution + absorbance by the substance to be quantified)

If the standard curves were prepared in water, then the blanc for the standard curve is the absorbance of the well containing only water of that curve. If it was prepared with the extractant, then the blanc is the mean of the values of the wells where the extractant was added.

[**To be thought through:**]{.underline}

-   If it becomes relevant: make some sort of if condition, based on plate information (`blanc_id` and `std_id`)

-   Make sure that the slice-min part in 2nd next chunk behaves as expected in the case of a tie

### 3.1 - Correct std curves for blanc

For now, this is a separate process to account for the fact that the standard curve was prepared in H2O, not in the extractant (K2SO4 or KCl).

First, we extract the rows containing data related to Standard curves only

```{r}
# done now in next function
```

A typical pipetting error with the automated pipette is to forget to expell the first bit (containing air) before the "real" pipetting starts. In this case, the first well to be pipetted (typically well A1) will receive a wrong amount of reagent, which in turn may impact stoechiometry and volume, thus absorbance reads. The next chunk allows the identification of minimum values within a standard curve that are not situated in the first or last row of the plate (usually the standard curve is pipetted in ascending or descending order).

With this information, for example, we can exclude wells where first row is bigger than second row (issue in pipetting).

--\> In the chunk below we get 3 suspicious curves. So we can exclude those combinations of plate and column from the computation of the average blanc. So we will only take the value from the other std curve. Here, we always pipetted 2 per plate (yay!).

--\> This option of course is not valid in the case where only one standard curve is pipetted per plate. In that case, one option is to check whether absorbance values are fairly constant between plates. If so, it is a fair correction to take the inter-plate average value (or a standardized version of it... we'll cross that bridge when we get to it)

```{r}
# Identify plates, wells, columns where there was an issue: the min value for the Std curve is not in row A or row H (in case pipetting was in the opposite direction...)

#suspicious_blancs <- # in case we need to store it somewhere
# std_data |> 
#   group_by(plate_id, column) |> 
#   slice_min(absorbance) |> 
#   filter(row %ni% c("A", "H")) 
```

Now, we can compute the average blanc values, but disregard those suspicious wells

First, we check that we indeed have 2 columns with std curve on every plate

```{r}

# check that we have 2 columns with Std per plate --> option to remove suspicious blancs
qc3_nb_curve_per_plate(Nmin_data, nb_std = 2)

```

Second, we can take a subset of `std_data` that contains only the rows with blancs, and only those that we trust (normally row A or H only)

```{r}

QC4 <- qc4_std_blanc_variation(Nmin_data, nb_std = 2)
lapply(QC4, print)


#** DECIDE WHAT TO DO WITH THIS INFO. SO FAR, I HAVEN'T HAD THE CASE THAT I NEED TO REMOVE FURTHER WELLS BEYOND THE "UNTRUSTED" ONES. SO CODE FOR THIS STILL REMAINS TO BE WRITTEN, ALTHOUGH INSPIRATION CAN BE TAKEN FROM FURTHER QC STEPS * 
```

Third, we compute the blanc value (average) and return a warning if blanc values show too much variation (in the case of several plate-columns with standard curves)

```{r}


```

If we are troubled by the big variation within plate, we can check out the identified suspicious plates. In this case, I find it not so dramatic. We are just dealing with small values which tend to artificially increase the coefficient of variation (division by small numbers...).

We can look at suspicious blancs in their plate context to decide how bad the situation is.

```{r}


```

Now we can correct the absorbance values for the standard curves.

[**!! HEREUNDER: CANDIDATE FOR TURNING INTO A FUNCTION !! - CORRECTION OF BAD WELLS FOR BLANC OF STD CURVE**]{.underline}

```{r}
std_corrected <- correct_abs_std(Nmin_data)


```

We could add those corrected values back into the main data table, but actually those numbers are only useful to compute the regression equation between corrected absorbance and concentration. For thematic clarity purpose, this will be done in a later section (to keep all work on blancs in one place)

### 3.2 - Correct samples for blanc

First, we extract the rows with extractant

```{r}

```

Then we do some quality check: how big is the variation? Do we have suspicious wells?

Looking at the distribution of absorbance, it appears that some wells have very different scoring, see @fig-hist-extr-coeff-var

```{r}
#| label: fig-hist-extr-coeff-var
#| fig-cap: "Distribution of coefficient of variation of absorbance of extractant (blanc)"

#** Look at this (and the next chunk) iteratively a couple of times to decide where to put the threshold. *
  QC5 <- qc5_extr_blanc_variation(Nmin_data,max_coeff = 3)
  
  QC5$distrib_coeff +
  labs(subtitle = "anything above 5% seems to be an outlier (even above 3%)")


```

Let's prepare a warning for plates containing these outliers

```{r}

```

Let's now have a look at those suspicious plates

```{r}
#| label: fig-suspicious-plates
#| fig-cap: "Distribution of absorbance in suspicious plates. From this we can manually identify then remove outliers"

#** To play around with the maximum coefficient of variation, change argument max_coeff in the function call *
#*
QC5$multiple_plot 


```

We have to remove outliers manually: it is really a personal appreciation that works. Watch out, in the case of plates with several blancs like here, that a bimodal distribution might not be an issue (e.g., plate NO2_R4R5 in @fig-suspicious-plates, although in this case the bimodal aspect comes split accross extractants, but actually with very close values).

Based on visual appreciation, here is the list of plates that we want to correct. One way to do it is to impose, for each plate, a threshold value that we can later use to filter out outlier wells.

(FYI: by default, the function `wrap_plots()` orders plots by row, so that the order of plates in `extr_suspicious` corresponds to the plots read from left to right, then next row, etc.)

In the next chunk, we manually enter a vector of values to use as max threshold (exclude values above it). For plates where we decide to keep all values, we put a value of 1 as threshold.

**!! In case you want to exclude lower values only, then just change the `>` into `<`. But if you want to exclude some upper values, and some lower values, consider updating the code.**

```{r}
#** !! Manually input the threshold values of your choosing (read plots from left to right, then from up to down (rowise reading)) *

# print it to check in which order the plates are
# suspicious_plates

# Or manual input
cut_threshold <- c(0.040, 0.041, 0.039, 0.038, 0.038, 0.041, 0.042, 0.072, 0.09, 0.09)


QC6 <- qc6_extr_trusted(QC5, cut_threshold = cut_threshold)

```

```{r}

```

Now we can use the list of untrusted wells to filter them out of the extractant data, and look at the improved distribution of intra-plate variation.

```{r}
#| label: fig-distrib-variation-extr-improved
#| fig-cap: "Distribution of variation of absorbance of extractant (blanc) after removal of outliers"

QC6$extr_distrib_coeff +
  labs(subtitle = "1 plate is still outlier (above 3%), but much less so than before")
```

Now that we have computed a trusted version of the average of absorbance per plate per "blanc", we can correct sample absorbance values.

```{r}

corrected_data <- correct_abs_samples(Nmin_data, QC6 = QC6)


```

# °^°°°^ Milestone : blanc-corrected data °^°°°^

# 4 - Compute regression equation btw absorbance and concentration

## 4.1 - Quality check of standard curves (NH4 only for now)

First we need to do some quality check of the Standard curve

-   checking that metadata and data have the same nb of plates

-   check that there are no negative values for the corrected absorbance

```{r}
# First check: do we have the same number of plates in data and in metadata?
n_groups(std_corrected) == nrow(Nmin_metadata) # yes :-)

# do we have negative values? (check last column)
std_corrected |> 
  arrange(abs_corrected) # no, nice :-)
```

-   Is it indeed a curve? If yes –\> proceed. But quite probably that some curves are not monotonous (stricly increasing or decreasing)

```         
-   First, identify those curves with `unsorted_curves`.

-   
```

```{r}

# extracted unsorted curves
unsorted_curves <- std_corrected |> 
  group_by(plate_id, column) |> 
  arrange(row) |> 
  summarise(
    suspicious = is.unsorted(abs_corrected),
    .groups = "keep"
  ) |> 
  filter(suspicious) |> 
  mutate(unique_curve_id = paste0(plate_id, "_c", column))

# plot them all
#i = 1
file = Nmin_metadata
plots <- list()
for (i in 1:nrow(unsorted_curves)) {
  
  curve <- std_corrected |> 
    #filter(unique_curve_id == unsorted_curves$unique_curve_id[i]) |> 
    filter(plate_id == unsorted_curves$plate_id[i]) 
  
  conc <- tibble(
    conc = extract_curve(file, N_sp = curve$N_sp[1])[2:8],
    row = pipetting_direction
  )
  
  curve <- curve |> 
    left_join(conc)
  nudge <- (max(curve$conc) - min(curve$conc))/30
  
  plot <- 
    curve |> 
    ggplot(aes(x = abs_corrected, y = conc)) + 
    theme_minimal() +
    geom_smooth(
      method = "lm", se = FALSE, formula = y~x-1,
      color = "grey70") +
    geom_point(color = "grey30", alpha = 1) + 
      annotate(geom = "text", x = curve$abs_corrected, y = curve$conc-nudge,
               label = curve$well_id, size = 3) +
    labs(title = curve$plate_id[1]) 
    
  plots[[i]] <- plot
}

wrap_plots(plots,axis_titles = "collect")


```

Now we can manually encode a vector containing all outlier wells, based on visual appraisal of graphs

```{r}

outlier_wells <- c(NA, "E12", "C1", "C1", "E12", NA, NA, NA)

#unsorted_curves$outlier <- outlier_wells
outlier_curves <- 
  unsorted_curves |> 
    ungroup() |> 
    mutate(
      outliers = outlier_wells,
      unique_well_id = case_when(
        is.na(outliers) ~ NA,
        .default = paste0(plate_id, "_", outliers))
    )
outlier_curves

# correct std data
std_tidy <- std_corrected |> 
  filter(unique_well_id %ni% outlier_curves$unique_well_id)

```

Now that we have a new version of the std data, we can look at all curves again –\> make a function!

```{r}
unsorted_curves <- std_tidy |> 
  group_by(plate_id, column) |> 
  arrange(row) |> 
  summarise(
    suspicious = is.unsorted(abs_corrected),
    .groups = "keep"
  ) |> 
  filter(suspicious) |> 
  mutate(unique_curve_id = paste0(plate_id, "_c", column))

# plot them all
#i = 1
file = Nmin_metadata
plots <- list()
for (i in 1:nrow(unsorted_curves)) {
  
  curve <- std_corrected |> 
    #filter(unique_curve_id == unsorted_curves$unique_curve_id[i]) |> 
    filter(plate_id == unsorted_curves$plate_id[i]) 
  
  conc <- tibble(
    conc = extract_curve(file, N_sp = curve$N_sp[1])[2:8],
    row = pipetting_direction
  )
  
  curve <- curve |> 
    left_join(conc)
  nudge <- (max(curve$conc) - min(curve$conc))/30
  
  plot <- 
    curve |> 
    ggplot(aes(x = abs_corrected, y = conc)) + 
    theme_minimal() +
    geom_smooth(method = "lm", se = FALSE, color = "grey70") +
    geom_point(color = "grey30", alpha = 1) + 
      annotate(geom = "text", x = curve$abs_corrected, y = curve$conc-nudge,
               label = curve$well_id, size = 3) +
    labs(title = curve$plate_id[1]) 
    
  plots[[i]] <- plot
}

wrap_plots(plots,axis_titles = "collect")
```

At this stage, we are satisfied with the curves.

We can visualize all curves for a given N species, to have an idea of inter-curve variability

The next chunk works, but it produces a very difficult to read plot with multiple panels (`r nrow(Nmin_metadata)` actually).

```{r}

# conc <- tibble(
#     conc_nh4 = extract_curve(Nmin_metadata, N_sp = "NH4")[2:8],
#     conc_no2 = extract_curve(Nmin_metadata, N_sp = "NO2")[2:8],
#     conc_no3 = extract_curve(Nmin_metadata, N_sp = "NO3")[2:8],
#     row = pipetting_direction)
# conc

i = 1
metadata = Nmin_metadata
data = std_tidy
plots = list()
for (i in 1:nrow(metadata)) {
  
  plate <- metadata$plate_id[i]
  N_sp <- str_split_i(plate, "_", 1)
  
  conc <- tibble(
    conc = extract_curve(metadata, N_sp = N_sp)[2:8],
    row = pipetting_direction
  )
  conc
  
  curve <- data |> 
    filter(plate_id == plate) |> 
    left_join(conc) 
  
  plot <- curve |> 
    ggplot(aes(x = abs_corrected, y = conc, color = column)) + 
    theme_minimal() +
    geom_smooth(method = "lm", se = TRUE, formula = y~x-1, color = "grey70", alpha = 0.2) +
    geom_point(color = "grey30", alpha = 1) + 
      annotate(geom = "text", x = curve$abs_corrected, y = curve$conc-nudge,
               label = curve$well_id, size = 3) +
    labs(title = curve$plate_id[1])
  
  plots[[i]] <- plot
}

wrap_plots(plots,axis_titles = "collect")
```

Let's find a neater way to look at it by overplotting, using the function `plot_qc_std_all`.

First, for NH4+ (@fig-QC-std-all-nh4), then for NO2- (@fig-QC-std-all-no2) and for NO3- (@fig-QC-std-all-no3)

```{r}
#| label: fig-QC-std-all-nh4
#| fig-cap: "QC for Standard curves. We se low intra-batch but higher inter-batch variability. Seeing this, I'd actually recommend considering increasing incubation time or concentrations: absorbance is very low"

# Choice of color palette

#display.brewer.all(n = 3)
color_time <- brewer.pal(n = 3, "Accent")
names(color_time) <- c("t1", "t2", "t3")

plot_qc_std_all(
  data = std_tidy |> filter(N_sp == "NH4"),
  metadata = Nmin_metadata |> filter(std_sp == "NH4"),
  color_time = color_time,
  pipetting_direction = top_down_pipetting)

```

```{r}
#| label: fig-QC-std-all-no2
#| fig-cap: "QC for Standard curves. We se low intra- and interbatch variability. Seeing this, I'd actually recommend considering increasing incubation time or concentrations: absorbance is very low"


plot_qc_std_all(
  data = std_tidy |> filter(N_sp == "NO2"),
  metadata = Nmin_metadata |> filter(std_sp == "NO2"),
  color_time = color_time,
  pipetting_direction = top_down_pipetting)
```

```{r}
#| label: fig-QC-std-all-no3
#| fig-cap: "QC for Standard curves. We se low intra- and interbatch variability. Seeing this, I'd actually recommend considering increasing incubation time or concentrations: absorbance is very low"


plot_qc_std_all(
  data = std_tidy |> filter(N_sp == "NO3"),
  metadata = Nmin_metadata |> filter(std_sp == "NO3"),
  color_time = color_time,
  pipetting_direction = top_down_pipetting)
```

## 4.2 - Perform linear model and infer slope

-   Regression & compute concentrations

    -   loop per plate:

        -   extract plate

        -   compute regression

        -   store coefficients, R2 and p-val in a df

    -   QC after loop

```{r}
#i = 1
metadata = Nmin_metadata
data = Nmin_corrected
std_data = std_tidy
plots <- list()
lm_output <- tibble(
  plate_id = character(), slope = double(), p_val_slope = double(), r_squared_mult = double()
)
for (i in 1:nrow(metadata)) {
#for (i in 1:6) {

  plate <- metadata$plate_id[i]
  N_sp <- str_split_i(plate, "_", 1)
  
  conc <- tibble(
    conc = extract_curve(metadata, N_sp = N_sp)[2:8],
    row = pipetting_direction
  )
  #conc
  
  curve <- std_data |> 
    filter(plate_id == plate) |> 
    left_join(conc, by = join_by(row)) 
  #curve
  
  meta_line <- metadata |> filter(plate_id == plate)
  
  curve_lm <- lm(data = curve, conc ~ 0 + abs_corrected) |> summary() 
  #curve_lm

  lm_coeff <- curve_lm$coefficients |> as.data.frame() |> as_tibble()
  
  names(lm_coeff) <- c(
  #"rowname", # not needed when fitted to go through origin
  "Estimate", "std_error", "t_value", "p_value_slope")
  #lm_coeff
  
  slope = lm_coeff$Estimate |> signif(digits = 3)
  p_val_slope = lm_coeff$p_value_slope |> signif(digits = 3)
  r_squared_mult = curve_lm$r.squared |> signif(digits = 3)
  
  color_p_val <- case_when(
    p_val_slope > 0.05 ~ "red",
    .default = "black"
  )
  
  size_p_val <- case_when(
    p_val_slope > 0.05 ~ 4,
    .default = 2.8
  )
    unit <-  meta_line |> 
      select(std_unit) |> magrittr::extract2(1)
    extractant <- meta_line |> select(extractant_sp, extractant_unit, extractant_conc) 

  # Plot it
  std_curve <-  curve |> 
    ggplot(aes(x = abs_corrected, y = conc)) + 
    theme_minimal() +
    geom_smooth(method = "lm", color = "grey30", formula = y~x-1) +
    geom_jitter(alpha = 0.5) +
    labs(
      title = plate,
      subtitle = paste("slope = ", slope, "\nMultiple R-squared = ", r_squared_mult),
      caption = paste0(
        "extracted in ", 
        extractant$extractant_conc[1], extractant$extractant_unit[1],
        " ",
        extractant)) +
    ylab(paste0("Concentration of ", curve$N_sp[1], "\n[", unit, "]")) +
    xlab("Blanc-corrected absorbance") +
    annotate(
      geom = "text", 
      x = median(curve$abs_corrected),
      y = max(curve$conc),
      label = paste0("p-value of\nslope = ", p_val_slope),
      color = color_p_val,
      fontface = "bold",
      size = size_p_val)
  std_curve
  
  plots[[plate]] <- std_curve 
  #plots[[plate]]
  
  lm_output <- bind_rows(
    lm_output,
    tibble(
      plate_id = plate,
      slope = slope,
      p_val_slope = p_val_slope,
      r_squared_mult = r_squared_mult
      )
  )
#i = i +1
}

max_nb_plots <- 16
nb_of_iterations <- length(plots) %/% max_nb_plots
rest_to_plot <- length(plots) %% max_nb_plots

#i = 1
multi_plots = list()
if (nb_of_iterations > 0) {
  for (i in 1:nb_of_iterations) {
    first_plot_index <- (i-1)*max_nb_plots + 1
    last_plot_index <- first_plot_index + max_nb_plots -1
    multi_plots[[i]] <- wrap_plots(plots[first_plot_index:last_plot_index], axis_titles = "collect")
    #i = i +1
  } 
  if (rest_to_plot > 0) { 
    first_plot_index <- (i-1)*max_nb_plots + 1
    last_plot_index <- first_plot_index + rest_to_plot -1
  multi_plots[[i+1]] <- wrap_plots(plots[first_plot_index:last_plot_index], axis_titles = "collect")
  }
} else {
  first_plot_index <- 1
  last_plot_index <- rest_to_plot
  multi_plots[[1]] <- wrap_plots(plots[first_plot_index:last_plot_index], axis_titles = "collect")
}

# Now save the QC somewhere
filenames <- "QC_singles_page"
filepath <- "output/figures/QC/"

#i = 1
for (i in 1:length(multi_plots)) {
single_file <- paste0(filepath, filenames, i, ".pdf")

#dev.new()
#pdf(single_file, width = 8, height = 6)

ggsave(plot = multi_plots[[i]], single_file, width = 12, height = 8)

#dev.off()
}

# Validate p-values of all curves
bad_p_val <- lm_output |> arrange(desc(p_val_slope)) |> filter(p_val_slope >= 0.05)

if (nrow(bad_p_val) == 0) {
  message(paste0(
    "!! YAY !!\nThe linear model is significative for all plates (p-value < 0.05). You can proceed with the inference of concentrations."
  ))
} else {
  warning(paste0(
    "!!!!! Watch out !!!! The linear model is not significative at p-value < 0.05 for ",
    bad_p_val, 
    " curve",
    if (nrow(bad_p_val > 1)) {"s"},
    ".\nCheck out the table hereabove to identify suspicious plates and decide what to do."
  ))
}

```

## 4.3 - Compute concentrations in N species

Setting up parameters. We will need molar masses

```{r}
# N and N-species molar masses [g/mol]
# n_molar_g_mol <- 14.0069
# no3_molar_g_mol <- 62.0051
# no2_molar_g_mol <- 46.0057
# nh4_molar_g_mol <- 36.0775

molar_masses <- c(
  "N" = 14.0069,
  "NO3" = 62.0051,
  "NO2" = 46.0057,
  "NH4" = 36.0775
)


```

First, we compute `Nmin_regressed`, a data frame with the computed concentration expressed in mg Nsp per L (eg., mg NH4+ per L).

**!! the concentration in mg N/L for NO3 is as this stage still a gross measurement that also contains the amouns of NO2 that was present in the sample but was oxidised to NO3. In theory we have to make a substraction (NO3 neat = NO3 gross - NO2). But**

-   **we can only do this once we've agregated data on each sample (so an average of the 4 wells = technical replicates)**

-   in practice we see that concentrations in NO2 are so low that it's ok, in first approximation, to have a look at this value for now.

```{r}

#i = 1
metadata <- Nmin_metadata
data <- Nmin_corrected |> group_by(plate_id)

# create a table to be iteratively completed
Nmin_regressed <- data |> filter(FALSE) |> mutate(conc_mgNsp_L = double(), conc_N_L = double())
for (i in 1:n_groups(data)) {
  # get plate details
  plate <- metadata$plate_id[i]
  N_sp <- str_split_i(plate, "_", 1)
  slope <- lm_output |> 
    filter(plate_id == plate) |> 
    select(slope) |> magrittr::extract2(1)
  
  # compute concentration in mg Nsp per L
  plate_data <- data |> 
    filter(plate_id == plate) |> 
    mutate(
      conc_mgNsp_L = slope * abs_corrected,
      conc_N_L = conc_mgNsp_L * molar_masses["N"] / molar_masses[N_sp]
        )
  Nmin_regressed <- bind_rows(Nmin_regressed, plate_data)
  #i = i+1
}

```

To finally convert these numbers from mg N /L into mg N / g dry soil, we need to integrate the 2 variables from external data sets: soil dry matter and the soil:exctractant ratio. This is thus something for another script. We can nevertheless have a first look at plots and see that indeed NO2 is mostly at 0, there is some noise in NH4, and clear variations in NO3.

```{r}
Nmin_regressed |> 
  ggplot(aes(x = plate_map, y = conc_N_L)) +
  theme_minimal()+
  geom_boxplot() +
  facet_wrap(~N_sp, nrow = 3)

Nmin_regressed |> 
  ggplot(aes(x = conc_N_L)) +
  theme_minimal()+
  geom_density() +
  facet_wrap(~N_sp, nrow = 3, scales = "free_y", strip.position = "right")

```

Let us now format then export the table for later use. We will need the last step of computation to happen per sample, so that technical reps (wells) should be pivotted onto a single line. We'll have a lot of NAs bc each plate will only have values at either A, B, C, D or E, F, G, H. We could find a way to fuse it probably...

```{r}

Nmin_conc <- Nmin_regressed |> 
  mutate(
    rep_tech = case_when(
      row %in% c("A", "E") ~ "rt1",
      row %in% c("B", "F") ~ "rt2",
      row %in% c("C", "G") ~ "rt3",
      row %in% c("D", "H") ~ "rt4"
      )
    ) |> 
  #filter(plate_id == "NH4_1F1") |> 
  select(!(c(well_id:unique_well_id, absorbance:conc_mgNsp_L))) |> 
  
  # the next 3 lines should be activated in case we get an error saying that there are duplicates observations (typically there are either 2x the same sample in one plate, or there is a mistake in the plate map). Once the problem is solved, these 3 lines can be deactivated again.
  # ungroup() |> 
  # dplyr::summarise(n = dplyr::n(), .by = c(plate_id, N_sp, plate_map, rep_tech)) |>
  # dplyr::filter(n > 1L)
  
  pivot_wider(
    id_cols = c(plate_id, N_sp, plate_map),
    names_from = rep_tech,
    values_from = conc_N_L,
    names_prefix = "conc_N_L"
    
  )


```

Export

```{r}
#| label: export

lm_output |> write_rds("output/data/Nmin_std_curves_lm.rds")
Nmin_conc |> write_rds("output/data/Nmin_conc.rds")

```

# °^°°°^ Milestone : all data ready for downstream analysis °^°°°^

# °°° —- START HERE — °°°°

# °°° —- Below this: draft, to be picked up — °°°°

At this point, each plate needs to be evaluated. This could go in another script. In the case where there is a standard curve (anything but MicroResp), we could store everything above in one or more function, then code an iterative process to go through each plate with those functions while

-   storing the corrected absorbance data and append it to a central data table per manip for downstrem computation

-   storing the slope, R-squared and p-values of the models in the original "plate-id" data frame

    -   this could be added to the corrected dataframe, but it adds about 72 times as much data, so better to have just one line per plate as this is per plate information

    -   We could consider adding other information like suspicious wells and so on

-   storing the graphs which probably is the quickest way for a quick assessment

    -   the plots are made so that p-values higher than 0.05 should be spotted directly bc the annotation will appear bigger and in red

Then, after this iterative process, we have everything that we need for the computation of the concentrations and other downstrem calculations

[**To be thought through:**]{.underline}

-   I haven't really considered in great depths how the downstream pipeline would look like for the MicroResp experiment. To be defined. For my data analysis, it is meant for later, so I'll get back to it, but not super soon *a priori*

-   We could consider computing the concentration already at this step, but I like to have a cut here where we first assess all the things to look at (suspicious wells, suspicious standard curves, etc.), before we move on. This is kind of a failsafe to avoid blindly going through the analysis without considering potential issues

## 6 - Exporting data

Until I go through upstream steps (extracting "real" data from original files) and consider the downstream steps more concretely, it is hard to be sure about how / in which format, etc, to export the data. Still, here is a list of items that need to be exported one way or another

# Algorithm in natural language {#algo_natural}

-   Plate info list:

    -   plate number - element \<chr\>

    -   column(s) with std curve(s) - vector \<chr\>

    -   standard identity & unit - vector \<chr\> (element1 = name, element2 = unit)

    -   range of std concentration - vector \<num\>

    -   column(s) with blanc - vector \<chr\>

    -   blanc identity & unit of concentration - vector \<chr\> (element1 = name, element2 = unit)

    -   concentration of blanc - element \<num\>

    -   timestamp - date-time format (?)

    -   wavelength in nm - element \<int\>

-   vectorization of absorbance data, vectors are:

    -   plate number

    -   row

    -   column

    -   raw absorbance

    -   legend (or ID)

-   Correct absorbances for blanc

    -   Return a warning message

        -   when absorbances are below 0.05 or above 1.5

        -   give the number of wells concerned,

        -   give the max and min value of those wells

    -   case when row concerns std curve: correct std absorbances for std blanc

        -   find rows where concentration is zero (slice.min?)

        -   take average of absorbance from those rows

        -   substract to all absorbances of the std the average "zero" value and store it in a new column = "corrected_abs"

    -   case when row concerns samples: Correct sample absorbances for blanc

        -   find rows with blanc

        -   compute variation coefficient

            -   return a warning message when var.coeff \> 30%?

            -   exclude wells based on this? Or human decision?

        -   compute average of blancs

        -   substract that value to all non-standard rows and store it in column "corrected_abs"

-   Compute standard curve

    -   Operation per plate

    -   in plate df, filter rows corresponding to the std
