---
title: "V. Pipeline for Absorbance data"
format: gfm
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
toc: true
---

# To Do

-   export + downstream steps (different options based on analysis)

# Intro

For an explanation of the pipeline in English, see last section "Algorithm in natural language". It is not 100% up to date, but it shows the main steps of the pipeline

# Code

## 1 - Set up

Loading packages

```{r}
library(tidyverse)
library(roperators) # to be able to add %ni% for "not in"
library(patchwork) # for wrap_plots
```

Prep template data: fake tables for the sake of building the next steps of the code. Once that code is working, then we can figure out how to extract real plate data instead of this fake model one.

```{r}

# import tidy data
Nmin_data <- read_rds("output/data/Nmin_tidy.rds")

# remove empty wells
Nmin_full <- Nmin_data |> filter(plate_map != "empty")

```

## 2 - QC suspicious wells

**--\> Issues a warning if absorbances not in specified range, e.g. \[0.03,1.1\]**

The ideal range for absorbance readings (Beer-Lambert in linear range of relationship between concentration and absorbance) is between 0.1 and 1. But these are not super strict borders. I don't want to send out a warning message too soon, so we take higher values.

This chunk filters out only rows where absorbance is out of range, and returns either a warning (when there are out-of-range values) or a happy message (when there are none). In case of a warning, it also shares the table with suspicious wells, so that the user can take an informed decision.

[**To be thought through:**]{.underline}

-   What are options then? Remove suspicious wells (replace by NAs?) –\> deal with it if we are confronted with it

-   another option could be, instead of returning a table, to return only the min and max values of absorbance and/or the number of wells that are out of range

-   

```{r}
#** Make sure empty wells contain NA, otherwise, lots of warning messages? To be tested *
#*

# the threshold can be moved a little
min_abs <- 0.03
max_abs <- 1.1

# initiate data frame that will contain suspicious well ids
suspicious_rows <- c() 
for (i in 1:nrow(Nmin_full)) {
  if (Nmin_full$absorbance[i] < min_abs || Nmin_full$absorbance[i] > max_abs) {
    print(Nmin_full$absorbance[i])
    suspicious_rows <- append(suspicious_rows, i)
    }
}
# Send a warning message
if (!is.null(suspicious_rows)) {
  warning(paste0("Some wells are out of range for absorbance, i.e., not in [", min_abs, "; ",max_abs, "] allowed \nSee table hereabove to identify suspicious wells"))
  
  Nmin_full |> filter(row_number() %in% suspicious_rows)
} else {
  message(paste0("°^° !! YAY !! °^° All wells are in range for absorbance between ", min_abs, " and ", max_abs))
}

# ok with a threshold min of 0.03, but more than 4000 when threshold of 0.05
#out_of_range <- Nmin_full |> filter(row_number() %in% suspicious_rows) 

# but they're all NH4 or NO2 --> acceptable!
# out_of_range |> filter(
#   str_split_i(plate_id, pattern = "_", 1) %ni% c("NH4", "NO2")
#   )
```

Other approach: Report on min and max well values + distribution so user can evaluate, see @fig-plot_QC_wells

```{r}
#| label: fig-plot_QC_wells
#| fig-cap: "QC of Suspicious wells - Distribution of raw absorbance per N species"
#| fig-subcap: "This figures shows raw numbers of uncorrected absorbance (no blanc correction). It contains all values, including values of blanc wells and values of the standard curve. Low values are expected for NO2-, less so for NH4+"
plot_QC_wells <- Nmin_full |> 
  ggplot(aes(x = absorbance)) +
  theme_minimal() +
  #geom_boxplot(aes(x = N_sp, y = absorbance))
  geom_histogram(binwidth = 0.01) +
  geom_vline(aes(xintercept = min_abs), color = "red", alpha = 0.5) +
  geom_vline(aes(xintercept = 0.1), color = "black", alpha = 0.5) +
  geom_vline(aes(xintercept = max_abs), color = "red", alpha = 0.5) +
  annotate(geom = "label", x = min_abs, y = -150, label = min_abs, color = "red", size = 2.5) +
  annotate(geom = "label", x = max_abs, y = -150, label = max_abs, color = "red", size = 2.5) +
  annotate(geom = "label", x = 0.1, y = -150, label = "0.1", color = "black", size = 2.5) +
 # geom_density() +
  facet_wrap(~N_sp, nrow = 3) +
  labs(
    title = "Distribution of absorbance per N species",
    subtitle = "Many low values, especially for NH4+ and NO2-")

plot_QC_wells
```

## 3 - Correct absorbance values

Now we correct absorbance values by subtracting blanc values from raw values (absorbance of the light by the solution = absorbance by the blank solution + absorbance by the substance to be quantified)

If the standard curves were prepared in water, then the blanc for the standard curve is the absorbance of the well containing only water of that curve. If it was prepared with the extractant, then the blanc is the mean of the values of the wells where the extractant was added.

[**To be thought through:**]{.underline}

-   If it becomes relevant: make some sort of if condition, based on plate information (`blanc_id` and `std_id`)

-   Make sure that the slice-min part in next chunk behaves as expected in the case of a tie

### 3.1 - Correct std curves for blanc

For now, this is a separate process to account for the fact that the standard curve was prepared in H2O, not in the extractant (K2SO4 or KCl).

First, we extract the rows containing data related to Standard curves only

```{r}
std_data <- Nmin_full |> 
  # take only plate-columns with standard curves
  filter(
    plate_map == "Std") |> 
  group_by(plate_id)
std_data
```

A typical pipetting error with the automated pipette is to forget to expell the first bit (containing air) before the "real" pipetting starts. In this case, the first well to be pipetted (typically well A1) will receive a wrong amount of reagent, which in turn may impact stoechiometry and volume, thus absorbance reads. The next chunk allows the identification of minimum values within a standard curve that are not situated in the first or last row of the plate (usually the standard curve is pipetted in ascending or descending order).

With this information, for example, we can exclude wells where first row is bigger than second row (issue in pipetting).

--\> In the chunk below we get 3 suspicious curves. So we can exclude those combinations of plate and column from the computation of the average blanc. So we will only take the value from the other std curve. Here, we always pipetted 2 per plate.

--\> This option of course is not valid in the case where only one standard curve is pipetted per plate. In that case, one option is to check whether absorbance values are fairly constant between plates. If so, it is a fair correction to take the inter-plate average value (or a standardized version of it... we'll cross that bridge when we get to it)

```{r}
# Identify plates, wells, columns where there was an issue: the min value for the Std curve is not in row A or row H (in case pipetting was in the opposite direction...)

#suspicious_blancs <- # in case we need to store it somewhere
std_data |> 
  group_by(plate_id, column) |> 
  slice_min(absorbance) |> 
  filter(row %ni% c("A", "H")) 
```

Now, we can compute the average blanc values, but disregard those suspicious wells

First, we check that we indeed have 2 columns with std curve on every plate

```{r}

# check that we have 2 columns with Std per plate --> option to remove suspicious blancs
# nb of std columns to check
nb_std <- 2

# sum for next code returns a number that is double the nb of rows
nb_std_per_plate <- std_data |> 
  summarise(
    min(absorbance),
    n_std = n()/8
    ) 
# check if true
nrow(nb_std_per_plate) == sum(nb_std_per_plate$n_std)/nb_std # TRUE --> ok!
nb_std_per_plate |> ggplot(aes(x = n_std)) + geom_histogram() + labs(title = "if only 1 bin, good news :-)") # also good check
```

Second, we can take a subset of `std_data` that contains only the rows with blancs, and only those that we trust (normally row A or H only)

```{r}

# extract ("slice") only rows with the smallest absorbance
std_blanc_all <- std_data |> 
  slice_min(
    absorbance, # slice min according to the valus in abs
    n = nb_std,  # pick as many rows as the nb of columns with standard curve
    with_ties = FALSE # in case there are ties, it will add extra rows
    ) 
std_blanc_all

# check that we will remove the "correct" suspicious blancs in a moment
std_blanc_trusted <- std_blanc_all |> 
  # remove anything not first or last row of plate <=> suspicious
  filter(row %in% c("A", "H"))

# see that we get 3 less rows now
std_blanc_trusted
```

Third, we compute the blanc value (average) and return a warning if blanc values show too much variation (in the case of several plate-columns with standard curves)

```{r}


#** !! Adapt threshold parameter *

# Change parameter "2" in something computed?

std_blanc_avg <-  std_blanc_trusted |> 
  summarise(
    blanc_avg = mean(absorbance), 
    blanc_sdev = sd(absorbance)) |> 
  mutate(
    blanc_coeff_var_percent = 100 * blanc_sdev / blanc_avg)
std_blanc_avg

# Do we accept this "worse" level of variation?
# Find an automatized way to look at it?
std_blanc_avg |> 
  slice_max(blanc_coeff_var_percent, n = 10)
# coeff variation above 3% are only a few, all below 10%, and only for NO2 and NH4, acceptable

# Warning if values are too divergent (decide what "threshold" is for the coefficient of variation ?)

threshold <- 5 # max coeff_var that we accept [%] 

# in case of several values...
if (#length(std_column) != 1 # replace by something else
  TRUE) {
  # ... and of coefficient of variation > set threshold
  if (max(std_blanc_avg$blanc_coeff_var_percent,na.rm = TRUE) > threshold) {
    # send a warning
    warning(paste0("There are plates showing a big variation in absorbance values the blanc of the standard curve (more than ", threshold, "%).\nPick the most likely values / remove outliers manually.\nSee tables above to judge on values and find suspicious wells"))
    # and show suspicious wells
    std_blanc_avg |> 
      filter(blanc_coeff_var_percent > threshold)
      #slice_max(blanc_coeff_var_percent, n = 10)
  }
}

std_blanc_avg |> 
  slice_max(blanc_coeff_var_percent, n = 10)
```

If we are troubled by the big variation within plate, we can check out the identified suspicious plates. In this case, I find it not so dramatic. We are just dealing with small values...

We can look at suspicious blancs in their plate context.

[**To be thought through:**]{.underline}

-   We could maybe use `patchwork::wrap_table()` to visualize them in the nice way. Maybe to do later...

```{r}

nb_of_plates_to_look_at <- 3

std_blanc_big_coeff <- std_blanc_avg |> 
  slice_max(blanc_coeff_var_percent, n = nb_of_plates_to_look_at)

for (plate in 1:length(std_blanc_big_coeff)) {
  print(std_data |> filter(plate_id == std_blanc_big_coeff$plate_id[plate]))
}
```

Now we can correct the absorbance values for the standard curves.

```{r}


std_corrected <- 
  std_data |>  
    # keep only data that is not from blanc wells
  filter(unique_well_id %ni% std_blanc_all$unique_well_id) |> 
  select(plate_id, well_id, absorbance) |> 
    pivot_wider(names_from = well_id, values_from = absorbance) |> 
    left_join(std_blanc_avg |> select(plate_id, blanc_avg)) |> 
    relocate(blanc_avg, .before = 2) |> 
    pivot_longer(
      cols = !c(plate_id,blanc_avg), 
      names_to = "well_id",
      values_to = "absorbance",
      values_drop_na = TRUE
      ) |> 
  mutate(abs_corrected = absorbance - blanc_avg, .keep = "unused") |> 
  # this will the rows containing the blancs, but with "NA" for obs_corrected (so the right_join will not drop observations when they are missing)
  right_join(std_data) |> 
  # not vital, just for readibility: rearrange column order
  relocate(row, column, well_id, plate_id, unique_well_id, N_sp, plate_map, absorbance)

#std_corrected |> filter(is.na(abs_corrected))

```

We could add those corrected values back into the main data table, but actually those numbers are only useful to compute the regression equation between corrected absorbance and concentration. For thematic clarity purpose, this will be done in a later section (to keep all work on blancs in one place)

### 3.2 - Correct samples for blanc

First, we extract the raws with extractant

```{r}
extr_data <- Nmin_full |> 
  filter(str_split_i(plate_map, "_", 1) == "extr") |> 
  group_by(plate_id, plate_map)
extr_data
```

Then we do some quality check: how big is the variation? Do we have suspicious wells?

Looking at the distribution of absorbance, it appears that some wells have very different scoring.

```{r}

extr_avg <- extr_data |> 
  summarise(
    extr_avg = mean(absorbance),
    extr_sdev = sd(absorbance)) |> 
  mutate(extr_coeff_var_percent = 100 * extr_sdev / extr_avg)
extr_avg

extr_avg |> 
  ggplot(aes(x = extr_coeff_var_percent)) +
  geom_histogram(bins = 100) +
  #geom_density() +
  #geom_boxplot() +
  labs(
    title = "Distribution of absorbance of extractant (blanc)",
    subtitle = "anything above 5% seems to be an outlier (even above 3%)") +
  xlab("intra-plate coefficient of variation [%]")


```

Let's prepare a warning for plates containing these outliers

```{r}

# set coeff max that we want to accept
threshold_coeff <- 3

  if (max(extr_avg$extr_coeff_var_percent,na.rm = TRUE) > threshold_coeff) {
    # store id of problematic plates
    suspicious_plates <- extr_avg |> 
      filter(extr_coeff_var_percent > threshold_coeff) |> 
      select(plate_id) |> as.vector() |> magrittr::extract2(1)
    # send a warning
    warning(paste0("There is a big variation in absorbance values for the blanc  (more than ", threshold_coeff, "%).\nRemove the most unlikely values / remove outliers manually.\nSee tables above to judge on values. \nSuspicious plates are stored in vector called suspicious_plates"))
    # and show suspicious wells
    extr_avg |> 
      filter(extr_coeff_var_percent > threshold_coeff) |> 
      arrange(desc(extr_coeff_var_percent))
  }

suspicious_plates

```

Let's now have a look at those plates

```{r}
#| label: fig-suspicious-plates
#| fig-cap: "Distribution of absorbance in suspicious plates. From this we can manually identify then remove outliers"
extr_suspicious <- extr_data |> 
  filter(plate_id %in% suspicious_plates)
extr_suspicious

plots <- extr_suspicious |> 
  group_map(
    .f = ~ggplot(.x, aes(x = absorbance)) + 
      geom_histogram() +
      theme_minimal() +
      labs(title = .y)
    ) 
wrap_plots(plots, axis_titles = "collect")
```

We have to remove outliers manually: it is really a personal appreciation that works. Watch out, in the case of plates with several blancs like here, that a bimodal distribution might not be an issue (e.g., plate NO2_R4R5 in @fig-suspicious-plates, although in this case the bimodal aspect comes split accross extractants, but actually with very close values).

Based on visual appreciation, here is the list of plates that we want to correct. One way to do it is to impose, for each plate, a threshold value that we can later use to filter out outlier wells.

(FYI: by default, the function `wrap_plots()` orders plots by row, so that the order of plates in `extr_suspicious` corresponds to the plots read from left to right, then next row, etc.)

In the next chunk, we manually enter a vector of values to use as max threshold (exclude values above it). For plates where we decide to keep all values, we put a value of 1 as threshold.

**!! In case you want to exclude lower values only, then just change the `>` into `<`. But if you want to exclude some upper values, and some lower values, consider updating the code.**

```{r}
# print it to check in which order the plates are
suspicious_plates

cut_threshold <- c(0.040, 0.041, 0.039, 0.038, 0.038, 0.041, 0.042, 0.072, 0.09, 0.09)

grouped_rows_suspicious <- group_rows(extr_suspicious)
grouped_rows_suspicious

rep_threshold <- c()
#i = 1
for (i in 1:length(grouped_rows_suspicious)) {
  to_append <- rep(cut_threshold[i], length(grouped_rows_suspicious[[i]]))
  rep_threshold <- append(rep_threshold, to_append)
}

extr_untrusted <- extr_suspicious |> 
  ungroup() |> 
  mutate(cut_threshold = rep_threshold) |> 
  filter(absorbance > cut_threshold)

```

Now we can use the list of untrusted wells to filter them out of the extractant data, and look at the improved distribution of intra-plate variation.

```{r}
#| label: fig-distrib-variation-extr-improved
#| fig-cap: "Distribution of variation of absorbance of extractant (blanc) after removal of outliers"


# filtering out untrusted values
extr_trusted <- extr_data |> 
  ungroup() |> 
  filter(unique_well_id %ni% extr_untrusted$unique_well_id) 
extr_trusted

warning(paste0(
  "From ", 
  nrow(extr_data), 
  " wells in total for extractant, ", 
  nrow(extr_data) - nrow(extr_trusted),
  " have been removed because their absorbance value appeared to be an outlier from a within-plate perspective. \nThis amounts to a removal of ",
  round(100*(nrow(extr_data) - nrow(extr_trusted))/nrow(extr_data), digits = 1),
  "% of extractant wells based on an intervention tolerance threshold of ",
  threshold_coeff,
  "% for the intra-plate coefficient of variation"))

# re-compute extr_average per plate
extr_avg_trusted <- extr_trusted |> 
  summarise(
    extr_avg = mean(absorbance),
    extr_sdev = sd(absorbance),
    .by = c(plate_id, plate_map)) |> 
  mutate(
    extr_coeff_var_percent = round((100 * extr_sdev / extr_avg), digits = 1),
    extr_avg = round(extr_avg, digits = 3),
    extr_sdev = round(extr_sdev, digits = 3)) |> 
  arrange(desc(extr_coeff_var_percent))

plot <- extr_avg_trusted |> 
  ggplot(aes(x = extr_coeff_var_percent)) +
  geom_histogram(bins = 100) +
  #geom_density() +
  #geom_boxplot() +
  labs(
    title = "Distribution of absorbance of extractant (blanc) after removal of outliers",
    subtitle = "2 plates are still outliers (above 3%), but much less so than before") +
  xlab("intra-plate coefficient of variation [%]")

extr_avg_trusted_prettier <- extr_avg_trusted |> 
  mutate(
    absorbance_avg = round(extr_avg, digits = 3),
    coeff_var = extr_coeff_var_percent,
    .keep = "unused") |> 
  select(!extr_sdev)

plot + wrap_table(
  extr_avg_trusted_prettier |> filter(coeff_var > 2.5),
  panel = "full", space = "full")

#extr_avg_trusted

```

Now that we have computed a trusted version of the average of absorbance per plate per "blanc", we can correct sample absorbance values.

```{r}

Nmin_corrected <- 
  Nmin_full |> 
  # keep only sample data
  filter(
    # filter out wells containing Std
    plate_map != "Std",
    # filtr out extractant data
    str_split_i(plate_map, "_", 1) != "extr"
    ) |> 
  select(plate_id, well_id, absorbance) |> 
  # make it wider to have only one row per plate
  pivot_wider(names_from = well_id, values_from = absorbance) |> 
  # so that we can add average of extractant that is valid for all wells of the plate concerned
  left_join(extr_avg_trusted |> select(plate_id, extr_avg)) |> 
  # relocate extractant for easier manipulation afterwards
  relocate(extr_avg, .before = 2) |> 
  # repivot so that computation of corrected absorbance is easy (work with columns)
  pivot_longer(
    cols = !c(plate_id, extr_avg),
    names_to = "well_id",
    values_to = "absorbance",
    values_drop_na = TRUE
  ) |> 
  # correct absorbance, with the keep = "unused" argument, we prevent the 
  # probable downstrem miskate of using absorbance instead of corrected absorbance
  mutate(abs_corrected = absorbance - extr_avg, .keep = "unused") |> 
  # get the rest of the well and plate data again
  right_join(Nmin_full) |> 
  # now we don't need data from std wells or extractant wells, so we can remove them
  filter(
    plate_map != "Std",
    str_split_i(plate_map, "_", 1) != "extr")


```

# °^°°°^ Milestone : corrected data ready for downstream analysis °^°°°^

At this point, we could export the data table for downstream analysis (Microresp, etc), although for most applications, the next step is still needed: computing regression equation

# 4 - Compute regression equation btw absorbance and concentration

# °°° —- START HERE — °°°°

First, visualize the relationship

[**To be thought through:**]{.underline}

-   Now, to make sure that there is no inversion of the standard curve (e.g., we write from smallest to biggest, but we pipette the biggest in row A), I sort both vectors (concentration ans absorbances). But in the case of a pipetting mistake where one value of the curve would be off, we may not realize the issue if it means that the order of wells is reshuffled

-   Propose something more elegant

-   **Maybe best to not reorganize anything. If the correlation is negative, we will spot it in the graph!**

-   Don't have an optimal way (or I am not yet convinced that it is correct) to access the p-value of the model / the slope. The problem is: the number that comes out is not the same as displayed in the summary. But it would be quite long to check summaries of each individual plate...

```{r}
#| eval: false
#| include: false
# create data frame

# first, reorder std curve absorbance as follows
# first per plate-column, then in increasing fashion
#***

regression_data <- tibble(
  # sort concentration in increasing values, repeat it as many times as there are columns with std curve
  conc = rep(std_conc[2:8], length(std_column)),
  abs = std_data$abs[c(2:8, 10:16)]
  #abs = std_corrected$abs_corrected
  )

# Compute linear model. The "0 + " forces lm to be fitted to pass through the origin
lm_data <- lm(regression_data$abs ~ 0 +regression_data$conc) |> summary()

lm_coeff <- lm_data$coefficients |> 
  as.data.frame() |> 
 # rownames_to_column() |> # not needed when fitted to go through origin
  as_tibble()

names(lm_coeff) <- c(
  #"rowname", # not needed when fitted to go through origin
  "Estimate", "std_error", "t_value", "p_value_slope")

slope <- lm_coeff$Estimate |> signif(digits = 3)
p_val_slope <- lm_coeff$p_value_slope |> signif(digits = 3)
r_squared <- lm_data$r.squared |> signif(digits = 3)

# Unsure of this value. But as long as it's low... who cares?  
p_val_lm <- pf(lm_data$fstatistic[1], 
   df1 = lm_data$fstatistic[2],
   df2 = lm_data$fstatistic[3],
  lower.tail = FALSE) |> glimpse()
 
color_p_val <- case_when(
  p_val_slope > 0.05 ~ "red",
  .default = "black"
)

size_p_val <- case_when(
  p_val_slope > 0.05 ~ 5,
  .default = 3.5
)

# Plot it
std_curve <-  regression_data |> 
  ggplot(aes(x = abs, y = conc)) + 
  theme_minimal() +
  geom_smooth(method = "lm", color = "grey30") +
  geom_jitter(alpha = 0.5) +
  labs(
    title = plate_id,
    subtitle = paste("slope = ", slope, "\nMultiple R-squared = ", r_squared),
    caption = paste0("extracted in ", extractant_id[1])) +
  ylab(paste0("Concentration of ", std_id[1], "\n[", std_id[2], "]")) +
  xlab("Absorbance") +
  annotate(
    geom = "text", 
    x = median(regression_data$abs),
    y = max(regression_data$conc),
    label = paste0("p-value of slope = ", p_val_slope),
    color = color_p_val,
    fontface = "bold",
    size = size_p_val)
#std_curve
```

# °^°°°^ Milestone : all data ready for downstream analysis °^°°°^

At this point, each plate needs to be evaluated. This could go in another script. In the case where there is a standard curve (anything but MicroResp), we could store everything above in one or more function, then code an iterative process to go through each plate with those functions while

-   storing the corrected absorbance data and append it to a central data table per manip for downstrem computation

-   storing the slope, R-squared and p-values of the models in the original "plate-id" data frame

    -   this could be added to the corrected dataframe, but it adds about 72 times as much data, so better to have just one line per plate as this is per plate information

    -   We could consider adding other information like suspicious wells and so on

-   storing the graphs which probably is the quickest way for a quick assessment

    -   the plots are made so that p-values higher than 0.05 should be spotted directly bc the annotation will appear bigger and in red

Then, after this iterative process, we have everything that we need for the computation of the concentrations and other downstrem calculations

[**To be thought through:**]{.underline}

-   I haven't really considered in great depths how the downstream pipeline would look like for the MicroResp experiment. To be defined. For my data analysis, it is meant for later, so I'll get back to it, but not super soon *a priori*

-   We could consider computing the concentration already at this step, but I like to have a cut here where we first assess all the things to look at (suspicious wells, suspicious standard curves, etc.), before we move on. This is kind of a failsafe to avoid blindly going through the analysis without considering potential issues

## 6 - Exporting data

Until I go through upstream steps (extracting "real" data from original files) and consider the downstream steps more concretely, it is hard to be sure about how / in which format, etc, to export the data. Still, here is a list of items that need to be exported one way or another

```{r}
#| eval: false
#| include: false
abs_corrected
slope
r_squared
p_val_slope # and possible p_val_lm if we get satisfied about this
# the last parameters, possibly in the form of an appended version of a table with all plate information (actually 1 line per plate, so I guess it can be seen as a vector)
std_curve

```

# Algorithm in natural language {#algo_natural}

-   Plate info list:

    -   plate number - element \<chr\>

    -   column(s) with std curve(s) - vector \<chr\>

    -   standard identity & unit - vector \<chr\> (element1 = name, element2 = unit)

    -   range of std concentration - vector \<num\>

    -   column(s) with blanc - vector \<chr\>

    -   blanc identity & unit of concentration - vector \<chr\> (element1 = name, element2 = unit)

    -   concentration of blanc - element \<num\>

    -   timestamp - date-time format (?)

    -   wavelength in nm - element \<int\>

-   vectorization of absorbance data, vectors are:

    -   plate number

    -   row

    -   column

    -   raw absorbance

    -   legend (or ID)

-   Correct absorbances for blanc

    -   Return a warning message

        -   when absorbances are below 0.05 or above 1.5

        -   give the number of wells concerned,

        -   give the max and min value of those wells

    -   case when row concerns std curve: correct std absorbances for std blanc

        -   find rows where concentration is zero (slice.min?)

        -   take average of absorbance from those rows

        -   substract to all absorbances of the std the average "zero" value and store it in a new column = "corrected_abs"

    -   case when row concerns samples: Correct sample absorbances for blanc

        -   find rows with blanc

        -   compute variation coefficient

            -   return a warning message when var.coeff \> 30%?

            -   exclude wells based on this? Or human decision?

        -   compute average of blancs

        -   substract that value to all non-standard rows and store it in column "corrected_abs"

-   Compute standard curve

    -   Operation per plate

    -   in plate df, filter rows corresponding to the std
