---
title: "2 - Npools convert mg N /L to ppm"
author: "Morgane de Toeuf"
format: 
  gfm:
    toc: true
    code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# To Do

Greenhouse:

- Move the PMN curve to script nb 5 !

- Deal with standard soils.

  - Problem: they have same biological unit nb. –\> a preliminary outlier removal needs to be ran separately, then average per biol_unit_nb, then it can be rejoined...

Field

- Incorporate

  - PNR

  - Nmin t3?

  - MicroResp

- Deal with standard soils.

  - Problem: they have same biological unit nb. –\> a preliminary outlier removal needs to be ran separately, then average per biol_unit_nb, then it can be rejoined...

- Move the PMN curve to script nb 5 ?

# Intro

What failed:

- yield at t2

  - quadrat: mess in the data - misunderstanding from colleagues at the field

  - plant-level: was interesting to compare to RS measurements, but possibly biased as an absolute measure of yield bc in IC: plants were selected for being close to each other -\> possibly not representative of the field

# Set up

```{r}
#| message: false
#| warning: false

# clean environment
#rm(list = ls())

# packages
library(tidyverse)
library(janitor)
```

Load data

```{r}
# Npool concentration data
conc_Nmin_field <- read_rds("output/data/1_field_t2_Nmin_clean.rds")
conc_PMN_field <- read_rds("output/data/1_field_PMN_clean.rds")
conc_TDN <- read_rds("output/data/1_field_TDN_clean.rds")
conc_Nmin_pot <- read_rds("output/data/1_greenhouse_t2_Nmin_clean.rds")
conc_PMN_pot <- read_rds("output/data/1_greenhouse_PMN_clean.rds")

# Lab data containing info on wc, soil:solution ratio, ...
flush_clean <- read_rds("output/data/1_flush_clean.rds")


flush_clean_greenhouse <- read_rds("output/data/1_greenhouse_flush_clean.rds")
flush_clean_field <- read_rds("output/data/1_field_flush_clean.rds")

# raw data
PMN_fw <- read_csv("raw_data/PMN/PMN_fw.csv") |> clean_names()

```

# 1 - Nmin (greenhouse & field)

Join field and greenhouse data to work on them at once

```{r}
conc_Nmin <- conc_Nmin_pot |> bind_rows(conc_Nmin_field)
```

Get data, separate samples from std soil data

```{r}
# Nmin_sample_greenhouse <- conc_Nmin_pot |> filter(biol_unit_nb < 100)
# Nmin_std_greenhouse <- conc_Nmin_pot |> filter_out(biol_unit_nb < 100)
# 
# flush_sample_greenhouse <- flush_clean_greenhouse |> filter(biol_unit_nb < 100)
# flush_std_greenhouse <- flush_clean_greenhouse |> filter_out(biol_unit_nb < 100)
```

Convert samples to ppm (Deal with std soils later)

First, we

- join absorbance data and flush data,

- Compute the N-sp N concentration in ppm (e.g. mg NO3-N / kg dry soil)

- do some data housecleaning

```{r}
#Nmin_sample_greenhouse |> arrange(biol_unit_nb) 

Nmin_ppm <- conc_Nmin |> 
  # join both data tables
  left_join(flush_clean) |> 
  # compute concentration in ppm
  mutate(conc_ppm = conc_mgN_L / (ratio_nf * dm)) |> 
  # remove again the bare soils
  filter_out(is.na(sample_short)) |> 
  ungroup() |> 
  # select relevant columns
  select(
    expe, dataset, std_sp, 
    biol_unit_nb, cs, soil, crop_diversity, bloc, zone,
    dm, wc, conc_ppm) |> 
  arrange(biol_unit_nb)

# Nmin_ppm_sample_greenhouse <- Nmin_sample_greenhouse |> 
#   # join both data tables
#   left_join(flush_sample_greenhouse) |> 
#   # compute concentration in ppm
#   mutate(conc_ppm = conc_mgN_L / (ratio_nf * dm)) |> 
#   # remove again the bare soils
#   filter_out(is.na(sample_short)) |> 
#   ungroup() |> 
#   # select relevant columns
#   select(
#     expe, dataset, std_sp, 
#     biol_unit_nb, cs, soil, crop_diversity, bloc, zone,
#     dm, wc, conc_ppm) |> 
#     # 
#     # !c(
#     # #dataset, 
#     # #expe, 
#     # target_sp:lm_p, 
#     # starts_with("sampl"), starts_with("ratio"),
#     # map, plate_id, conc_mgN_L,
#     # st_dev, coef_var)) 
#   arrange(biol_unit_nb)
# 
# # Check it out
# Nmin_ppm_sample_greenhouse
```

Now, we pivot the data so that each N-sp receives its own column

```{r}
Nmin_ppm_wider <- Nmin_ppm |>
  ungroup() |>  
  pivot_wider(
    names_from = std_sp,
    values_from = conc_ppm,
    names_prefix = "ppm_"
  ) |> 
  # to make obvious that NO3 is still uncorrected
  rename(ppm_NO3_uncorrected = ppm_NO3)

# Nmin_ppm_wider <- Nmin_ppm_sample |>
#   ungroup() |>  
#   select(!c(plate_id, map, st_dev, coef_var)) |> 
#   pivot_wider(
#     names_from = std_sp,
#     values_from = conc_ppm,
#     names_prefix = "ppm_"
#   ) |> 
#   # to make obvious that NO3 is still uncorrected
#   rename(ppm_NO3_uncorrected = ppm_NO3)
```

Then we can compute final Nmin variables:

- correct NO3 value (what is measured is the sum of NO3 and NO2, bc we measure NO2 after reduction of NO3 to NO2)

- Compute Nmin (= NO3 + NO2 + NH4)

- Compute ratio, e.g., NO3/Nmin, NH4/Nmin, NO3/NH4

```{r}
Nmin_all_variables <- Nmin_ppm_wider |> 
  mutate(
    ppm_NO3 = ppm_NO3_uncorrected - ppm_NO2,
    ppm_Nmin = ppm_NO3_uncorrected + ppm_NH4,
    NO3_Nmin = ppm_NO3 / ppm_Nmin,
    NH4_Nmin = ppm_NH4 / ppm_Nmin,
    NO3_NH4 = ppm_NO3 / ppm_NH4
  ) |> 
  # remove uncorrected NO3
  select(!ppm_NO3_uncorrected)

Nmin_all_variables |> select(biol_unit_nb) |> unique()
```

### TO DO: deal with standard soils (missing here)

# 2 - PMN

The PMN data is missing 1 key information: weight of fresh soil in the sub-sample that undergoes the KCl extraction. That is why we imported `PMN_fw` in the set up section.

First, we join field and greenhouse data

```{r}
conc_PMN <- conc_PMN_pot |> 
  # tweak it to get same structure as conc_PMN_field
  select(!c(expe, soil)) |> 
  separate_wider_delim(
    map, delim = "_", 
    names = c("expe", "soil", "incubation_time", "tech_rep"), 
    cols_remove = FALSE
  ) |> 
  mutate(dataset = "PMN") |> 
  bind_rows(conc_PMN_field)
```

## 2.1 - Compute new variables

Here, we

- join PMN_fw (data set containing fresh weight of the subsample that underwent the incubation) with the absorbance derived concentration data in mg N / L
- compute the ratio between weight of fresh soil and volume of extractant (150ml) –\> \~30g/150ml
- Derive the concentration in ppm from mg N/L, ratio and dry matter

```{r}
volume_kcl <- 150

PMN_ppm <- conc_PMN |> 
  left_join(PMN_fw) |> 
  mutate(
    ratio = fw / volume_kcl,
    conc_ppm = conc_mgN_L / (ratio * dm)
  )
```

Now, we want the concentration in separate columns as above for Nmin,

```{r}
PMN_wider <- PMN_ppm |> 
  pivot_wider(
  names_from = std_sp,
  names_prefix = "ppm_",
  values_from = conc_ppm,
  id_cols = c(map, biol_unit_nb, soil, dm, wc, expe, incub_time, tech_rep, fw, ratio)
) |> 
  rename(ppm_NO3_uncorrected = ppm_NO3)
```

Then, we can do the NO3 correction and other computation of variables

```{r}
# We need to quantify the timeline with nb of days
dates_field <- c("2023-12-12", "2023-12-17", "2023-12-26", "2024-01-02", "2024-01-09") |> as.Date()
dates_pot <- c("2023-12-5", "2023-12-12", "2023-12-17", "2023-12-26", "2024-01-02") |> as.Date()

incub_days_field <- dates_field - dates_field[1]
incub_days_pot <- dates_pot - dates_pot[1]
names(incub_days_field) <- PMN_wider$incub_time |> unique() |> sort()
names(incub_days_pot) <- PMN_wider$incub_time |> unique() |> sort()

# Compute all variables
PMN_all_variables <- PMN_wider |> 
  mutate(
    ppm_NO3 = ppm_NO3_uncorrected - ppm_NO2,
    ppm_Nmin = ppm_NO3_uncorrected + ppm_NH4,
    NO3_Nmin = ppm_NO3 / ppm_Nmin,
    NH4_Nmin = ppm_NH4 / ppm_Nmin,
    NO3_NH4 = ppm_NO3 / ppm_NH4,
    incub_day = case_when(
      expe == "Field" ~ incub_days_field[incub_time],
      expe == "Pot" ~ incub_days_pot[incub_time]
    ),
    soil = factor(soil, levels = c("Conv", "Ref", "Auto", "ABC"))
  ) |> 
  # remove uncorrected NO3
  select(!ppm_NO3_uncorrected) #|> 
  # test with only last 3 incub times
  #filter(incub_time %in% c("i2", "i3", "i4"))

#check correspondence incubation vs dates
#PMN_all_variables |> ungroup() |> select(incub_time, incub_day) |> unique()
```

## 2.2 - Plot the curves –\> move this to next script?

Finally, we can plot the curve of the potential mineralisation of Nitrogen (PMN)

```{r}
soil = c("#F8766D", "#7CAE00", "#00BFC4", "#C77CFF")

PMN_plot <- PMN_all_variables |> 
  ggplot(aes(x = incub_day, y = ppm_Nmin, group = soil, colour = soil, fill = soil)) +
  theme_minimal() + 
  geom_point() +
  geom_smooth(method = "lm") +
  xlab("Time since incubation at 28°C [days]") +
  ylab("Mineral Nitrogen in the soil [ppm]") +
  facet_wrap(~expe) +
  scale_color_discrete(soil) +
  scale_fill_discrete(soil) +
  labs(
    title = "Potential Mineralisation of Nitrogen (PMN)",
    subtitle = "Nmin as sum of NO2-, NO3-, NH4+")

PMN_plot

# PMN_plot <- PMN_all_variables |> 
#   ggplot(aes(x = incub_day, y = ppm_Nmin, group = soil, colour = soil, fill = soil)) +
#   theme_minimal() +
#   geom_point() +
#   geom_smooth(method = "lm") +
#   labs(
#     title = "Potential Mineralization of Nitrogen (PMN)",
#     subtitle = "Nmin as sum of NO2-, NO3-, NH4+") +
#   xlab("Time since incubation at 28°C [days]") +
#   ylab("Mineral Nitrogen in the soil [ppm]")

# for annotation: extract data from smooth curve
smooth_data <- ggplot_build(PMN_plot)$data[[2]]

smooth_data |> str()
# get its maximum value to anker the annotation
annotations <- smooth_data |> 
  slice_max(x, by = c(PANEL,group)) |> 
  mutate(
    soil = c(
      levels(PMN_all_variables$soil)[2:4], 
      levels(PMN_all_variables$soil)),
    expe = c(rep("Field", 3), rep("Pot", 4))) 
#c("#F8766D", "#7CAE00", "#00BFC4", "#C77CFF") 

# Also annotate date of humidification
day_water <- as.Date("2023-12-20")
  
PMN_plot_all <- PMN_plot + 
  geom_text(
    data = annotations,
    aes(x = x+0.5, y = y, colour = soil, label = soil), 
    hjust = 0, size = 5) +
  xlim(c(0,30)) +
  theme(legend.position = "none") 

PMN_plot_all
```

Looks nice, but if we facet... The curves... are not linear at all. Probably due to re-humidification in-between:

```{r}

PMN_plot_all + facet_grid(expe~soil)
```

## 2.3 - Corrected PMN

After discussion with PI: keep only points that we trust: 1st one and post-humidification –\> i0, i3, i4

```{r}
# subset data
PMN_trusted <- PMN_all_variables |> 
  filter(incub_time %in% c("i0", "i3", "i4"))
```

[**Plot upcoming in later section?**]{.underline}

# 3 - Derive MBN from TDN

First, as for the other data sets, we need to derive N concentration in ppm, using `conc_mgN_L`, `ratio` and `dm`. For this, we can re-use the flush data

```{r}
# no ratio nor dm in TDN data set
conc_TDN

# ratio and dm in flush data
flush_clean

# make it a field-sample vs field-std version

```

Because samples work by zone, but standard soils do not, we have to "trick" a zone (=bloc) for standard data.

### TO DO: check dilution etc. in the TDN protocol: diluted 2x at hood, but also headspace etc. –\> think it through!

```{r}
TDN <- conc_TDN |> 
  # take only sample data
  #filter(biol_unit_nb < 112) |> 
  # join to flush data to add dm and ratios
  #left_join(flush_sample) |> 
  left_join(
    flush_clean |> 
      mutate(zone = case_when(
        biol_unit_nb == 112 ~ bloc,
        .default = zone))) |> #view()
  # remove bloc 1 again
  filter_out(is.na(bloc)) |> #view()
  # declutter, keep only relevant columns
  select(!c(map, st_dev, coef_var, sample_name)) |> 
  ungroup() |> 
  # translate concentration to ppm
  # here NO3 was dosed, after oxidation of all N-compounds to NO3, so TDN is indeed measured
  # multiply by 2 because samples were diluted 2x before dosage
  mutate(
    conc_ppm = case_when(
      fumigation == "CFE" ~ 2 * conc_mgN_L / (ratio_cfe * dm),
      fumigation == "NF" ~ 2 * conc_mgN_L / (ratio_nf * dm),
      .default = NA)
    ) |> 
  # drop useless columns
  select(!c(conc_mgN_L, dm, starts_with("ratio"), wc)) |> 
  # pivot to get ppm of CFE and ppm of NF in separate columns
  pivot_wider(
    values_from = conc_ppm,
    names_from = fumigation,
    names_prefix = "ppm_"
  ) |> 
  # compute MBN
  mutate(
    MBN_ppm = ppm_CFE - ppm_NF,
    TDN_ppm = ppm_CFE,
    .keep = "unused")

# Check it out
TDN
```

# 4 - Join & complete Npools

Now we

- join both datasets: TDN and Nmin (FYI: no TDN for greenhouse data –\> lots of NA, not to worry)

- Compute new Npool variables

```{r}
Npools_t2 <- TDN |> 
  # remove dataset info because both tables have different info stored
  select(!dataset) |> 
  full_join(
  Nmin_all_variables |> select(!c(dataset, expe)), 
  by = join_by(biol_unit_nb, zone, soil, crop_diversity, cs, bloc)) |> 
  mutate(
    expe = case_when(biol_unit_nb < 80 ~ "Pot", .default = expe),
    Norg_tot = TDN_ppm - ppm_Nmin,
    Norg_non_micro = Norg_tot - MBN_ppm)
```

# 5 - Export

[**!! Check if complete !!**]{.underline}

```{r}
Npools_t2 |> write_rds("output/data/2_Npools.rds")
PMN_trusted |> write_rds("output/data/2_PMN.rds")
```

```{r}
#| eval: false
#| include: false
#*** SAFEKEEP Greenhouse script nb 4 --> TO DO : join yield and Npools *
data_to_export <- Nmin_all_variables |> left_join(yield_clean)
data_to_export |> write_rds("output/data/4_t2_greenhouse_transformed.rds")

```
