---
title: "3.1 Field, t2 - Import and tidy data"
format: 
  gfm:
    toc: true
    code-fold: true
# format: 
#   html:
#     embed-resources: true
#     toc: true
#     toc_float: true
#     code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# To Do

!!! DISPATCH THIS DOC INTO

- CRA yield data –\> join Lab script? Or a different one?

Still missing:

- soil CRA

# Set up

```{r}
#| message: false
#| warning: false

rm(list = ls())

library(plate2N) # for remove_wells
library(tidyverse)
library(janitor)
library(roperators) # for %ni%
library(ggrepel) # for geom_text_repel()
library(ggridges) # for geom_density_ridges()
library(patchwork) # for the "+" layout and plot_layout()

# functions
source("functions/plot_qc_sample_conc.R")
```

# 1 - Import data

## 1.4 - Data from the CRA-W

### 1.4.1 - Yield data

In the row data, there is a column under the headers containing units. So we skip this in the import, re-adding column names

```{r}
colnames <- read_csv("../raw_data/SYCBIO_Froment-feverole_2024_MDeToeuf.csv",n_max = 0) |> clean_names() |> names()
yield_raw <- read_csv(
  "../raw_data/SYCBIO_Froment-feverole_2024_MDeToeuf.csv", skip = 2, col_names = colnames, show_col_types = FALSE)
```

Now we tidy this data set:

- rename –\> translate headers + add unit

- Remove empty rows (missing IC samples in Ref soil)

- Recompute (check) computed values

  - For yield: from kg / plot to t/ha :

    - /1000 : kg –\> t

    - X 10000 : m2 –\> ha (in denominator)

    - /(10 X plot_width) : nb of square meters per plot (width x 10m of harvesting)

- (then keep only recomputed ones, remove original raw)

```{r}
yield_full <- yield_raw |> 
  rename(
    year = annee,
    crop = culture,
    biol_unit_nb = code_placeau,
    harvest_date = date_de_recolte,
    yield_per_plot = rdt_brut_parcelle,
    harvester_width_m = largeur_de_mesure,
    water_content_1_percent = humidite_1,
    water_content_2_percent = humidite_2,
    test_weight_1_kg_per_hl = ps_1,
    test_weight_2_kg_per_hl = ps_2,
    protein_percent_dw = mpt_nx5_7
  ) |> 
  filter_out(is.na(yield_per_plot)) |> 
  rowwise() |> 
  mutate(
    soil = case_when(sd_c == 3 ~ "ABC", sd_c == 2 ~ "Auto", sd_c == 1 ~ "Ref", .default = "which_soil?" ),
    crop = case_when(crop == "Froment" ~ "Wheat", crop == "Féverole" ~ "Faba bean", .default = "which crop?"),
    test_weight_kg_per_hl = mean(c(test_weight_1_kg_per_hl, test_weight_2_kg_per_hl), na.rm = TRUE),
    water_content_percent = mean(c(water_content_1_percent, water_content_2_percent), na.rm = TRUE),
    grain_yield_t_per_ha = yield_per_plot/harvester_width_m,
    grain_yield_15p_t_per_ha = grain_yield_t_per_ha * (100 - water_content_percent) / 85) 

# Check computed values

```

Then check re-computed values: all rounding errors. Still, prefer to compute it here

```{r}
# Water content
sum(yield_full$hum_moy != yield_full$water_content_percent)
yield_full |> filter(hum_moy != water_content_percent) |> select(biol_unit_nb, hum_moy, water_content_percent)

# test weight
sum(yield_full$ps_moy != yield_full$test_weight_kg_per_hl)
yield_full |> filter(ps_moy != test_weight_kg_per_hl) |> select(biol_unit_nb, ps_moy, test_weight_kg_per_hl)

# yield in tons per ha
sum(yield_full$rdt_brut_ha != yield_full$grain_yield_t_per_ha)
yield_full |> filter(rdt_brut_ha != grain_yield_t_per_ha) |> select(biol_unit_nb, rdt_brut_ha, grain_yield_t_per_ha)

#yield 15% humidity
sum(yield_full$rdt_15_percent_ha != yield_full$grain_yield_15p_t_per_ha)
yield_full |> filter(rdt_15_percent_ha != grain_yield_15p_t_per_ha) |> select(biol_unit_nb, rdt_15_percent_ha, grain_yield_15p_t_per_ha)

```

Now, keep only useful columns

```{r}
yield_tidy <- yield_full |> 
  select(
    year, crop, soil, biol_unit_nb, harvest_date, 
    grain_yield_15p_t_per_ha, protein_percent_dw,
    grain_yield_t_per_ha, water_content_percent, test_weight_kg_per_hl)
```

# 4 - Export

```{r}
yield_tidy |> write_rds("output/data/3_field_yield_clean.rds")
```
