---
title: "3.1 Field, t2 - Import and tidy data"
format: 
  gfm:
    toc: true
    code-fold: true
# format: 
#   html:
#     embed-resources: true
#     toc: true
#     toc_float: true
#     code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# To Do

- Import MicroResp
- Import PNR
- Nmin t3?
- soil CRA
- 

# Set up

```{r}
#| message: false
#| warning: false

rm(list = ls())

library(plate2N) # for remove_wells
library(tidyverse)
library(janitor)
library(roperators) # for %ni%
library(ggrepel) # for geom_text_repel()
library(ggridges) # for geom_density_ridges()
library(patchwork) # for the "+" layout and plot_layout()

# functions
source("functions/plot_qc_sample_conc.R")
```

# 1 - Import data

## 1.1 - Non-absorbance data from AELab

### 1.1.1 - Raw data

```{r}
# import other "wet lab" raw data
raw_data_field <- read_csv(
  "../raw_data/2024_raw.csv", show_col_types = TRUE,
  col_types = list(
    Soil = col_factor(),
    crop_diversity = col_factor(),
    CS = col_factor(),
    bloc = col_factor()
  ),
  na = c("", "NA", "Na")
  ) |> clean_names() |> 
  rename(sample_short = short) |> 
  filter(expe == "Field")
```

### 1.1.2 - Lab data on a per plot basis

(biological units = plots of t2, with 3 zones per plot)

```{r}
raw_field_t2_lab <- raw_data_field |> 
  filter(sampling_time == "t2") |> 
  arrange(biol_unit_nb, zone) |> 
  # remove useless columns
  select(
    biol_unit_nb:sample_short, 
    soil:zone,
    sample_name, #useful?
   starts_with(c("whc", "run", "flush", "yd_rs", "rt"))
    )

# Check it out
raw_field_t2_lab
```

### 1.1.3 - Lab data to get water content needed for PMN.

Unfortunately, WC was not computed separately, but can only be derived from the WHC manip.

! Although WC was computed per bloc (4 blocs per soil), PMN was done on a pooled sample from the 4 blocs. So we compute here only one mean value of wc per soil, that is the mean of the 4 blocs.

```{r}
pmn_wc <- raw_data_field |> 
  select(biol_unit_nb, expe, soil, bloc, sampling_time, starts_with("whc")) |> 
  filter(sampling_time == "t0", !is.na(whc_tare_tube_g)) |> 
  #dm = g dry soil / g fresh soil       
  # wc = 1 - dm
  mutate(
    dm = (whc_g_dw_g - whc_tare_dish_g) / (whc_g_fw_g - whc_tare_tube_g),
    wc = 1-dm,
    .keep = "unused",
    .after = sampling_time
  ) |> 
  group_by(soil) |> 
  summarize(
    dm = mean(dm),
    wc = mean(wc)
  )
```

## 1.2 - Microresp Data (TO DO?)

Then, MicroResp data may need to be added, probably with its own pipeline

## 1.3 - Absorbance data

This data needs tidying, because we still have 4 values per sample (corresponding to the 4 wells given to each sample for analytical replicates on the absorbance plate.

First, we import the dataset

```{r}
raw_Nmin <- read_rds("output/data/2_mgNL_noTDN.rds")
raw_TDN <- read_rds("output/data/2_mgNL_TDN.rds")
```

Then we make sure that each data set contains relevant categorical data:

- biological unit nb

- experiment (to filter on)

- sampling time (idem)

- cs (except PMN)

- soil

- zone

### 1.3.1 - Nmin t2: Subset t2, field data set

Then we take a subset (t2 only, field only).

This complex pipeline is only necessary because I was not very consistent in plate-naming structure.

```{r}
cs_map <- raw_field_t2_lab |> 
  select(biol_unit_nb, cs, soil) |> 
  unique() 


raw_field_t2_Nmin <- raw_Nmin |> 
  filter(dataset == "Nmint1t2") |> #select(plate_id) |> unique() |> print(n = 40)
  # create sampling_time and expe variables from plate_ids (first number and first letter after N species)
  mutate(
    # paste "t" and the sampling time as number: for t1 and t2: is stored in plate data (--> str_extract)
    sampling_time = paste0(
      "t", 
      str_extract(plate_id, "^\\w\\w\\d_(\\d)(\\w).*", group = 1)
      ),
    # for expe: will work for t1 and t2, 
    expe = str_extract(plate_id, "^\\w\\w\\d_(\\d)(\\w).*", group = 2),
    # rephrase "F" into "Field" 
    expe = case_when(expe %in% c("F") ~ "Field", .default = expe),
    .before = plate_id,
    zone = str_extract(map, ".*_(\\w\\d)$", group = 1)
      ) |> 
  # filter based on sampling_time
  filter(sampling_time == "t2", expe == "Field") |> 
  separate_wider_delim(
    cols = map,
    names = c("biol_unit_nb"),
    delim = "_",
    too_many = "drop", 
    cols_remove = FALSE
  ) |> 
  mutate(
    biol_unit_nb = as.double(biol_unit_nb),
    # for some reason, biol_unit_nb of Std soil is 112 in lab data vs 110 in abs data
    biol_unit_nb = case_when(biol_unit_nb == 110 ~ 112, .default = biol_unit_nb)) |> 
  # add info on crop stand and soil
  left_join(cs_map) |> 
  relocate(cs, soil, .before = sampling_time)

# Check it out
raw_field_t2_Nmin
```

### 1.3.2 - PMN: Subset field

```{r}
raw_field_PMN <- raw_Nmin |> 
  filter(dataset == "PMN") |> 
  separate_wider_delim(
    cols = map,
    names = c("expe", "soil", "incubation_time", "tech_rep"),
    delim = "_",
    cols_remove = FALSE
  ) |> 
  mutate(
    sampling_time = rep("t0"),
    biol_unit_nb = paste0(expe, "_", soil),
    .before = well_id) |> 
  filter_out(expe == "Pot")
```

Check out both subsets. First, Nmin data

```{r}
raw_field_t2_Nmin
```

Then, PMN data

```{r}
raw_field_PMN
```

### 1.3.3 - TDN: Subset field

First, keep only field data

```{r}
field_TDN <- raw_TDN |> 
  # get only field data
  mutate(
    plate_nb = str_extract(plate_id, ".*_TDN_(\\d\\d).*", group = 1) |> as.double(),
    .before = map) |> 
  filter((std_sp == "NO3" & plate_nb < 17) | (std_sp == "NO2" & plate_nb < 9)) |> 
  mutate(expe = rep("Field")) |> 
  # make a separation btw Standard soils and samples bc different mapping logic
  mutate(
    sample_std = case_when(
      str_extract(map, "(Field)_.*", group = 1) == "Field" ~ "std_soil",
      .default = "sample"),
    .before = map
  ) 
```

Then, separate between sample and standard soil:

```{r}
field_TDN_sample <- field_TDN |> filter(sample_std == "sample")
field_TDN_std <- field_TDN |> filter(sample_std == "std_soil")
```

For those 2 subsets, different logic to extract biol_unit_nb, sampling_time, zone, CFE/NF, dilution

```{r}
# First, samples
field_TDN_sample_separated <- field_TDN_sample |> 
  separate_wider_delim(
    map, delim = c("_"), cols_remove = FALSE ,
    names = c("biol_unit_nb", "sampling_time", "zone", "fumigation_dilution")
  ) |> 
  mutate(biol_unit_nb = as.double(biol_unit_nb))

# Then, Std soil
field_TDN_std_separated <- field_TDN_std |> 
  separate_wider_delim(
    map, delim = "_", cols_remove = FALSE,
    names = c("field", "sampling_time", "std", "zone", "fumigation_dilution")
  ) |> 
  select(!c(field, std)) |> 
  mutate(biol_unit_nb = 112, .before = sampling_time)
```

Then we rejoin them, separate fumigation_dilution and fetch the categorical data on crop stand (cs) and soil from cs_map

```{r}

field_TDN_clean <- field_TDN_sample_separated |> 
  bind_rows(field_TDN_std_separated) |> 
  separate_wider_delim(
    fumigation_dilution, delim = ".", 
    names = c("fumigation", "dilution")
  ) |> 
  left_join(cs_map) 

# Check it out
field_TDN_clean
```

### 1.3.4 - TO DO: PNR

## 1.4 - Data from the CRA-W

### 1.4.1 - Yield data

In the row data, there is a column under the headers containing units. So we skip this in the import, re-adding column names

```{r}
colnames <- read_csv("../raw_data/SYCBIO_Froment-feverole_2024_MDeToeuf.csv",n_max = 0) |> clean_names() |> names()
yield_raw <- read_csv(
  "../raw_data/SYCBIO_Froment-feverole_2024_MDeToeuf.csv", skip = 2, col_names = colnames, show_col_types = FALSE)
```

Now we tidy this data set:

- rename –\> translate headers + add unit

- Remove empty rows (missing IC samples in Ref soil)

- Recompute (check) computed values

  - For yield: from kg / plot to t/ha :

    - /1000 : kg –\> t

    - X 10000 : m2 –\> ha (in denominator)

    - /(10 X plot_width) : nb of square meters per plot (width x 10m of harvesting)

- (then keep only recomputed ones, remove original raw)

```{r}
yield_full <- yield_raw |> 
  rename(
    year = annee,
    crop = culture,
    biol_unit_nb = code_placeau,
    harvest_date = date_de_recolte,
    yield_per_plot = rdt_brut_parcelle,
    harvester_width_m = largeur_de_mesure,
    water_content_1_percent = humidite_1,
    water_content_2_percent = humidite_2,
    test_weight_1_kg_per_hl = ps_1,
    test_weight_2_kg_per_hl = ps_2,
    protein_percent_dw = mpt_nx5_7
  ) |> 
  filter_out(is.na(yield_per_plot)) |> 
  rowwise() |> 
  mutate(
    soil = case_when(sd_c == 3 ~ "ABC", sd_c == 2 ~ "Auto", sd_c == 1 ~ "Ref", .default = "which_soil?" ),
    crop = case_when(crop == "Froment" ~ "Wheat", crop == "Féverole" ~ "Faba bean", .default = "which crop?"),
    test_weight_kg_per_hl = mean(c(test_weight_1_kg_per_hl, test_weight_2_kg_per_hl), na.rm = TRUE),
    water_content_percent = mean(c(water_content_1_percent, water_content_2_percent), na.rm = TRUE),
    grain_yield_t_per_ha = yield_per_plot/harvester_width_m,
    grain_yield_15p_t_per_ha = grain_yield_t_per_ha * (100 - water_content_percent) / 85) 

# Check computed values

```

Then check re-computed values: all rounding errors. Still, prefer to compute it here

```{r}
# Water content
sum(yield_full$hum_moy != yield_full$water_content_percent)
yield_full |> filter(hum_moy != water_content_percent) |> select(biol_unit_nb, hum_moy, water_content_percent)

# test weight
sum(yield_full$ps_moy != yield_full$test_weight_kg_per_hl)
yield_full |> filter(ps_moy != test_weight_kg_per_hl) |> select(biol_unit_nb, ps_moy, test_weight_kg_per_hl)

# yield in tons per ha
sum(yield_full$rdt_brut_ha != yield_full$grain_yield_t_per_ha)
yield_full |> filter(rdt_brut_ha != grain_yield_t_per_ha) |> select(biol_unit_nb, rdt_brut_ha, grain_yield_t_per_ha)

#yield 15% humidity
sum(yield_full$rdt_15_percent_ha != yield_full$grain_yield_15p_t_per_ha)
yield_full |> filter(rdt_15_percent_ha != grain_yield_15p_t_per_ha) |> select(biol_unit_nb, rdt_15_percent_ha, grain_yield_15p_t_per_ha)

```

Now, keep only useful columns

```{r}
yield_tidy <- yield_full |> 
  select(
    year, crop, soil, biol_unit_nb, harvest_date, 
    grain_yield_15p_t_per_ha, protein_percent_dw,
    grain_yield_t_per_ha, water_content_percent, test_weight_kg_per_hl)
```

# 2 - Tidy data: per-sample outlier removal

::: callout-note
## Not all graphs are displayed

Many graphs are created to do this analysis. There is quite some computing needed to display each graph, so I set as "comment" those that are part of the iterative process of outlier removal, and only display "satisfying" graphs
:::

## 2.1 - Nmin

So we will compute the per-sample average, for the field dataset. But first, let's have a look at the distribution of concentrations before taking the average: are there clear outliers?

### 2.1.1 - NO3

```{r}
boxplot_no3 <- raw_field_t2_Nmin |> 
  filter(biol_unit_nb != 112, std_sp == "NO3") |> 
  boxplot_conc(x = "zone") + labs(title = "NO3") + 
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(soil~biol_unit_nb, nrow = 3, scales = "free_x")

ridges_no3 <- raw_field_t2_Nmin |> 
  filter(biol_unit_nb != 112, std_sp == "NO3") |>  # exclude sand and conv soil std
  plot_ridges_conc(y = "map",colour = "zone", groups = "map") + 
  facet_wrap(~soil, ncol = 3) + labs(title = "NO3")

#boxplot_no3 + ridges_no3
```

Those are obvious outliers (wave really bimodal, outlier point clear)

- 93_t2_z1: D6

- 89_t2_z3: F11

```{r}
to_remove <- raw_field_t2_Nmin |> 
  filter(std_sp == "NO3" &
    ((map == "93_t2_z1" & well_id == "D6") | (map == "89_t2_z3" & well_id == "F11")))

Nmin_wash1 <- raw_field_t2_Nmin |> remove_wells(to_remove) 

boxplot_no3_outlierfree <- Nmin_wash1 |> 
  filter(biol_unit_nb != 112, std_sp == "NO3") |> 
  boxplot_conc(x = "zone") + labs(title = "NO3, outliers removed") + 
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(soil~biol_unit_nb, nrow = 3, scales = "free_x")

boxplot_no3 + boxplot_no3_outlierfree

ridges_no3_outlierfree <- Nmin_wash1 |> 
  filter(biol_unit_nb != 112, std_sp == "NO3") |>  # exclude sand and conv soil std
  plot_ridges_conc(y = "map",colour = "zone", groups = "map") + 
  facet_wrap(~soil, ncol = 3) + labs(title = "NO3, outliers removed")

ridges_no3 + ridges_no3_outlierfree + plot_layout(guides = "collect")
```

### 2.1.2 - NH4

Now same, for NH4.

```{r}
boxplot_nh4 <- Nmin_wash1 |> 
  filter(biol_unit_nb != 112, std_sp == "NH4") |> 
  boxplot_conc(x = "zone") + labs(title = "NH4") + 
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(soil~biol_unit_nb, nrow = 3, scales = "free_x")
  
ridges_nh4 <- Nmin_wash1 |> 
  filter(biol_unit_nb != 112, std_sp == "NH4") |>  # exclude sand and conv soil std
  plot_ridges_conc(y = "map",colour = "zone", groups = "map") + 
  facet_wrap(~soil, ncol = 3) + labs(title = "NH4")

#boxplot_nh4 + ridges_nh4
```

Here are those where I would remove

- 95_t2_z1: F11

- 92_t2_z2: D9

- 89_t2_z3: G11

- 99_t2_z3: E10

- 82_t2_z2: H9

```{r}
to_remove <- Nmin_wash1 |> 
  filter(std_sp == "NH4" & (
           (map == "95_t2_z1" & well_id == "F11") |
           (map == "92_t2_z2" & well_id == "D9") |
           (map == "89_t2_z3" & well_id == "G11") |
           (map == "99_t2_z3" & well_id == "E10") |
           (map == "82_t2_z2" & well_id == "H9")))

Nmin_wash2 <- Nmin_wash1 |> remove_wells(to_remove) 

boxplot_nh4_outlierfree <- Nmin_wash2 |> 
  filter(biol_unit_nb != 112, std_sp == "NH4") |> 
  boxplot_conc(x = "zone") + labs(title = "NH4, outliers removed") + 
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(soil~biol_unit_nb, nrow = 3, scales = "free_x")

ridges_nh4_outlierfree <- Nmin_wash2 |> 
  filter(biol_unit_nb != 112, std_sp == "NH4") |>  # exclude sand and conv soil std
  plot_ridges_conc(y = "map",colour = "zone", groups = "map") + 
  facet_wrap(~soil, ncol = 3) + labs(title = "NH4, outliers removed")

boxplot_nh4 + boxplot_nh4_outlierfree

ridges_nh4 + ridges_nh4_outlierfree + plot_layout(guides = "collect")
```

### 2.1.3 - NO2

Because NO2 readings are virtually zero, there is only little point in removing outliers. Nevertheless, we can have a look at the data

Per sample

```{r}
boxplot_no2 <- Nmin_wash2 |> 
  filter(biol_unit_nb != 112, std_sp == "NO2") |> 
  boxplot_conc(x = "zone") + labs(title = "NO2") + 
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(soil~biol_unit_nb, nrow = 3, scales = "free_x")

ridges_no2 <- Nmin_wash2 |> 
  filter(biol_unit_nb != 112, std_sp == "NO2") |>  # exclude sand and conv soil std
  plot_ridges_conc(y = "map",colour = "zone", groups = "map") + 
  facet_wrap(~soil, ncol = 3) + labs(title = "NO2")

#boxplot_no2 + ridges_no2 + plot_layout(guides = "collect")
```

- 86_t2_z3: H4

```{r}
to_remove <- Nmin_wash2 |> 
  filter(std_sp == "NO2" & map == "86_t2_z3" & well_id == "H4")

Nmin_wash3 <- Nmin_wash2 |> remove_wells(to_remove)

boxplot_no2_2 <- Nmin_wash3 |> 
  filter(biol_unit_nb != 112, std_sp == "NO2") |> 
  boxplot_conc(x = "zone") + labs(title = "NO2, outliers removed") + 
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(soil~biol_unit_nb, nrow = 3, scales = "free_x")

ridges_no2_2 <- Nmin_wash3 |> 
  filter(biol_unit_nb != 112, std_sp == "NO2") |>  # exclude sand and conv soil std
  plot_ridges_conc(y = "map",colour = "zone", groups = "map") + 
  facet_wrap(~soil, ncol = 3) + labs(title = "NO2, outliers removed")

#boxplot_no2_2 + ridges_no2_2 + plot_layout(guides = "collect")

boxplot_no2 + boxplot_no2_2 
ridges_no2 + ridges_no2_2 + plot_layout(guides = "collect")
```

### 2.1.4 - Standard Soils

Here we look at all Nmin species in one go because there are a lot less "samples" to look at

```{r}
boxplot_std <- Nmin_wash3 |> 
  filter(biol_unit_nb == 112) |> 
  boxplot_conc(x = "zone") + labs(title = "Standard Soil") +
  facet_wrap(~std_sp, nrow = 3, scales = "free_y")

ridges_std <- Nmin_wash3 |> 
  filter(biol_unit_nb == 112) |>  # exclude sand and conv soil std
  plot_ridges_conc(y = "map",colour = "zone", groups = "map") + 
  facet_wrap(~std_sp, ncol = 3, scales = "free_x") + labs(title = "Standard Soil")

boxplot_std + ridges_std
```

Nothing to modify!

### 2.1.5 - All outliers removed Nmin

Visually, I am satisfied with this outlier removal, so I save this cleaned table

```{r}
Field_t2_Nmin_clean <- Nmin_wash3
```

## 2.3 - PMN

### 2.3.1 - NO3

```{r}
boxplot_pmn_no3 <- raw_field_PMN |> 
  filter(std_sp == "NO3") |> 
  boxplot_conc(x = "tech_rep") + labs(title = "NO3") +
  facet_wrap(soil~incubation_time, scales = "free_y", ncol = 5)  

ridges_pmn_no3 <- raw_field_PMN |> 
  filter(std_sp == "NO3") |> 
  plot_ridges_conc(groups = "soil", colour = "tech_rep", y = "map") + 
  facet_wrap(~incubation_time, nrow = 1) + labs(title = "NO3")

#boxplot_pmn_no3 + ridges_pmn_no3 
```

- Field_Ref_i3_rt2: E2

- Field_Ref_i2_rt4: B9

- Field_Auto_i2_rt2: A10

- Field_Auto_i2_rt1: A10

- Field_ABC_i3_rt3: E4

```{r}
to_remove <- raw_field_PMN |> 
  filter((std_sp == "NO3") & 
    ((map == "Field_Ref_i3_rt2" & well_id == "E2") | 
       (map == "Field_Ref_i2_rt4" & well_id == "B9") |
       (map == "Field_Auto_i2_rt2" & well_id == "A10") |
       (map == "Field_Auto_i2_rt1" & well_id == "A10") |
       (map == "Field_ABC_i3_rt3" & well_id == "E4")) 
  )

PMN_wash1 <- raw_field_PMN |> remove_wells(to_remove)

# check it out again
boxplot_pmn_no3_outlierfree <- PMN_wash1 |> 
  filter(std_sp == "NO3") |> 
  boxplot_conc(x = "tech_rep") + 
  labs(title = "NO3, outlier removed") +
  facet_wrap(soil~incubation_time, scales = "free_y", ncol = 5) 

ridges_pmn_no3_outlierfree <- PMN_wash1 |> 
  filter(std_sp == "NO3") |> 
  plot_ridges_conc(groups = "soil", colour = "tech_rep", y = "map") + 
  facet_wrap(~incubation_time, nrow = 1) + labs(title = "NO3, outlier removed")

#boxplot_pmn_no3_outlierfree + ridges_pmn_no3_outlierfree
```

Compare before/after

```{r}
boxplot_pmn_no3 + boxplot_pmn_no3_outlierfree 

ridges_pmn_no3 + ridges_pmn_no3_outlierfree + plot_layout(guides = "collect")
```

### 2.3.2 - NH4

```{r}
boxplot_pmn_nh4 <- PMN_wash1 |> 
  filter(std_sp == "NH4") |> 
  boxplot_conc(x = "tech_rep") + facet_wrap(soil~incubation_time, ncol = 5) + 
  labs(title = "NH4")

ridges_pmn_nh4 <- PMN_wash1 |> 
  filter(std_sp == "NH4") |> 
  plot_ridges_conc(groups = "soil", colour = "tech_rep", y = "map") + 
  facet_wrap(~incubation_time, nrow = 1) + labs(title = "NH4")


#boxplot_pmn_nh4 + ridges_pmn_nh4 
```

- Field_Auto_i0_rt3: B3

```{r}
to_remove <- PMN_wash1 |> filter(
  std_sp == "NH4"& map == "Field_Auto_i0_rt3" & well_id == "B3")

PMN_wash2 <- PMN_wash1 |> remove_wells(to_remove)

boxplot_pmn_nh4_outlierfree <- PMN_wash2 |> 
  filter(std_sp == "NH4") |> 
  boxplot_conc(x = "tech_rep") + facet_wrap(soil~incubation_time, ncol = 5) + 
  labs(title = "NH4, outliers removed")

ridges_pmn_nh4_outlierfree <- PMN_wash2 |> 
  filter(std_sp == "NH4") |> 
  plot_ridges_conc(groups = "soil", colour = "tech_rep", y = "map") + 
  facet_wrap(~incubation_time, nrow = 1) + labs(title = "NH4, outliers removed")


#boxplot_pmn_nh4_outlierfree + ridges_pmn_nh4_outlierfree

```

Compare before/after

```{r}
boxplot_pmn_nh4 + boxplot_pmn_nh4_outlierfree 

ridges_pmn_nh4 + ridges_pmn_nh4_outlierfree + plot_layout(guides = "collect")
```

### 2.3.3 - NO2

```{r}
boxplot_pmn_no2 <- PMN_wash2 |> 
  filter(std_sp == "NO2") |> 
  boxplot_conc(x = "tech_rep") + facet_wrap(soil~incubation_time, ncol = 5) + 
  labs(title = "NO2")

ridges_pmn_no2 <- PMN_wash2 |> 
  filter(std_sp == "NO2") |> 
  plot_ridges_conc(groups = "soil", colour = "tech_rep", y = "map") + 
  facet_wrap(~incubation_time, nrow = 1) + labs(title = "NO2")


#boxplot_pmn_no2 + ridges_pmn_no2 
```

Possibly masking other outliers, so maybe a second round is needed

- Field_Auto_i0_rt2: A3

```{r}
to_remove <- PMN_wash2 |> filter(
  std_sp == "NO2" & map == "Field_Auto_i0_rt2" & well_id == "A3")

PMN_wash3 <- PMN_wash2 |> remove_wells(to_remove)
```

Run it once more

```{r}
boxplot_pmn_no2 <- PMN_wash3 |> 
  filter(std_sp == "NO2") |> 
  boxplot_conc(x = "tech_rep") + facet_wrap(soil~incubation_time, ncol = 5) + 
  labs(title = "NO2")

ridges_pmn_no2 <- PMN_wash3 |> 
  filter(std_sp == "NO2") |> 
  plot_ridges_conc(groups = "soil", colour = "tech_rep", y = "map") + 
  facet_wrap(~incubation_time, nrow = 1) + labs(title = "NO2")


#boxplot_pmn_no2 + ridges_pmn_no2 
```

- Field_Ref_i4_rt3: H5

- Field_Auto_i1_rt4: B6

```{r}
to_remove <- PMN_wash3 |> filter(
  (std_sp == "NO2") &
    ((map == "Field_Ref_i4_rt3" & well_id == "H5") |
    (map == "Field_Auto_i1_rt4" & well_id == "B6")))

PMN_wash4 <- PMN_wash3 |> remove_wells(to_remove)

boxplot_pmn_no2_outlierfree <- PMN_wash4 |> 
  filter(std_sp == "NO2") |> 
  boxplot_conc(x = "tech_rep") + facet_wrap(soil~incubation_time, ncol = 5) + 
  labs(title = "NO2, outliers removed")

ridges_pmn_no2_outlierfree <- PMN_wash4 |> 
  filter(std_sp == "NO2") |> 
  plot_ridges_conc(groups = "soil", colour = "tech_rep", y = "map") + 
  facet_wrap(~incubation_time, nrow = 1) + labs(title = "NO2, outliers removed")


#boxplot_pmn_no2_outlierfree + ridges_pmn_no2_outlierfree 
```

Compare before / after

```{r}
boxplot_pmn_no2 + boxplot_pmn_no2_outlierfree 

ridges_pmn_no2 + ridges_pmn_no2_outlierfree + plot_layout(guides = "collect")
```

### 2.3.4 - All outliers removed PMN

Visually, I am satisfied with this outlier removal, so I save this cleaned table

```{r}
field_t2_PMN_clean <- PMN_wash4
```

## 2.4 - TDN

Tiny correction to the dataframe to help readibility of graphs

```{r}
tdn_simpler <- field_TDN_clean |> 
  filter(dilution == "2x", std_sp == "NO3") |> 
  mutate(sample = paste0(biol_unit_nb, "_", zone))
```

### 2.4.1 - NO3

```{r}
boxplot_tdn_no3 <- tdn_simpler |> 
  filter(biol_unit_nb != 112) |> 
  boxplot_conc(x = "sample", colour = "zone") + labs(title = "NO3") + 
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(fumigation~soil, ncol = 3, scales = "free")

ridges_tdn_no3 <- tdn_simpler |> 
  filter(biol_unit_nb != 112) |> 
  plot_ridges_conc(y = "sample",colour = "zone", groups = "map") + 
  facet_wrap(fumigation~soil, ncol = 3, scales = "free") + labs(title = "NO3")

#boxplot_tdn_no3 + ridges_tdn_no3 + plot_layout(guides = "collect")
```

- 95_t2_z2_CFE.2x: D11,

- 100_t2_z1_CFE.2x: C5

- 83_t2_z2_CFE.2x: C10

- 82_t2_z3_CFE.2x: C5

- 96_t2_z3_NF.2x: C10

- 94_t2_z3_NF.2x: B4

- 90_t2_z3_NF.2x: D2

- 99_t2_z2_NF.2x: D2

- 86_t2_z3_NF.2x: B7

- 82_t2_z1_NF.2x: B4

```{r}
to_remove <- tdn_simpler |> 
  filter((std_sp == "NO3") &(
    (map == "95_t2_z2_CFE.2x" & well_id == "D11") | 
      (map == "100_t2_z1_CFE.2x" & well_id == "C5") | 
      (map == "83_t2_z2_CFE.2x" & well_id == "C10") | 
      (map == "82_t2_z3_CFE.2x" & well_id == "C5") | 
      (map == "96_t2_z3_NF.2x" & well_id == "C10") | 
      (map == "94_t2_z3_NF.2x" & well_id == "B4") | 
      (map == "90_t2_z3_NF.2x" & well_id == "D2") | 
      (map == "99_t2_z2_NF.2x" & well_id == "D2") | 
      (map == "86_t2_z3_NF.2x" & well_id == "B7") | 
      (map == "82_t2_z1_NF.2x" & well_id == "B4")))

TDN_wash1 <- tdn_simpler |> remove_wells(to_remove)

boxplot_tdn_no3_outlierfree <- TDN_wash1 |> 
  filter(biol_unit_nb != 112) |> 
  boxplot_conc(x = "sample", colour = "zone") + labs(title = "NO3, outliers removed") + 
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(fumigation~soil, ncol = 3, scales = "free")

ridges_tdn_no3_outlierfree <- TDN_wash1 |> 
  filter(biol_unit_nb != 112) |> 
  plot_ridges_conc(y = "sample",colour = "zone", groups = "map") + 
  facet_wrap(fumigation~soil, ncol = 3, scales = "free") + labs(title = "NO3, outliers removed")

boxplot_tdn_no3 + boxplot_tdn_no3_outlierfree + plot_layout(guides = "collect")
ridges_tdn_no3 + ridges_tdn_no3_outlierfree + plot_layout(guides = "collect")
```

### 2.4.2 - NO2

We skip this for now as I don't think that NO2 is relevant!

### 2.4.3 - Standard soil

```{r}
boxplot_tdn_std <- TDN_wash1 |> 
  filter(biol_unit_nb == 112) |> 
  boxplot_conc(x = "sample") + labs(title = "Standard soil") + 
  theme(axis.text.x = element_text(angle = 90)) +
  facet_wrap(~fumigation, scales = "free")

ridges_tdn_std <- TDN_wash1 |> 
  filter(biol_unit_nb == 112) |> 
  plot_ridges_conc(y = "sample",colour = "zone", groups = "map") + 
  facet_wrap(~fumigation, ncol = 3, scales = "free") + labs(title = "Standard soil")

boxplot_tdn_std + ridges_tdn_std + plot_layout(guides = "collect")
```

All good.

### 2.4.3 - All outliers removed TDN

Save the new data

```{r}
Field_t2_TDN_clean <- TDN_wash1
```

## 2.5 - PNR - TO DO

### 2.5.1 - NO3

### 2.5.2 - NO2

### 2.5.3 - All outliers removed PNR

# 3 - Per-sample mean

## 3.1 - Nmin

```{r}
conc_mean_Nmin <- Field_t2_Nmin_clean |> 
  select(map, plate_id, biol_unit_nb, zone, std_sp, conc_mgN_L) |> 
  group_by(plate_id, map, biol_unit_nb, zone, std_sp) |> 
  summarise(
    mean = mean(conc_mgN_L),
    st_dev = sd(conc_mgN_L)) |> 
  mutate(coef_var = 100*st_dev / mean) |> 
  rename(conc_mgN_L = mean) |> ungroup()

#biggest coef var quite high
conc_mean_Nmin |> arrange(desc(coef_var))

# but not too bad when look only at NO3
conc_mean_Nmin |> arrange(desc(std_sp), desc(coef_var))

conc_mean_Nmin |> filter(coef_var > 10) |> 
  ggplot(aes(x = coef_var)) + geom_histogram() + facet_wrap(~std_sp, ncol = 1)
```

We still have quite a few samples with a high between-wells coefficient of variation, but mostly with NO2 and NH4 (very low values). With only 3 to 4 wells and sometimes very low values, this is unavoidable, so we move on

Now, finally, we re-join this mean value to the rest of the relevant information from the absorbance dataset

```{r}
conc_Nmin_export_ready <- conc_mean_Nmin |> 
  select(!st_dev:coef_var) |> 
  inner_join(
    raw_field_t2_Nmin |> 
      select(!c(well_id:abs_corrected, starts_with("conc"), slope:lm_p, target_sp)) |> 
      unique())
```

## 3.2 - PMN

```{r}
conc_mean_PMN <- field_t2_PMN_clean |> 
  select(map, plate_id, biol_unit_nb, std_sp, conc_mgN_L) |> 
  group_by(map, biol_unit_nb, std_sp) |> 
  summarise(
    mean = mean(conc_mgN_L),
    st_dev = sd(conc_mgN_L)) |> 
  mutate(
    coef_var = 100*st_dev / mean,
    soil = str_extract(biol_unit_nb, "_(\\w*)", group = 1)) |> 
  rename(conc_mgN_L = mean)

conc_mean_PMN |> arrange(desc(std_sp), desc(coef_var))
field_t2_PMN_clean |> filter(map == "Field_ABC_i1_rt4", std_sp == "NO3")

conc_mean_PMN |> filter(coef_var > 10) |> 
  ggplot(aes(x = coef_var)) + geom_histogram() + xlim(0,100)
```

Same here, there are still quite a few samples with a high coefficient of variation, but with only 3 to 4 wells and sometimes very low values, this is unavoidable, so we move on

Now, we can add the data on wc to get a full dataset

```{r}
conc_PMN_export_ready <- conc_mean_PMN |> 
  # get dm and wc
  left_join(pmn_wc) |> 
  # get important categorical variables
  left_join(
    field_t2_PMN_clean |> 
      select(map, dataset, expe, incubation_time, tech_rep, sampling_time) |> 
      unique())
```

## 3.3 - TDN

```{r}
conc_mean_TDN <- Field_t2_TDN_clean |> 
  select(map, biol_unit_nb, std_sp, conc_mgN_L, fumigation) |> 
  group_by(std_sp, fumigation, biol_unit_nb, map) |> 
  summarise(
    mean = mean(conc_mgN_L),
    st_dev = sd(conc_mgN_L)) |> 
  mutate(
    coef_var = 100*st_dev / mean) |> 
  rename(conc_mgN_L = mean)

conc_mean_TDN |> arrange(desc(std_sp), desc(coef_var))
conc_mean_TDN |> arrange(desc(coef_var))


conc_mean_TDN |> 
  ggplot(aes(x = coef_var)) + geom_histogram() 
```

For once, it's looking pretty good!

```{r}
conc_TDN_export_ready <- conc_mean_TDN |> 
  # get relevant categorical data
  left_join(
    Field_t2_TDN_clean |> 
      select(map, dataset, sampling_time, zone, fumigation, dilution) |> 
      unique())
```

## 3.4 - PNR

# 4 - Export

```{r}
raw_field_t2_lab |> write_rds("output/data/3_field_t2_raw_lab.rds")
conc_Nmin_export_ready |> write_rds("output/data/3_field_t2_Nmin_clean.rds")
conc_PMN_export_ready |> write_rds("output/data/3_field_PMN_clean.rds")
conc_TDN_export_ready |> write_rds("output/data/3_field_TDN_clean.rds")
yield_tidy |> write_rds("output/data/3_field_yield_clean.rds")
```
