---
title: "III. Data Tidying"
format: 
  gfm:
    toc: true
    code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

In this script we tidy data sets into the shape that we need for data transformation and following analysis and visualization

# Set up

```{r}
rm(list = ls())

library(tidyverse)
library(janitor)
```

# 1 - Import data

```{r}
#lm_output <- read_rds("output/data/Nmin_std_curves_lm.rds")

# import absorbance data + initiate an empty column for data set
N_data <- read_rds("output/data/Nmin_conc.rds") |> 
  # add any column to be taken from metadata
  mutate(
    dataset = NA,
    sampling_time = NA) |> 
  rename(sample_short = plate_map)
N_metadata <- read_rds("output/data/N_all_metadata.rds")


# import other "wet lab" raw data
lab_data <- read_rds("output/data/2_lab_data.rds") 
```

# 2 - Compute new variables on per-sample basis

This is the place to compute new variables that will be needed for several downstream analyses

# 2 - Greenhouse, t2

We first focus on data from the greenhouse trial at t2. It follows a complete randomized block design, with modalities: 4 soils x 3 crop stands x 5 blocs

## 1.1 - Extract subsets of data: Nmin, TDN (incl. MBN)

### 1.2.1 - Retrieve dataset information into absorbance dataframe

```{r}

for (i in 1:nrow(N_data)) {
  plate <- N_data$plate_id[i]
  line <- N_metadata |> filter(plate_id == plate)
  
  N_data$dataset[i] <- line$dataset[1]
  N_data$sampling_time[i] <- line$sampling_time[1]
    
}

#check that works:
N_data$dataset |> unique()
N_data$sampling_time |> unique()
```

### 1.2.2 - Store subsets in new objects

Now that each row has been given the corresponding dataset, we can take a subset based on dataset

#### Nmin

For Nmin --\> we don't take t3 for now, we'll still have to filter based on sampling_time

```{r}
# plate data
Nmin_plate_data <- N_data |> 
  filter(dataset == "Nmint1t2", sampling_time == "t2")

# metadata
Nmin_metadata <- N_metadata |> 
  filter(dataset == "Nmint1t2", sampling_time == "t2")

# raw data
Nmin_raw <- lab_data |> 
  # keep only t2 and pot, remove bare soil
  filter(sampling_time == "t2", expe == "Pot", cs != "B") |> 
  # tidy: 
  # For now, only actual samples (no std soil with weird number)
  filter(soil %in% c("Conv", "Ref", "Auto", "ABC")) |> 
  # sample number was made unique, not practical here
  mutate(
    sample_short = paste0(biol_unit_nb - 200, "_", sampling_time)
  ) |> 
  select(!starts_with(c("amf", "rt")))
```

Now we join those tables in one big data set

```{r}

# join
Nmin <- left_join(
  Nmin_raw, Nmin_plate_data, 
  by = join_by(sample_short, sampling_time)) |>
  left_join(
    Nmin_metadata,
    by = join_by(sampling_time, plate_id, dataset)) 

```

#### TDN

Same principle. First subset

```{r}

# For TDN and MBN
TDN_plate_data <- N_data |> 
  filter(dataset == "TDN", sampling_time == "t2")
TDN_metadata <- N_metadata |> 
  filter(dataset == "TDN", sampling_time == "t2")
TDN_raw <- Nmin_raw

```

Then join

```{r}
# join
TDN <- left_join(
  TDN_raw, TDN_plate_data, 
  by = join_by(sample_short, sampling_time)) |>
  left_join(
    TDN_metadata,
    by = join_by(sampling_time, plate_id, dataset)) 
```

## 1.3 - Export

```{r}
Nmin |>  write_rds("output/data/Nmin_tidy.rds")
TDN |> write_rds("output/data/TDN_tidy.rds")
```
