---
title: "3 Integrated Data Analysis"
author: "Morgane de Toeuf"
format: 
  gfm:
    toc: true
    code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# To Do

Field:

- Correct LER with seeding rates (now relative values ok, absolute are off)

- Review notes PI meeting validation stat: 2 modèles

  - effet de système sur froment seul (1 facteur = 3 sols)

  - effet du labour sur effet de IC (Auto et ABC uniquement) –\> 2 facteurs 2 niveaux (2 sols, 2 cs)

  - SO: remove all Ref & IC !

Still missing:

- PNR

- MicroResp

- Nmin t3 ?

# Set up

```{r}
#| message: false
#| warning: false

# clean environment
rm(list = ls())

# packages
library(tidyverse)
library(vegan) # for rda()
library(patchwork)
library(RColorBrewer)
library(emmeans) # for post-hoc test (after ANCOVA of PMN)

# function
source("functions/plot_pca.R")

```

# 1 - Introduction

## 1.1 - Greenhouse

For the greenhouse data sets, many bits and pieces are dropped along the way, so that only few variables are left in the end:

- The whole t1 sampling is problematic due to the scattered sampling over time

- For t2:

  - MicroResp experiment: still needs calibration, but responses were really low, probably due to high sand levels. We'll look into it anyway

  - PMN: samples had to be re-humidified because the container system was not optimal and samples had become basically dry. See script on data wrangling: curves show interesting trends, but incubation times 2 and 3 problematic. After discussion with PI: removed those points, curves are better now. Still a lot of noise, which could come from the added variability of the re-humidification + the missing data.

  - TDN & MBN: never got the protocol from Sang, and did a mistake by not adjusting the serial dilution of the standard curve –\> samples out of range for Greenhouse

  - WHC is not intrinsically interesting, though we have the data as well (was needed for MicroResp)

In the end, what is left here is less than anticipated:

- Yield data is trustworthy (plant biomass, plant height, nb of stems / tillers)

  - LER can be derived from yield data (but = aggregated data –\> no incorporation in raw data PCA possible). But PCA with variables with ratio IC/SC

- Npools are incomplete:

  - Nmin data at t2 (NO2, NO3, NH4 –\> Nmin and ratios)

  - Potential mineralization of N (at sowing time)

  - But [NO]{.underline} Total Dissolved N, Microbial biomass N and organic N

- MicroResp data give an idea of soil biomass, soil metabolism potential, functional diversity

What is still dormant is weed data. Relevant? not sure...

## 1.2 - Field

For the field data sets, some bits and pieces are dropped along the way:

- MicroResp experiment: still needs calibration (maybe)

- PMN: samples had to be re-humidified because the container system was not optimal and samples had become basically dry (see script on data wrangling: curves show interesting trends, but don't know what to do with them)

- TDN & MBN: I believe that the blank was not digested

- WHC is not intrinsically interesting, though we have the data as well (was needed for MicroResp)

In the end, what is left here is:

- yield data (harvest: grain, grain protein, PS)

- Nmin data at t2 (NO2, NO3, NH4 –\> Nmin and ratios)

- LER can be derived from yield data (but = aggregated data –\> no incorporation in PCA possible)

- Question mark on:

  - slurry test (upcoming)

  - AMF quantification

  - Leghemoglobin quantification

What is still dormant is weed data. Relevant? not sure...

## 1.3 - Experimental design and questions asked

Whereas the greenhouse had a full design with both sole crops and intercrop, and all modalitites present, this is not the case in the field.

- First, there was no sole Faba Bean, so we cannot approach data with the same ratios IC/SC as in the greenhouse (or then only for wheat, which still bears some interest).

- Second, the Faba Bean did not take in the Reference Soil in the field, so that

  - for all 3 soils, we can compare sole wheat, that is, it is an experimental design with 1 factor (soil) and 3 levels (Ref, Auto, ABC), with factorial block design and a system trial (it's the whole farming system that changes)

  - for only 2 soils (Auto and ABC), we can also compare the response of wheat to IC from one soil to the other. But here, it is no longer a system trial because only ploughing is different (well, and maybe weeding? to be checked). And it is a split plot design, with soil as the main factor (2 levels) and crop diversity as the nested factor (2 levels: sole wheat and intercropped wheat)

## 1.4 - Setting general parameters

### 1.4.1 - Significance

```{r}
p_signif <- tribble(
  ~ signif_stars, ~ max_p_value,
  "",    1,
  ".",   0.1,
  "*",   0.05,
  "**",  0.01,
  "***", 0.001
)
```

### 1.4.2 - Graphical parameters

```{r}
library(RColorBrewer)
display.brewer.all(n = 3, type = "qual")
display.brewer.pal(n = 3, name = "Accent")
cs_colors <- brewer.pal(n = 3, name = "Accent")
names(cs_colors) <- c("FB", "IC", "W")

# display.brewer.all(n = 4, type = "qual")
# display.brewer.pal(n = 4, name = "Set1")
# soil_colors <- brewer.pal(n = 4, name = "Set1")
# names(soil_colors) <- c("Conv", "Ref", "Auto", "ABC")

# Palette "Badlands" from package nationalparkcolors
soil_colors <- c(
  "Conv" = "#5495CFFF", "Ref" = "#F5AF4DFF", 
  "Auto" = "#DB4743FF", "ABC" = "#7C873EFF")
# soil_colors <- c("#5495CFFF", "#F5AF4DFF", "#DB4743FF", "#7C873EFF", "#FEF4D5FF")
# names(soil_colors) <- c("Conv", "Ref", "Auto", "ABC", "Std")
# soil_colors <- tibble(
#   colours = c("#5495CFFF", "#F5AF4DFF", "#DB4743FF", "#7C873EFF", "#FEF4D5FF") ,
#   soil = c("Conv", "Ref", "Auto", "ABC", "Std"))
```

# 2 - Analysis on Greenhouse Data

## 2.1 - Load and clean data

[**! Data from the standard soils is not in `Npools_t2`. They will need to be addressed separately, if at all.**]{.underline}

```{r}

Npools_t2_greenhouse <- read_rds("output/data/2_Npools.rds") |> 
  filter(expe == "Pot")
yield_clean_greenhouse <- read_rds("output/data/1_greenhouse_yield_clean.rds")
MR_greenhouse <- read_rds("output/data/1_MR_data.rds") |> 
  filter(expe == "Pot") |> 
  filter_out(soil %in% c("Std", "Sand") | cs == "B") |> 
  mutate(biol_unit_nb = as.numeric(sample_id), .keep = "unused") |> 
  select(!c(dataset, zone))

PMN_greenhouse <- read_rds("output/data/2_PMN.rds") |> filter(expe == "Pot")
```

First, we join all pot-level data, i.e., Npool data, yield data and MicroResp data. PMN cannot be joined because it is a measure at soil-level.

```{r}
pot_level_data <- Npools_t2_greenhouse |> 
  select(
    biol_unit_nb, expe, soil, crop_diversity, cs, bloc, 
    dm:last(names(Npools_t2_greenhouse))
    ) |> 
  left_join(yield_clean_greenhouse) |> 
  left_join(MR_greenhouse) |> 
  select(!starts_with("Norg")) |> 
  mutate(soil = factor(soil, levels = c("Conv", "Ref", "Auto", "ABC")))
```

Problem: for now, the statistical unit is the pot (1 row per pot). But some variables cannot be regrouped for both plant species. Consequently, there are some missing data, see here:

```{r}
pot_level_data |> select(cs, fw_fb_per_pot:dw_w_per_plant)
```

A PCA cannot be ran with missing data.

A solution is to change the statistical unit to be the plant, and have 2 data sets: dataset fababean (`data_fb`) and dataset wheat (`data_w`). Anyway, it makes little sense to compare the 2 plant species, but it is interesting to see if effects are similar or not. To be considered: bulk soil data of intercrops will appear twice, so that the 2 datasets are not to be combined in downstream analyses.

So we do that subsetting

```{r}
data_fb_pot <- pot_level_data |> 
  select(!contains("_w")) |> 
  filter(cs != "W") |> 
  arrange(biol_unit_nb)

data_w_pot <- pot_level_data |> 
  select(!contains("_f")) |> 
  filter(cs != "F") |> 
  arrange(biol_unit_nb)

# check it out
str(data_fb_pot)
str(data_w_pot)
```

## 2.2 - Compute LER

This also cannot go into the regular PCA since we get only data that agregates on SC + IC.

So we can have:

- a regular PCA with Npools, MicroResp and yield data on a pot level (one plant at the time)

- A reduced PCA with Npools and microresp expressed as a ratio of value for IC / value for SC, together with LER, also one plant at the time

```{r}
#| message: false
ler_fb_pot <- data_fb_pot |> 
  select(crop_diversity, soil, bloc, dw_fb_per_plant) |> 
  pivot_wider(
    names_from = crop_diversity,
    values_from = dw_fb_per_plant) |> 
  mutate(
    ler_fb = IC / SC
  )
  
ler_w_pot <- data_w_pot |> 
  select(crop_diversity, soil, bloc, dw_w_per_plant) |> 
  pivot_wider(
    names_from = crop_diversity,
    values_from = dw_w_per_plant) |> 
  mutate(
    ler_w = IC / SC
  )

ler_total_pot <- ler_w_pot |> 
  select(!SC:IC) |> 
  left_join(ler_fb_pot |> select(!SC:IC)) |> 
  mutate(ler_total = (ler_w + ler_fb)/2)
```

Plot it

```{r}
#| warning: false
plot_ler_total_pot <- ler_total_pot |> 
  ggplot(aes(x = soil, y = ler_total)) +
  theme_minimal() +
  geom_boxplot() +
  labs(title = "LER total", subtitle = "Pot experiment")

plot_ler_fb_pot <- ler_total_pot |> 
  ggplot(aes(x = soil, y = ler_fb)) +
  theme_minimal() +
  geom_boxplot() +
  labs(title = "LER Faba Bean", subtitle = "Pot experiment")

plot_ler_w_pot <- ler_total_pot |> 
  ggplot(aes(x = soil, y = ler_w)) +
  theme_minimal() +
  geom_boxplot() +
  labs(title = "LER Wheat", subtitle = "Pot experiment")

plot_ler_total_pot + plot_ler_fb_pot + plot_ler_w_pot
```

[**!! TO DO: check those outliers: could it be some of those pots where some plants were not as should be? Or anything? Can I remove them?**]{.underline}

## 2.3 - PMN

### 2.3.1 - ANCOVA

With the PMN data, we can compute the slope of the curve.

::: callout-important
## Statistical issue

We have 4 technical replicates per incubation time, but rep 1 of incubation time 1 has no particular relationship to rep 1 of incubation time 2. It would make little sense to compute 4 different slopes and then take the mean of the 4 slopes to get a boxplot and the option of an anova with comparison of means (of slopes).
:::

According to AI (see full doc):

- Analysis of CoVariance, ANCOVA, (independant measurements bc subsamples evolve separately) .

  - linear model with interaction term: outcome \~ group \* predictor

  - --\> try with lm(ppm_Nmin \~ soil \* incub_days)

  - Details see doc

```{r}
PMN_continuous <- PMN_greenhouse |> mutate(incub_day = as.integer(incub_day))

my_lm <- lm(data = PMN_continuous, ppm_Nmin ~ incub_day * soil)
summary(my_lm)

anova(my_lm)
```

The Anova (in this case ANCOVA) says:

- Effect of incubation time (duh)

- effect of soil

- interaction effect –\> reject (H0 = slopes are equal)

—\> need of a post-hoc test, with Tukey adjustment

```{r}
slope_trends <- emtrends(my_lm, pairwise ~ soil, var = "incub_day")

summary(slope_trends)

# see the slopes
slope_trends$emtrends

# see the pairwise contrasts with adjusted p-values (Tukey method)
slope_trends$contrasts
```

- conv ≠ Ref, ABC

- Conv = Auto

- Ref = ABC, Ref = Auto

- Auto ≠ ABC

```{r}
ancova_signif <- slope_trends$contrasts |> as_tibble() |> 
  mutate(signif = c("**", "", "***", "", "", "*"))

```

Checking model assumptions

```{r}
library(performance)
plot_check_ancova <- check_model(my_lm) 
plot <- plot(plot_check_ancova) + patchwork::plot_annotation(title = "QC: Checking model assumptions for ANCOVA of PMN - Greenhouse data")

# check it out
plot

# save it
ggsave("output/figures/greenhouse/qc_pmn_ancova_greenhouse.pdf", plot,height = 8, width = 13.5)
```

### 2.3.2 - Plot PMN

```{r}
dates <- c("2023-12-5", "2023-12-26", "2024-01-02") |> as.Date()
incub_days <- dates - dates[1]
names(incub_days) <- PMN_greenhouse$incub_time |> unique() |> sort()

# plot again
PMN_plot <- PMN_greenhouse |> 
 # left_join(soil_colors) |> 
  ggplot(aes(x = incub_day, y = ppm_Nmin, group = soil, colour = soil, fill = soil)) +
  theme_minimal() +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    title = "Potential Mineralization of Nitrogen (PMN) - Greenhouse",
    subtitle = "Mineral N as sum of NO2-, NO3-, NH4+\nCurves are statistically different (ANCOVA)",
    caption = "Modelled with lm(ppm_Nmin ~ incub_time * soil)\nPost-Hoc with emtrends() and Tukey correction") +
  xlab("Time since incubation at 28°C [days]") +
  ylab("Mineral Nitrogen in the soil [ppm]") +
  scale_color_manual(values = soil_colors) +
  scale_fill_manual(values = soil_colors)

# for annotation: extract data from smooth curve
smooth_data <- ggplot_build(PMN_plot)$data[[2]]

# get its maximum value to anker the annotation
annotations <- smooth_data |> 
  slice_max(x, by = group) |> 
  mutate(soil = levels(PMN_greenhouse$soil))
#c("#F8766D", "#7CAE00", "#00BFC4", "#C77CFF") 

PMN_plot_all <- 
  PMN_plot + 
  geom_text(
    data = annotations,
    aes(x = x+0.5, y = y, colour = soil, label = soil), 
    hjust = 0, size = 5) +
  xlim(c(0,45)) +
  theme(legend.position = "none") #+

signif_annotations <- ancova_signif |> 
  filter(signif %in% c(".", "*", "**", "***")) |> 
  select(contrast, p.value, signif) |> 
  mutate(
    start = str_extract(contrast, "(\\w*) - (\\w*)", group = 1),
    end = str_extract(contrast, "(\\w*) - (\\w*)", group = 2)
  ) |> 
  left_join(
    annotations |> select(soil, x, y) |> rename(start = soil)
  ) |> 
  left_join(
    annotations |> select(soil, x, y) |> rename(end = soil, yend = y)
  ) |> 
  mutate(
    x = x + c(11,7,15),
    y = y - 0.1,
    yend = yend + 0.1) |> 
  mutate(soil = start)

plot_signif <- PMN_plot_all +
  # vertical segments
  geom_segment(data = signif_annotations,
    aes(x = x, y = y, yend = yend), colour = "grey30") +
  # significance stars
  geom_text(data = signif_annotations,
    aes(x = x + 1, y = (y + yend)/2, label = signif), colour = "grey30") +
  # top-horizontal segments
  geom_segment(data = signif_annotations,
    aes(x = x, xend = x - 1, y = yend, yend = yend), colour = "grey30") +
  # bottom-horizontal segments
  geom_segment(data = signif_annotations,
    aes(x = x, xend = x - 1, y = y, yend = y), colour = "grey30")
```

Check it out & save

```{r}
plot_signif  

ggsave("output/figures/greenhouse/plot_pmn_signif_greenhouse.pdf", plot = plot_signif, height = 5, width = 7)
```

## 2.4 - PCA

### 2.4.1 - Check assumptions

.... UPCOMING?

### 2.4.2 - Subset PCA data

PCA requires only numerical data, so we first get rid on any variable that should not enter the pca

[**!! MISSING SINGLE SIRs, will get them back in**]{.underline}

```{r}
data_fb_pca <- data_fb_pot |> 
  select(
    biol_unit_nb, 
    ppm_NH4, ppm_NO3, #ppm_Nmin, 
    stem_per_fb, height_per_fb, dw_fb_per_plant,
    basal_resp, MBC, qCO2, MSIR, shannon) |> 
  column_to_rownames("biol_unit_nb") |> 
  drop_na()

data_w_pca <- data_w_pot |> 
  select(
    biol_unit_nb, 
    ppm_NH4, ppm_NO3, ppm_Nmin, 
    till_per_w, height_per_w, dw_w_per_plant,
    basal_resp, MBC, qCO2, MSIR, shannon) |> 
  column_to_rownames("biol_unit_nb") |> 
  drop_na()
```

### 3.2.2 - Faba Bean

```{r}

# rda_matrix_fb <- data_fb_pca |> 
#   column_to_rownames("biol_unit_nb") |> 
#   drop_na()
  
pca_fb <- pca(data_fb_pca, scale = TRUE)

summary(pca_fb)

screeplot(pca_fb, bstick = TRUE, npcs = length(pca_fb$CA$eig))

#biplot(pca_fb, scaling = 1, main = "PCA - scaling 1 = object relationships preserved")
biplot(pca_fb, main = "PCA FB - scaling 2 = variable relationships preserved")


```

Prepare data for plotting

```{r}
#non_pca <- greenhouse |> select(!ppm_NH4:NO3_NH4)
pot_scores <- scores(pca_fb, display = "sites", scaling = 2) 
biol_unit <- rownames(pot_scores) |> as.double()
pot_scores <- pot_scores |> 
  as_tibble() |> 
  mutate(biol_unit_nb = biol_unit)

var_scores <- scores(pca_fb, display = "species", scaling = 2)
variables <- rownames(var_scores)
var_scores <- var_scores |> 
  as_tibble() |> 
  mutate(variable = variables)

PC_percent <- pca_fb$CA$eig / sum(pca_fb$CA$eig)

data_fb_pca <- left_join(data_fb_pot, pot_scores, by = join_by(biol_unit_nb))
```

Plot it

```{r}
#| warning: false
#dev.new()
plot_pca_fb_cs <- plot_pca(data_fb_pca, var_scores, PC_percent, colour_point = "cs") +
  #geom_point(aes( colour = cs)) +
  labs(title = "PCA - Faba Bean") +
  scale_color_discrete(palette = c(cs_colors["FB"], cs_colors["IC"])) 

plot_pca_fb_cs

plot_pca_fb_soil <- plot_pca(data_fb_pca, var_scores, PC_percent, colour_point = "soil") +
 # geom_point(aes( colour = soil, shape = cs)) +
  labs(title = "PCA - Faba Bean") +
  scale_color_discrete(palette = c(soil_colors)) 

plot_pca_fb_soil
```

### 3.2.3 - Wheat

```{r}

# rda_matrix_w <- data_w |> 
#   select(
#     biol_unit_nb,
#     ppm_NO3, ppm_NH4, 
#     #till_per_w, 
#     height_per_w, dw_w_per_plant) |> 
#   column_to_rownames("biol_unit_nb") |> 
#   drop_na()
  
pca_w <- pca(data_w_pca, scale = TRUE)

summary(pca_w)

screeplot(pca_w, bstick = TRUE, npcs = length(pca_w$CA$eig))

#biplot(pca_fb, scaling = 1, main = "PCA - scaling 1 = object relationships preserved")
biplot(pca_w, main = "PCA Wheat - scaling 2 = variable relationships preserved")
```

Prepare data for plotting

```{r}
#non_pca <- greenhouse |> select(!ppm_NH4:NO3_NH4)
pot_scores <- scores(pca_w, display = "sites", scaling = 2) 
biol_unit <- rownames(pot_scores) |> as.double()
pot_scores <- pot_scores |> 
  as_tibble() |> 
  mutate(biol_unit_nb = biol_unit)

var_scores <- scores(pca_w, display = "species", scaling = 2)
variables <- rownames(var_scores)
var_scores <- var_scores |> 
  as_tibble() |> 
  mutate(variable = variables)

PC_percent <- pca_w$CA$eig / sum(pca_w$CA$eig)

data_w_pca <- left_join(data_w_pot, pot_scores, by = join_by(biol_unit_nb))
```

Plot it

```{r}
#dev.new()
plot_pca_w_cs <- plot_pca(data_w_pca, var_scores, PC_percent, colour_point = "cs") +
  geom_point(aes( colour = cs)) +
  scale_color_discrete(palette = c(cs_colors["IC"], cs_colors["W"])) +
  labs(title = "PCA - Wheat")
  
#plot_pca_w_cs

plot_pca_w_soil <- plot_pca(data_w_pca, var_scores, PC_percent, colour_point = "soil") +
 # geom_point(aes(shape = cs)) +
  labs(title = "PCA - Wheat") +
  scale_color_discrete(palette = c(soil_colors)) 

plot_pca_w_soil
```

### 3.2.4 - All plots

```{r}
#| warning: false
plot_pca_fb_cs + plot_pca_w_cs + plot_layout(guides = "collect")

plot_pca_fb_soil + plot_pca_w_soil + plot_layout(guides = "collect")
```

# What now?

Wait until we include single SIRs

# 3 - Analysis on Field Data

## 3.1 - Load and clean data

[**! Yield data to come: script 1_CRA_Data still needs to be worked on**]{.underline}

```{r}
Npools_t2_field <- read_rds("output/data/2_Npools.rds") |> 
  filter(expe == "Field")
yield_field <- read_rds("output/data/1_field_yield_clean.rds")
MR_field <- read_rds("output/data/1_MR_data.rds") |> 
  filter(expe == "Field") |> 
  filter_out(soil == c("Std")) |> 
  mutate(biol_unit_nb = as.numeric(str_extract(sample_id, "t2_(\\d*)_z\\d", group = 1))) |> 
  select(!c(dataset))

PMN_field <- read_rds("output/data/2_PMN.rds") |> filter(expe == "Field")




# field_Npools <- read_rds("output/data/4.1_t2_field_Npools.rds") |> 
#   select(!c(expe, sampling_time, dilution)) 
# 
# field_yield <- read_rds("output/data/3_field_yield_clean.rds")
# 
# PMN <- read_rds("output/data/4.1_t2_field_PMN.rds")
# 
# MR <- read_rds("output/data/1_MR_data.rds") |> filter(dataset == "MR_field")
# 
# #str(field_Npools)
```

```{r}
# check out graphically that not too many outliers
Npools_t2_field |> 
  ggplot(aes(x = soil, colour = cs, y = NO3_Nmin)) + 
  facet_wrap(~bloc) +
  geom_boxplot() +
  geom_point()

Npools_t2_field |> 
  ggplot(aes(x = soil, colour = cs, y = NH4_Nmin)) + 
  facet_wrap(~bloc) +
  geom_boxplot() +
  geom_point()

yield_field |>
  ggplot(aes(x = soil, colour = crop, y = grain_yield_15p_t_per_ha)) +   geom_boxplot() +
  geom_point()

yield_field |>
  ggplot(aes(x = soil, colour = crop, y = protein_percent_dw)) +
  geom_boxplot() +
  geom_point()
```

# °°° !!! START HERE !!! °°°

Now we want to get soil and yield data together, per crop (starting with Faba Bean, then Wheat). For that, we need to compute the mean soil values across the 3 zones (bc yield was taken once per plot).

```{r}
field_Npools_mean <- field_Npools |> 
  group_by(soil, cs, bloc, biol_unit_nb) |> 
  summarise(
    MBN_ppm = mean(MBN_ppm, na.rm = TRUE),
    TDN_ppm = mean(TDN_ppm, na.rm = TRUE),
    ppm_NH4 = mean(ppm_NH4, na.rm = TRUE),
    ppm_NO3 = mean(ppm_NO3, na.rm = TRUE),
    ppm_NO2 = mean(ppm_NO2, na.rm = TRUE)
  ) |> 
  mutate(
    Nmin_ppm = ppm_NH4 + ppm_NO3 + ppm_NO2,
    Norg_ppm = TDN_ppm - Nmin_ppm,
    NO3_NH4 = ppm_NO3 / ppm_NH4,
    NO3_Nmin = ppm_NO3 / Nmin_ppm,
    Nmin_TDN = Nmin_ppm / TDN_ppm) |> 
  ungroup()

#pairs(field_Npools_mean |> select(!c(soil,cs, biol_unit_nb)))
```

We subset yield per plant and join the corresponding soil data

```{r}
data_fb <- field_yield |> 
  filter(crop == "Faba bean") |> 
  arrange(biol_unit_nb) |> 
  left_join(field_Npools_mean)

data_w <- field_yield |> 
  filter(crop == "Wheat") |> 
  arrange(biol_unit_nb) |> 
  left_join(field_Npools_mean)

# check it out
data_fb
data_w
```

# 
