---
title: "3 Integrated Data Analysis"
author: "Morgane de Toeuf"
format: 
  gfm:
    toc: true
    code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# To Do

# Set up

```{r}
#| message: false
#| warning: false

# clean environment
rm(list = ls())

# packages
library(tidyverse)
library(vegan) # for rda()
library(patchwork)
library(RColorBrewer)
library(emmeans) # for post-hoc test (after ANCOVA of PMN)

# function
source("functions/plot_pca.R")

```

# 1 - Introduction

## 1.1 - Greenhouse

For the greenhouse data sets, many bits and pieces are dropped along the way, so that only few variables are left in the end:

- The whole t1 sampling is problematic due to the scattered sampling over time

- For t2:

  - MicroResp experiment: still needs calibration, but responses were really low, probably due to high sand levels. We'll look into it anyway

  - PMN: samples had to be re-humidified because the container system was not optimal and samples had become basically dry. See script on data wrangling: curves show interesting trends, but incubation times 2 and 3 problematic. After discussion with PI: removed those points, curves are better now. Still a lot of noise, which could come from the added variability of the re-humidification + the missing data.

  - TDN & MBN: never got the protocol from Sang, and did a mistake by not adjusting the serial dilution of the standard curve –\> samples out of range for Greenhouse

  - WHC is not intrinsically interesting, though we have the data as well (was needed for MicroResp)

In the end, what is left here is less than anticipated:

- Yield data is trustworthy (plant biomass, plant height, nb of stems / tillers)

  - LER can be derived from yield data (but = aggregated data –\> no incorporation in raw data PCA possible). But PCA with variables with ratio IC/SC

- Npools are incomplete:

  - Nmin data at t2 (NO2, NO3, NH4 –\> Nmin and ratios)

  - Potential mineralization of N (at sowing time)

  - But [NO]{.underline} Total Dissolved N, Microbial biomass N and organic N

- MicroResp data give an idea of soil biomass, soil metabolism potential, functional diversity

What is still dormant is weed data. Relevant? not sure...

## 1.2 - Field

## 1.3 - Experimental design and questions asked

Whereas the greenhouse had a full design with both sole crops and intercrop, and all modalitites present, this is not the case in the field.

- First, there was no sole Faba Bean, so we cannot approach data with the same ratios IC/SC as in the greenhouse (or then only for wheat, which still bears some interest).

- Second, the Faba Bean did not take in the Reference Soil in the field, so that

  - for all 3 soils, we can compare sole wheat, that is, it is an experimental design with 1 factor (soil) and 3 levels (Ref, Auto, ABC), with factorial block design and a system trial (it's the whole farming system that changes)

  - for only 2 soils (Auto and ABC), we can also compare the response of wheat to IC from one soil to the other. But here, it is no longer a system trial because only ploughing is different (well, and maybe weeding? to be checked). And it is a split plot design, with soil as the main factor (2 levels) and crop diversity as the nested factor (2 levels: sole wheat and intercropped wheat)

## 1.4 - Setting general parameters

```{r}
p_signif <- tribble(
  ~ signif_stars, ~ max_p_value,
  "",    1,
  ".",   0.1,
  "*",   0.05,
  "**",  0.01,
  "***", 0.001
)
```

# 2 - Analysis on Greenhouse Data

## 2.1 - Load and clean data

[**! Data from the standard soils is not in `Npools_t2`. They will need to be addressed separately, if at all.**]{.underline}

```{r}

Npools_t2_greenhouse <- read_rds("output/data/2_Npools.rds") |> 
  filter(expe == "Pot")
yield_clean_greenhouse <- read_rds("output/data/1_greenhouse_yield_clean.rds")
MR_greenhouse <- read_rds("output/data/1_MR_data.rds") |> 
  filter(expe == "Pot") |> 
  filter_out(soil %in% c("Std", "Sand") | cs == "B") |> 
  mutate(biol_unit_nb = as.numeric(sample_id), .keep = "unused") |> 
  select(!c(dataset, zone))

PMN_greenhouse <- read_rds("output/data/2_PMN.rds") |> filter(expe == "Pot")
```

First, we join all pot-level data, i.e., Npool data, yield data and MicroResp data. PMN cannot be joined because it is a measure at soil-level.

```{r}
pot_level_data <- Npools_t2_greenhouse |> 
  select(
    biol_unit_nb, expe, soil, crop_diversity, cs, bloc, 
    dm:last(names(Npools_t2_greenhouse))
    ) |> 
  left_join(yield_clean_greenhouse) |> 
  left_join(MR_greenhouse) |> 
  select(!starts_with("Norg")) |> 
  mutate(soil = factor(soil, levels = c("Conv", "Ref", "Auto", "ABC")))
```

Problem: for now, the statistical unit is the pot (1 row per pot). But some variables cannot be regrouped for both plant species. Consequently, there are some missing data, see here:

```{r}
pot_level_data |> select(cs, fw_fb_per_pot:dw_w_per_plant)
```

A PCA cannot be ran with missing data.

A solution is to change the statistical unit to be the plant, and have 2 data sets: dataset fababean (`data_fb`) and dataset wheat (`data_w`). Anyway, it makes little sense to compare the 2 plant species, but it is interesting to see if effects are similar or not. To be considered: bulk soil data of intercrops will appear twice, so that the 2 datasets are not to be combined in downstream analyses.

So we do that subsetting

```{r}
data_fb_pot <- pot_level_data |> 
  select(!contains("_w")) |> 
  filter(cs != "W") |> 
  arrange(biol_unit_nb)

data_w_pot <- pot_level_data |> 
  select(!contains("_f")) |> 
  filter(cs != "F") |> 
  arrange(biol_unit_nb)

# check it out
str(data_fb_pot)
str(data_w_pot)
```

## 2.2 - Compute LER

This also cannot go into the regular PCA since we get only data that agregates on SC + IC.

So we can have:

- a regular PCA with Npools, MicroResp and yield data on a pot level (one plant at the time)

- A reduced PCA with Npools and microresp expressed as a ratio of value for IC / value for SC, together with LER, also one plant at the time

```{r}
#| message: false
ler_fb_pot <- data_fb_pot |> 
  select(crop_diversity, soil, bloc, dw_fb_per_plant) |> 
  pivot_wider(
    names_from = crop_diversity,
    values_from = dw_fb_per_plant) |> 
  mutate(
    ler_fb = IC / SC
  )
  
ler_w_pot <- data_w_pot |> 
  select(crop_diversity, soil, bloc, dw_w_per_plant) |> 
  pivot_wider(
    names_from = crop_diversity,
    values_from = dw_w_per_plant) |> 
  mutate(
    ler_w = IC / SC
  )

ler_total_pot <- ler_w_pot |> 
  select(!SC:IC) |> 
  left_join(ler_fb_pot |> select(!SC:IC)) |> 
  mutate(ler_total = (ler_w + ler_fb)/2)
```

Plot it

```{r}
#| warning: false
plot_ler_total_pot <- ler_total_pot |> 
  ggplot(aes(x = soil, y = ler_total)) +
  theme_minimal() +
  geom_boxplot() +
  labs(title = "LER total", subtitle = "Pot experiment")

plot_ler_fb_pot <- ler_total_pot |> 
  ggplot(aes(x = soil, y = ler_fb)) +
  theme_minimal() +
  geom_boxplot() +
  labs(title = "LER Faba Bean", subtitle = "Pot experiment")

plot_ler_w_pot <- ler_total_pot |> 
  ggplot(aes(x = soil, y = ler_w)) +
  theme_minimal() +
  geom_boxplot() +
  labs(title = "LER Wheat", subtitle = "Pot experiment")

plot_ler_total_pot + plot_ler_fb_pot + plot_ler_w_pot
```

[**!! TO DO: check those outliers: could it be some of those pots where some plants were not as should be? Or anything? Can I remove them?**]{.underline}

## 2.3 - PMN

### 2.3.1 - ANCOVA

With the PMN data, we can compute the slope of the curve.

::: callout-important
## Statistical issue

We have 4 technical replicates per incubation time, but rep 1 of incubation time 1 has no particular relationship to rep 1 of incubation time 2. It would make little sense to compute 4 different slopes and then take the mean of the 4 slopes to get a boxplot and the option of an anova with comparison of means (of slopes).
:::

According to AI (see full doc):

- Analysis of CoVariance, ANCOVA, (independant measurements bc subsamples evolve separately) .

  - linear model with interaction term: outcome \~ group \* predictor

  - --\> try with lm(ppm_Nmin \~ soil \* incub_days)

  - Details see doc

```{r}
PMN_continuous <- PMN_greenhouse |> mutate(incub_day = as.integer(incub_day))

my_lm <- lm(data = PMN_continuous, ppm_Nmin ~ incub_day * soil)
summary(my_lm)

anova(my_lm)
```

The Anova (in this case ANCOVA) says:

- Effect of incubation time (duh)

- effect of soil

- interaction effect –\> reject (H0 = slopes are equal)

—\> need of a post-hoc test, with Tukey adjustment

```{r}
slope_trends <- emtrends(my_lm, pairwise ~ soil, var = "incub_day")

summary(slope_trends)

# see the slopes
slope_trends$emtrends

# see the pairwise contrasts with adjusted p-values (Tukey method)
slope_trends$contrasts
```

- conv ≠ Ref, ABC

- Conv = Auto

- Ref = ABC, Ref = Auto

- Auto ≠ ABC

```{r}
ancova_signif <- slope_trends$contrasts |> as_tibble() |> 
  mutate(signif = c("**", "", "***", "", "", "*"))

```

Checking model assumptions

```{r}
library(performance)
plot_check_ancova <- check_model(my_lm) 
plot <- plot(plot_check_ancova) + patchwork::plot_annotation(title = "QC: Checking model assumptions for ANCOVA of PMN - Greenhouse data")

# check it out
plot

# save it
ggsave("output/figures/greenhouse/qc_pmn_ancova_greenhouse.pdf", plot,height = 8, width = 13.5)
```

### 2.3.2 - Plot PMN

```{r}
dates <- c("2023-12-5", "2023-12-26", "2024-01-02") |> as.Date()
incub_days <- dates - dates[1]
names(incub_days) <- PMN$incub_time |> unique() |> sort()

# Plot
# plot again
PMN_plot <- PMN_greenhouse |> 
  ggplot(aes(x = incub_day, y = ppm_Nmin, group = soil, colour = soil, fill = soil)) +
  theme_minimal() +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    title = "Potential Mineralization of Nitrogen (PMN) - Greenhouse",
    subtitle = "Mineral N as sum of NO2-, NO3-, NH4+\nSlopes are statistically different (Conv ≠ Ref, ABC; Auto ≠ ABC)",
    caption = "Modelled with lm(ppm_Nmin ~ incub_time * soil)\nPost-Hoc with emtrends() and Tukey correction") +
  xlab("Time since incubation at 28°C [days]") +
  ylab("Mineral Nitrogen in the soil [ppm]")

# for annotation: extract data from smooth curve
smooth_data <- ggplot_build(PMN_plot)$data[[2]]

# get its maximum value to anker the annotation
annotations <- smooth_data |> 
  slice_max(x, by = group) |> 
  mutate(soil = levels(PMN_greenhouse$soil))
#c("#F8766D", "#7CAE00", "#00BFC4", "#C77CFF") 

PMN_plot_all <- 
  PMN_plot + 
  geom_text(
    data = annotations,
    aes(x = x+0.5, y = y, colour = soil, label = soil), 
    hjust = 0, size = 5) +
  xlim(c(0,45)) +
  theme(legend.position = "none") #+

signif_annotations <- ancova_signif |> 
  filter(signif %in% c(".", "*", "**", "***")) |> 
  select(contrast, p.value, signif) |> 
  mutate(
    start = str_extract(contrast, "(\\w*) - (\\w*)", group = 1),
    end = str_extract(contrast, "(\\w*) - (\\w*)", group = 2)
  ) |> 
  left_join(
    annotations |> select(soil, x, y) |> rename(start = soil)
  ) |> 
  left_join(
    annotations |> select(soil, x, y) |> rename(end = soil, yend = y)
  ) |> 
  mutate(
    x = x + c(11,7,15),
    y = y - 0.1,
    yend = yend + 0.1) |> 
  mutate(soil = start)

plot_signif <- PMN_plot_all +
  geom_segment(
    data = signif_annotations,
    aes(x = x, y = y, yend = yend),
    colour = "grey30"
  ) +
  geom_text(
    data = signif_annotations,
    aes(x = x + 1, y = (y + yend)/2, label = signif),
    colour = "grey30"
  ) +
  geom_segment(
    data = signif_annotations,
    aes(x = x, xend = x - 1, y = yend, yend = yend),
    colour = "grey30"
  ) +
  geom_segment(
    data = signif_annotations,
    aes(x = x, xend = x - 1, y = y, yend = y),
    colour = "grey30"
  )
  

```

Check it out & save

```{r}
plot_signif  

ggsave("output/figures/greenhouse/plot_pmn_signif_greenhouse.pdf", plot = plot_signif, height = 5, width = 7)
```
