---
title: "3 Greenhouse, t2 - Import and tidy data"
format: 
  gfm:
    toc: true
    code-fold: true
# format: 
#   html:
#     embed-resources: true
#     toc: true
#     toc_float: true
#     code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# To Do

- Link to a MicroResp pipeline!

# Set up

```{r}
#| message: false
#| warning: false

rm(list = ls())

library(plate2N) # for remove_wells
library(tidyverse)
library(janitor)
library(roperators) # for %ni%
library(ggrepel) # for geom_text_repel()
library(ggridges) # for geom_density_ridges()
library(patchwork) # for the "+" layout and plot_layout()

# functions
source("functions/plot_qc_sample_conc.R")
```

# 1 - Import data

## 1.1 - Non-absorbance data from AELab

```{r}
# import other "wet lab" raw data
raw_data_pot <- read_csv(
  "../raw_data/2024_raw.csv", show_col_types = TRUE,
  col_types = list(
    Soil = col_factor(),
    crop_diversity = col_factor(),
    CS = col_factor(),
    bloc = col_factor()
  ),
  na = c("", "NA", "Na")
  ) |> clean_names() |> 
  rename(sample_short = short) |> 
  filter(expe == "Pot")

raw_greenhouse_t2 <- raw_data_pot |> 
  filter(sampling_time == "t2") |> 
  # remove bare soil
  filter_out(cs == "B") |> 
  # correct biological unit
  mutate(biol_unit_nb = case_when(
    biol_unit_nb < 200 ~ biol_unit_nb,
    biol_unit_nb > 200 ~ biol_unit_nb - 200
  )) |> 
  arrange(biol_unit_nb) |> 
  # remove useless columns
  select(
    biol_unit_nb:sampling_time, 
    sample_name, #useful?
   starts_with(c("whc", "flush", "yd_rs"))
    )

# Check it out
raw_greenhouse_t2
```

## 1.2 - Microresp Data (TO DO?)

Then, MicroResp data may need to be added, probably with its own pipeline

## 1.3 - Absorbance data

This data needs tidying, because we still have 4 values per sample (corresponding to the 4 wells given to each sample for analytical replicates on the absorbance plate.

First, we import this dataset

```{r}
raw_Nmin <- read_rds("output/data/2_mgNL_noTDN.rds") |> filter_out(dataset == "Nmint3")
```

### 1.3.1 - Nmin: Subset t2, greenhouse data set, no bare soil

Then we take a subset (t2 only, greenhouse only, no bare soil).

This complex pipeline is only necessary because I was not very consistent in plate-naming structure.

```{r}
cs_map <- raw_greenhouse_t2 |> select(biol_unit_nb, cs, soil) |> unique()


raw_greenhouse_t2_Nmin <- raw_Nmin |> 
  filter_out(dataset == "PMN") |> 
  # create sampling_time and expe variables from plate_ids (first number and first letter after N species)
  mutate(
    # paste "t" and the sampling time as number: for t1 and t2: is stored in plate data (--> str_extract)
    sampling_time = paste0(
      "t", 
      str_extract(plate_id, "\\w_(\\d)(\\w).*", group = 1)
      ),
    # for expe: will work for t1 and t2, 
    expe = str_extract(plate_id, "\\w_(\\d)(\\w).*", group = 2),
    # rephrase "P", "G" into "Pot" (G was for greenhouse)
    expe = case_when(expe %in% c("P", "G") ~ "Pot", .default = expe),
    .before = plate_id) |> 
  # filter based on sampling_time
  filter(sampling_time == "t2", expe == "Pot") |> 
  separate_wider_delim(
    cols = map,
    names = c("biol_unit_nb"),
    delim = "_",
    too_many = "drop", 
    cols_remove = FALSE
  ) |> 
  mutate(
    biol_unit_nb = as.double(biol_unit_nb)) |> 
  # remove bare soils (multiples of 4)
  filter_out(biol_unit_nb %ni% cs_map$biol_unit_nb) |> 
  # add info on crop stand and soil
  left_join(cs_map)
```

### 1.3.2 - PMN: Subset greenhouse

```{r}
raw_greenhouse_PMN <- raw_Nmin |> 
  filter(dataset == "PMN") |> 
  separate_wider_delim(
    cols = map,
    names = c("expe", "soil", "incubation_time", "tech_rep"),
    delim = "_",
    cols_remove = FALSE
  ) |> 
  mutate(
    sampling_time = rep("t0"),
    biol_unit_nb = paste0(expe, "_", soil),
    .before = well_id) |> 
  filter_out(expe == "Field")
```

Check out both subsets. First, Nmin data

```{r}
raw_greenhouse_t2_Nmin
```

Then, PMN data

```{r}
raw_greenhouse_PMN
```

# 2 - Tidy data: per-sample outlier removal

## 2.1 - Nmin (samples)

So we will compute the per-sample average, for the greenhouse dataset. But first, let's have a look at the distribution of concentrations before taking the average: are there clear outliers?

### 2.1.1 - NO3

```{r}
boxplot_no3 <- raw_greenhouse_t2_Nmin |> 
  filter(biol_unit_nb < 100, std_sp == "NO3") |> 
  boxplot_conc() + labs(title = "NO3")

ridges_no3 <- raw_greenhouse_t2_Nmin |> 
  filter(biol_unit_nb < 100, std_sp == "NO3") |>  # exclude sand and conv soil std
  plot_ridges_conc() + facet_wrap(~soil, ncol = 4) + labs(title = "NO3")
```

I see some points that are a little bit off compared to the other three.

- samples where the highest concentration seems to be an outlier:

  - 21(H7), 54(H3), 15(A3), 61(C4), 77(H11), 69(H3), 10(A9), 41(A9), 74(A9)

- samples where the lowest concentration seems to be an outlier:

  - 49(E10), 6(D11), 38(E10), 53(E10), 71(F9), 42(C9)

Manually removing those points, then re-running the plotting

```{r}
biol_unit <- c(21, 54, 15, 61, 77, 69, 10, 41, 74, 49, 6, 38, 53, 71, 42)
wells <- c("H7", "H3", "A3", "C4", "H11", "H3", "A9", "A9", "A9", "E10", "D11", "E10", "E10", "F9", "C9")
names(wells) <- biol_unit

to_remove <- tibble(
  biol_unit_nb = biol_unit,
  well_id = wells
) |> left_join(
  raw_greenhouse_t2_Nmin |> 
    filter(std_sp == "NO3") |> 
    select(biol_unit_nb, plate_id, dataset) |> 
    unique())

Nmin_wash1 <- remove_wells(
  table_to_clean = raw_greenhouse_t2_Nmin,
  well_table = to_remove) 



boxplot_no3_outlierfree <- Nmin_wash1 |> 
  filter(biol_unit_nb < 100, std_sp == "NO3") |> # exclude sand and conv soil std
  boxplot_conc(x = "biol_unit_nb", y = "conc_mgN_L") + labs(title = "NO3, outliers removed")

boxplot_no3 + boxplot_no3_outlierfree

ridges_no3_outlierfree <- Nmin_wash1 |> 
  filter(biol_unit_nb < 100, std_sp == "NO3") |> # exclude sand and conv soil std
  plot_ridges_conc() + facet_wrap(~soil, ncol = 4) + labs(title = "NO3, outliers removed")

ridges_no3 + ridges_no3_outlierfree + plot_layout(guides = "collect")
```

### 2.1.2 - NH4

Now same, for NH4.

```{r}
boxplot_nh4 <- Nmin_wash1 |> 
  filter(biol_unit_nb < 100, std_sp == "NH4") |> # exclude sand and conv soil std
  boxplot_conc() + labs(title = "NH4")
  
ridges_nh4 <- Nmin_wash1 |> 
  filter(biol_unit_nb < 100, std_sp == "NH4") |> # exclude sand and conv soil std
  plot_ridges_conc() + facet_wrap(~soil, ncol = 4) + labs(title = "NH4")
```

Here are those where I would remove

- the highest concentration:

  - 6(C11), 15(C3), 23(E2), 37(A7), 45(D10), 53(F10), 66(F4), 70(G10)

- the lowest concentration:

  - 18(G9)

- those where I am really unsure:

  - 33, 75 –\> should I remove those samples all together?

```{r}
biol_unit <- c(6, 15, 23, 37, 45, 53, 66, 70, 18)
wells <- c("C11", "C3", "E2", "A7", "D10", "F10", "F4", "G10", "G9")
names(wells) <- biol_unit

to_remove <- tibble(
  biol_unit_nb = biol_unit,
  well_id = wells
) |> left_join(
  Nmin_wash1 |> 
    filter(std_sp == "NH4") |> 
    select(biol_unit_nb, plate_id, dataset) |> 
    unique())

Nmin_wash2 <- remove_wells(
  table_to_clean = Nmin_wash1,
  well_table = to_remove) 

boxplot_nh4_outlierfree <- Nmin_wash2 |> 
  filter(biol_unit_nb < 100, std_sp == "NH4") |> # exclude sand and conv soil std
  boxplot_conc() + labs(title = "NH4, outliers removed")

ridges_nh4_outlierfree <- Nmin_wash2 |> 
  filter(biol_unit_nb < 100, std_sp == "NH4") |> # exclude sand and conv soil std
  plot_ridges_conc() + facet_wrap(~soil, ncol = 4) + labs(title = "NH4, outliers removed")

boxplot_nh4 + boxplot_nh4_outlierfree

ridges_nh4 + ridges_nh4_outlierfree + plot_layout(guides = "collect")
```

### 2.1.3 - NO2

Because NO2 readings are virtually zero, there is only little point in removing outliers. Nevertheless, we can have a look at the data

Per sample

```{r}
boxplot_no2 <- Nmin_wash2 |> 
  filter(biol_unit_nb < 100, std_sp == "NO2") |> # exclude sand and conv soil std
  boxplot_conc() + labs(title = "NO2")

ridges_no2 <- Nmin_wash2 |> 
  filter(biol_unit_nb < 100, std_sp == "NO2") |> # exclude sand and conv soil std
  plot_ridges_conc() + facet_wrap(~soil, ncol = 4) + labs(title = "NO2")

boxplot_no2 + ridges_no2 + plot_layout(guides = "collect")
```

Few to remove

```{r}
biol_unit <- c(37, 42, 49, 54)
wells <- c("D7", "C9", "E10", "H3")

to_remove <- tibble(
  biol_unit_nb = biol_unit,
  well_id = wells, 
  std_sp = rep("NO2")
) |> left_join(Nmin_wash2)

Nmin_wash3 <- Nmin_wash2 |> remove_wells(to_remove)

boxplot_no2_2 <- Nmin_wash3 |> 
  filter(biol_unit_nb < 100, std_sp == "NO2") |> # exclude sand and conv soil std
  boxplot_conc() + labs(title = "NO2, outliers removed")

ridges_no2_2 <- Nmin_wash3 |> 
  filter(biol_unit_nb < 100, std_sp == "NO2") |> # exclude sand and conv soil std
  plot_ridges_conc() + facet_wrap(~soil, ncol = 4) + labs(title = "NO2, outliers removed")

boxplot_no2_2 + ridges_no2_2 + plot_layout(guides = "collect")

boxplot_no2 + boxplot_no2_2 + plot_layout(guides = "collect")
ridges_no2 + ridges_no2_2 + plot_layout(guides = "collect")
```

## 2.2 - Nmin (Standard Soils)

### 2.2.1 - NO3

For NO3,

```{r}
boxplot_std_no3 <- Nmin_wash3 |> 
  filter(biol_unit_nb > 100, std_sp == "NO3") |> # 
  boxplot_conc(x = "soil", colour = "soil") + labs(title = "Standard soil NO3")

ridges_std_no3 <- Nmin_wash3 |> 
  filter(biol_unit_nb > 100, std_sp == "NO3") |> # exclude sand and conv soil std
  plot_ridges_conc(colour = "soil") + labs(title = "Standard soil NO3")

boxplot_std_no3 + ridges_std_no3
```

I find that it looks good on a per-standard basis. But each had many samples, let's look at a per-sample basis

```{r}
boxplot_std_no3_2 <- Nmin_wash3 |> 
  filter(biol_unit_nb > 100, std_sp == "NO3") |> # 
  boxplot_conc(x = "map", colour = "soil") + 
  labs(title = "Standard soil NO3 per sample") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ridges_std_no3_2 <- Nmin_wash3 |> 
  filter(biol_unit_nb > 100, std_sp == "NO3") |> # exclude sand and conv soil std
  plot_ridges_conc(groups = "map", colour = "soil") + 
  facet_wrap(~map) + labs(title = "Standard soil NO3 per sample") 

boxplot_std_no3_2 + ridges_std_no3_2 + plot_layout(guides = "collect")
```

Still looks good, we keep it.

### 2.2.2 - NH4

For NH4, same, first per standard

```{r}
boxplot_std_nh4 <- Nmin_wash3 |> 
  filter(biol_unit_nb > 100, std_sp == "NH4") |> # 
  boxplot_conc(x = "soil", colour = "soil") + labs(title = "Standard soil NH4")

ridges_std_nh4 <- Nmin_wash3 |> 
  filter(biol_unit_nb > 100, std_sp == "NH4") |> # exclude sand and conv soil std
  plot_ridges_conc(colour = "soil") + labs(title = "Standard soil NH4")

boxplot_std_nh4 + ridges_std_nh4 + plot_layout(guides = "collect")
```

Multimodal curve –\> dig into per-sample view

```{r}
boxplot_std_nh4_2 <- Nmin_wash3 |> 
  filter(biol_unit_nb > 100, std_sp == "NH4") |> # 
  boxplot_conc(x = "map", colour = "soil") + 
  labs(title = "Standard soil NH4, per sample") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ridges_std_nh4_2 <- Nmin_wash3 |> 
  filter(biol_unit_nb > 100, std_sp == "NH4") |> # exclude sand and conv soil std
  plot_ridges_conc(groups = "map", colour = "soil") + facet_wrap(~map) + 
  labs(title = "Standard soil NH4, per sample")

boxplot_std_nh4_2 + ridges_std_nh4_2 + plot_layout(guides = "collect")
```

I find 2 wells to remove:

- plate 109_t2_MR_R8, well A5

- plate 110_t2_MR_R7, well G9

Let's remove them and re-run the analysis

```{r}
samples <- c("109_t2_MR_R8", "110_t2_MR_R7")
wells <- c("A5", "G9")

to_remove <- Nmin_wash3 |> filter(
  ((map == samples[1] & well_id == wells[1]) |
    (map == samples[2] & well_id == wells[2])) &
    std_sp == "NH4"
)

Nmin_wash4 <- Nmin_wash3 |> remove_wells(to_remove)

boxplot_std_nh4_3 <- Nmin_wash4 |> 
  filter(biol_unit_nb > 100, std_sp == "NH4") |> 
  boxplot_conc(x = "map", colour = "soil") + 
  labs(title = "Standard soil NH4\nPer sample, outliers removed") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ridges_std_nh4_3 <- Nmin_wash4 |> 
  filter(biol_unit_nb > 100, std_sp == "NH4") |> 
  plot_ridges_conc(groups = "map", colour = "soil") + 
  facet_wrap(~map) + labs(title = "Standard soil NH4\nPer sample, outliers removed")

boxplot_std_nh4_3 + ridges_std_nh4_3 + plot_layout(guides = "collect")
```

Compare before/after outlier removal

```{r}
boxplot_std_nh4_2 + boxplot_std_nh4_3 + plot_layout(guides = "collect")

ridges_std_nh4_2 + ridges_std_nh4_3 + plot_layout(guides = "collect")
```

### 2.2.3 - NO2

For NO2, per standard

```{r}
boxplot_std_no2 <- Nmin_wash4 |> 
  filter(biol_unit_nb > 100, std_sp == "NO2") |> 
  boxplot_conc(x = "soil", colour = "soil") + labs(title = "Standard soil NO2")

ridges_std_no2 <- Nmin_wash4 |> 
  filter(biol_unit_nb > 100, std_sp == "NO2") |> 
  plot_ridges_conc(colour = "soil") + labs(title = "Standard soil NO2")

boxplot_std_no2 + ridges_std_no2 + plot_layout(guides = "collect")
```

Then per sample

```{r}
boxplot_std_no2_2 <- Nmin_wash4 |> 
  filter(biol_unit_nb > 100, std_sp == "NO2") |> 
  boxplot_conc(x = "map", colour = "soil") + 
  labs(title = "Standard soil NO2 per sample") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ridges_std_no2_2 <- Nmin_wash4 |> 
  filter(biol_unit_nb > 100, std_sp == "NO2") |> 
  plot_ridges_conc(groups = "map", colour = "soil") + facet_wrap(~map) + 
  labs(title = "Standard soil NO2 per sample")

boxplot_std_no2_2 + ridges_std_no2_2 + plot_layout(guides = "collect")
```

to remove: plate 110_t2_MR_R8, well H6

```{r}
to_remove <- Nmin_wash4 |> 
  filter(
    map == "110_t2_MR_R8",
    well_id == "H6",
    std_sp == "NO2")

Nmin_wash5 <- Nmin_wash4 |> remove_wells(to_remove)

boxplot_std_no2_3 <- Nmin_wash5 |> 
  filter(biol_unit_nb > 100, std_sp == "NO2") |> 
  boxplot_conc(x = "map", colour = "soil") + 
  labs(title = "Standard soil NO2\nPer sample, outliers removed") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ridges_std_no2_3 <- Nmin_wash5 |> 
  filter(biol_unit_nb > 100, std_sp == "NO2") |> 
  plot_ridges_conc(groups = "map", colour = "soil") + facet_wrap(~map) + 
  labs(title = "Standard soil NO2\nPer sample, outliers removed")

boxplot_std_no2_2 + boxplot_std_no2_3 + plot_layout(guides = "collect")
ridges_std_no2_2 + ridges_std_no2_3 + plot_layout(guides = "collect")
```

### 2.2.4 - All outliers removed Nmin

Visually, I am satisfied with this outlier removal, so I save this cleaned table

```{r}
greenhouse_t2_Nmin_clean <- Nmin_wash5
```

## 2.3 - PMN

### 2.3.1 - NO3

```{r}
boxplot_pmn_no3 <- raw_greenhouse_PMN |> 
  filter(std_sp == "NO3") |> 
  boxplot_conc(x = "tech_rep") + facet_wrap(soil~incubation_time, scales = "free_y") + 
  labs(title = "NO3")

ridges_pmn_no3 <- raw_greenhouse_PMN |> 
  filter(std_sp == "NO3") |> 
  plot_ridges_conc(groups = "soil", colour = "tech_rep", y = "map") + 
  facet_wrap(~incubation_time, nrow = 1) + labs(title = "NO3")

# to be sure which well was out
# raw_greenhouse_PMN |> 
#   filter(std_sp == "NO3", soil == "Ref", incubation_time == "i4", tech_rep == "rt2") |> 
#   boxplot_conc(x = "tech_rep") + facet_wrap(soil~incubation_time) + labs(title = "NO3")


boxplot_pmn_no3 + ridges_pmn_no3 
```

- Pot_ABC_i1_rt4: C7

- Pot_Ref_i4_rt2: H5

- Pot_Auto_i2_rt4: A10

```{r}
to_remove <- raw_greenhouse_PMN |> 
  filter((std_sp == "NO3") & 
    ((map == "Pot_ABC_i1_rt4" & well_id == "C7") | 
       (map == "Pot_Ref_i4_rt2" & well_id == "H5") |
       (map == "Pot_Auto_i2_rt4" & well_id == "A10")) 
  )

PMN_wash1 <- raw_greenhouse_PMN |> remove_wells(to_remove)

# check it out again
boxplot_pmn_no3_outlierfree <- PMN_wash1 |> 
  filter(std_sp == "NO3") |> 
  boxplot_conc(x = "tech_rep") + facet_wrap(soil~incubation_time, scales = "free_y") + 
  labs(title = "NO3, outlier removed")

ridges_pmn_no3_outlierfree <- PMN_wash1 |> 
  filter(std_sp == "NO3") |> 
  plot_ridges_conc(groups = "soil", colour = "tech_rep", y = "map") + 
  facet_wrap(~incubation_time, nrow = 1) + labs(title = "NO3, outlier removed")

boxplot_pmn_no3_outlierfree + ridges_pmn_no3_outlierfree
```

Compare before/after

```{r}
boxplot_pmn_no3 + boxplot_pmn_no3_outlierfree + plot_layout(guides = "collect")

ridges_pmn_no3 + ridges_pmn_no3_outlierfree + plot_layout(guides = "collect")
```

### 2.3.2 - NH4

```{r}
boxplot_pmn_nh4 <- PMN_wash1 |> 
  filter(std_sp == "NH4") |> 
  boxplot_conc(x = "tech_rep") + facet_wrap(soil~incubation_time) + 
  labs(title = "NH4")

ridges_pmn_nh4 <- PMN_wash1 |> 
  filter(std_sp == "NH4") |> 
  plot_ridges_conc(groups = "soil", colour = "tech_rep", y = "map") + 
  facet_wrap(~incubation_time, nrow = 1) + labs(title = "NH4")


boxplot_pmn_nh4 + ridges_pmn_nh4 
```

- Pot_ABC_i2_rt3: A11

- Pot_ABC_i4_rt2: H7

- Pot_Ref_i1_rt2: C5

```{r}
to_remove <- PMN_wash1 |> filter(
  (std_sp == "NH4") & 
    ((map == "Pot_ABC_i2_rt3" & well_id == "A11") |
    (map == "Pot_ABC_i4_rt2" & well_id == "H7") |
    (map == "Pot_Ref_i1_rt2" & well_id == "C5")))

PMN_wash2 <- PMN_wash1 |> remove_wells(to_remove)

boxplot_pmn_nh4_outlierfree <- PMN_wash2 |> 
  filter(std_sp == "NH4") |> 
  boxplot_conc(x = "tech_rep") + facet_wrap(soil~incubation_time) + 
  labs(title = "NH4, outliers removed")

ridges_pmn_nh4_outlierfree <- PMN_wash2 |> 
  filter(std_sp == "NH4") |> 
  plot_ridges_conc(groups = "soil", colour = "tech_rep", y = "map") + 
  facet_wrap(~incubation_time, nrow = 1) + labs(title = "NH4, outliers removed")


boxplot_pmn_nh4_outlierfree + ridges_pmn_nh4_outlierfree

```

Compare before/after

```{r}
boxplot_pmn_nh4 + boxplot_pmn_nh4_outlierfree + plot_layout(guides = "collect")

ridges_pmn_nh4 + ridges_pmn_nh4_outlierfree + plot_layout(guides = "collect")
```

### 2.3.3 - NO2

```{r}
boxplot_pmn_no2 <- PMN_wash2 |> 
  filter(std_sp == "NO2") |> 
  boxplot_conc(x = "tech_rep") + facet_wrap(soil~incubation_time) + 
  labs(title = "NO2")

ridges_pmn_no2 <- PMN_wash2 |> 
  filter(std_sp == "NO2") |> 
  plot_ridges_conc(groups = "soil", colour = "tech_rep", y = "map") + 
  facet_wrap(~incubation_time, nrow = 1) + labs(title = "NO2")


boxplot_pmn_no2 + ridges_pmn_no2 
```

- Pot_ABC_i0_rt4: C4

- Pot_ABC_i1_rt4: B7

- Pot_Auto_i0_rt4: C3

- Pot_Auto_i1_rt4: B6

- Pot_Ref_i0_rt4: C2 & B2

```{r}
to_remove <- PMN_wash2 |> filter(
  (std_sp == "NO2") &
    ((map == "Pot_ABC_i0_rt4" & well_id == "C4") |
    (map == "Pot_ABC_i1_rt4" & well_id == "B7") |
    (map == "Pot_Auto_i0_rt4" & well_id == "C3") |
    (map == "Pot_Auto_i1_rt4" & well_id == "B6") |
    (map == "Pot_Ref_i0_rt4" & well_id %in% c("B2", "C2")))
)

PMN_wash3 <- PMN_wash2 |> remove_wells(to_remove)

boxplot_pmn_no2_outlierfree <- PMN_wash3 |> 
  filter(std_sp == "NO2") |> 
  boxplot_conc(x = "tech_rep") + facet_wrap(soil~incubation_time) + 
  labs(title = "NO2, outliers removed")

ridges_pmn_no2_outlierfree <- PMN_wash3 |> 
  filter(std_sp == "NO2") |> 
  plot_ridges_conc(groups = "soil", colour = "tech_rep", y = "map") + 
  facet_wrap(~incubation_time, nrow = 1) + labs(title = "NO2, outliers removed")


boxplot_pmn_no2_outlierfree + ridges_pmn_no2_outlierfree 
```

Compare before / after

```{r}
boxplot_pmn_no2 + boxplot_pmn_no2_outlierfree + plot_layout(guides = "collect")

ridges_pmn_no2 + ridges_pmn_no2_outlierfree + plot_layout(guides = "collect")
```

### 2.3.4 - All outliers removed PMN

Visually, I am satisfied with this outlier removal, so I save this cleaned table

```{r}
greenhouse_t2_PMN_clean <- PMN_wash3
```

# 3 - Per-sample mean

## 3.1 - Nmin

```{r}
conc_mean_Nmin <- greenhouse_t2_Nmin_clean |> 
  select(map, plate_id, biol_unit_nb, std_sp, conc_mgN_L) |> 
  group_by(plate_id, map, biol_unit_nb, std_sp) |> 
  summarise(
    mean = mean(conc_mgN_L),
    st_dev = sd(conc_mgN_L)) |> 
  mutate(coef_var = 100*st_dev / mean) |> 
  rename(conc_mgN_L = mean)

conc_mean_Nmin |> arrange(desc(std_sp), desc(coef_var))

conc_mean_Nmin |> filter(coef_var > 10) |> 
  ggplot(aes(x = coef_var)) + geom_histogram()
```

We still have quite a few samples with a high between-wells coefficient of variation. But with only 3 to 4 wells and sometimes very low values, this is unavoidable, so we move on

Now, finally, we re-join this mean value to the rest of the relevant information from the absorbance dataset

```{r}
conc_Nmin_export_ready <- conc_mean_Nmin |> 
  inner_join(
    raw_greenhouse_t2_Nmin |> 
      select(!c(well_id:abs_corrected, starts_with("conc"))) |> 
      unique())
```

## 3.2 - PMN

```{r}
conc_mean_PMN <- greenhouse_t2_PMN_clean |> 
  select(map, plate_id, biol_unit_nb, std_sp, conc_mgN_L) |> 
  group_by(map, biol_unit_nb, std_sp) |> 
  summarise(
    mean = mean(conc_mgN_L),
    st_dev = sd(conc_mgN_L)) |> 
  mutate(coef_var = 100*st_dev / mean) |> 
  rename(conc_mgN_L = mean)

conc_mean_PMN |> arrange(desc(std_sp), desc(coef_var))

conc_mean_PMN |> filter(coef_var > 10) |> 
  ggplot(aes(x = coef_var)) + geom_histogram()
```

Same here, there are still quite a few samples with a high coefficient of variation, but with only 3 to 4 wells and sometimes very low values, this is unavoidable, so we move on

Because there is no equivalence with biological unit nb of the rest of the data set, we cannot yet rejoin data sets, PMN will keep separate for now.

# 4 - Export

```{r}
raw_greenhouse_t2 |> write_rds("output/data/3_greenhouse_t2_raw_lab.rds")
conc_Nmin_export_ready |> write_rds("output/data/3_greenhouse_t2_Nmin_clean.rds")
conc_mean_PMN |> write_rds("output/data/3_greenhouse_PMN_clean.rds")
```
