---
title: "3 Greenhouse, t2 - Import non-absorbance data"
format: 
  gfm:
    toc: true
    code-fold: true
# format: 
#   html:
#     embed-resources: true
#     toc: true
#     toc_float: true
#     code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# To Do

- Link to a MicroResp pipeline!

# Set up

```{r}
#| message: false
#| warning: false

rm(list = ls())

library(plate2N) # for remove_wells
library(tidyverse)
library(janitor)
library(roperators) # for %ni%
library(ggrepel) # for geom_text_repel()
library(ggridges) # for geom_density_ridges()
library(patchwork) # for the "+" layout and plot_layout()
```

# 1 - Non-absorbance data from AELab

```{r}
# import other "wet lab" raw data
raw_data_pot <- read_csv(
  "../raw_data/2024_raw.csv", show_col_types = TRUE,
  col_types = list(
    Soil = col_factor(),
    crop_diversity = col_factor(),
    CS = col_factor(),
    bloc = col_factor()
  )
  ) |> clean_names() |> 
  rename(sample_short = short) |> 
  filter(expe == "Pot")

raw_greenhouse_t2 <- raw_data_pot |> 
  filter(sampling_time == "t2") |> 
  # remove bare soil
  filter_out(cs == "B") |> 
  # correct biological unit
  mutate(biol_unit_nb = case_when(
    biol_unit_nb < 200 ~ biol_unit_nb,
    biol_unit_nb > 200 ~ biol_unit_nb - 200
  )) |> 
  arrange(biol_unit_nb) |> 
  # remove useless columns
  select(
    biol_unit_nb:sampling_time, 
    sample_name, #useful?
   starts_with(c("whc", "flush", "yd_rs"))
    )

# Check it out
raw_greenhouse_t2
```

# 2 - Microresp Data

Then, MicroResp data may need to be added, probably with its own pipeline

# 3 - Absorbance data

This data needs tidying, because we still have 4 values per sample (corresponding to the 4 wells given to each sample for analytical replicates on the absorbance plate.

First, we import this dataset

```{r}
raw_Nmin <- read_rds("output/data/2_mgNL_noTDN.rds") |> filter_out(dataset == "Nmint3")
```

## 3.1 - Subset t2, greenhouse data set, no bare soil

Then we take a subset (t2 only, greenhouse only, no bare soil)

```{r}

cs_map <- raw_greenhouse_t2 |> select(biol_unit_nb, cs, soil) |> unique()

raw_greenhouse_t2_Nmin <- raw_Nmin |> 
  # create sampling_time and expe variables from plate_ids (first number and first letter after N species)
  mutate(
    sampling_time = paste0(
      "t", 
      str_extract(plate_id, "\\w_(\\d)(\\w).*", group = 1)),
    expe = str_extract(plate_id, "\\w_(\\d)(\\w).*", group = 2),
    expe = case_when(expe %in% c("P", "G") ~ "Pot", .default = expe),
    .before = plate_id) |>
  # filter based on sampling_time
  filter(sampling_time == "t2", expe == "Pot") |> 
  separate_wider_delim(
    cols = map,
    names = c("biol_unit_nb"),
    delim = "_",
    too_many = "drop", 
    cols_remove = FALSE
  ) |> 
  mutate(biol_unit_nb = as.double(biol_unit_nb)) |> 
  # remove bare soils (multiples of 4)
  filter_out(biol_unit_nb %ni% cs_map$biol_unit_nb) |> 
  # add info on crop stand and soil
  left_join(cs_map) #|> 
  # remove 
```

## 3.2 - Per-sample outlier removal

So we will compute the per-sample average, for the greenhouse dataset. But first, let's have a look at the distribution of concentrations before taking the average: are there clear outliers?

### 3.2.1 - NO3

```{r}
boxplot_no3 <- raw_greenhouse_t2_Nmin |> 
  filter(biol_unit_nb < 100) |> # exclude sand and conv soil std
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NO3") |> 
  ggplot(aes(x = as.factor(biol_unit_nb), y = conc_mgN_L)) + 
  theme_minimal() +
  #geom_violin() +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.4, colour = "purple") +
  geom_text_repel(
    aes(label = well_id), 
    size = 2,
    colour = "purple", alpha = 1, 
    min.segment.length = 1)  + labs(title = "NO3")

ridges_no3 <- raw_greenhouse_t2_Nmin |> 
  filter(biol_unit_nb < 100) |> # exclude sand and conv soil std
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NO3") |> 
  ggplot(aes(x = conc_mgN_L, groups = as.factor(biol_unit_nb), color = cs, fill = cs)) + 
  theme_minimal() +
  geom_density_ridges(aes(y = biol_unit_nb), alpha = 0.3) + facet_wrap(~soil, ncol = 4) + labs(title = "NO3")
  
```

I see some points that are a little bit off compared to the other three.

- samples where the highest concentration seems to be an outlier:

  - 21(H7), 54(H3), 15(A3), 61(C4), 77(H11), 69(H3), 10(A9), 41(A9), 74(A9)

- samples where the lowest concentration seems to be an outlier:

  - 49(E10), 6(D11), 38(E10), 53(E10), 71(F9), 42(C9)

Manually removing those points, then re-running the plotting

```{r}
biol_unit <- c(21, 54, 15, 61, 77, 69, 10, 41, 74, 49, 6, 38, 53, 71, 42)
wells <- c("H7", "H3", "A3", "C4", "H11", "H3", "A9", "A9", "A9", "E10", "D11", "E10", "E10", "F9", "C9")
names(wells) <- biol_unit

to_remove <- tibble(
  biol_unit_nb = biol_unit,
  well_id = wells
) |> left_join(
  raw_greenhouse_t2_Nmin |> 
    filter(std_sp == "NO3") |> 
    select(biol_unit_nb, plate_id, dataset) |> 
    unique())


test_outlier_no3 <- remove_wells(
  table_to_clean = raw_greenhouse_t2_Nmin,
  well_table = to_remove) 

boxplot_no3_outlierfree <- test_outlier_no3 |> 
  filter(biol_unit_nb < 100) |> # exclude sand and conv soil std
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NO3") |> 
  ggplot(aes(x = as.factor(biol_unit_nb), y = conc_mgN_L)) + 
  theme_minimal() +
  #geom_violin() +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.4, colour = "purple") +
  geom_text_repel(
    aes(label = well_id), 
    size = 2,
    colour = "purple", alpha = 1, 
    min.segment.length = 1) + labs(title = "NO3, outliers removed")

boxplot_no3 + boxplot_no3_outlierfree

ridges_no3_outlierfree <- test_outlier_no3 |> 
  filter(biol_unit_nb < 100) |> # exclude sand and conv soil std
  #filter(biol_unit_nb < 5) |> 
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NO3") |> 
  ggplot(aes(x = conc_mgN_L, groups = as.factor(biol_unit_nb), colour = cs, fill = cs )) + 
  theme_minimal() +
  geom_density_ridges(aes(y = biol_unit_nb), alpha = 0.3) +
  #geom_density(position = "stack")
  #geom_boxplot(outliers = FALSE) +
  #geom_point(alpha = 0.4, colour = "purple")# +
  facet_wrap(~soil, ncol = 4) + labs(title = "NO3, outliers removed")


ridges_no3 + ridges_no3_outlierfree + plot_layout(guides = "collect")
```

### 3.2.2 - NH4

Now same, for NH4.

```{r}
boxplot_nh4 <- test_outlier_no3 |> 
  filter(biol_unit_nb < 100) |> # exclude sand and conv soil std
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NH4") |> 
  ggplot(aes(x = as.factor(biol_unit_nb), y = conc_mgN_L)) +
  theme_minimal() +
  #geom_violin() +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.4, colour = "purple") +
  geom_text_repel(
    aes(label = well_id), 
    size = 2,
    colour = "purple", alpha = 1, 
    min.segment.length = 1)  + labs(title = "NH4")

ridges_nh4 <- test_outlier_no3 |> 
  filter(biol_unit_nb < 100) |> # exclude sand and conv soil std
  #filter(biol_unit_nb < 5) |> 
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NH4") |> 
  ggplot(aes(x = conc_mgN_L, groups = as.factor(biol_unit_nb), colour = cs, fill = cs )) + 
  theme_minimal() +
  geom_density_ridges(aes(y = biol_unit_nb), alpha = 0.3) +
  #geom_density(position = "stack")
  #geom_boxplot(outliers = FALSE) +
  #geom_point(alpha = 0.4, colour = "purple")# +
  facet_wrap(~soil, ncol = 4) + labs(title = "NH4")
```

Here are those where I would remove

- the highest concentration:

  - 6(C11), 15(C3), 23(E2), 37(A7), 45(D10), 53(F10), 66(F4), 70(G10)

- the lowest concentration:

  - 18(G9)

- those where I am really unsure:

  - 33, 75 –\> should I remove those samples all together?

```{r}
biol_unit <- c(6, 15, 23, 37, 45, 53, 66, 70, 18)
wells <- c("C11", "C3", "E2", "A7", "D10", "F10", "F4", "G10", "G9")
names(wells) <- biol_unit

to_remove <- tibble(
  biol_unit_nb = biol_unit,
  well_id = wells
) |> left_join(
  test_outlier_no3 |> 
    filter(std_sp == "NH4") |> 
    select(biol_unit_nb, plate_id, dataset) |> 
    unique())

test_outlier_nh4 <- remove_wells(
  table_to_clean = test_outlier_no3,
  well_table = to_remove) 

boxplot_nh4_outlierfree <- test_outlier_nh4 |> 
  filter(biol_unit_nb < 100) |> # exclude sand and conv soil std
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NH4") |> 
  ggplot(aes(x = as.factor(biol_unit_nb), y = conc_mgN_L)) +
  theme_minimal() +
  #geom_violin() +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.4, colour = "purple") +
  geom_text_repel(
    aes(label = well_id), 
    size = 2,
    colour = "purple", alpha = 1, 
    min.segment.length = 1)  + labs(title = "NH4, outliers removed")

ridges_nh4_outlierfree <- test_outlier_nh4 |> 
  filter(biol_unit_nb < 100) |> # exclude sand and conv soil std
  #filter(biol_unit_nb < 5) |> 
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NH4") |> 
  ggplot(aes(x = conc_mgN_L, groups = as.factor(biol_unit_nb), colour = cs, fill = cs )) + 
  theme_minimal() +
  geom_density_ridges(aes(y = biol_unit_nb), alpha = 0.3) +
  #geom_density(position = "stack")
  #geom_boxplot(outliers = FALSE) +
  #geom_point(alpha = 0.4, colour = "purple")# +
  facet_wrap(~soil, ncol = 4) + labs(title = "NH4, outliers removed")

boxplot_nh4 + boxplot_nh4_outlierfree

ridges_nh4 + ridges_nh4_outlierfree + plot_layout(guides = "collect")
```

### 3.2.4 - NO2

Because NO2 readings are virtually zero, there is only little point in removing outliers. Nevertheless, we can have a look at the data

Per sample

```{r}
boxplot_no2 <- test_outlier_nh4 |> 
  filter(biol_unit_nb < 100) |> # exclude sand and conv soil std
  filter(std_sp == "NO2") |> 
  ggplot(aes(x = as.factor(biol_unit_nb), y = conc_mgN_L)) + 
  theme_minimal() +
  #geom_violin() +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.4, colour = "purple") +
  geom_text_repel(
    aes(label = well_id), 
    size = 2,
    colour = "purple", alpha = 1, 
    min.segment.length = 1)  + labs(title = "NO2")

ridges_no2 <- test_outlier_nh4 |> 
  filter(biol_unit_nb < 100) |> # exclude sand and conv soil std
  filter(std_sp == "NO2") |> 
  ggplot(aes(x = conc_mgN_L, groups = as.factor(biol_unit_nb), color = cs, fill = cs)) + 
  theme_minimal() +
  geom_density_ridges(aes(y = biol_unit_nb), alpha = 0.3) + facet_wrap(~soil, ncol = 4) + labs(title = "NO2")

boxplot_no2 + ridges_no2 + plot_layout(guides = "collect")
```

Few to remove

```{r}
biol_unit <- c(37, 42, 49, 54)
wells <- c("D7", "C9", "E10", "H3")

to_remove <- tibble(
  biol_unit_nb = biol_unit,
  well_id = wells, 
  std_sp = rep("NO2")
) |> left_join(test_outlier_nh4)

test_outlier_no2 <- test_outlier_nh4 |> remove_wells(to_remove)

boxplot_no2_2 <- test_outlier_no2 |> 
  filter(biol_unit_nb < 100) |> # exclude sand and conv soil std
  filter(std_sp == "NO2") |> 
  ggplot(aes(x = as.factor(biol_unit_nb), y = conc_mgN_L)) + 
  theme_minimal() +
  #geom_violin() +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.4, colour = "purple") +
  geom_text_repel(
    aes(label = well_id), 
    size = 2,
    colour = "purple", alpha = 1, 
    min.segment.length = 1)  + labs(title = "NO2, outliers removed")

ridges_no2_2 <- test_outlier_no2 |> 
  filter(biol_unit_nb < 100) |> # exclude sand and conv soil std
  filter(std_sp == "NO2") |> 
  ggplot(aes(x = conc_mgN_L, groups = as.factor(biol_unit_nb), color = cs, fill = cs)) + 
  theme_minimal() +
  geom_density_ridges(aes(y = biol_unit_nb), alpha = 0.3) + facet_wrap(~soil, ncol = 4) + labs(title = "NO2, outliers removed")

boxplot_no2_2 + ridges_no2_2 + plot_layout(guides = "collect")

boxplot_no2 + boxplot_no2_2 + plot_layout(guides = "collect")
ridges_no2 + ridges_no2_2 + plot_layout(guides = "collect")
```

### 3.2.3 - Standard soils

For NO3,

```{r}
boxplot_std_no3 <- test_outlier_no2 |> 
  filter(biol_unit_nb > 100) |> # 
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NO3") |> 
  ggplot(aes(x = soil, y = conc_mgN_L, colour = soil)) +
  theme_minimal() +
  #geom_violin() +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.4, aes(colour = soil)) +
  geom_text_repel(
    aes(label = well_id, colour = soil), 
    size = 2, alpha = 1, 
    min.segment.length = 1)  + labs(title = "Standard soil NO3")

ridges_std_no3 <- test_outlier_no2 |> 
  filter(biol_unit_nb > 100) |> # exclude sand and conv soil std
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NO3") |> 
  ggplot(aes(x = conc_mgN_L, groups = as.factor(biol_unit_nb), colour = soil, fill = soil)) + 
  theme_minimal() +
  geom_density_ridges(aes(y = biol_unit_nb), alpha = 0.3) + labs(title = "Standard soil NO3")

boxplot_std_no3 + ridges_std_no3
```

I find that it looks good on a per-standard basis. But each had many samples, let's look at a per-sample basis

```{r}
boxplot_std_no3_2 <- test_outlier_no2 |> 
  filter(biol_unit_nb > 100) |> # 
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NO3") |> 
  ggplot(aes(x = map, y = conc_mgN_L, colour = soil)) +
  theme_minimal() +
  #geom_violin() +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.4, aes(colour = soil)) +
  geom_text_repel(
    aes(label = well_id, colour = soil), 
    size = 2, alpha = 1, 
    min.segment.length = 1,)  + labs(title = "Standard soil NO3 per sample") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

ridges_std_no3_2 <- test_outlier_no2 |> 
  filter(biol_unit_nb > 100) |> # exclude sand and conv soil std
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NO3") |> 
  ggplot(aes(x = conc_mgN_L, groups = as.factor(map), colour = soil, fill = soil)) + 
  theme_minimal() +
  geom_density_ridges(aes(y = biol_unit_nb), alpha = 0.3) + facet_wrap(~map) + labs(title = "Standard soil NO3 per sample")

boxplot_std_no3_2 + ridges_std_no3_2 + plot_layout(guides = "collect")
```

Still looks good, we keep it.

For NH4, same, first per standard

```{r}
boxplot_std_nh4 <- test_outlier_no2 |> 
  filter(biol_unit_nb > 100) |> # 
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NH4") |> 
  ggplot(aes(x = soil, y = conc_mgN_L, colour = soil)) +
  theme_minimal() +
  #geom_violin() +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.4, aes(colour = soil)) +
  geom_text_repel(
    aes(label = well_id, colour = soil), 
    size = 2, alpha = 1, 
    min.segment.length = 1)  + labs(title = "Standard soil NH4")

ridges_std_nh4 <- test_outlier_no2 |> 
  filter(biol_unit_nb > 100) |> # exclude sand and conv soil std
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NH4") |> 
  ggplot(aes(x = conc_mgN_L, groups = as.factor(biol_unit_nb), colour = soil, fill = soil)) + 
  theme_minimal() +
  geom_density_ridges(aes(y = biol_unit_nb), alpha = 0.3) + labs(title = "Standard soil NH4")

boxplot_std_nh4 + ridges_std_nh4 + plot_layout(guides = "collect")
```

Multimodal curve –\> dig into per-sample view

```{r}
boxplot_std_nh4_2 <- test_outlier_no2 |> 
  filter(biol_unit_nb > 100) |> # 
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NH4") |> 
  ggplot(aes(x = map, y = conc_mgN_L, colour = soil)) +
  theme_minimal() +
  #geom_violin() +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.4, aes(colour = soil)) +
  geom_text_repel(
    aes(label = well_id, colour = soil), 
    size = 2, alpha = 1, 
    min.segment.length = 1,)  + labs(title = "Standard soil NH4, per sample") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

ridges_std_nh4_2 <- test_outlier_no2 |> 
  filter(biol_unit_nb > 100) |> # exclude sand and conv soil std
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NH4") |> 
  ggplot(aes(x = conc_mgN_L, groups = as.factor(map), colour = soil, fill = soil)) + 
  theme_minimal() +
  geom_density_ridges(aes(y = biol_unit_nb), alpha = 0.3) + facet_wrap(~map) + labs(title = "Standard soil NH4, per sample")

boxplot_std_nh4_2 + ridges_std_nh4_2 + plot_layout(guides = "collect")
```

I find 2 wells to remove:

- plate 109_t2_MR_R8, well A5

- plate 110_t2_MR_R7, well G9

Let's remove them and re-run the analysis

```{r}
samples <- c("109_t2_MR_R8", "110_t2_MR_R7")
wells <- c("A5", "G9")

to_remove <- test_outlier_no2 |> filter(
  ((map == samples[1] & well_id == wells[1]) |
    (map == samples[2] & well_id == wells[2])) &
    std_sp == "NH4"
)

test_outlier_nh4_std <- test_outlier_no2 |> remove_wells(to_remove)

boxplot_std_nh4_3 <- test_outlier_nh4_std |> 
  filter(biol_unit_nb > 100) |> # 
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NH4") |> 
  ggplot(aes(x = map, y = conc_mgN_L, colour = soil)) +
  theme_minimal() +
  #geom_violin() +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.4, aes(colour = soil)) +
  geom_text_repel(
    aes(label = well_id, colour = soil), 
    size = 2, alpha = 1, 
    min.segment.length = 1,)  + labs(title = "Standard soil NH4\nPer sample, outliers removed") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

ridges_std_nh4_3 <- test_outlier_nh4_std |> 
  filter(biol_unit_nb > 100) |> # exclude sand and conv soil std
  filter_out(std_sp == "NO2") |> 
  filter(std_sp == "NH4") |> 
  ggplot(aes(x = conc_mgN_L, groups = as.factor(map), colour = soil, fill = soil)) + 
  theme_minimal() +
  geom_density_ridges(aes(y = biol_unit_nb), alpha = 0.3) + facet_wrap(~map) + labs(title = "Standard soil NH4\nPer sample, outliers removed")

boxplot_std_nh4_3 + ridges_std_nh4_3 + plot_layout(guides = "collect")
```

Compare before/after outlier removal

```{r}
boxplot_std_nh4_2 + boxplot_std_nh4_3 + plot_layout(guides = "collect")

ridges_std_nh4_2 + ridges_std_nh4_3 + plot_layout(guides = "collect")
```

For NO2, per standard

```{r}
boxplot_std_no2 <- test_outlier_nh4_std |> 
  filter(biol_unit_nb > 100) |> # 
  filter(std_sp == "NO2") |> 
  ggplot(aes(x = soil, y = conc_mgN_L, colour = soil)) +
  theme_minimal() +
  #geom_violin() +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.4, aes(colour = soil)) +
  geom_text_repel(
    aes(label = well_id, colour = soil), 
    size = 2, alpha = 1, 
    min.segment.length = 1)  + labs(title = "Standard soil NO2")

ridges_std_no2 <- test_outlier_nh4_std |> 
  filter(biol_unit_nb > 100) |> # exclude sand and conv soil std
  filter(std_sp == "NO2") |> 
  ggplot(aes(x = conc_mgN_L, groups = as.factor(biol_unit_nb), colour = soil, fill = soil)) + 
  theme_minimal() +
  geom_density_ridges(aes(y = biol_unit_nb), alpha = 0.3) + labs(title = "Standard soil NO2")

boxplot_std_no2 + ridges_std_no2 + plot_layout(guides = "collect")
```

Then per sample

```{r}
boxplot_std_no2_2 <- test_outlier_nh4_std |> 
  filter(biol_unit_nb > 100) |> # 
  filter(std_sp == "NO2") |> 
  ggplot(aes(x = map, y = conc_mgN_L, colour = soil)) +
  theme_minimal() +
  #geom_violin() +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.4, aes(colour = soil)) +
  geom_text_repel(
    aes(label = well_id, colour = soil), 
    size = 2, alpha = 1, 
    min.segment.length = 1,)  + labs(title = "Standard soil NO2 per sample") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

ridges_std_no2_2 <- test_outlier_nh4_std |> 
  filter(biol_unit_nb > 100) |> # exclude sand and conv soil std
  filter(std_sp == "NO2") |> 
  ggplot(aes(x = conc_mgN_L, groups = as.factor(map), colour = soil, fill = soil)) + 
  theme_minimal() +
  geom_density_ridges(aes(y = biol_unit_nb), alpha = 0.3) + facet_wrap(~map) + labs(title = "Standard soil NO2 per sample")

boxplot_std_no2_2 + ridges_std_no2_2 + plot_layout(guides = "collect")
```

to remove: plate 110_t2_MR_R8, well H6

```{r}
to_remove <- test_outlier_nh4_std |> 
  filter(
    map == "110_t2_MR_R8",
    well_id == "H6",
    std_sp == "NO2")

test_outlier_no2_std <- test_outlier_nh4_std |> remove_wells(to_remove)

boxplot_std_no2_3 <- test_outlier_no2_std |> 
  filter(biol_unit_nb > 100) |> # 
  filter(std_sp == "NO2") |> 
  ggplot(aes(x = map, y = conc_mgN_L, colour = soil)) +
  theme_minimal() +
  #geom_violin() +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.4, aes(colour = soil)) +
  geom_text_repel(
    aes(label = well_id, colour = soil), 
    size = 2, alpha = 1, 
    min.segment.length = 1,)  + labs(title = "Standard soil NO2\nPer sample, outliers removed") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

ridges_std_no2_3 <- test_outlier_no2_std |> 
  filter(biol_unit_nb > 100) |> # exclude sand and conv soil std
  filter(std_sp == "NO2") |> 
  ggplot(aes(x = conc_mgN_L, groups = as.factor(map), colour = soil, fill = soil)) + 
  theme_minimal() +
  geom_density_ridges(aes(y = biol_unit_nb), alpha = 0.3) + facet_wrap(~map) + labs(title = "Standard soil NO2\nPer sample, outliers removed")

boxplot_std_no2_2 + boxplot_std_no2_3 + plot_layout(guides = "collect")
ridges_std_no2_2 + ridges_std_no2_3 + plot_layout(guides = "collect")
```

Visually, I am satisfied with this outlier removal, so I save this cleaned table

```{r}
greenhouse_t2_Nmin_clean <- test_outlier_no2_std
```

## 3.3 - Per-sample mean

```{r}
conc_mean <- greenhouse_t2_Nmin_clean |> 
  select(biol_unit_nb, std_sp, conc_mgN_L) |> 
  group_by(biol_unit_nb, std_sp) |> 
  summarise(
    mean = mean(conc_mgN_L),
    st_dev = sd(conc_mgN_L)) |> 
  mutate(coef_var = 100*st_dev / mean) |> 
  rename(conc_mgN_L = mean)

conc_mean |> arrange(desc(std_sp), desc(coef_var))

conc_mean |> filter(coef_var > 10) |> 
  ggplot(aes(x = coef_var)) + geom_histogram()
```

We still have quite a few samples with a high between-wells coefficient of variation. But with only 3 to 4 wells and sometimes very low values, this is unavoidable, so we move on

# 3 - Export

```{r}
raw_greenhouse_t2 |> write_rds("output/data/3_greenhouse_t2_raw_lab.rds")
conc_mean |> write_rds("output/data/3_greenhouse_t2_conc_clean.rds")

```
