---
title: "IV. Data Transformation"
format: 
  gfm:
    toc: true
    code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# Set up

```{r}
rm(list = ls())

library(tidyverse)
```

# 1 - Nmin

## 1.1 - Import tidy data set

```{r}
Nmin <- read_rds("output/data/Nmin_tidy.rds") |> 
  select(
    !starts_with(c("yd", "whc", "nod", "pmn", "dna", "zone", "incub")))
#Nmin |> str()
```

## 1.2 - Compute new variables

### 1.2.1 - dry matter content

In this data set, dry matter and water content have not yet been computed. The raw data with 3 technical replicates is still encoded, in separate columns `sse_dm...`\`. In this section, we compute average dry matter and water content for the sample (across technical replicates). We add additional columns with intermediary results:

-   FW~neat~ = FW~gross~ - tare; DW~neat~ = DW~gross~ - tare

-   average (FW, then DW) = sum(3 technical replicates) / 3

-   DM = avg~DW~ / avg~FW~

-   WC = 1 - DM

-   With

    -   FW = fresh weight \[g\]

    -   DW = dry weight \[g\],

    -   DM = dry matter content \[g dry soil / g fresh soil\],

    -   WC = water content \[g water / g [**fresh**]{.underline} soil\]

-   ratio = fresh wait of soil added into tubes for extraction (\~10g) / volume of extractant added (=20ml)

Then, we can finally convert concentrations in mg/L into concentrations in ppm (mg/kg dry soil)

-   see <https://mycloud.ulb.be/index.php/f/30359374> for equation

```{r}
# Volume of extractant used in the K2SO4 extraction
vol_extr <- 20

# 
Nmin_tfo <- Nmin |> 
  # Compute dm
  mutate(
    # subtract tare from gross fresh weight
    flush_dm_neat_fw_tr1 = flush_dm_g_fw_tr1 - flush_dm_tare_tr1,
    flush_dm_neat_fw_tr2 = flush_dm_g_fw_tr2 - flush_dm_tare_tr2,
    flush_dm_neat_fw_tr3 = flush_dm_g_fw_tr3 - flush_dm_tare_tr3,
    # subtract tare from gross dry weight
    flush_dm_neat_dw_tr1 = flush_dm_g_dw_tr1 - flush_dm_tare_tr1,
    flush_dm_neat_dw_tr2 = flush_dm_g_dw_tr2 - flush_dm_tare_tr2,
    flush_dm_neat_dw_tr3 = flush_dm_g_dw_tr3 - flush_dm_tare_tr3,
    # compute average fresh and dry weight
    flush_dm_avg_fw = (
      flush_dm_neat_fw_tr1 + flush_dm_neat_fw_tr2 + flush_dm_neat_fw_tr3
      ) / 3,
    flush_dm_avg_dw = (
      flush_dm_neat_dw_tr1 + flush_dm_neat_dw_tr2 + flush_dm_neat_dw_tr3) / 3,
    # compute dry matter content
    dm = flush_dm_avg_dw / flush_dm_avg_fw,
    # compute water content
    wc = 1- dm,
    # compute each sample's individual ratio (~10g moist per 20 ml)
    ratio = flush_fw_nf_g / vol_extr,
    .keep = "unused"
  ) |> 
  # Convert concentrations in mg/L to mg/kg
  mutate(
    conc_N_ppm_rt1 = unlist(conc_N_Lrt1) / (ratio*dm),
    conc_N_ppm_rt2 = unlist(conc_N_Lrt2) / (ratio*dm),
    conc_N_ppm_rt3 = unlist(conc_N_Lrt3) / (ratio*dm),
    conc_N_ppm_rt4 = unlist(conc_N_Lrt4) / (ratio*dm)
    ) |> 
  rowwise() |> 
  mutate(
    conc_N_ppm_avg = mean(c_across(c("conc_N_ppm_rt1", "conc_N_ppm_rt2", "conc_N_ppm_rt3", "conc_N_ppm_rt4")))
  ) |> 
 # select(!starts_with("flush")) |> 
  select(
    expe, sample_short, sd_c, soil, crop_diversity, cs, bloc, sampling_time, 
    plate_id, N_sp, dataset, std_sp, 
    wc, dm, starts_with("conc")) |> 
  select(!contains("N_L"))

```

## 1.3 - Export

```{r}
Nmin_tfo |> write_rds("output/data/4_Nmin_tfo.rds")
```

# 2 - TDN

## 2.1 - Import tidy data set

```{r}
TDN <- read_rds("output/data/TDN_tidy.rds")
```
