---
title: "IV. Data Transformation"
format: 
  gfm:
    toc: true
    code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# Set up

```{r}
rm(list = ls())

library(tidyverse)
```

# 1 - Nmin

## 1.1 - Import tidy data set

```{r}
Nmin <- read_rds("output/data/Nmin_tidy.rds")
```

## 1.2 - Compute new variables

### 1.2.1 - dry matter content 

In this data set, dry matter and water content have not yet been computed. The raw data with 3 technical replicates is still encoded, in separate columns `sse_dm...`\`. In this section, we compute average dry matter and water content for the sample (across technical replicates). We add additional columns with intermediary results:

-   FW~neat~ = FW~gross~ - tare; DW~neat~ = DW~gross~ - tare

-   average (FW, then DW) = sum(3 technical replicates) / 3

-   DM = avg~DW~ / avg~FW~

-   WC = 1 - DM

-   With

    -   FW = fresh weight \[g\]

    -   DW = dry weight \[g\],

    -   DM = dry matter content \[g dry soil / g fresh soil\],

    -   WC = water content \[g water / g [**fresh**]{.underline} soil\]

```{r}
# 
Nmin_tfo <- Nmin |> 
  # Compute dm
  mutate(
    # subtract tare from gross fresh weight
    flush_dm_neat_fw_tr1 = flush_dm_g_fw_tr1 - flush_dm_tare_tr1,
    flush_dm_neat_fw_tr2 = flush_dm_g_fw_tr2 - flush_dm_tare_tr2,
    flush_dm_neat_fw_tr3 = flush_dm_g_fw_tr3 - flush_dm_tare_tr3,
    # subtract tare from gross dry weight
    flush_dm_neat_dw_tr1 = flush_dm_g_dw_tr1 - flush_dm_tare_tr1,
    flush_dm_neat_dw_tr2 = flush_dm_g_dw_tr2 - flush_dm_tare_tr2,
    flush_dm_neat_dw_tr3 = flush_dm_g_dw_tr3 - flush_dm_tare_tr3,
    # compute average fresh and dry weight
    flush_dm_avg_fw = (
      flush_dm_neat_fw_tr1 + flush_dm_neat_fw_tr2 + flush_dm_neat_fw_tr3
      ) / 3,
    flush_dm_avg_dw = (
      flush_dm_neat_dw_tr1 + flush_dm_neat_dw_tr2 + flush_dm_neat_dw_tr3) / 3,
    # compute dry matter content
    dm = flush_dm_avg_dw / flush_dm_avg_fw,
    # compute water content
    wc = 1- dm,
    .keep = "unused"
  ) 
```

# °°° !!! START HERE !!! °°°

To DO: Add further computations!

# 2 - TDN

## 2.1 - Import tidy data set

```{r}
TDN <- read_rds("output/data/TDN_tidy.rds")
```
