---
title: "IV. Data Transformation"
format: 
  gfm:
    toc: true
    code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# Set up

```{r}
rm(list = ls())

library(tidyverse)
```

# 1 - Nmin

## 1.1 - Import tidy data set

```{r}
Nmin <- read_rds("output/data/3_Nmin_tidy.rds") |> 
  select(
    !starts_with(c("yd", "whc", "nod", "pmn", "dna", "zone", "incub")))
#Nmin |> str()
```

## 1.2 - Compute new variables

### 1.2.1 - dry matter content

```{r}

# 
Nmin_tfo1 <- Nmin |> 
  
  #** °°°°° first, update format of variables °°°°° *
  mutate(
    soil = factor(soil, levels = c("Conv", "Ref", "Auto", "ABC")),
    crop_diversity = factor(crop_diversity, levels = c("SC", "IC")),
    cs = factor(cs, levels = c("F", "IC", "W")),
    bloc = as.factor(bloc)
  ) |> 
  #** °°°°° Convert concentrations in mg/L to mg/kg °°°°° *
  mutate(
    conc_N_ppm_rt1 = unlist(conc_N_Lrt1) / (ratio_nf*dm),
    conc_N_ppm_rt2 = unlist(conc_N_Lrt2) / (ratio_nf*dm),
    conc_N_ppm_rt3 = unlist(conc_N_Lrt3) / (ratio_nf*dm),
    conc_N_ppm_rt4 = unlist(conc_N_Lrt4) / (ratio_nf*dm)
    ) |> 
  rowwise() |> 
  # compute mean between technical replicates 
  mutate(
    conc_N_ppm_avg = mean(c_across(c("conc_N_ppm_rt1", "conc_N_ppm_rt2", "conc_N_ppm_rt3", "conc_N_ppm_rt4")))
  ) |> 
 # select(!starts_with("flush")) |> 
  select(
    expe, sample_short, sd_c, soil, crop_diversity, cs, bloc, sampling_time, 
    plate_id, N_sp, dataset, std_sp, 
    wc, dm, starts_with("conc")) |> 
  select(!contains("N_L"))



```

Then, we correct NO3 (subtract NO2) and compute new variables: Nmin (ppm) NO3/Nmin ratio and NH4/NO3 ratio. For that, we need to pivot wider so that different N_sp are in different columns.

```{r}
Nmin_tfo2 <- Nmin_tfo1 |> 
  #** °°°°° select relevant columns and pivot wider °°°°° *
  select(expe:sampling_time, N_sp, conc_N_ppm_avg) |> 
  pivot_wider(
    names_from = N_sp,
    names_prefix = "conc_",
    values_from = conc_N_ppm_avg
  ) |> 
  #** °°°°° Compute new variables °°°°° *
  mutate(
    conc_NO3_neat = conc_NO3-conc_NO2,
    conc_Nmin = conc_NO3 + conc_NH4, # NO2 not included here bc it is already included in NO3
    conc_NO3_Nmin = conc_NO3_neat / conc_Nmin,
    conc_NH4_NO3 = conc_NH4 / conc_NO3_neat) |> 
  #** °°°°° remove NO3 so no mistake is made, and rename NO3_neat °°°°° *
  select(!conc_NO3) |> 
  rename(conc_NO3 = conc_NO3_neat) 
```

## 1.3 - Export

It would have made no biological sense to compute these new variables on the technical replicates which represent the noise from 4 wells of the 96-well plate, but the first well of NO3 has no direct relation to the first well of NO2, etc. Thus we pivot from the average value. This means that in the pivot table, single values of individual technical replicates get lost. In order to not lose that information (just in case), we will export both tables (pivoted and unpivoted)

```{r}
Nmin_tfo1 |> write_rds("output/data/4_Nmin_tfo1.rds")
Nmin_tfo2 |> write_rds("output/data/4_Nmin_tfo2.rds")
```

# 2 - TDN

## 2.1 - Import tidy data set

```{r}
TDN <- read_rds("output/data/TDN_tidy.rds") |> 
  select(
    !starts_with(c("yd", "whc", "nod", "pmn", "dna", "zone", "incub")))

```

## 2.2 - Compute new variables

### 2.2.1 - dry matter content
