---
title: "IV. Data Transformation"
format: 
  gfm:
    toc: true
    code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# Set up

```{r}
rm(list = ls())

library(tidyverse)
```

# 1 - Nmin

## 1.1 - Import tidy data set

```{r}
Nmin <- read_rds("output/data/Nmin_tidy.rds") |> 
  select(
    !starts_with(c("yd", "whc", "nod", "pmn", "dna", "zone", "incub")))
#Nmin |> str()
```

## 1.2 - Compute new variables

### 1.2.1 - dry matter content

In this data set, dry matter and water content have not yet been computed. The raw data with 3 technical replicates is still encoded, in separate columns `sse_dm...`\`. In this section, we compute average dry matter and water content for the sample (across technical replicates). We add additional columns with intermediary results:

-   FW~neat~ = FW~gross~ - tare; DW~neat~ = DW~gross~ - tare

-   average (FW, then DW) = sum(3 technical replicates) / 3

-   DM = avg~DW~ / avg~FW~

-   WC = 1 - DM

-   With

    -   FW = fresh weight \[g\]

    -   DW = dry weight \[g\],

    -   DM = dry matter content \[g dry soil / g fresh soil\],

    -   WC = water content \[g water / g [**fresh**]{.underline} soil\]

-   ratio = fresh wait of soil added into tubes for extraction (\~10g) / volume of extractant added (=20ml)

Then, we can finally convert concentrations in mg/L into concentrations in ppm (mg/kg dry soil)

-   see <https://mycloud.ulb.be/index.php/f/30359374> for equation

```{r}
# Volume of extractant used in the K2SO4 extraction
vol_extr <- 20

# 
Nmin_tfo1 <- Nmin |> 
  
  #** °°°°° first, update format of variables °°°°° *
  mutate(
    soil = factor(soil, levels = c("Conv", "Ref", "Auto", "ABC")),
    crop_diversity = factor(crop_diversity, levels = c("SC", "IC")),
    cs = factor(cs, levels = c("F", "IC", "W")),
    bloc = as.factor(bloc)
  ) |> 
  #** °°°°° Compute dm °°°°° *
  mutate(
    # subtract tare from gross fresh weight
    flush_dm_neat_fw_tr1 = flush_dm_g_fw_tr1 - flush_dm_tare_tr1,
    flush_dm_neat_fw_tr2 = flush_dm_g_fw_tr2 - flush_dm_tare_tr2,
    flush_dm_neat_fw_tr3 = flush_dm_g_fw_tr3 - flush_dm_tare_tr3,
    # subtract tare from gross dry weight
    flush_dm_neat_dw_tr1 = flush_dm_g_dw_tr1 - flush_dm_tare_tr1,
    flush_dm_neat_dw_tr2 = flush_dm_g_dw_tr2 - flush_dm_tare_tr2,
    flush_dm_neat_dw_tr3 = flush_dm_g_dw_tr3 - flush_dm_tare_tr3,
    # compute average fresh and dry weight
    flush_dm_avg_fw = (
      flush_dm_neat_fw_tr1 + flush_dm_neat_fw_tr2 + flush_dm_neat_fw_tr3
      ) / 3,
    flush_dm_avg_dw = (
      flush_dm_neat_dw_tr1 + flush_dm_neat_dw_tr2 + flush_dm_neat_dw_tr3) / 3,
    # compute dry matter content
    dm = flush_dm_avg_dw / flush_dm_avg_fw,
    # compute water content
    wc = 1- dm,
    
    #** °°°°° compute each sample's individual ratio (~10g moist per 20 ml) °°°°° *
    ratio = flush_fw_nf_g / vol_extr,
    .keep = "unused"
  ) |> 
  #** °°°°° Convert concentrations in mg/L to mg/kg °°°°° *
  mutate(
    conc_N_ppm_rt1 = unlist(conc_N_Lrt1) / (ratio*dm),
    conc_N_ppm_rt2 = unlist(conc_N_Lrt2) / (ratio*dm),
    conc_N_ppm_rt3 = unlist(conc_N_Lrt3) / (ratio*dm),
    conc_N_ppm_rt4 = unlist(conc_N_Lrt4) / (ratio*dm)
    ) |> 
  rowwise() |> 
  # compute mean between technical replicates 
  mutate(
    conc_N_ppm_avg = mean(c_across(c("conc_N_ppm_rt1", "conc_N_ppm_rt2", "conc_N_ppm_rt3", "conc_N_ppm_rt4")))
  ) |> 
 # select(!starts_with("flush")) |> 
  select(
    expe, sample_short, sd_c, soil, crop_diversity, cs, bloc, sampling_time, 
    plate_id, N_sp, dataset, std_sp, 
    wc, dm, starts_with("conc")) |> 
  select(!contains("N_L"))



```

Then, we correct NO3 (subtract NO2) and compute new variables: Nmin (ppm) NO3/Nmin ratio and NH4/NO3 ratio. For that, we need to pivot wider so that different N_sp are in different columns.

```{r}
Nmin_tfo2 <- Nmin_tfo1 |> 
  #** °°°°° select relevant columns and pivot wider °°°°° *
  select(expe:sampling_time, N_sp, conc_N_ppm_avg) |> 
  pivot_wider(
    names_from = N_sp,
    names_prefix = "conc_",
    values_from = conc_N_ppm_avg
  ) |> 
  #** °°°°° Compute new variables °°°°° *
  mutate(
    conc_NO3_neat = conc_NO3-conc_NO2,
    conc_Nmin = conc_NO3 + conc_NH4, # NO2 not included here bc it is already included in NO3
    conc_NO3_Nmin = conc_NO3_neat / conc_Nmin,
    conc_NH4_NO3 = conc_NH4 / conc_NO3_neat) |> 
  #** °°°°° remove NO3 so no mistake is made, and rename NO3_neat °°°°° *
  select(!conc_NO3) |> 
  rename(conc_NO3 = conc_NO3_neat) 
```

## 1.3 - Export

It would have made no biological sense to compute these new variables on the technical replicates which represent the noise from 4 wells of the 96-well plate, but the first well of NO3 has no direct relation to the first well of NO2, etc. Thus we pivot from the average value. This means that in the pivot table, single values of individual technical replicates get lost. In order to not lose that information (just in case), we will export both tables (pivoted and unpivoted)

```{r}
Nmin_tfo1 |> write_rds("output/data/4_Nmin_tfo1.rds")
Nmin_tfo2 |> write_rds("output/data/4_Nmin_tfo2.rds")
```

# 2 - TDN

## 2.1 - Import tidy data set

```{r}
TDN <- read_rds("output/data/TDN_tidy.rds") |> 
  select(
    !starts_with(c("yd", "whc", "nod", "pmn", "dna", "zone", "incub")))

```

## 2.2 - Compute new variables

### 2.2.1 - dry matter content
