---
title: "4 Greenhouse t2 - Data wrangling"
format: 
  gfm:
    toc: true
    code-fold: true
# format: 
#   html:
#     embed-resources: true
#     toc: true
#     toc_float: true
#     code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# TO DO

Move the PMN curve to script nb 5 !

Deal with standard soils.

- Problem: they have same biological unit nb. –\> a preliminary outlier removal needs to be ran separately, then average per biol_unit_nb, then it can be rejoined...

# Set up

```{r}
#| message: false
#| warning: false

# clean environment
rm(list = ls())

# packages
library(tidyverse)
library(ggridges) # for geom_density_ridges()
library(ggrepel) # for geom_text_repel()
library(patchwork)
library(janitor)

# data
#lab <- read_rds("output/data/3_greenhouse_t2_raw_lab.rds")
Nmin <- read_rds("output/data/3_greenhouse_t2_Nmin_clean.rds") 
PMN <- read_rds("output/data/3_greenhouse_PMN_clean.rds")
PMN_fw <- read_csv("raw_data/PMN/PMN_fw.csv") |> clean_names()
#lm_output <- read_rds("output/data/2_lm_output_noTDN.rds")
```

# 1 - Subset data

## 1.1 - Yield data

```{r}
# yield_data <- lab |> 
#   select(biol_unit_nb:sampling_time, starts_with("yd")) 
```

## 1.2 - Flush data (from K2SO4 extraction)

```{r}
# raw_flush <- lab |> 
#   select(
#     !starts_with(c("yd", "whc")) &
#       !cra_trial:sd_c)
```

## 1.3 - Nmin data (from absorbance pipeline)

```{r}

Nmin_sample <- Nmin |> filter(biol_unit_nb < 100)
Nmin_std <- Nmin |> filter_out(biol_unit_nb < 100)
```

## 1.4 - PMN data (from absorbance pipeline)

The PMN data is missing 1 key information: weight of fresh soil in the sub-sample that undergoes the KCl extraction.

```{r}

PMN_fw_greenhouse <- PMN_fw |> filter_out(expe == "Field")
```

# 2 - Compute new variables

## 2.1 - dry matter content (t2 samples)

```{r}
 
```

## 2.2 - Nmin concentrations in ppm (t2 samples)

First, we

- join absorbance data and flush data,

- Compute the N-sp N concentration in ppm (e.g. mg NO3-N / kg dry soil)

- do some data housecleaning

```{r}
Nmin_sample |> arrange(biol_unit_nb) 

Nmin_ppm_sample <- Nmin_sample |> 
  # join both data tables
  left_join(flush_sample) |> 
  # compute concentration in ppm
  mutate(conc_ppm = conc_mgN_L / (ratio_nf * dm)) |> 
  # remove again the bare soils
  filter_out(is.na(sample_short)) |> 
  # select relevant columns
  select(!c(
    #dataset, 
    expe, 
    target_sp:lm_p, 
    starts_with("sampl"), starts_with("ratio"),
    map, plate_id, conc_mgN_L)) |> 
  arrange(biol_unit_nb)

# Check it out
Nmin_ppm_sample
```

Now, we pivot the data so that each N-sp receives its own column

```{r}
Nmin_ppm_wider <- Nmin_ppm_sample |>
  ungroup() |>  
  select(!c(plate_id, map, st_dev, coef_var)) |> 
  pivot_wider(
    names_from = std_sp,
    values_from = conc_ppm,
    names_prefix = "ppm_"
  ) |> 
  # to make obvious that NO3 is still uncorrected
  rename(ppm_NO3_uncorrected = ppm_NO3)
```

Then we can compute final Nmin variables:

- correct NO3 value (what is measured is the sum of NO3 and NO2, bc we measure NO2 after reduction of NO3 to NO2)

- Compute Nmin (= NO3 + NO2 + NH4)

- Compute ratio, e.g., NO3/Nmin, NH4/Nmin, NO3/NH4

```{r}
Nmin_all_variables <- Nmin_ppm_wider |> 
  mutate(
    ppm_NO3 = ppm_NO3_uncorrected - ppm_NO2,
    ppm_Nmin = ppm_NO3_uncorrected + ppm_NH4,
    NO3_Nmin = ppm_NO3 / ppm_Nmin,
    NH4_Nmin = ppm_NH4 / ppm_Nmin,
    NO3_NH4 = ppm_NO3 / ppm_NH4
  ) |> 
  # remove uncorrected NO3
  select(!ppm_NO3_uncorrected)
```

## 2.3 - PMN concentration in ppm

### 2.3.1 - Compute new variables

Here, we

- join PMN_fw (data set containing fresh weight of the subsample that underwent the incubation) with the absorbance derived concentration data in mg N / L
- compute the ratio between weight of fresh soil and volume of extractant (150ml) –\> \~30g/150ml
- Derive the concentration in ppm from mg N/L, ratio and dry matter

```{r}
volume_kcl <- 150
PMN_ppm <- PMN |> 
  left_join(PMN_fw_greenhouse) |> 
  mutate(
    ratio = fw / 150,
    conc_ppm = conc_mgN_L / (ratio * dm)
  )
```

Now, we want the concentration in separate columns as above for Nmin,

```{r}
PMN_wider <- PMN_ppm |> 
  pivot_wider(
  names_from = std_sp,
  names_prefix = "ppm_",
  values_from = conc_ppm,
  id_cols = c(map, biol_unit_nb, soil, dm, wc, expe, incub_time, rep_tech, fw, ratio)
) |> 
  rename(ppm_NO3_uncorrected = ppm_NO3)
```

Then, we can do the NO3 correction and other computation of variables

```{r}
dates <- c("2023-12-5", "2023-12-12", "2023-12-17", "2023-12-26", "2024-01-02") |> as.Date()
incub_days <- dates - dates[1]
names(incub_days) <- PMN_wider$incub_time |> unique() |> sort()

PMN_all_variables <- PMN_wider |> 
  mutate(
    ppm_NO3 = ppm_NO3_uncorrected - ppm_NO2,
    ppm_Nmin = ppm_NO3_uncorrected + ppm_NH4,
    NO3_Nmin = ppm_NO3 / ppm_Nmin,
    NH4_Nmin = ppm_NH4 / ppm_Nmin,
    NO3_NH4 = ppm_NO3 / ppm_NH4,
    incub_day = incub_days[incub_time],
    soil = factor(soil, levels = c("Conv", "Ref", "Auto", "ABC"))
  ) |> 
  # remove uncorrected NO3
  select(!ppm_NO3_uncorrected) 

#check correspondence incubation vs dates
PMN_all_variables |> ungroup() |> select(incub_time, incub_day) |> unique()
```

### 2.3.2 - Plot it

Finally, we can plot the curve of the potential mineralization of Nitrogen (PMN)

```{r}
PMN_plot <- PMN_all_variables |> 
  ggplot(aes(x = incub_day, y = ppm_Nmin, group = soil, colour = soil, fill = soil)) +
  theme_minimal() +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    title = "Potential Mineralization of Nitrogen (PMN)",
    subtitle = "Nmin as sum of NO2-, NO3-, NH4+") +
  xlab("Time since incubation at 28°C [days]") +
  ylab("Mineral Nitrogen in the soil [ppm]")

# for annotation: extract data from smooth curve
smooth_data <- ggplot_build(PMN_plot)$data[[2]]

# get its maximum value to anker the annotation
annotations <- smooth_data |> 
  slice_max(x, by = group) |> 
  mutate(soil = levels(PMN_all_variables$soil))
#c("#F8766D", "#7CAE00", "#00BFC4", "#C77CFF") 

# Also annotate date of humidification
day_water <- as.Date("2023-12-20")
  
PMN_plot_all <- PMN_plot + 
  geom_text(
    data = annotations,
    aes(x = x+0.5, y = y, colour = soil, label = soil), 
    hjust = 0, size = 5) +
  xlim(c(0,30)) +
  theme(legend.position = "none") 
```

Looks nice, but if we facet... The curves... are not linear at all. Probably due to re-humidification in-between:

```{r}
PMN_plot_all_annotated <- PMN_plot_all +
  annotate(
    geom = "segment", colour = "red",
    x = day_water - dates[1] - 2, xend = day_water - dates[1],
    y = 7, yend = 5,
    arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate(
    geom = "text", colour = "red",
    x = day_water - dates[1] - 2,
    y = 7, hjust = 1.05,
    label = "Re-humidification of samples"
  )

#PMN_plot_all_annotated
```

Look at all 3 versions

```{r}
PMN_plot_all

PMN_plot_all_annotated

PMN_plot_all_annotated + facet_wrap(~soil)
```

### 2.3.3 - Corrected PMN

After discussion with PI: keep only points that we trust: 1st one and post-humidification –\> i0, i3, i4

```{r}
# subset data
PMN_trusted <- PMN_all_variables |> 
  filter(incub_time %in% c("i0", "i3", "i4"))
```

## 2.4 - Yield variables

## 2.5 - Join all data (except PMN)

```{r}

data_to_export <- Nmin_all_variables |> left_join(yield_clean)
```

# 3 - Deal with standard soils

# **°°° !!! TO DO !!! °°°**

# 4 - Export

We don't export PMN data for now, as it seems like a dead-end..

```{r}
data_to_export |> write_rds("output/data/4_t2_greenhouse_transformed.rds")
PMN_trusted |> write_rds("output/data/4_greehouse_PMN.rds")
```
