---
title: "4 Greenhouse t2 - Data wrangling"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# TO DO

!!! Imported data incomplete

- Forgot mean per sample... in abs pipeline !

For PMN:

- water content can only be obtained from the WHC experiment (CHECK!)

For Nmin:

- Deal with standard soils. Problem: they have same biological unit nb. Their analysis needs to run separately

# Set up

```{r}
#| message: false
#| warning: false

# clean environment
rm(list = ls())

# packages
library(tidyverse)

# data
raw_greenhouse_t2 <- read_rds("output/data/3_raw_greenhouse_t2.rds")
raw_Nmin <- read_rds("output/data/2_mgNL_noTDN.rds") |> filter_out(dataset == "Nmint3")

```

# 1 - Subset data

Flush data (from K2SO4 extraction)

```{r}
raw_greenhouse_t2_flush <- raw_greenhouse_t2 |> 
  select(
    !starts_with(c("yd", "whc")) &
      !cra_trial:sd_c)
```

From Nmin data (from absorbance pipeline)

```{r}

raw_greenhouse_t2_Nmin <- raw_Nmin |> 
  # create sampling_time and expe variables from plate_ids (first number and first letter after N species)
  mutate(
    sampling_time = paste0(
      "t", 
      str_extract(plate_id, "\\w_(\\d)(\\w).*", group = 1)),
    expe = str_extract(plate_id, "\\w_(\\d)(\\w).*", group = 2),
    expe = case_when(expe %in% c("P", "G") ~ "Pot", .default = expe),
    .before = plate_id) |>
  # filter based on sampling_time
  filter(sampling_time == "t2", expe == "Pot") |> 
  separate_wider_delim(
    cols = map,
    names = c("biol_unit_nb"),
    delim = "_",
    too_many = "drop", 
    cols_remove = FALSE
  ) |> 
  mutate(biol_unit_nb = as.double(biol_unit_nb))
  
# Check it out
raw_greenhouse_t2_Nmin

greenhouse_t2_Nmin_sample <- raw_greenhouse_t2_Nmin |> filter(biol_unit_nb < 100)
greenhouse_t2_Nmin_std <- raw_greenhouse_t2_Nmin |> filter_out(biol_unit_nb < 100)
```

# 1 - Compute new variables

## 1.1 - dry matter content

### 1.1.1 - Principle and equations

In this data set, dry matter and water content have not yet been computed. The raw data with 3 technical replicates is encoded, in separate columns starting with `flush_dm...`\`. In this section, we compute average dry matter and water content for the sample (across technical replicates). We add additional columns with intermediary results:

- FW~neat~ = FW~gross~ - tare ; DW~neat~ = DW~gross~ - tare

- average (FW, then DW) = sum(3 technical replicates) / 3

- DM = avg~DW~ / avg~FW~

- WC = 1 - DM

- With

  - FW = fresh weight \[g\]

  - DW = dry weight \[g\],

  - DM = dry matter content \[g~dry\ soil~ / g~fresh\ soil~\],

  - WC = water content \[g~water~ / g~fresh\ soil~\] [**!! UNIT: per g fresh soil !!**]{.underline}

- ratio = fresh weight of soil added into tubes for extraction (\~10g) / volume of extractant added (=20ml)

Then, we can finally convert concentrations in mg/L into concentrations in ppm (mg/kg dry soil)

- see <https://mycloud.ulb.be/index.php/s/9ZKaJg9rXJg4iB7> for equation

### 1.1.2 - Pivot data so that technical replicates are above each other, not next to each other

First, we select relevant columns and pivot to get

- one column respectively for tare, gross fresh weight (g_fw) and gross dry weight (g_dw)

- 1 row per technical replicate

```{r}
raw_greenhouse_t2_subset_wc <- raw_greenhouse_t2_flush |> 
  select(sample_short, flush_dm_tare_tr1:flush_dm_g_dw_tr3) |> 
  pivot_longer(
    cols = !sample_short,
    names_pattern = "flush_dm_(g_fw|g_dw|tare)_tr(\\d)",
    names_to = c(".value", "tech_rep"),
    values_to = "weight") 

# check it out
raw_greenhouse_t2_subset_wc 
```

Then, we

- compute the neat fresh weight and dry weight for each row, and the dry matter content and water content

- Compute the average per sample (over the technical triplicate)

```{r}
greenhouse_t2_wc <- raw_greenhouse_t2_subset_wc |> 
  # compute neat values
  mutate(
    n_fw = g_fw - tare,
    n_dw = g_dw - tare,
    dm = n_dw / n_fw,
    wc = 1 - dm,
    .keep = "unused"
  ) |> 
  # compute mean per sample
  summarize(
    .by = sample_short,
    dm = mean(dm),
    wc = mean(wc)
  ) 

# check it out
greenhouse_t2_wc 
```

Now that we have that clean data, we can

- remove useless columns (raw data used above)

- rejoin it to the complete greenhouse data (with all columns)

- compute each sample's individual "ratio" (\~10g/20ml)

```{r}
# Volume of extractant used in the K2SO4 extraction
vol_extr <- 20 

greenhouse_t2_flush_clean <- raw_greenhouse_t2_flush |> 
  select(
    !flush_dm_tare_tr1:flush_dm_g_dw_tr3 & 
      !ends_with("comment")) |> 
  left_join(greenhouse_t2_wc) |> 
  mutate(
    ratio_nf = flush_fw_nf_g / vol_extr,
    ratio_cfe = flush_fw_cfe_g / vol_extr,
    .keep = "unused") 
  
greenhouse_t2_flush_clean
```

Now, we divide the data in 2 subsets: sample data and standard soil data (several reps, messes up the analysis (for now))

```{r}
greenhouse_t2_flush_sample <- greenhouse_t2_flush_clean |> filter(biol_unit_nb < 100)
greenhouse_t2_flush_std <- greenhouse_t2_flush_clean |> filter_out(biol_unit_nb < 100)
```

## 1.2 - Nmin concentrations in ppm

First, we

- join absorbance data and flush data,

- Compute the N-sp N concentration in ppm (e.g. mg NO3-N / kg dry soil)

- do some data housecleaning

```{r}
greenhouse_t2_Nmin_sample |> arrange(biol_unit_nb) |> view()

greenhouse_t2_Nmin_ppm_sample <- greenhouse_t2_Nmin_sample |> 
  # join both data tables
  left_join(greenhouse_t2_flush_sample) |> 
  # compute concentration in ppm
  mutate(conc_ppm = conc_mgN_L / (ratio_nf * dm)) |> 
  # remove again the bare soils
  filter_out(is.na(sample_short)) |> 
  # select relevant columns
  select(!c(
    dataset, expe, 
    target_sp:lm_p, 
    starts_with("sampl"), starts_with("ratio"),
    map, plate_id, abs_corrected, conc_mgNsp_L, conc_mgN_L)) |> 
  arrange(biol_unit_nb)

# Check it out
greenhouse_t2_Nmin_ppm_sample
```

Now, all that is left to do is to pivot the data so that

# 3 - Export

```{r}

```
