---
title: "5.1 exploratory analysis for t2 field"
format: 
  gfm:
    toc: true
   # code-fold: true
# format: 
#   html:
#     embed-resources: true
#     toc: true
#     toc_float: true
#     code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# Set up

```{r}
#| message: false
#| warning: false

# clean environment
rm(list = ls())

# packages
library(tidyverse)
library(vegan) # for rda()
library(patchwork)
library(RColorBrewer)

# function
source("functions/plot_pca.R")

```

# 2 - Compute LER

Only possible for wheat, because we don't have sole faba bean in the field.

But this needs to consider seeding rates, so I'm not sure it will be correct in terms of absolute values (yet).

[**!! TO DO: correct LER based on seeding rates !!**]{.underline}

```{r}
#| message: false

  
ler_w <- data_w |> 
  select(cs, soil, bloc, grain_yield_15p_t_per_ha, protein_percent_dw) |> 
  filter_out(is.na(bloc)) |> # remove empty blocs (= bloc 1, still here bc info on bloc was not stored in yield data --> not removed yet)
  pivot_wider(
    names_from = cs,
    values_from = c(grain_yield_15p_t_per_ha, protein_percent_dw)) |> 
  mutate(
    ler_w_off = grain_yield_15p_t_per_ha_IC / grain_yield_15p_t_per_ha_W,
    ler_w_n = protein_percent_dw_IC / protein_percent_dw_W
  )


```

Plot it

```{r}
#| warning: false
plot_ler_yield <- ler_w |> 
  ggplot(aes(x = soil, y = ler_w_off)) +
  theme_minimal() +
  geom_boxplot() +
  geom_point() +
  labs(title = "LER wheat", subtitle = "Field, harvest\n! Uncorrected values (seeding rates missing)")

plot_ler_protein <- ler_w |> 
  ggplot(aes(x = soil, y = ler_w_n)) +
  theme_minimal() +
  geom_boxplot() +
  geom_point() +
  labs(title = "LER wheat, grain protein", subtitle = "Field, harvest")

plot_ler_yield + plot_ler_protein 
```

We shouldn't look at the Ref in theory, because the Faba Bean did not take, although it was re-sown. Still, it was harvested. Interesting that it shows the highest gain in protein in the case of intercrops. Could it be that because the Fababean was way overshadowed, it really over-regulated the N fixation? Or is it just the slighly lower seeding density of the wheat that gave it a highest share of the available soil N? Possibly a bit of both?

Though of course: Ref has lower yield (per ha) in IC bc of the lower density?

# 3 - PMN

With the PMN data, we can compute the slope of the curve.

::: callout-important
## Statistical issue

We have 4 technical replicates per incubation time, but rep 1 of incubation time 1 has no particular relationship to rep 1 of incubation time 2. It would make little sense to compute 4 different slopes and then take the mean of the 4 slopes to get a boxplot and the option of an anova with comparison of means (of slopes).
:::

According to AI (see full doc):

- ANCOVA, (independant measurements bc subsamples evolve separately) .

  - linear model with interaction term: outcome \~ group \* predictor

  - --\> try with lm(ppm_Nmin \~ soil \* incub_days)

  - Details see doc

```{r}
my_lm <- lm(data = PMN, ppm_Nmin ~ soil * incub_time)
summary(my_lm)

anova(my_lm)
```

The Anova (in this case ANCOVA) says:

- effect of soil

- effect of incub_time ++ (duh)

- but no effect of the interaction –\> slopes are not different (no rejection of H0 = slopes are the same)

```{r}

dates <- c("2023-12-12", "2024-01-02", "2024-01-09") |> as.Date()
incub_days <- dates - dates[1]
names(incub_days) <- PMN$incub_time |> unique() |> sort()

# plot again
PMN_plot <- PMN |> 
  ggplot(aes(x = incub_day, y = ppm_Nmin, group = soil, colour = soil, fill = soil)) +
  theme_minimal() +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    title = "Potential Mineralization of Nitrogen (PMN) - Field",
    subtitle = "Mineral N as sum of NO2-, NO3-, NH4+\nNo statistical difference between slopes, only intercept",
    caption = "Modelled with lm(ppm_Nmin ~ soil * incub_time)") +
  xlab("Time since incubation at 28°C [days]") +
  ylab("Mineral Nitrogen in the soil [ppm]")

# for annotation: extract data from smooth curve
smooth_data <- ggplot_build(PMN_plot)$data[[2]]

# get its maximum value to anker the annotation
annotations <- smooth_data |> 
  slice_max(x, by = group) |> 
  mutate(soil = levels(PMN$soil))
#c("#F8766D", "#7CAE00", "#00BFC4", "#C77CFF") 

PMN_plot_all <- PMN_plot + 
  geom_text(
    data = annotations,
    aes(x = x+0.5, y = y, colour = soil, label = soil), 
    hjust = 0, size = 5) +
  xlim(c(0,30)) +
  theme(legend.position = "none") 

```

Look at both plots

```{r}
PMN_plot_all

PMN_plot_all + facet_wrap(~soil) +
  xlim(c(0,38))
```

Save PMN plot

```{r}
ggsave("output/figures/field/PMN.pdf", plot = PMN_plot_all)
```

# 4 - MicroResp

To do: per microplot average (across zones)?, definitely: remove bloc 1 and remove Ref-IC

```{r}
MR |> 
  filter_out(soil == "Std") |> 
  ggplot(aes(x = soil, y = shannon, colour = cs)) +
  theme_minimal() +
  geom_boxplot() +
  facet_wrap(~bloc, nrow = 1) +
  geom_point() 

MR |> 
  filter_out(soil == "Std") |>
  ggplot(aes(x = soil, y = MSIR, colour = cs)) +
  theme_minimal() +
  geom_boxplot() +
  facet_wrap(~bloc, nrow = 1) +
  geom_point()

MR |> 
  filter_out(soil == "Std") |>
  ggplot(aes(x = soil, y = MBC, colour = cs)) +
  theme_minimal() +
  geom_boxplot() +
  facet_wrap(~bloc, nrow = 1) +
  geom_point()

MR |> 
  filter_out(soil == "Std") |>
  ggplot(aes(x = soil, y = basal_resp, colour = cs)) +
  theme_minimal() +
  geom_boxplot() +
  facet_wrap(~bloc, nrow = 1) +
  geom_point()

MR |> 
  filter_out(soil == "Std") |>
  ggplot(aes(x = soil, y = qCO2, colour = cs)) +
  theme_minimal() +
  geom_boxplot() +
  facet_wrap(~bloc, nrow = 1) +
  geom_point()
```

# °°° !!! FROM HERE - TO ADAPT FROM GREENHOUSE !!! °°°

# 5 - PCA

## 4.1 - Check assumptions

.... UPCOMING?

## 4.2 - Run PCA

### 4.2.1 - Graphical parameters

```{r}
library(RColorBrewer)
display.brewer.all(n = 2, type = "qual")
display.brewer.pal(n = 3, name = "Accent")
cs_colors <- brewer.pal(n = 3, name = "Accent")[2:3]
names(cs_colors) <- c("IC", "W")

display.brewer.all(n = 4, type = "qual")
display.brewer.pal(n = 4, name = "Set1")
soil_colors <- brewer.pal(n = 4, name = "Set1")[2:4]
names(soil_colors) <- c("Ref", "Auto", "ABC")
```

### 4.2.2 - Faba Bean

```{r}

rda_matrix_fb <- data_fb |> 
  select(
    biol_unit_nb,
    ppm_NO3, ppm_NH4, 
    #stem_per_fb, 
    grain_yield_15p_t_per_ha, protein_percent_dw) |> 
  column_to_rownames("biol_unit_nb") |> 
  drop_na()
  
pca_fb <- pca(rda_matrix_fb, scale = TRUE)

summary(pca_fb)

screeplot(pca_fb, bstick = TRUE, npcs = length(pca_fb$CA$eig))

#biplot(pca_fb, scaling = 1, main = "PCA - scaling 1 = object relationships preserved")
biplot(pca_fb, main = "PCA FB - scaling 2 = variable relationships preserved")


```

Prepare data for plotting

```{r}
#non_pca <- greenhouse |> select(!ppm_NH4:NO3_NH4)
pot_scores <- scores(pca_fb, display = "sites", scaling = 2) 
biol_unit <- rownames(pot_scores) |> as.double()
pot_scores <- pot_scores |> 
  as_tibble() |> 
  mutate(biol_unit_nb = biol_unit)

var_scores <- scores(pca_fb, display = "species", scaling = 2)
variables <- rownames(var_scores)
var_scores <- var_scores |> 
  as_tibble() |> 
  mutate(variable = variables)

PC_percent <- pca_fb$CA$eig / sum(pca_fb$CA$eig)

data_fb_pca <- left_join(data_fb, pot_scores, by = join_by(biol_unit_nb))
```

Plot it

```{r}
#| warning: false
#dev.new()
plot_pca_fb_cs <- plot_pca(data_fb_pca, var_scores, PC_percent) +
  geom_point(aes( colour = cs)) +
  labs(title = "PCA - Faba Bean") +
  scale_color_discrete(palette = c(cs_colors["FB"], cs_colors["IC"])) 

plot_pca_fb_cs

plot_pca_fb_soil <- plot_pca(data_fb_pca, var_scores, PC_percent) +
  geom_point(aes( colour = soil, shape = cs)) +
  labs(title = "PCA - Faba Bean") +
  scale_color_discrete(palette = c(soil_colors)) 

plot_pca_fb_soil
```

### 3.2.3 - Wheat

```{r}

rda_matrix_w <- data_w |> 
  select(
    biol_unit_nb,
    ppm_NO3, ppm_NH4, 
    #till_per_w, 
    grain_yield_15p_t_per_ha, protein_percent_dw) |> 
  column_to_rownames("biol_unit_nb") |> 
  drop_na()
  
pca_w <- pca(rda_matrix_w, scale = TRUE)

summary(pca_w)

screeplot(pca_w, bstick = TRUE, npcs = length(pca_w$CA$eig))

#biplot(pca_fb, scaling = 1, main = "PCA - scaling 1 = object relationships preserved")
biplot(pca_w, main = "PCA Wheat - scaling 2 = variable relationships preserved")


```

Prepare data for plotting

```{r}
#non_pca <- greenhouse |> select(!ppm_NH4:NO3_NH4)
pot_scores <- scores(pca_w, display = "sites", scaling = 2) 
biol_unit <- rownames(pot_scores) |> as.double()
pot_scores <- pot_scores |> 
  as_tibble() |> 
  mutate(biol_unit_nb = biol_unit)

var_scores <- scores(pca_w, display = "species", scaling = 2)
variables <- rownames(var_scores)
var_scores <- var_scores |> 
  as_tibble() |> 
  mutate(variable = variables)

PC_percent <- pca_w$CA$eig / sum(pca_w$CA$eig)

data_w_pca <- left_join(data_w, pot_scores, by = join_by(biol_unit_nb))
```

Plot it

```{r}
#dev.new()
plot_pca_w_cs <- plot_pca(data_w_pca, var_scores, PC_percent) +
  geom_point(aes( colour = cs)) +
  scale_color_discrete(palette = c(cs_colors["IC"], cs_colors["W"])) +
  labs(title = "PCA - Wheat")
  
#plot_pca_w_cs

plot_pca_w_soil <- plot_pca(data_w_pca, var_scores, PC_percent) +
  geom_point(aes( colour = soil, shape = cs)) +
  labs(title = "PCA - Wheat") +
  scale_color_discrete(palette = c(soil_colors)) 

#plot_pca_w_soil
```

### 3.2.4 - All plots

```{r}
#| warning: false
plot_pca_fb_cs + plot_pca_w_cs + plot_layout(guides = "collect")

plot_pca_fb_soil + plot_pca_w_soil + plot_layout(guides = "collect")
```

# What now?

According to the screeplots, I shouldn't even represent the first PCA (at least not with this version with only 4 variables).

```{r}
# TEST CHUNK



```
