---
title: "V. Exploratory data analysis"
format: 
  gfm:
    toc: true
    code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# To Do

-   move transformation to pivot_wider (compute new variables) to the transformation script & adapt this one

# Set up

```{r}
rm(list = ls())

library(tidyverse)
library(roperators)
library(RColorBrewer)
```

# 1 - Nmin

## 1.1 - import data

```{r}
Nmin_tfo <- read_rds("output/data/4_Nmin_tfo.rds")
Nmin_tfo
```

## 1.2 - plots

### 1.2.1 - Raw distributions

First, distribution of values (taking all technical reps = wells and of average) per N species

! NO3 was not yet corrected for NO2

```{r}
# need to pivot to get 4 tech rep in one column. But !! also average will be taken
Nmin_tfo_longer <- Nmin_tfo |> pivot_longer(
  cols = starts_with("conc"),
  values_to = "conc_N_ppm",
  names_to = "rep_tech"
) 

# plot with all rep techs
Nmin_tfo_longer |> 
  # filter out the average
  filter(rep_tech != "conc_N_ppm_avg") |> 
  ggplot(aes(x = conc_N_ppm)) +
  theme_minimal() +
  labs(caption = "!! NO3 not yet corrected for NO2 !!") +
  #geom_histogram(fill = "white", color = "grey70", bins = 50) +
  geom_density() +
  facet_wrap(~N_sp, nrow = 3, strip.position = "right", scales = "free_y")

# plot with averages
Nmin_tfo_longer |> 
  # filter out the average
  filter(rep_tech == "conc_N_ppm_avg") |> 
  ggplot(aes(x = conc_N_ppm)) +
  theme_minimal() +
  labs(caption = "!! NO3 not yet corrected for NO2 !!") +
#  geom_histogram(fill = "white", color = "grey70", bins = 50) +
  geom_density() +
  facet_wrap(~N_sp, nrow = 3, strip.position = "right", scales = "free_y")

```

### 1.2.2 - Distributions of Nmin and N species

```{r}
# To correct NO3, we need to pivot wider so that different N_sp are in different columns
Nmin_tfo_wider <- Nmin_tfo |> 
  select(expe:sampling_time, N_sp, conc_N_ppm_avg) |> 
  pivot_wider(
    names_from = N_sp,
    names_prefix = "conc_",
    values_from = conc_N_ppm_avg
  ) |> 
  mutate(
    conc_NO3_neat = conc_NO3-conc_NO2,
    conc_Nmin = conc_NO3 + conc_NH4, # NO2 not included here bc it is already included in NO3
    conc_NO3_Nmin = conc_NO3_neat / conc_Nmin,
    conc_NH4_NO3 = conc_NH4 / conc_NO3_neat) |> 
  # remove NO3 so no mistake is made, and rename NO3_neat
  select(!conc_NO3) |> 
  rename(conc_NO3 = conc_NO3_neat) 

Nmin_tfo_corrected <- Nmin_tfo_wider |> 
  # then we re-pivot so plotting with wrap is easy (and without also --° use filter)
  pivot_longer(
    cols = starts_with("conc"),
    names_to = "N_sp",
    names_prefix = "conc_",
    values_to = "conc_N_ppm") |> 
  # make N_sp a factor with correct order
  mutate(N_sp = factor(N_sp, levels = c("NO2", "NH4", "NO3", "Nmin", "NO3_Nmin", "NH4_NO3")))

Nmin_tfo_corrected  

Nmin_tfo_corrected |> 
  filter(N_sp %ni% c("NO2", "NO3_Nmin", "NH4_NO3")) |> # comment this line to add NO2 again
  ggplot(aes(x = conc_N_ppm)) +
  theme_minimal() +
  geom_density(aes(color = N_sp, fill = N_sp), alpha = 0.2) #+
#  facet_wrap(~N_sp, nrow = 4, strip.position = "right", scales = "free_y")
```

### 1.2.3 - 

Look at NO3 and NH4

```{r}
Nmin_tfo_corrected |> 
  filter(N_sp %in% c("NH4", "NO3")) |> 
  ggplot(aes(x = soil, y = conc_N_ppm)) +
  theme_minimal() +
  geom_boxplot(aes(fill = cs)) +
  facet_wrap(~N_sp, nrow = 2, scales = "free_y") +
  scale_fill_discrete(palette = "Accent")
  

```

Look at Nmin and ratios

```{r}
Nmin_tfo_corrected |> 
  filter(N_sp %in% c("Nmin", "NO3_Nmin")) |> 
  ggplot(aes(x = soil, y = conc_N_ppm)) +
  theme_minimal() +
  geom_boxplot(aes(fill = cs)) +
  facet_wrap(~N_sp, nrow = 2, scales = "free_y") +
  scale_fill_discrete(palette = "Accent")
  
```

Look at relationship between NO3 and NH4

```{r}
Nmin_tfo_wider |>
  ggplot(aes(x = conc_NH4, y = conc_NO3)) +
  theme_minimal() +
  geom_point() #+
#  facet_wrap(~cs)

  
```

And between NO3 and Nmin

```{r}
Nmin_tfo_wider |>
  ggplot(aes(x = conc_NO3, y = conc_Nmin)) +
  theme_minimal() +
  geom_point() #+
#  facet_wrap(~cs)
```
