---
title: "V. Exploratory data analysis"
format: 
  gfm:
    toc: true
    code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# To Do

-   move transformation to pivot_wider (compute new variables) to the transformation script & adapt this one

# Set up

```{r}
rm(list = ls())

library(tidyverse)
library(roperators)
library(RColorBrewer)
```

# 1 - Nmin

## 1.1 - import data

```{r}
Nmin_tfo1 <- read_rds("output/data/4_Nmin_tfo1.rds")
Nmin_tfo2 <- read_rds("output/data/4_Nmin_tfo2.rds")

#Nmin_tfo1
```

## 1.2 - plots

### 1.2.1 - Raw distributions

First, distribution of values (taking all technical reps = wells and of average) per N species

! NO3 was not yet corrected for NO2

```{r}
# need to pivot to get 4 tech rep in one column. But !! also average will be taken
#** °°°°° Pivot according to rep tech °°°°° *
Nmin_rep_tech_longer <- Nmin_tfo1 |> pivot_longer(
  cols = starts_with("conc"),
  values_to = "conc_N_ppm",
  names_to = "rep_tech"
) 

# plot with all rep techs
Nmin_rep_tech_longer |> 
  # filter out the average
  filter(rep_tech != "conc_N_ppm_avg") |> 
  ggplot(aes(x = conc_N_ppm)) +
  theme_minimal() +
  labs(caption = "!! NO3 not yet corrected for NO2 !!") +
  #geom_histogram(fill = "white", color = "grey70", bins = 50) +
  geom_density() +
  facet_wrap(~N_sp, nrow = 3, strip.position = "right", scales = "free_y")

# plot with averages
Nmin_rep_tech_longer |> 
  # filter out the average
  filter(rep_tech == "conc_N_ppm_avg") |> 
  ggplot(aes(x = conc_N_ppm)) +
  theme_minimal() +
  labs(caption = "!! NO3 not yet corrected for NO2 !!") +
#  geom_histogram(fill = "white", color = "grey70", bins = 50) +
  geom_density() +
  facet_wrap(~N_sp, nrow = 3, strip.position = "right", scales = "free_y")

```

### 1.2.2 - Distributions of Nmin and N species

```{r}


Nmin_N_variable_longer <- Nmin_tfo2 |> 
  # then we re-pivot so plotting will wrap is easily (and without also --° use filter)
  #** °°°°° Pivot according to N_variable
  pivot_longer(
    cols = starts_with("conc"),
    names_to = "N_variable",
    names_prefix = "conc_",
    values_to = "conc_N_ppm") |> 
  # make N_sp a factor with correct order
  mutate(N_variable = factor(N_variable, levels = c("NO2", "NH4", "NO3", "Nmin", "NO3_Nmin", "NH4_NO3")))

Nmin_N_variable_longer  

Nmin_N_variable_longer |> 
  filter(N_variable %ni% c("NO2", "NO3_Nmin", "NH4_NO3")) |> # comment this line to add NO2 again
  ggplot(aes(x = conc_N_ppm)) +
  theme_minimal() +
  geom_density(aes(color = N_variable, fill = N_variable), alpha = 0.2) #+
#  facet_wrap(~N_sp, nrow = 4, strip.position = "right", scales = "free_y")
```

### 1.2.3 -

Look at NO3 and NH4

```{r}
Nmin_N_variable_longer |> 
  filter(N_variable %in% c("NH4", "NO3")) |> 
  ggplot(aes(x = soil, y = conc_N_ppm)) +
  theme_minimal() +
  geom_boxplot(aes(fill = cs)) +
  facet_wrap(~N_variable, nrow = 2, scales = "free_y") +
  scale_fill_discrete(palette = "Accent")
  

```

Look at Nmin and ratios

```{r}
Nmin_N_variable_longer |> 
  filter(N_variable %in% c("Nmin", "NO3_Nmin")) |> 
  ggplot(aes(x = soil, y = conc_N_ppm)) +
  theme_minimal() +
  geom_boxplot(aes(fill = cs)) +
  facet_wrap(~N_variable, nrow = 2, scales = "free_y") +
  scale_fill_discrete(palette = "Accent")
  
```

Look at relationship between NO3 and NH4

```{r}
Nmin_tfo2 |>
  ggplot(aes(x = conc_NH4, y = conc_NO3)) +
  theme_minimal() +
  geom_point() #+
#  facet_wrap(~cs)

  
```

And between NO3 and Nmin

```{r}
Nmin_tfo2 |>
  ggplot(aes(x = conc_NO3, y = conc_Nmin)) +
  theme_minimal() +
  geom_point() #+
#  facet_wrap(~cs)
```
