---
title: "5 PCA for t2 greenhouse"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# TO DO

# Set up

```{r}
#| message: false
#| warning: false

# clean environment
rm(list = ls())

# packages
library(tidyverse)
library(vegan) # for rda()

# data

```

# 1 - Load and clean data

```{r}
greenhouse <- read_rds("output/data/4_t2_greenhouse_transformed.rds") |> 
  select(!c(
    dataset, expe:sampling_time
  )) |> 
  relocate(yd_rs_comment, .after = last_col())

str(greenhouse)
```

Problem: for now, the statistical unit is the pot (1 row per pot). But some variables cannot be regrouped for both plant species. Consequently, there are some missing data, see here:

```{r}
greenhouse |> select(cs, fw_fb_per_pot:dw_w_per_plant)
```

A PCA cannot be ran with missing data.

A solution is to change the statistical unit to be the plant, and have 2 data sets: dataset fababean (`data_fb`) and dataset wheat (`data_w`). Anyway, it makes little sense to compare the 2 plant species, but it is interesting to see if effects are similar or not. To be considered: bulk soil data of intercrops will appear twice, so that the 2 datasets are not to be combined in downstream analyses.

So we do that subsetting

```{r}
data_fb <- greenhouse |> 
  select(!contains("_w")) |> 
  filter(cs != "W")

data_w <- greenhouse |> 
  select(!contains("_f")) |> 
  filter(cs != "F")

# check it out
str(data_fb) ; str(data_w)
```

# 2 - Compute LER

This also cannot go into the PCA since we get only data that agregates on SC + IC

```{r}
ler_fb <- data_fb |> 
  select(crop_diversity, soil, bloc, dw_fb_per_plant) |> 
  pivot_wider(
    names_from = crop_diversity,
    values_from = dw_fb_per_plant) |> 
  mutate(
    ler_fb = IC / SC
  )
  
ler_w <- data_w |> 
  select(crop_diversity, soil, bloc, dw_w_per_plant) |> 
  pivot_wider(
    names_from = crop_diversity,
    values_from = dw_w_per_plant) |> 
  mutate(
    ler_w = IC / SC
  )

ler_total <- ler_w |> 
  select(!SC:IC) |> 
  left_join(ler_fb |> select(!SC:IC)) |> 
  mutate(ler_total = (ler_w + ler_fb)/2)
```

Plot it

```{r}
plot_ler_total <- ler_total |> 
  ggplot(aes(x = soil, y = ler_total)) +
  theme_minimal() +
  geom_boxplot() +
  labs(title = "LER total - pot experiment, t2")

plot_ler_fb <- ler_total |> 
  ggplot(aes(x = soil, y = ler_fb)) +
  theme_minimal() +
  geom_boxplot() +
  labs(title = "LER Faba Bean - pot experiment, t2")

plot_ler_w <- ler_total |> 
  ggplot(aes(x = soil, y = ler_w)) +
  theme_minimal() +
  geom_boxplot() +
  labs(title = "LER Wheat - pot experiment, t2")

plot_ler_total + plot_ler_fb + plot_ler_w
```

# 2 - PCA

## 2.1 - Check asumptions

.... UPCOMING?

## 2.2 - Run PCA

```{r}
rda_matrix <- greenhouse |> 
  select(
    ppm_NH4:NO3_NH4,
  )
Nmin_pca <- rda(rda_matrix, scale = TRUE)

summary(Nmin_pca)

screeplot(Nmin_pca, bstick = TRUE, npcs = length(Nmin_pca$CA$eig))

biplot(Nmin_pca, scaling = 1, main = "PCA - scaling 1 = object relationships preserved")
biplot(Nmin_pca, main = "PCA - scaling 2 = variable relationships preserved")
```

project categ variables

```{r}
non_pca <- greenhouse |> select(!ppm_NH4:NO3_NH4)
site_scores <- scores(Nmin_pca, display = "wa", scaling = 1)

Nmin_w <- hclust(dist(scale(rda_matrix)), "ward.D")
cutree(Nmin_w, k = 4)

variable <- greenhouse$soil
var_lev <- levels(factor(variable))

data <- left_join(greenhouse, 
                  site_scores |> as_tibble() |> mutate(biol_unit_nb = seq(1,60))
                  )

data |> 
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point(aes( colour = cs))


plot(
  Nmin_pca,
  display = "wa",
  scaling = 1,
  type = "n",
  main = "PCA correlation"
)
abline(v = 0, lty = "dotted")
abline(h = 0, lty = "dotted")
for (level in levels(variable)) {
  points(site_scores[variable == level,],
         pch = (14 + i),
         cex = 2,
         col = i + 1)
}

```
