---
title: "5 PCA and LER for t2 greenhouse"
format: 
  gfm:
    toc: true
   # code-fold: true
# format: 
#   html:
#     embed-resources: true
#     toc: true
#     toc_float: true
#     code-fold: true
editor: visual
editor_options: 
  chunk_output_type: console
message: true
warning: true
echo: true
output: true
---

# TO DO

Still missing:

- PMN

- Possibly MR, but probably not so relevant bc very low activity

- Not TDN (because incomplete)

# 0 - Intro

For the greenhouse data sets, many bits and pieces are dropped along the way, so that only few variables are left in the end:

- The whole t1 sampling is problematic due to the scattered sampling over time

- For t2:

  - MicroResp experiment: still needs calibration, but responses were really low, probably due to high sand levels

  - PMN: samples had to be re-humidified because the container system was not optimal and samples had become basically dry (see script on data wrangling: curves show interesting trends, but don't know what to do with them)

  - TDN & MBN: never got the protocol from Sang, and did a mistake by not adjusting the serial dilution of the standard curve –\> samples out of range

  - WHC is not intrinsically interesting, though we have the data as well (was needed for MicroResp)

In the end, what is left here is not much to work with:

- yield data (plant biomass, plant height, nb of stems / tillers)

- Nmin data at t2 (NO2, NO3, NH4 –\> Nmin and ratios)

- LER can be derived from yield data (but = aggregated data –\> no incorporation in PCA possible)

What is still dormant is weed data. Relevant? not sure...

# Set up

```{r}
#| message: false
#| warning: false

# clean environment
rm(list = ls())

# packages
library(tidyverse)
library(vegan) # for rda()
library(patchwork)
library(RColorBrewer)
library(emmeans) # for post-hoc test (after ANCOVA of PMN)

# function
source("functions/plot_pca.R")

```

# 1 - Load and clean data

```{r}
greenhouse <- read_rds("output/data/4_t2_greenhouse_transformed.rds") |> 
  select(!c(
    dataset, expe:sampling_time
  )) |> 
  relocate(yd_rs_comment, .after = last_col())

PMN <- read_rds("output/data/4_greehouse_PMN.rds")
```

Problem: for now, the statistical unit is the pot (1 row per pot). But some variables cannot be regrouped for both plant species. Consequently, there are some missing data, see here:

```{r}
greenhouse |> select(cs, fw_fb_per_pot:dw_w_per_plant)
```

A PCA cannot be ran with missing data.

A solution is to change the statistical unit to be the plant, and have 2 data sets: dataset fababean (`data_fb`) and dataset wheat (`data_w`). Anyway, it makes little sense to compare the 2 plant species, but it is interesting to see if effects are similar or not. To be considered: bulk soil data of intercrops will appear twice, so that the 2 datasets are not to be combined in downstream analyses.

So we do that subsetting

```{r}
data_fb <- greenhouse |> 
  select(!contains("_w")) |> 
  filter(cs != "W") |> 
  arrange(biol_unit_nb)

data_w <- greenhouse |> 
  select(!contains("_f")) |> 
  filter(cs != "F") |> 
  arrange(biol_unit_nb)

# check it out
str(data_fb) ; str(data_w)
```

# 2 - Compute LER

This also cannot go into the PCA since we get only data that agregates on SC + IC

```{r}
#| message: false
ler_fb <- data_fb |> 
  select(crop_diversity, soil, bloc, dw_fb_per_plant) |> 
  pivot_wider(
    names_from = crop_diversity,
    values_from = dw_fb_per_plant) |> 
  mutate(
    ler_fb = IC / SC
  )
  
ler_w <- data_w |> 
  select(crop_diversity, soil, bloc, dw_w_per_plant) |> 
  pivot_wider(
    names_from = crop_diversity,
    values_from = dw_w_per_plant) |> 
  mutate(
    ler_w = IC / SC
  )

ler_total <- ler_w |> 
  select(!SC:IC) |> 
  left_join(ler_fb |> select(!SC:IC)) |> 
  mutate(ler_total = (ler_w + ler_fb)/2)
```

Plot it

```{r}
#| warning: false
plot_ler_total <- ler_total |> 
  ggplot(aes(x = soil, y = ler_total)) +
  theme_minimal() +
  geom_boxplot() +
  labs(title = "LER total", subtitle = "Pot experiment, t2")

plot_ler_fb <- ler_total |> 
  ggplot(aes(x = soil, y = ler_fb)) +
  theme_minimal() +
  geom_boxplot() +
  labs(title = "LER Faba Bean", subtitle = "Pot experiment, t2")

plot_ler_w <- ler_total |> 
  ggplot(aes(x = soil, y = ler_w)) +
  theme_minimal() +
  geom_boxplot() +
  labs(title = "LER Wheat", subtitle = "Pot experiment, t2")

plot_ler_total + plot_ler_fb + plot_ler_w
```

# 3 - PMN

## 3.1 - ANCOVA

With the PMN data, we can compute the slope of the curve.

::: callout-important
## Statistical issue

We have 4 technical replicates per incubation time, but rep 1 of incubation time 1 has no particular relationship to rep 1 of incubation time 2. It would make little sense to compute 4 different slopes and then take the mean of the 4 slopes to get a boxplot and the option of an anova with comparison of means (of slopes).
:::

According to AI (see full doc):

- Analysis of CoVariance, ANCOVA, (independant measurements bc subsamples evolve separately) .

  - linear model with interaction term: outcome \~ group \* predictor

  - --\> try with lm(ppm_Nmin \~ soil \* incub_days)

  - Details see doc

```{r}
PMN_continuous <- PMN |> mutate(incub_day = as.integer(incub_day))

my_lm <- lm(data = PMN_continuous, ppm_Nmin ~ incub_day * soil)
summary(my_lm)

anova(my_lm)
```

The Anova (in this case ANCOVA) says:

- Effect of incubation time (duh)

- effect of soil

- interaction effect –\> reject H0 = slopes are equal

—\> need of a post-hoc test, with Tukey adjustment

- conv ≠ Ref, ABC

- Ref = ABC, Ref = Auto

- Auto ≠ ABC

```{r}
slope_trends <- emtrends(my_lm, pairwise ~ soil, var = "incub_day")

summary(slope_trends)

# see the slopes
slope_trends$emtrends

# see the pairwise contrasts with adjusted p-values (Tukey method)
slope_trends$contrasts
```

Checking model assumptions

```{r}
library(performance)
plot_check_ancova <- check_model(my_lm) 
plot <- plot(plot_check_ancova) + patchwork::plot_annotation(title = "QC: Checking model assumptions for ANCOVA of PMN - Greenhouse data")

# check it out
plot

# save it
ggsave("output/figures/greenhouse/qc_pmn_ancova.pdf", plot,height = 8, width = 13.5)
```

## 3.2 - Plot PMN

```{r}
dates <- c("2023-12-5", "2023-12-26", "2024-01-02") |> as.Date()
incub_days <- dates - dates[1]
names(incub_days) <- PMN$incub_time |> unique() |> sort()

# Plot
# plot again
PMN_plot <- PMN |> 
  ggplot(aes(x = incub_day, y = ppm_Nmin, group = soil, colour = soil, fill = soil)) +
  theme_minimal() +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    title = "Potential Mineralization of Nitrogen (PMN) - Greenhouse",
    subtitle = "Mineral N as sum of NO2-, NO3-, NH4+\nSlopes are statistically different (Conv ≠ Ref, ABC; Auto ≠ ABC)",
    caption = "Modelled with lm(ppm_Nmin ~ incub_time * soil)\nPost-Hoc with emtrends() and Tukey correction") +
  xlab("Time since incubation at 28°C [days]") +
  ylab("Mineral Nitrogen in the soil [ppm]")

# for annotation: extract data from smooth curve
smooth_data <- ggplot_build(PMN_plot)$data[[2]]

# get its maximum value to anker the annotation
annotations <- smooth_data |> 
  slice_max(x, by = group) |> 
  mutate(soil = levels(PMN$soil))
#c("#F8766D", "#7CAE00", "#00BFC4", "#C77CFF") 

PMN_plot_all <- PMN_plot + 
  geom_text(
    data = annotations,
    aes(x = x+0.5, y = y, colour = soil, label = soil), 
    hjust = 0, size = 5) +
  xlim(c(0,30)) +
  theme(legend.position = "none") 

```

Check them out

```{r}
PMN_plot_all

PMN_plot_all + facet_wrap(~soil) +
  xlim(c(0,33))
```

Save it

```{r}
ggsave("output/figures/greenhouse/PMN.pdf", plot = PMN_plot_all)
```

# 4 - PCA

## 3.1 - Check assumptions

.... UPCOMING?

## 3.2 - Run PCA

### 3.2.1 - Graphical parameters

```{r}
library(RColorBrewer)
display.brewer.all(n = 3, type = "qual")
display.brewer.pal(n = 3, name = "Accent")
cs_colors <- brewer.pal(n = 3, name = "Accent")
names(cs_colors) <- c("FB", "IC", "W")

display.brewer.all(n = 4, type = "qual")
display.brewer.pal(n = 4, name = "Set1")
soil_colors <- brewer.pal(n = 4, name = "Set1")
names(soil_colors) <- c("Conv", "Ref", "Auto", "ABC")
```

### 3.2.2 - Faba Bean

```{r}

rda_matrix_fb <- data_fb |> 
  select(
    biol_unit_nb,
    ppm_NO3, ppm_NH4, 
    #stem_per_fb, 
    height_per_fb, dw_fb_per_plant) |> 
  column_to_rownames("biol_unit_nb") |> 
  drop_na()
  
pca_fb <- pca(rda_matrix_fb, scale = TRUE)

summary(pca_fb)

screeplot(pca_fb, bstick = TRUE, npcs = length(pca_fb$CA$eig))

#biplot(pca_fb, scaling = 1, main = "PCA - scaling 1 = object relationships preserved")
biplot(pca_fb, main = "PCA FB - scaling 2 = variable relationships preserved")


```

Prepare data for plotting

```{r}
#non_pca <- greenhouse |> select(!ppm_NH4:NO3_NH4)
pot_scores <- scores(pca_fb, display = "sites", scaling = 2) 
biol_unit <- rownames(pot_scores) |> as.double()
pot_scores <- pot_scores |> 
  as_tibble() |> 
  mutate(biol_unit_nb = biol_unit)

var_scores <- scores(pca_fb, display = "species", scaling = 2)
variables <- rownames(var_scores)
var_scores <- var_scores |> 
  as_tibble() |> 
  mutate(variable = variables)

PC_percent <- pca_fb$CA$eig / sum(pca_fb$CA$eig)

data_fb_pca <- left_join(data_fb, pot_scores, by = join_by(biol_unit_nb))
```

Plot it

```{r}
#| warning: false
#dev.new()
plot_pca_fb_cs <- plot_pca(data_fb_pca, var_scores, PC_percent) +
  geom_point(aes( colour = cs)) +
  labs(title = "PCA - Faba Bean") +
  scale_color_discrete(palette = c(cs_colors["FB"], cs_colors["IC"])) 

plot_pca_fb_cs

plot_pca_fb_soil <- plot_pca(data_fb_pca, var_scores, PC_percent) +
  geom_point(aes( colour = soil, shape = cs)) +
  labs(title = "PCA - Faba Bean") +
  scale_color_discrete(palette = c(soil_colors)) 

plot_pca_fb_soil
```

### 3.2.3 - Wheat

```{r}

rda_matrix_w <- data_w |> 
  select(
    biol_unit_nb,
    ppm_NO3, ppm_NH4, 
    #till_per_w, 
    height_per_w, dw_w_per_plant) |> 
  column_to_rownames("biol_unit_nb") |> 
  drop_na()
  
pca_w <- pca(rda_matrix_w, scale = TRUE)

summary(pca_w)

screeplot(pca_w, bstick = TRUE, npcs = length(pca_w$CA$eig))

#biplot(pca_fb, scaling = 1, main = "PCA - scaling 1 = object relationships preserved")
biplot(pca_w, main = "PCA Wheat - scaling 2 = variable relationships preserved")


```

Prepare data for plotting

```{r}
#non_pca <- greenhouse |> select(!ppm_NH4:NO3_NH4)
pot_scores <- scores(pca_w, display = "sites", scaling = 2) 
biol_unit <- rownames(pot_scores) |> as.double()
pot_scores <- pot_scores |> 
  as_tibble() |> 
  mutate(biol_unit_nb = biol_unit)

var_scores <- scores(pca_w, display = "species", scaling = 2)
variables <- rownames(var_scores)
var_scores <- var_scores |> 
  as_tibble() |> 
  mutate(variable = variables)

PC_percent <- pca_w$CA$eig / sum(pca_w$CA$eig)

data_w_pca <- left_join(data_w, pot_scores, by = join_by(biol_unit_nb))
```

Plot it

```{r}
#dev.new()
plot_pca_w_cs <- plot_pca(data_w_pca, var_scores, PC_percent) +
  geom_point(aes( colour = cs)) +
  scale_color_discrete(palette = c(cs_colors["IC"], cs_colors["W"])) +
  labs(title = "PCA - Wheat")
  
#plot_pca_w_cs

plot_pca_w_soil <- plot_pca(data_w_pca, var_scores, PC_percent) +
  geom_point(aes( colour = soil, shape = cs)) +
  labs(title = "PCA - Wheat") +
  scale_color_discrete(palette = c(soil_colors)) 

plot_pca_w_soil
```

### 3.2.4 - All plots

```{r}
#| warning: false
plot_pca_fb_cs + plot_pca_w_cs + plot_layout(guides = "collect")

plot_pca_fb_soil + plot_pca_w_soil + plot_layout(guides = "collect")
```

# What now?

According to the screeplots, I shouldn't even represent the first PCA (at least not with this version with only 4 variables).

```{r}
# TEST CHUNK



```
